### 2013-09-04 rscripts/manuscript.Rnw
manuscript components

\documentclass{article}

% PLOS One TEMPLATE
\usepackage{amsmath,amssymb,graphicx,cite,setspace}
\usepackage[textwidth=1.75in]{todonotes}
\usepackage[total={6in,9.5in},top=0.5in,left=0.5in,includefoot]{geometry}
\usepackage{listings}
\usepackage{url}
\usepackage{color}

% define Scala
\lstdefinelanguage{scala}{morekeywords={class,object,trait,extends,with,new,if,while,for,def,val,var,this},
otherkeywords={->,=>},
%procnamekeys={def,.},
sensitive=true,
keywordstyle=\color{magenta},
commentstyle=\color{green},
identifierstyle=\color{blue},
%procnamestyle=\color{black},
morecomment=[l]{//},
morecomment=[s]{/*}{*/},
morestring=[b]"}
% Default settings for code listings
\lstset{frame=tb,language=scala,aboveskip=3mm,belowskip=3mm,
showstringspaces=false,columns=flexible,basicstyle={\small\ttfamily}}
% handle tilde characters
\lstset{literate=%
{~}{\url{~}}1
}
% define no break code listing
\lstnewenvironment{code}[1][]%
{
   \noindent
   \minipage{\linewidth} 
   \vspace{0.5\baselineskip}
   \lstset{basicstyle=\ttfamily\footnotesize,frame=single,#1}}
{\endminipage}

\newenvironment{rnwfig}[0]{\begin{figure}\begin{center}}{\end{center}\end{figure}}

%\renewcommand{\lstlistingname}{Scala Code}

%\doublespacing

% Text layout
%\topmargin 0.0cm
%\oddsidemargin 0.5cm
%\evensidemargin 0.5cm
%\textwidth 16cm 
%\textheight 21cm

\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

\newcommand{\Hub}[0]{\ensuremath{\mathbf{H}}}
\newcommand{\C}[1]{\ensuremath{\mathbf{C}_{#1}}}
\newcommand{\Obs}[0]{\ensuremath{\mathbf{O}}}
\newcommand{\todoCP}[1]{\todo{CP, #1}}

\begin{document}
\SweaveOpts{concordance=TRUE}
\begin{flushleft}
{\Large
\textbf{Detecting Covert Groups Embedded in a Population}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carl A. B. Pearson$^{1,\ast}$, 
Edo Airoldi$^{2}$, 
Edward Kao$^{2}$,
Burton Singer$^{1}$, 
\\
\bf{1} Emerging Pathogens Institute, University of Florida, Gainesville, FL, USA
\\
\bf{2} Statistics, Harvard University, Cambridge, MA, USA
\\
$\ast$ E-mail: cap10@ufl.edu
\end{flushleft}
% Please keep the abstract between 250 and 300 words
\section*{Abstract}
We outline the problem of characterizing strategies for detection and concealment of clandestine coordination within a broader population, in terms of network models.  Specifically, we propose means to accommodate the uncertainty about behavior and capability from both ends, a general plan for developing such models, some best practices for model parameterization and data gathering, and finally some particularly pernicious pitfalls.

As a practical, but mostly pedagogical demonstration, we specify a graph-based model of simple communication across a procedurally generated population, with an embedded, relatively small module representing a clandestine group, pitted against surveillance systems.  We discuss measuring performance of the opposing sides (e.g. Receiver Operator Characteristic), fitting the model against real data, and finally how this model can be extended.

%\todoCP{{\em I think not for this round, but maybe:} Finally, we consider the implications of {\em forged} messages.  In the basic model, we consider incomplete information about the communications, but the available information is always accurate.  In this extension, we allow the Observer and the clandestine group to forge messages.  We again measure various Observer performance traits relative to properties of the observed network.}

\newpage

\section*{Introduction}
For investigators ranging from anthropologists to law enforcement, the need to identify groups operating in secret -- through deliberate action or otherwise -- is paramount.  In particular, the need for intelligence organizations to identify terrorist cells and defuse their violent plots is a matter of increasing import.  Symmetrically, being able to operate clandestinely in an age of ubiquitous monitoring is invaluable, for criminal organizations certainly, but also for groups subject to government abuses or businesses targeted by espionage.

In the modern era, the underlying drive for these opposed efforts (leaving aside ideology and the cases of passively hidden cohorts) is the implacable expansion of the byte trail.  Compared to past ages where the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including suspected cause of death -- the tools of the information era produce a near endless supply.  Cellular phones transmit constant location data, transactions with even the most remote subsidiary of a company leave trails in their logistics records, and of course any use of the internet produces veritable contrails of bits.  Their rate of production far exceeds the direct processing capability of any practically-sized team of analysts.

Hence, these teams employ computer-based, heuristic filtration to decide which data to record, to review, and to obtain.  We avoid saying ``algorithmic'' at this point, though these teams may themselves use that term.  ``Algorithmic'' implies a false certainty about patterns in and quality of the data associated with these analysis activities.

Given the real uncertainty, what these filters call for is testing and validation, but those present their own difficulties.  Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' is either non-existent or deceptive, and experiments could have dangerous side effects.  Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior or technology, even assuming the historical data are more than the victor's retelling.

Generating synthetic data seems like the obvious alternative.  It allows for comparison across both detection and masking strategies, consideration of multiple background contexts, forecasting of risks and tradeoffs in a way that allows uncertainties, and in general providing a framework for imaginative assessment.  Like all such flexible tools producing quantitative results, it has the subtle downsides of analyst biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable evidence for a particular time and situation -- the would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.

In the following sections, we layout the uses and abuses of such a framework.  What makes for useful synthetic data sets? What are the appropriate measures for detection strategies on them?  We motivate that discussion by inspiration from a simple, network-based model of terrorism -- a sub-group of the Salafi jihad networks as described by Sageman {\em et al.}\cite{sageman} -- and community organization and communication.  Whether or not that work is accurate description of that group and its associated events\todoCP{connective tissue missing here.} though we will point out where assumptions can be modified to identify different kinds of groups against a background population, since the tactics of these organizations are constantly evolving.

\section*{Framing a Covert Group Model}
To test strategies from either end, one must have a model capable of representing those strategies.  That means modeling entities that take action, modeling how that action is observed (including if it is observed correctly, or at all), and modeling how those observations are digested into reactions.  Notably, the entities must include some sort of background -- if the only data being simulated is to do with the covert entities, they are hardly covert within that simulation.  From here, we will focus on network based models of these components; networks seem like a natural tool, given the role of individually-based action, discrete events, and the relatively small number of participants in these groups.  Though we do not do so in our example, the background population might be more tractably modeled with continuous phenomena, given its large size and potentially more homogeneous behavior.

\subsection*{Modeling the Entities}
For our motivating example, we divide the population into three types, two of which belong to the covert group -- management and subordinates -- and a third representing ordinary individuals in the background population.  We choose this number of types because we are using that many models of activity, though different degree nodes will present somewhat differently.  The background population may be less homogenous in types, or perhaps types may be better modeled as being selected from distribution of features.  What must be guarded against here is over fitting by mechanism -- essentially the problem of choosing a polynomial of power equal to the available data, but more subtle.  Fishing with different mechanics may yield a better historical fit, but not necessarily a better forecast.

\subsubsection*{Background Population}
Most observable action will be that of the general populace surrounding the clandestine group.  This population has some structural component -- {\em e.g.} family sizes, typical numbers of working members, tendency towards assortativity -- though not necessarily well-known when a given investigation begins (and perhaps even assumed to be something it is not).  This structure may also be dynamic due to natural evolution (or activity on the structure may change pattern dynamically, those perspectives not being easy to disentangle), or possibly in response to the investigation.  For example, the ongoing revelations about the NSA will no doubt influence the behavior of the technical elite and percolate into the general public.  Therefore, assessment of any particular pair of opposed strategies should cut across multiple models of the background, each independently parametrized around what data is available.

As an example background population, we have the ordinary individuals form small groups, which in turn connect into larger groups, those groups into still larger groups, and so on until the background consists of single component.  If one were inclined to require that this description corresponded to a particular mechanism, this might loosely be interpreted as individuals forming households, households forming blocks, blocks forming neighborhoods, {\em ad nauseum}.  However, here it is only an academic fiction -- a compact, algorithmically and analytically convenient expression, without any connection to well established mechanics or data.  If demographic data for households were available, then we could plausibly parameterize the lowest level, then possibly combine that with mortality and mobility data to characterize how closely connected households remained, and so one.

Independently, we establish a second set of edges with a different flavor.  The previously described edges we label ``Familial'', these we call ``Economic''.  We will generate these in an identical fashion.  Again, if one were inclined to propose an explanation, one might call these small businesses or groups within a business, those forming collaborating businesses or whole firms, and so on hierarchically.  Again, we emphasize: this choice is purely an academic fiction, where we have added this extra fiction purely to highlight the need for multiple dimensions to represent different kinds of relationships in the population.

For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with some allowances to handle an arbitrary total population size).  That is:\begin{enumerate}
\item\label{grouping_init} divide the population $P_0$ into equal groups of size $n$, randomly assorting them;
\item for each group $i$, completely connect the individuals, and label that group $C_i^0$
\item\label{grouping_end} form a population from the $C_i^0$
\item repeat steps \ref{grouping_init} to \ref{grouping_end} with the $C_i^0$ connecting each edge between the $C_i^0$ to a uniformly drawn individual within the group, then with the $C_i^1$, etc until a single component is obtained
\end{enumerate}

\begin{code}[title=Pseudo-Code: Hierarchical Cliques]
// form a clique from a vertex collection, col:
def clique(col : Collection[V]) =
    for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
      left <~> right // form a bidirectional edge across the pair

def hierarchicalClique(col: Collection[V], size:Int) = 
   val thisLvl : Collection[Collection[V]] =
     // make the lowest level cliques 
     for ( subGroup <- groupBySize(col, size) ) yield {
       clique(subGroup)
       subGroup
     }
   cliqueGroups(thisLvl, size)

@recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
  val thisLvl : Collection[Collection[V]] =
    // gather the groups being made into a higher level cliques
    for ( subGroups <- groupBySize(col, size) ) yield {
      // clique a group of groups
      for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
        leftGroup.randomMember <~> rightGroup.randomMember
      }
      // merge this set of subgroups into a single new group
      merge(subGroups)
    }
  // repeat with the cliqued cliques, unless everything has been connected
  if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
\end{code}

Lastly for this model, we establish a final set of edges with a third flavor: ``Religious''.  These edges occur between members with a probability based the distance between the individuals on the ``Familial'' graph.  That is -- for those wanting to assign a meaning -- members of the same family are most likely to observably interact in a religious capacity, then immediate relatives or neighbors, and so on.  Of course, this is again only a convenient fiction, chosen to illustrate that other generation algorithms are possible, even with dependence between dimensions.  The detailed algorithm is:\begin{enumerate}
\item\label{religious_init} for each individual $i$:
\item assign $d_i=1$ and $F_i$ to the set of all their familial connections, excluding individuals already considered
\item with probability $p^{d_i}$ connect $i$ to the members of $F_i$
\item\label{religious_end} increment $d_i$, move all of $F_i$ to $P_i$, then add all of the familial connections of $P_i$ that are not in $P_i$ (or previously considered) to $F_i$.
\item repeat steps \ref{religious_init} to \ref{religious_end} until $F_i$ has no members added in step \ref{religious_end} 
\end{enumerate}

<<readInCliqued,echo=FALSE>>=
library(igraph)
pree <- read.table("../1-EL.txt",col.names=c("sender_id","recipient_id","type"))
pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
edgeinfo <- t(pree)
# hack: doesn't seem reasonable to have sim output other vertex labels
#edgeinfo[1,which(edgeinfo[1,] == 107)] <- 103
#edgeinfo[2,which(edgeinfo[2,] == 107)] <- 103
#edgeinfo[1,which(edgeinfo[1,] == 112)] <- 104
#edgeinfo[2,which(edgeinfo[2,] == 112)] <- 104

gWhole <- graph(edgeinfo[-3,])
V(gWhole)$size<-2
V(gWhole)[c(102,103,104)]$size<-6
E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
E(gWhole)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
E(gWhole)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
E(gWhole)[which(edgeinfo[3,]=="Plot")]$color <- "red"
#gAlt <- delete.vertices(gWhole, which(degree(gWhole) < 1)-1)
l<-layout.auto(gWhole)
@

\begin{rnwfig}
<<all, fig=TRUE, width=7.5, echo=FALSE>>=
plot(gWhole, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
\end{rnwfig}

\begin{rnwfig}
<<religion, fig=TRUE, width=7.5, echo=FALSE>>=
E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- NA
plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
\end{rnwfig}

\begin{rnwfig}
<<work, fig=TRUE, width=7.5, echo=FALSE>>=
E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- NA
plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{Now showing work (green) links.}
\end{rnwfig}

\begin{rnwfig}
<<family, fig=TRUE, width=7.5, echo=FALSE>>=
E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{Now showing family (blue) links.}
\end{rnwfig}

We also consider an alternative background: individuals form trees for ``Familial'' and ``Economic'' ties, and then the same treatment for ``Religious'' ties.  The trees have a similar $n$ parameter for branching, and form bidirectional links.  This is obviously less sophisticated than the aforementioned hierarchical structures.  Is it less plausible as well?  Certainly, it seems off relative to, say, a NATO country: no interaction loops in a particular social context seems unimaginable.  Perhaps this is more reasonable in a setting where there is less information technology penetration or less observable interaction.  Perhaps its more plausible if the trees instead had siblings form cliques as well.  All of that is supposition: we picked another academic fiction to compare with our previous one.

<<readInTree,echo=FALSE>>=
pree <- read.table("../1-EL-alt.txt",col.names=c("sender_id","recipient_id","type"))
pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
edgeinfo <- t(pree)

gWholeAlt <- graph(edgeinfo[-3,])
V(gWholeAlt)$size<-2
V(gWholeAlt)[c(102,103,104)]$size<-6
E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
E(gWholeAlt)[which(edgeinfo[3,]=="Plot")]$color <- "red"
l<-layout.auto(gWholeAlt)
@

\begin{rnwfig}
<<allAlt, fig=TRUE, width=7.5, echo=FALSE>>=
plot(gWholeAlt, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
\end{rnwfig}

\begin{rnwfig}
<<religionAlt, fig=TRUE, width=7.5, echo=FALSE>>=
E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- NA
plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
\end{rnwfig}

\begin{rnwfig}
<<workAlt, fig=TRUE, width=7.5, echo=FALSE>>=
E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- NA
plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{Now showing work (green) links.}
\end{rnwfig}

\begin{rnwfig}
<<familyAlt, fig=TRUE, width=7.5, echo=FALSE>>=
E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
@
\caption{Now showing family (blue) links.}
\end{rnwfig}

\subsubsection*{Covert Actors}
Sageman, Qin, et al. describe the structure of the Salafi networks as comprising a few key individuals with links to a large group of lieutenants -- the middle management of terror -- that are each connected to several tightly clustered subordinate groups -- terrorist cells -- that execute plots.  The lieutenants typically integrate with the regular population, while the subordinate groups are largely cloistered.

So we consider a covert organization consisting of two types, those directing a plot, {\em Management}, and those carrying out the day-to-day details, {\em Subordinates}.  For this demonstration model, we consider a ``small enough'' plot with a ``narrow enough'' schedule, such that only a single plot will occur during the simulation and that only a single manager is necessary to run that plot.

This manager, $M$, exists along side the background population.  Binomially sample the $C_i^0$ for ``Familial'' (with probability $p_F$) and ``Economic'' (with probability $p_E$) graphs and add $M$ to the selected $C_i^0$.  Then create ``Religious'' edges according to the same procedure used by ordinary individuals.

The subordinates, however, are isolated from the background population.  The manager and subordinates form a clique with a new type of edge: ``Plot''.

We also consider an alternative model: $M$ joins the background population as described above, but connects to the subordinate group via a hierarchical tree.

\subsection*{Modeling Action \&\ Observation}
Our proposed types have differences in their structural organization, but we also use those types to distinguish activity by those types on their related structure.  In this assessment, we represent activity only as monitorable communications, and those communications have their content flattened into two categories: ``Good'' and ``Bad''.  This is obviously a gross simplification of individual behavior (or over-estimation of analyst categorization capabilities); a potentially more appropriate version would be to have an abstract vocabulary with usage distinctions between the background and the clandestine group (e.g., uniform use in the background versus enriched in a subset in the clandestine group).  However, as we no doubt boringly emphasize: there is no particular basis for informing this model.  A time and group sensitive partitioning of intercepts for variety and distribution could plausibly form a basis for such a fit; one would have to consider, however, the distinction between the open source background communications (i.e., generally known to be public) versus the intercepted communications of the clandestine group (generally assumed private).

One last technical issue before proceeding to our example implementation: activity models may obviously differ in the data they generate, both structurally and semantically.  The underlying masking and investigating strategies may be framed relative to a particular action model, which may then require an adaptation of the activity data to be consistent with those adjacent models.

\subsubsection*{Background Action}
Each simulation time step, individuals in the background population generate messages by binomially sampling from all of their available connections.  They exhibit no preference for the type of those connections (beyond the structural consequence of having different numbers of different types).  If one interprets each message as a whole conversation, initiated by the sending party, then one implication of this model is that their messaging activity occupies an inconsequential period of real time relative to the real time equivalent of simulation step.  If one believed it was useful to model conversations explicitly -- i.e., each message is a word or phrase -- then one obvious change would be that individuals would only be selecting one target at a time, as well as modeling the speaker switch versus terminating communication.  This highlights a previously mentioned problem for the observation strategy -- there must exist an adapter for strategies between data types.  In this proposed continuous model, the aggregation for a strategy that only considers whole messages might be to average (on some specific real time scale) the total vocabulary in each direction of the conversation and then send one message from each party.

As for message content, the background population sends ``Good'' messages with a higher probability than ``Bad'' messages.

\subsubsection*{Covert Action}
The clandestine group's manager, $M$, behaves much like background population relative to his non-``Plot'' edges.  His tendency to send ``Bad'' messages to the background population is defined relative to the background probability.

However, he additionally sends direction to the subordinate group via the ``Plot'' connections with some probability.  Presumably, he would balance execution rate with secrecy.

The subordinates, however, do not interact with the background population.  They randomly speak with each other, with an enhanced probability of sending ``Bad'' messages after having received them from another member of the clandestine group.

\subsection*{Modeling Observation}
Incomplete information is the norm in these investigations, much like any work in the non-physical sciences.  However, it is typically possible to modulate what information is available (by investing more resources, by redirecting assets, etc).  We do not illustrate any strategies for either side in terms of modulating the flow of information, though an obvious addition dimension to add to both parties would be some resource pool that can be applied to modifying what information arrives at the observer (e.g., modulating the probability of detecting ``Bad'' messages from the covert group, faking messages in the background or to suspect members of the clandestine group).  For our simple model, we have different detection rates for the various edge flavors: ``Economic'' being highest, then ``Religious'', then ''Familial'', then ''Plot''.  The type of edge used to transmit the message is also not disclosed.

\subsection*{Modeling Reaction}
We rate investigating entities that implement a few different strategies.  One strategy is purely structural based on degrees, another purely content-oriented based on ``Bad'' frequency, and the third mixes the first two.

\subsubsection*{Structural Strategy}
Structural identification strategies range from having strong prior belief about a particular feature -- e.g., degree distribution -- and a relatively simple detection computation to belief only that there is a meaningful structural distinction (and a covert group to detect at all), thus running up against practical computational considerations trying to analyze all possible distinctions.

We demonstrate a case of the former, positing that the unique structural feature is to do with degree distribution -- specifically, a relatively high degree distribution for the manager, a relatively low degree distribution for the subordinates, and connection between them.  The criteria for labeling a member of the covert group is then a matter of setting what slices of the distribution to take from the top and bottom, and then testing for an observed path between the manager and the subordinate.

\subsubsection*{Content Strategy}
A pure content strategy ignores details about sender and receiver arrangement, instead focusing on the sending and receiving of different types of messages.  For our simple model, we consider a strategy that measures relative in and outflow of messages and the frequency of ``Bad'' messages.

\subsubsection*{Combination Strategy}
For our example combination strategy, we simply require individuals pass both threshold measures.

\section*{Evaluation}
There are two basic aspects of evaluation, which correspond to the general scientific method questions of model fitting versus model selection.  Note that these are entirely separate from detailed application of these aspects to particular model components; those activities are certainly critical in narrow assessments, but we must again emphasize the ability of both sides to adapt their strategies, improve technological capabilities, and so on, all of which will present disruptive changes to any established model.

The aspect which corresponds to questions of fit is basically identifying, for a particular model context -- specific combination of opposed strategies and background activity -- parameter surfaces for optimal performance on the chosen metrics.

The question of model selection corresponds more to considering these performance surfaces across a wide breadth of combinations.  That is, how well does a particular covert strategy perform across multiple background population behaviors and against varying investigatory capabilities?  Vice versa for the investigatory system?

For our toy systems we consider a simple performance metric, the Receiver Operator Characteristic discrimination statistic, across a background parameter sweep and as a function of time duration.

<<preRes,echo=FALSE>>=
structureFiles <- sort(list.files("..","structure_\\d+\\.txt"))
contentFiles <- sort(list.files("..","content_\\d+\\.txt"))
sandcFiles <- sort(list.files("..","sandc_\\d+\\.txt"))
badFiles <- sort(list.files("..","bad_\\d+\\.txt"))
astructureFiles <- sort(list.files("..","astructure_\\d+\\.txt"))
acontentFiles <- sort(list.files("..","acontent_\\d+\\.txt"))
asandcFiles <- sort(list.files("..","asandc_\\d+\\.txt"))
abadFiles <- sort(list.files("..","abad_\\d+\\.txt"))
popsize <- 100
plotsize <- 6

qData<-function(src) {
  FPR<-sapply(src,function(fname) {
    read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$FPR
  },simplify="matrix") / popsize
  FPRq <- apply(FPR,1,quantile)
  TPR<-sapply(src,function(fname) {
    read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$TPR
  },simplify="matrix") / plotsize
  TPRq <- apply(TPR,1,quantile)
  list(TPRq=TPRq,FPRq=FPRq)
}
sQData <- qData(structureFiles)
sFPRq <- sQData$FPRq
sTPRq <- sQData$TPRq
asQData <- qData(astructureFiles)
asFPRq <- asQData$FPRq
asTPRq <- asQData$TPRq

cQData <- qData(contentFiles)
cFPRq <- cQData$FPRq
cTPRq <- cQData$TPRq
acQData <- qData(acontentFiles)
acFPRq <- acQData$FPRq
acTPRq <- acQData$TPRq

sandcQData <- qData(sandcFiles)
sandcFPRq <- sandcQData$FPRq
sandcTPRq <- sandcQData$TPRq
asandcQData <- qData(asandcFiles)
asandcFPRq <- asandcQData$FPRq
asandcTPRq <- asandcQData$TPRq

badMR <- sapply(badFiles,function(fname) {
  read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
},simplify="matrix")
badMR <- badMR / max(badMR)
badMRq <- apply(badMR,1,quantile)

abadMR <- sapply(abadFiles,function(fname) {
  read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
},simplify="matrix")
abadMR <- abadMR / max(abadMR)
abadMRq <- apply(abadMR,1,quantile)

timesteps <- 100
samples <- 100
t<-1:timesteps
tm <- matrix(t,nrow=timesteps,ncol=samples)
backpoly<-function(t,x,color,lb=0) {
  polygon( c(min(t), t, max(t)), c( lb, x, lb), col=color, border=NA )
}
liner<-function(t,TPRq,FPRq,TPRxy,FPRxy) {
  lines(x=t,y=TPRq[3,],col="green3",type="l",xlab="iterate",ylab="FPR",ylim=c(0,1))
  lines(x=t,y=TPRq[4,],col="darkolivegreen1")
  lines(x=t,y=TPRq[2,],col="darkolivegreen1")
  lines(x=t,y=FPRq[3,],col="red")
  lines(x=t,y=FPRq[4,],col="coral")
  lines(x=t,y=FPRq[2,],col="coral")
  text(TPRxy[1],TPRxy[2],"TPR",col="green3")
  text(FPRxy[1],FPRxy[2],"FPR",col="red")
}
plotter<-function(t,badMRq,TPRq,FPRq,TPRxy,FPRxy) {
  plot(t,badMRq[5,],type="l",col="grey93",ylab="",xlab="",yaxt="n",xaxt="n",bty="n")
  backpoly(t,badMRq[5,],col="grey93")
  backpoly(t,badMRq[4,],col="grey85")
  backpoly(t,badMRq[2,],col="grey93")
  backpoly(t,badMRq[1,],col="white",lb=-1)
  axis(side=1,at=c(min(t),max(t)),labels=c(expression(t[min]),expression(t[max])),tick=F,line=-1)
  text(75,0.5,"MR")
  liner(t,TPRq,FPRq,TPRxy,FPRxy)
}
@
\subsection*{Structural Performance}
\begin{rnwfig}
<<structure, fig=TRUE, width=7.5, echo=FALSE>>=
plotter(t,badMRq,sTPRq,sFPRq,c(20,0.5),c(80,.15))
@
\caption{Structure-Only performance versus clique-based background.}
\end{rnwfig}

\begin{rnwfig}
<<astructure, fig=TRUE, width=7.5, echo=FALSE>>=
plotter(t,abadMRq,asTPRq,asFPRq,c(20,0.5),c(80,.15))
@
\caption{Structure-Only performance versus tree-based background.}
\end{rnwfig}

\begin{rnwfig}
<<content, fig=TRUE, width=7.5, echo=FALSE>>=
plotter(t,badMRq,cTPRq,cFPRq,c(20,0.5),c(80,.15))
@
\caption{Content-Only performance versus clique-based background.}
\end{rnwfig}

\begin{rnwfig}
<<acontent, fig=TRUE, width=7.5, echo=FALSE>>=
plotter(t,abadMRq,acTPRq,acFPRq,c(20,0.5),c(80,.15))
@
\caption{Content-Only performance versus tree-based background.}
\end{rnwfig}

\begin{rnwfig}
<<sandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
plotter(t,badMRq,sandcTPRq,sandcFPRq,c(20,0.5),c(80,.15))
@
\caption{Structure and Content performance versus clique-based background.}
\end{rnwfig}

\begin{rnwfig}
<<asandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
plotter(t,abadMRq,asandcTPRq,asandcFPRq,c(20,0.5),c(80,.15))
@
\caption{Structure and Content performance versus clique-based background.}
\end{rnwfig}

\section*{Discussion}
\todoCP{Get figures to give some direction for this.  Probably going to see the interesting stuff in the differently organized pops.}

\end{document}

### 2013-09-04 rscripts/manuscript.tex
manuscript components

--- 2013-09-04 rscripts/manuscript.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{manuscript-concordance}
 \begin{flushleft}
@@
 
-@recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
   val thisLvl : Collection[Collection[V]] =
@@
 
-<<readInCliqued,echo=FALSE>>=
-library(igraph)
-pree <- read.table("../1-EL.txt",col.names=c("sender_id","recipient_id","type"))
-pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
-pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
-edgeinfo <- t(pree)
-# hack: doesn't seem reasonable to have sim output other vertex labels
-#edgeinfo[1,which(edgeinfo[1,] == 107)] <- 103
-#edgeinfo[2,which(edgeinfo[2,] == 107)] <- 103
-#edgeinfo[1,which(edgeinfo[1,] == 112)] <- 104
-#edgeinfo[2,which(edgeinfo[2,] == 112)] <- 104
-
-gWhole <- graph(edgeinfo[-3,])
-V(gWhole)$size<-2
-V(gWhole)[c(102,103,104)]$size<-6
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWhole)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
-E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWhole)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
-E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
-E(gWhole)[which(edgeinfo[3,]=="Plot")]$color <- "red"
-#gAlt <- delete.vertices(gWhole, which(degree(gWhole) < 1)-1)
-l<-layout.auto(gWhole)
-@
-
 \begin{rnwfig}
-<<all, fig=TRUE, width=7.5, echo=FALSE>>=
-plot(gWhole, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-all}
 \caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
@@
 \begin{rnwfig}
-<<religion, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
-E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- NA
-plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-religion}
 \caption{The same graph, showing the plot connections (red) and religious (purple) links.}
@@
 \begin{rnwfig}
-<<work, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- NA
-plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-work}
 \caption{Now showing work (green) links.}
@@
 \begin{rnwfig}
-<<family, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
-plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-family}
 \caption{Now showing family (blue) links.}
@@
 
-<<readInTree,echo=FALSE>>=
-pree <- read.table("../1-EL-alt.txt",col.names=c("sender_id","recipient_id","type"))
-pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
-pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
-edgeinfo <- t(pree)
-
-gWholeAlt <- graph(edgeinfo[-3,])
-V(gWholeAlt)$size<-2
-V(gWholeAlt)[c(102,103,104)]$size<-6
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
-E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
-E(gWholeAlt)[which(edgeinfo[3,]=="Plot")]$color <- "red"
-l<-layout.auto(gWholeAlt)
-@
-
 \begin{rnwfig}
-<<allAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-plot(gWholeAlt, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-allAlt}
 \caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
@@
 \begin{rnwfig}
-<<religionAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- NA
-plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-religionAlt}
 \caption{The same graph, showing the plot connections (red) and religious (purple) links.}
@@
 \begin{rnwfig}
-<<workAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- NA
-plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-workAlt}
 \caption{Now showing work (green) links.}
@@
 \begin{rnwfig}
-<<familyAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
-plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
+\includegraphics{manuscript-familyAlt}
 \caption{Now showing family (blue) links.}
@@
 
-<<preRes,echo=FALSE>>=
-structureFiles <- sort(list.files("..","structure_\\d+\\.txt"))
-contentFiles <- sort(list.files("..","content_\\d+\\.txt"))
-sandcFiles <- sort(list.files("..","sandc_\\d+\\.txt"))
-badFiles <- sort(list.files("..","bad_\\d+\\.txt"))
-astructureFiles <- sort(list.files("..","astructure_\\d+\\.txt"))
-acontentFiles <- sort(list.files("..","acontent_\\d+\\.txt"))
-asandcFiles <- sort(list.files("..","asandc_\\d+\\.txt"))
-abadFiles <- sort(list.files("..","abad_\\d+\\.txt"))
-popsize <- 100
-plotsize <- 6
-
-qData<-function(src) {
-  FPR<-sapply(src,function(fname) {
-    read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$FPR
-  },simplify="matrix") / popsize
-  FPRq <- apply(FPR,1,quantile)
-  TPR<-sapply(src,function(fname) {
-    read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$TPR
-  },simplify="matrix") / plotsize
-  TPRq <- apply(TPR,1,quantile)
-  list(TPRq=TPRq,FPRq=FPRq)
-}
-sQData <- qData(structureFiles)
-sFPRq <- sQData$FPRq
-sTPRq <- sQData$TPRq
-asQData <- qData(astructureFiles)
-asFPRq <- asQData$FPRq
-asTPRq <- asQData$TPRq
-
-cQData <- qData(contentFiles)
-cFPRq <- cQData$FPRq
-cTPRq <- cQData$TPRq
-acQData <- qData(acontentFiles)
-acFPRq <- acQData$FPRq
-acTPRq <- acQData$TPRq
-
-sandcQData <- qData(sandcFiles)
-sandcFPRq <- sandcQData$FPRq
-sandcTPRq <- sandcQData$TPRq
-asandcQData <- qData(asandcFiles)
-asandcFPRq <- asandcQData$FPRq
-asandcTPRq <- asandcQData$TPRq
-
-badMR <- sapply(badFiles,function(fname) {
-  read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
-},simplify="matrix")
-badMR <- badMR / max(badMR)
-badMRq <- apply(badMR,1,quantile)
-
-abadMR <- sapply(abadFiles,function(fname) {
-  read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
-},simplify="matrix")
-abadMR <- abadMR / max(abadMR)
-abadMRq <- apply(abadMR,1,quantile)
-
-timesteps <- 100
-samples <- 100
-t<-1:timesteps
-tm <- matrix(t,nrow=timesteps,ncol=samples)
-backpoly<-function(t,x,color,lb=0) {
-  polygon( c(min(t), t, max(t)), c( lb, x, lb), col=color, border=NA )
-}
-liner<-function(t,TPRq,FPRq,TPRxy,FPRxy) {
-  lines(x=t,y=TPRq[3,],col="green3",type="l",xlab="iterate",ylab="FPR",ylim=c(0,1))
-  lines(x=t,y=TPRq[4,],col="darkolivegreen1")
-  lines(x=t,y=TPRq[2,],col="darkolivegreen1")
-  lines(x=t,y=FPRq[3,],col="red")
-  lines(x=t,y=FPRq[4,],col="coral")
-  lines(x=t,y=FPRq[2,],col="coral")
-  text(TPRxy[1],TPRxy[2],"TPR",col="green3")
-  text(FPRxy[1],FPRxy[2],"FPR",col="red")
-}
-plotter<-function(t,badMRq,TPRq,FPRq,TPRxy,FPRxy) {
-  plot(t,badMRq[5,],type="l",col="grey93",ylab="",xlab="",yaxt="n",xaxt="n",bty="n")
-  backpoly(t,badMRq[5,],col="grey93")
-  backpoly(t,badMRq[4,],col="grey85")
-  backpoly(t,badMRq[2,],col="grey93")
-  backpoly(t,badMRq[1,],col="white",lb=-1)
-  axis(side=1,at=c(min(t),max(t)),labels=c(expression(t[min]),expression(t[max])),tick=F,line=-1)
-  text(75,0.5,"MR")
-  liner(t,TPRq,FPRq,TPRxy,FPRxy)
-}
-@
 \subsection*{Structural Performance}
 \begin{rnwfig}
-<<structure, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,badMRq,sTPRq,sFPRq,c(20,0.5),c(80,.15))
-@
+\includegraphics{manuscript-structure}
 \caption{Structure-Only performance versus clique-based background.}
@@
 \begin{rnwfig}
-<<astructure, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,abadMRq,asTPRq,asFPRq,c(20,0.5),c(80,.15))
-@
+\includegraphics{manuscript-astructure}
 \caption{Structure-Only performance versus tree-based background.}
@@
 \begin{rnwfig}
-<<content, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,badMRq,cTPRq,cFPRq,c(20,0.5),c(80,.15))
-@
+\includegraphics{manuscript-content}
 \caption{Content-Only performance versus clique-based background.}
@@
 \begin{rnwfig}
-<<acontent, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,abadMRq,acTPRq,acFPRq,c(20,0.5),c(80,.15))
-@
+\includegraphics{manuscript-acontent}
 \caption{Content-Only performance versus tree-based background.}
@@
 \begin{rnwfig}
-<<sandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,badMRq,sandcTPRq,sandcFPRq,c(20,0.5),c(80,.15))
-@
+\includegraphics{manuscript-sandcontent}
 \caption{Structure and Content performance versus clique-based background.}
@@
 \begin{rnwfig}
-<<asandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,abadMRq,asandcTPRq,asandcFPRq,c(20,0.5),c(80,.15))
-@
+\includegraphics{manuscript-asandcontent}
 \caption{Structure and Content performance versus clique-based background.}

### 2013-09-04 rscripts/manuscript.Rnw
draft updates

@@
 We outline the problem of characterizing strategies for detection and concealment of clandestine coordination within a broader population, in terms of network models.
-Specifically, we propose means to accommodate the uncertainty about behavior and capability from both ends, a general plan for developing such models, some best practices for model parameterization and data gathering, and finally some particularly pernicious pitfalls.
+Specifically, we present (i) means to accommodate the uncertainty about behavior and capability of both covert groups and survillience efforts, (ii) a general outline for developing models to describe to describe those entities and background populations, (iii) some best practices for model parameterization and data gathering, and (iv) finally some particularly pernicious pitfalls.
 
@@
 We discuss measuring performance of the opposing sides (e.g.
-Receiver Operator Characteristic), fitting the model against real data, and finally how this model can be extended.
+Receiver Operator Characteristic), fitting the model against to real data, and finally indiciating how this model can be extended.
 
@@
 In the modern era, the underlying drive for these opposed efforts (leaving aside ideology and the cases of passively hidden cohorts) is the implacable expansion of the byte trail.
-Compared to past ages where the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including suspected cause of death -- the tools of the information era produce a near endless supply.
-Cellular phones transmit constant location data, transactions with even the most remote subsidiary of a company leave trails in their logistics records, and of course any use of the internet produces veritable contrails of bits.
-Their rate of production far exceeds the direct processing capability of any practically-sized team of analysts.
+Compared to past ages where the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including morbidity\todoCP{the right word?} -- the tools of the information era produce a near endless supply.
+Cellular phones transmit constant location data, transactions with even the most remote subsidiary of a company leave trails in their logistics records, and of course any use of the internet produces veritable bit contrails.
+This rate of production far exceeds the direct processing capability of any practically-sized team of analysts.
 
@@
 Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' is either non-existent or deceptive, and experiments could have dangerous side effects.
-Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior or technology, even assuming the historical data are more than the victor's retelling.
+Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior or technological innovation, even assuming the historical data are more than victor's embellishment.
 
-Generating synthetic data seems like the obvious alternative.
+Generating synthetic data seems like an obvious alternative.
 It allows for comparison across both detection and masking strategies, consideration of multiple background contexts, forecasting of risks and tradeoffs in a way that allows uncertainties, and in general providing a framework for imaginative assessment.
-Like all such flexible tools producing quantitative results, it has the subtle downsides of analyst biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable evidence for a particular time and situation -- the would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.
+Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable evidence for a particular time and situation -- the would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.
 
-In the following sections, we layout the uses and abuses of such a framework.
+In the following sections, we lay out the uses and abuses of such a framework.
 What makes for useful synthetic data sets?
@@
 We motivate that discussion by inspiration from a simple, network-based model of terrorism -- a sub-group of the Salafi jihad networks as described by Sageman {\em et al.}\cite{sageman} -- and community organization and communication.
-Whether or not that work is accurate description of that group and its associated events\todoCP{connective tissue missing here.} though we will point out where assumptions can be modified to identify different kinds of groups against a background population, since the tactics of these organizations are constantly evolving.
+Whether or not that work is an accurate description is not of particular concern.
+Their qualitative properties are enough of a defensible testbed for modeling the communication patterns of one covert group.
+We will point out where assumptions can be modified to identify different kinds of groups against a background population, since the behavior and structure of both the background and covert organizations are constantly evolving.
 
 \section*{Framing a Covert Group Model}
-To test strategies from either end, one must have a model capable of representing those strategies.
+To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
 That means modeling entities that take action, modeling how that action is observed (including if it is observed correctly, or at all), and modeling how those observations are digested into reactions.
 Notably, the entities must include some sort of background -- if the only data being simulated is to do with the covert entities, they are hardly covert within that simulation.
-From here, we will focus on network based models of these components; networks seem like a natural tool, given the role of individually-based action, discrete events, and the relatively small number of participants in these groups.
+Here, we will focus on network based models of these components.
+Networks seem like a natural tool, given the role of individually-based action, discrete events, and the relatively small number of participants in these groups.
 Though we do not do so in our example, the background population might be more tractably modeled with continuous phenomena, given its large size and potentially more homogeneous behavior.
@@
 Most observable action will be that of the general populace surrounding the clandestine group.
-This population has some structural component -- {\em e.g.} family sizes, typical numbers of working members, tendency towards assortativity -- though not necessarily well-known when a given investigation begins (and perhaps even assumed to be something it is not).
-This structure may also be dynamic due to natural evolution (or activity on the structure may change pattern dynamically, those perspectives not being easy to disentangle), or possibly in response to the investigation.
+This population has some structural component -- {\em e.g.} family sizes, typical numbers of working members, tendency towards assortativity -- though not necessarily well-known when a given investigation begins and perhaps even assumed to be something it is not.
+This structure is also likely dynamic due to natural evolution (or activity on the structure may change pattern dynamically, those perspectives not being easy to disentangle), or possibly in response to the investigation.
 For example, the ongoing revelations about the NSA will no doubt influence the behavior of the technical elite and percolate into the general public.
@@
 
-As an example background population, we have the ordinary individuals form small groups, which in turn connect into larger groups, those groups into still larger groups, and so on until the background consists of single component.
+As an example background population, we have the ordinary individuals form small groups, which in turn connect into larger groups, those groups into still larger groups, and so on until the background consists of a single component.
 If one were inclined to require that this description corresponded to a particular mechanism, this might loosely be interpreted as individuals forming households, households forming blocks, blocks forming neighborhoods, {\em ad nauseum}.
-However, here it is only an academic fiction -- a compact, algorithmically and analytically convenient expression, without any connection to well established mechanics or data.
-If demographic data for households were available, then we could plausibly parameterize the lowest level, then possibly combine that with mortality and mobility data to characterize how closely connected households remained, and so one.
+However, here it is only an academic fiction -- a compact, algorithmically and analytically convenient expression, without any connection to well-established mechanics or data.
+If demographic data for households were available, then we could plausibly parameterize the lowest level, then possibly combine that with mortality and mobility data to characterize how closely connected households remained, and so on.
 
@@
 
-For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with some allowances to handle an arbitrary total population size).  That is:\begin{enumerate}
-\item\label{grouping_init} divide the population $P_0$ into equal groups of size $n$, randomly assorting them;
+For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with allowance to handle an arbitrary total population size).  That is:\begin{enumerate}
+\item create and randomly permute a population $P_0$, with $N$ members (later results use $N=100$).
+\item\label{grouping_init} divide $P_0$ into equal groups of size $n$ (later results use $n=3$);
 \item for each group $i$, completely connect the individuals, and label that group $C_i^0$
@@
 That is -- for those wanting to assign a meaning -- members of the same family are most likely to observably interact in a religious capacity, then immediate relatives or neighbors, and so on.
-Of course, this is again only a convenient fiction, chosen to illustrate that other generation algorithms are possible, even with dependence between dimensions.
+Of course, this is again academic, chosen to illustrate that other generation algorithms are possible, even with dependence between dimensions.
 The detailed algorithm is:\begin{enumerate}
@@
 
-We also consider an alternative background: individuals form trees for ``Familial'' and ``Economic'' ties, and then the same treatment for ``Religious'' ties.
+We also consider an alternative background: instead of cliques for ``Familial'' and ``Economic'' ties, individuals form trees.
+The treatment for ``Religious'' ties remains the same.
 The trees have a similar $n$ parameter for branching, and form bidirectional links.
@@
 Perhaps its more plausible if the trees instead had siblings form cliques as well.
-All of that is supposition: we picked another academic fiction to compare with our previous one.
+None of that supposition is particularly relevant: we picked another academic fiction to compare with our previous one.
 

### 2014-01-03 rscripts/manuscript.Rnw
updates to abstract

@@
 {\Large
-\textbf{Detecting Covert Groups Embedded in a Population}
+\textbf{Modeling Requirements for Simulating Covert Social Groups}
 }
@@
 \section*{Abstract}
-We outline the problem of characterizing strategies for detection and concealment of clandestine coordination within a broader population, in terms of network models.
-Specifically, we present (i) means to accommodate the uncertainty about behavior and capability of both covert groups and survillience efforts, (ii) a general outline for developing models to describe to describe those entities and background populations, (iii) some best practices for model parameterization and data gathering, and (iv) finally some particularly pernicious pitfalls.
+Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.
+Moreover, increasing active observation would be expected to perturb these social systems.
+However, these groups never have perfect cover, their secrecy may make them conspicuously absent from the everyday activity of the background population, and finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc.
+-- that might predict their actions.
 
-As a practical, but mostly pedagogical demonstration, we specify a graph-based model of simple communication across a procedurally generated population, with an embedded, relatively small module representing a clandestine group, pitted against surveillance systems.
-We discuss measuring performance of the opposing sides (e.g.
-Receiver Operator Characteristic), fitting the model against to real data, and finally indiciating how this model can be extended.
+This situation -- strong constraints on passively and actively obtained direct empirical data, overwhelming indirect data, and competing mechanical explanations -- recommends simulation.
+However, because the of the censoring issues in the data, that simulation requires particular attention to characterizing uncertainty.
+
+Herein, we discuss these issues in terms of network models.
+Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) finally some philosophy relative to implementation.
+
+As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.
+We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g.
+Receiver Operator Characteristic).
+Finally, we also demonstrate a small extension.
 

### 2014-01-03 rscripts/manuscript.tex
updates to abstract

@@
 \end{enumerate}
-
-\begin{rnwfig}
-\includegraphics{manuscript-all}
-\caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-religion}
-\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-work}
-\caption{Now showing work (green) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-family}
-\caption{Now showing family (blue) links.}
-\end{rnwfig}
-
-We also consider an alternative background: instead of cliques for ``Familial'' and ``Economic'' ties, individuals form trees.
-The treatment for ``Religious'' ties remains the same.
-The trees have a similar $n$ parameter for branching, and form bidirectional links.
-This is obviously less sophisticated than the aforementioned hierarchical structures.
-Is it less plausible as well?
-Certainly, it seems off relative to, say, a NATO country: no interaction loops in a particular social context seems unimaginable.
-Perhaps this is more reasonable in a setting where there is less information technology penetration or less observable interaction.
-Perhaps its more plausible if the trees instead had siblings form cliques as well.
-None of that supposition is particularly relevant: we picked another academic fiction to compare with our previous one.
-
-\begin{rnwfig}
-\includegraphics{manuscript-allAlt}
-\caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-religionAlt}
-\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-workAlt}
-\caption{Now showing work (green) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-familyAlt}
-\caption{Now showing family (blue) links.}
-\end{rnwfig}
-
-\subsubsection*{Covert Actors}
-Sageman, Qin, et al.
-describe the structure of the Salafi networks as comprising a few key individuals with links to a large group of lieutenants -- the middle management of terror -- that are each connected to several tightly clustered subordinate groups -- terrorist cells -- that execute plots.
-The lieutenants typically integrate with the regular population, while the subordinate groups are largely cloistered.
-
-So we consider a covert organization consisting of two types, those directing a plot, {\em Management}, and those carrying out the day-to-day details, {\em Subordinates}.
-For this demonstration model, we consider a ``small enough'' plot with a ``narrow enough'' schedule, such that only a single plot will occur during the simulation and that only a single manager is necessary to run that plot.
-
-This manager, $M$, exists along side the background population.
-Binomially sample the $C_i^0$ for ``Familial'' (with probability $p_F$) and ``Economic'' (with probability $p_E$) graphs and add $M$ to the selected $C_i^0$.
-Then create ``Religious'' edges according to the same procedure used by ordinary individuals.
-
-The subordinates, however, are isolated from the background population.  The manager and subordinates form a clique with a new type of edge: ``Plot''.
-
-We also consider an alternative model: $M$ joins the background population as described above, but connects to the subordinate group via a hierarchical tree.
-
-\subsection*{Modeling Action \&\ Observation}
-Our proposed types have differences in their structural organization, but we also use those types to distinguish activity by those types on their related structure.
-In this assessment, we represent activity only as monitorable communications, and those communications have their content flattened into two categories: ``Good'' and ``Bad''.
-This is obviously a gross simplification of individual behavior (or over-estimation of analyst categorization capabilities); a potentially more appropriate version would be to have an abstract vocabulary with usage distinctions between the background and the clandestine group (e.g., uniform use in the background versus enriched in a subset in the clandestine group).
-However, as we no doubt boringly emphasize: there is no particular basis for informing this model.
-A time and group sensitive partitioning of intercepts for variety and distribution could plausibly form a basis for such a fit; one would have to consider, however, the distinction between the open source background communications (i.e., generally known to be public) versus the intercepted communications of the clandestine group (generally assumed private).
-
-One last technical issue before proceeding to our example implementation: activity models may obviously differ in the data they generate, both structurally and semantically.
-The underlying masking and investigating strategies may be framed relative to a particular action model, which may then require an adaptation of the activity data to be consistent with those adjacent models.
-
-\subsubsection*{Background Action}
-Each simulation time step, individuals in the background population generate messages by binomially sampling from all of their available connections.
-They exhibit no preference for the type of those connections (beyond the structural consequence of having different numbers of different types).
-If one interprets each message as a whole conversation, initiated by the sending party, then one implication of this model is that their messaging activity occupies an inconsequential period of real time relative to the real time equivalent of simulation step.
-If one believed it was useful to model conversations explicitly -- i.e., each message is a word or phrase -- then one obvious change would be that individuals would only be selecting one target at a time, as well as modeling the speaker switch versus terminating communication.
-This highlights a previously mentioned problem for the observation strategy -- there must exist an adapter for strategies between data types.
-In this proposed continuous model, the aggregation for a strategy that only considers whole messages might be to average (on some specific real time scale) the total vocabulary in each direction of the conversation and then send one message from each party.
-
-As for message content, the background population sends ``Good'' messages with a higher probability than ``Bad'' messages.
-
-\subsubsection*{Covert Action}
-The clandestine group's manager, $M$, behaves much like background population relative to his non-``Plot'' edges.
-His tendency to send ``Bad'' messages to the background population is defined relative to the background probability.
-
-However, he additionally sends direction to the subordinate group via the ``Plot'' connections with some probability.  Presumably, he would balance execution rate with secrecy.
-
-The subordinates, however, do not interact with the background population.
-They randomly speak with each other, with an enhanced probability of sending ``Bad'' messages after having received them from another member of the clandestine group.
-
-\subsection*{Modeling Observation}
-Incomplete information is the norm in these investigations, much like any work in the non-physical sciences.
-However, it is typically possible to modulate what information is available (by investing more resources, by redirecting assets, etc).
-We do not illustrate any strategies for either side in terms of modulating the flow of information, though an obvious addition dimension to add to both parties would be some resource pool that can be applied to modifying what information arrives at the observer (e.g., modulating the probability of detecting ``Bad'' messages from the covert group, faking messages in the background or to suspect members of the clandestine group).
-For our simple model, we have different detection rates for the various edge flavors: ``Economic'' being highest, then ``Religious'', then ''Familial'', then ''Plot''.
-The type of edge used to transmit the message is also not disclosed.
-
-\subsection*{Modeling Reaction}
-We rate investigating entities that implement a few different strategies.
-One strategy is purely structural based on degrees, another purely content-oriented based on ``Bad'' frequency, and the third mixes the first two.
-
-\subsubsection*{Structural Strategy}
-Structural identification strategies range from having strong prior belief about a particular feature -- e.g., degree distribution -- and a relatively simple detection computation to belief only that there is a meaningful structural distinction (and a covert group to detect at all), thus running up against practical computational considerations trying to analyze all possible distinctions.
-
-We demonstrate a case of the former, positing that the unique structural feature is to do with degree distribution -- specifically, a relatively high degree distribution for the manager, a relatively low degree distribution for the subordinates, and connection between them.
-The criteria for labeling a member of the covert group is then a matter of setting what slices of the distribution to take from the top and bottom, and then testing for an observed path between the manager and the subordinate.
-
-\subsubsection*{Content Strategy}
-A pure content strategy ignores details about sender and receiver arrangement, instead focusing on the sending and receiving of different types of messages.
-For our simple model, we consider a strategy that measures relative in and outflow of messages and the frequency of ``Bad'' messages.
-
-\subsubsection*{Combination Strategy}
-For our example combination strategy, we simply require individuals pass both threshold measures.
-
-\section*{Evaluation}
-There are two basic aspects of evaluation, which correspond to the general scientific method questions of model fitting versus model selection.
-Note that these are entirely separate from detailed application of these aspects to particular model components; those activities are certainly critical in narrow assessments, but we must again emphasize the ability of both sides to adapt their strategies, improve technological capabilities, and so on, all of which will present disruptive changes to any established model.
-
-The aspect which corresponds to questions of fit is basically identifying, for a particular model context -- specific combination of opposed strategies and background activity -- parameter surfaces for optimal performance on the chosen metrics.
-
-The question of model selection corresponds more to considering these performance surfaces across a wide breadth of combinations.
-That is, how well does a particular covert strategy perform across multiple background population behaviors and against varying investigatory capabilities?
-Vice versa for the investigatory system?
-
-For our toy systems we consider a simple performance metric, the Receiver Operator Characteristic discrimination statistic, across a background parameter sweep and as a function of time duration.
-
-\subsection*{Structural Performance}
-\begin{rnwfig}
-\includegraphics{manuscript-structure}
-\caption{Structure-Only performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-astructure}
-\caption{Structure-Only performance versus tree-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-content}
-\caption{Content-Only performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-acontent}
-\caption{Content-Only performance versus tree-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-sandcontent}
-\caption{Structure and Content performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-asandcontent}
-\caption{Structure and Content performance versus clique-based background.}
-\end{rnwfig}
-
-\section*{Discussion}
-\todoCP{Get figures to give some direction for this.  Probably going to see the interesting stuff in the differently organized pops.}
-
-\end{document}

### 2014-01-13 rscripts/manuscript.Rnw
updates in the opening sections

@@
 This situation -- strong constraints on passively and actively obtained direct empirical data, overwhelming indirect data, and competing mechanical explanations -- recommends simulation.
-However, because the of the censoring issues in the data, that simulation requires particular attention to characterizing uncertainty.
+However, because the of the censoring issues in the data, that simulation requires particular attention to characterizing uncertainty and avoiding overfitting.
 
@@
 \section*{Introduction}
-For investigators ranging from anthropologists to law enforcement, the need to identify groups operating in secret -- through deliberate action or otherwise -- is paramount.
-In particular, the need for intelligence organizations to identify terrorist cells and defuse their violent plots is a matter of increasing import.
-Symmetrically, being able to operate clandestinely in an age of ubiquitous monitoring is invaluable, for criminal organizations certainly, but also for groups subject to government abuses or businesses targeted by espionage.
+For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior groups operating at the edge of the observable is a principle concern.
+In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.
+Of course that revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
+Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.
 
-In the modern era, the underlying drive for these opposed efforts (leaving aside ideology and the cases of passively hidden cohorts) is the implacable expansion of the byte trail.
-Compared to past ages where the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including morbidity\todoCP{the right word?} -- the tools of the information era produce a near endless supply.
-Cellular phones transmit constant location data, transactions with even the most remote subsidiary of a company leave trails in their logistics records, and of course any use of the internet produces veritable bit contrails.
-This rate of production far exceeds the direct processing capability of any practically-sized team of analysts.
+The underlying drive for these opposed efforts is the implacably expanding byte trail.
+Once, the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including morbidity -- but now information era tools produce a near endless supply or 1s and 0s.
+In some cases, this production is a side effect of organized commercial instinct -- monetizing search through targetted advertisement, e.g.
+In other cases, individuals are voluntarily advancing their own brand -- Instagram ``selfies'', e.g.
+However, in many cases this bitwise production is an engineering consequence of the associated technology.
+E.g., cellular phones transmit constant location data, financial systems are increasingly just data-based, and of course any use of the internet produces veritable bit contrails.
+This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.
 
 Hence, these teams employ computer-based, heuristic filtration to decide which data to record, to review, and to obtain.
-We avoid saying ``algorithmic'' at this point, though these teams may themselves use that term.
-``Algorithmic'' implies a false certainty about patterns in and quality of the data associated with these analysis activities.
+We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.
+Like ``algorithmic'' bubble prediction, we tend to side more with Fama than Shiller.
 
@@
 Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' is either non-existent or deceptive, and experiments could have dangerous side effects.
-Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior or technological innovation, even assuming the historical data are more than victor's embellishment.
+Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior and technological innovation (without even further abstraction), even assuming the historical data are more than victor's embellishment.
 
@@
 It allows for comparison across both detection and masking strategies, consideration of multiple background contexts, forecasting of risks and tradeoffs in a way that allows uncertainties, and in general providing a framework for imaginative assessment.
-Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable evidence for a particular time and situation -- the would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.
+Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable agreement for a particular time and situation -- there would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.
 
@@
 What are the appropriate measures for detection strategies on them?
-We motivate that discussion by inspiration from a simple, network-based model of terrorism -- a sub-group of the Salafi jihad networks as described by Sageman {\em et al.}\cite{sageman} -- and community organization and communication.
+We motivate that discussion by inspiration from an over-simplified, network-based model of terrorism -- a sub-group of the Salafi jihad networks as described by Sageman {\em et al.}\cite{sageman} -- and community organization and communication.
 Whether or not that work is an accurate description is not of particular concern.
@@
 
-\section*{Framing a Covert Group Model}
-To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
+\section*{Overview of the Simulation Problem}
+Detecting a covert group is fundamentally about distinguishing the trace observable activity of that group from the activity of the background, bootstrapping that into guesses about the structure of the group, focusing the observation and distinguishing process based on that structure, and so on iteratively.
+The purpose of simulation is ultimately to measure performance in the observer-covert group competition.
+
+Given that these aspects must be expressed in the model (with varying degrees of detail), and that this model is used by a team or communicated between researchers, the language for that expression must be both powerful (to capture complexity where necessary), but also comprehensible to communicate what in fact is the meaning of the universe in this model.
+Because these simulations are fundamentally about individuals and their behavior, we largely frame our discussion in terms of individuals, their internal state, and their interaction with others.
+
+We can of course think in different terms.
+For example, we are considering a network model of the population and covert group embedded within it, with edges representing relationships between people.
+We might think of those relationships as being external state and correspondingly represent those relationships as external state in the simulation.
+E.g., we might represent the population as a graph rather than a bag of individuals, and in the simulation the edges belong to the graph object rather than individuals.
+As long as we recognize the translation between these perspectives, the choice boils down to what is most pragmatic relative to the science and simulation.
+If most of the questions answered by the simulation relate to graph measures -- e.g., calculating various centrality measures -- then the simulation-as-graph rather than simulation-as-individuals probably makes more sense.
+
+\subsection*{Modeling the Covert Groups}
+Relative to the covert group, the simulation must model how the group members act and interact.
+Typically, that further entails a modeling few moving pieces: (i) the members internal state (including what other members they know and communicate with), (ii) how that internal state evolves relative to external forces or states, and (iii) what actions those members undertake based on their internal state (possibly in response to some specific external event).
+
+\subsection*{Modeling the Background}
+Like the covert group, the background population is a (larger) group of individuals that act and interact.
+They have the same basic modeling requirements as the covert group, but they behave differently.
+This difference may be as simple as being distinct relative to observation process -- e.g., their communications are more likely to be recorded -- or it may be more noticeable like having fundamentally more diverse contacts, taking different kinds of actions, etc.
+
+An alternative representation for the background would be to model it as a continuous entity vice individuals in a network.
+This could be used to reduce the simulation size.
+However, this model would still need to create activity data that was consistent (in format) to that created by the covert group.
+
+\subsection*{Modeling Activity}
+Individuals create modeled activity based on their internal state.
+These actions correspond to (i) the observable events of specific interest (e.g., calls, financial transactions), (ii) generally observable events (e.g., travel, work, religious observations), and (iii) interactions that, while not externally observable, play a role in the evolution of internal state in other individuals either directly or by modifying the state of the world.
+These actions should occur at specific times in the simulation, at whatever resolution is appropriate to the model.
+
+The covert group and background population should be different in their activity; either performing different activities or undertaking them at different rates or on different schedules.
+
+\subsection*{The Observation Model}
+
+%To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
 That means modeling entities that take action, modeling how that action is observed (including if it is observed correctly, or at all), and modeling how those observations are digested into reactions.
@@
 
-\subsection*{Modeling the Entities}
-For our motivating example, we divide the population into three types, two of which belong to the covert group -- management and subordinates -- and a third representing ordinary individuals in the background population.
+%\subsection*{Modeling the Entities}
+%For our motivating example, we divide the population into three types, two of which belong to the covert group -- management and subordinates -- and a third representing ordinary individuals in the background population.
 We choose this number of types because we are using that many models of activity, though different degree nodes will present somewhat differently.
@@
 
-\subsubsection*{Background Population}
-Most observable action will be that of the general populace surrounding the clandestine group.
+%\subsubsection*{Background Population}
+%Most observable action will be that of the general populace surrounding the clandestine group.
 This population has some structural component -- {\em e.g.} family sizes, typical numbers of working members, tendency towards assortativity -- though not necessarily well-known when a given investigation begins and perhaps even assumed to be something it is not.

### 2014-01-13 rscripts/manuscript.tex
updates in the opening sections

--- 2013-09-04 rscripts/manuscript.tex
@@
 \section*{Abstract}
-We outline the problem of characterizing strategies for detection and concealment of clandestine coordination within a broader population, in terms of network models.
-Specifically, we present (i) means to accommodate the uncertainty about behavior and capability of both covert groups and survillience efforts, (ii) a general outline for developing models to describe to describe those entities and background populations, (iii) some best practices for model parameterization and data gathering, and (iv) finally some particularly pernicious pitfalls.
+Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.
+Moreover, increasing active observation would be expected to perturb these social systems.
+However, these groups never have perfect cover, their secrecy may make them conspicuously absent from the everyday activity of the background population, and finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc.
+-- that might predict their actions.
 
-As a practical, but mostly pedagogical demonstration, we specify a graph-based model of simple communication across a procedurally generated population, with an embedded, relatively small module representing a clandestine group, pitted against surveillance systems.
-We discuss measuring performance of the opposing sides (e.g.
-Receiver Operator Characteristic), fitting the model against to real data, and finally indiciating how this model can be extended.
+This situation -- strong constraints on passively and actively obtained direct empirical data, overwhelming indirect data, and competing mechanical explanations -- recommends simulation.
+However, because the of the censoring issues in the data, that simulation requires particular attention to characterizing uncertainty and avoiding overfitting.
+
+Herein, we discuss these issues in terms of network models.
+Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) finally some philosophy relative to implementation.
+
+As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.
+We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g.
+Receiver Operator Characteristic).
+Finally, we also demonstrate a small extension.
 

### 2014-01-19 rscripts/manuscript.Rnw
more updates, won't show in compiled version

@@
 \subsection*{The Observation Model}
+Activity is understood by the observing entities by an imperfect process.
+Depending on the situation modeled, that process might simply miss events.
+Alternatively, the error could be misunderstanding the type or content of the activity, or between what individuals and interaction is occurring.
+In general, the observing entities should be able to modify what is observed at what rates.
+Also, the observation process may itself alter the state of the world -- i.e., the covert group may have the opportunity to respond to the observation process.
+
+\section*{A Language Matching the Simulation Requirements: Scala}
+Scala is an mixin inheritance object oriented, functional language, which is also includes an message-passing based model for parallel computing (which is transitioning from the initial implementation to that developed by the Akka project).
+Indeed, Scala is used under the hood for the network simulation framework NetLogo.
+
+Why do these traits make Scala particularly effective for the simulation requirements described above?
+
+First, the covert group members and the background population share behavior (in the real and simulation sense), but are not necessarily simple extensions, which makes a mixin inheritance object system a natural fit for the model subjects.
+
+Second, many interactions in an organization can be thought of as management directing labor to complete a task; in a functional paradigm, functions are themselves treated as objects.
+This translates natural to sending a function object from one simulated individual (in a leadership position) to another (in a subordinate position).
+Loosely adopting some of the other functional paradigms, as Scala does -- emphasis on immutability, principally -- provides practical advantages to concurrent programming, which is necessary for large scale simulations.
+
+Finally, the message-passing framework is an exact analogy to an individual / agent / actor - based model.
+Having this baked into the language, makes it a feature more like control loops or collections, than a more complicated syntax available only advanced programmers (sometimes available on via library support) rather than more phenomena and model focused scientists.
+
+\section*{A Practical Demonstration}
 

### 2014-01-22 rscripts/manuscript.Rnw
update, shifting direction of discussion
not yet reflected in code updates

@@
 % define Scala
-\lstdefinelanguage{scala}{morekeywords={class,object,trait,extends,with,new,if,while,for,def,val,var,this},
-otherkeywords={->,=>},
+\lstdefinelanguage{scala}{morekeywords={class,object,trait,extends,with,new,if,while,for,def,val,var,this,case,->,=>},
+%otherkeywords={->,=>},
 %procnamekeys={def,.},
@@
 Given the real uncertainty, what these filters call for is testing and validation, but those present their own difficulties.
-Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' is either non-existent or deceptive, and experiments could have dangerous side effects.
+Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' ranges from incomplete to deceptive, and experiments could have dangerous side effects.
 Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior and technological innovation (without even further abstraction), even assuming the historical data are more than victor's embellishment.
@@
 This could be used to reduce the simulation size.
-However, this model would still need to create activity data that was consistent (in format) to that created by the covert group.
+However, this model would still need to create activity data that was consistent (in format) to that created by the covert group, because of the simulation requirements for observation.
 
@@
 
-\section*{A Practical Demonstration}
+\section*{Practical Application}
+The practical application of these ideas is -- for a simple simulation -- as straightforward as stated.
 
-%To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
+Using the Scala actor framework, we simply define how each agent in the system responds to events, including the observing entities, and how they generate new events.
+We accomplish the first by appropriately defining {\texttt receive} and the handling of different {\texttt case}s of messages.
+We accomplish the second by having actors know about other actors, and having their internal state determine if they send events to these actors at any particular time.
+
+In considering the salafi network, we will observe a highly constrained set of activity: only communication, with all of that communication having been reduced to ``Good'' or ``Bad''.
+The communication connections between individuals will be static, therefore we may set the social network at the outset of the simulation.
+
+In the code, we could define different kinds of actors -- the covert leader, covert followers, background individuals -- as different classes of simulation object.
+However, we instead take advantage of the context switching in the newer Scala actor framework.
+By approaching the simulation this way, we lay a good foundation for observing the dynamics of individuals -- transitioning from background to covert member to covert leader -- as well as the dynamics of relationship and group formation and dissolution.
+
+One syntactical aside for the code the below sections: the Scala collections (like \texttt{Set}) define methods for adding and removing elements in terms of the natural mathematical operators (\texttt{+} and \texttt{-}) and for similar operations with groups of elements (\texttt{++} and \texttt{--}).
+Similarly, the names for message senders (\texttt{sender}), changing context (\texttt{context become ...}), et cetera are all quite clear.
+Given the informative, mixed-natural-and-mathematical-like syntax, we use actual code snippets instead of pseudo-code to illustrate.
+
+\subsection*{The Covert Group}
+First, we define the covert group context:
+
+\begin{code}
+  def plotter(
+    superiors:Set[Plotter] = Set.empty,
+    collaborators:Set[Plotter] = Set.empty,
+    subordinates:Set[Plotter] = Set.empty) : Receive = {
+      case _ => // do nothing with messages
+  }
+\end{code}
+
+We need to define how they adopt and evolve this context.
+Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though be taking this approach we have laid the ground for monitoring that activity as part of future detection scenarios.
+The simulation runner will send a message \texttt{Radicalize} to the initial cabal, containing all its members:
+
+\begin{code}
+  case class Radicalize(id:Long, leaders:Set[Plotter])
+
+  def receive = {
+    case Radicalize(_, cabal) =>
+      context become plotter(collaborators = (cabal - self))
+    case _ => // do nothing with other messages
+  }
+\end{code}
+
+These leaders then need to recruit subordinates.
+In this particular simulation, we are using exactly the network from Sageman et al., so we will have the simulation runner inform each plotter whom they are recruiting with a \texttt{Subordinates} message.
+The leads will then send \texttt{Recruit} messages to those individuals and they will adopt the plotter context.
+
+\begin{code}
+  case class Subordinates(id:Long, recruits:Set[Plotter])
+  case class Recruit(id:Long)
+  case class RecruitAck(id:Long) // acknowledge recruitment
+
+  def plotter(...) = {
+    case Subordinates(_, recruits) =>
+      recruits foreach { recruit =>
+        recruit ! Recruit(0)
+      }
+    case RecruitAck(_) =>
+      context become plotter(superiors, collaborators, subordinates + sender)
+    // ... other cases
+  }
+
+  def receive = {
+    case Recruit(id) =>
+      context become plotter(superiors = Set(sender))
+      sender ! RecruitAck(id)
+    // ... other cases
+  }
+\end{code}
+
+Some of these new recruits might be introduced to others as collaborators, so we need to need to handle collaboration messages, again assuming that the simulation runner is setting the particular covert structure.
+
+\begin{code}
+  case class Collaboration(id:Long, members:Set[Plotter])
+
+  def plotter(...) = {
+    case Collaboration(_, members) =>
+      context become plotter(superiors, collaborators ++ members, subordinates)
+    // ... other cases
+  }
+\end{code}
+
+Similarly, we can advise these recruits about other actors for them to recruit.
+
+Then we need to consider how these actors will know about each other in the simulation, how they will interact, etc.
+In this example, we consider a static group, and as such the simulation could simply initialize the group.
+However, we know that these groups do not really appear fully formed, and thus someday it will likely be desirable to include recruitment in the model.
+We can anticipate that day by building our static initialization in those terms.
+
+We will represent recruitment activity with two messages, which we add to the definition of the Leader:
+
+\begin{code}[title=Covert Leader Helper Object]
+object Leader {
+  case class Recruit(id:Long) // try to recuit a member
+  case class RecruitAck(id:Long) // join the group
+}
+\end{code}
+
+We then modify the definitions of Leader and Member to respond to these messages:
+
+\begin{code}[title=Using Messages]
+class Leader extends Actor {
+  var recruits = Set.empty[Member]
+
+  def receive = {
+    case RecruitAck(_) =>
+      recruits += sender
+    case _ =>
+  }
+}
+
+class Member extends Actor {
+  var leader;
+
+  def receive = {
+    case Recruit(id) =>
+      leader = sender
+    case _ =>
+  }
+
+}
+\end{code}
+% // form a clique from a vertex collection, col:
+% def clique(col : Collection[V]) =
+%     for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
+%       left <~> right // form a bidirectional edge across the pair
+%       
+% def hierarchicalClique(col: Collection[V], size:Int) = 
+%    val thisLvl : Collection[Collection[V]] =
+%      // make the lowest level cliques 
+%      for ( subGroup <- groupBySize(col, size) ) yield {
+%        clique(subGroup)
+%        subGroup
+%      }
+%    cliqueGroups(thisLvl, size)
+%    
+% @recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
+%   val thisLvl : Collection[Collection[V]] =
+%     // gather the groups being made into a higher level cliques
+%     for ( subGroups <- groupBySize(col, size) ) yield {
+%       // clique a group of groups
+%       for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
+%         leftGroup.randomMember <~> rightGroup.randomMember
+%       }
+%       // merge this set of subgroups into a single new group
+%       merge(subGroups)
+%     }
+%   // repeat with the cliqued cliques, unless everything has been connected
+%   if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
+% \end{code}
+% 
+% %To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
 That means modeling entities that take action, modeling how that action is observed (including if it is observed correctly, or at all), and modeling how those observations are digested into reactions.
@@
 Though we do not do so in our example, the background population might be more tractably modeled with continuous phenomena, given its large size and potentially more homogeneous behavior.
-
-%\subsection*{Modeling the Entities}
-%For our motivating example, we divide the population into three types, two of which belong to the covert group -- management and subordinates -- and a third representing ordinary individuals in the background population.
+% 
+% %\subsection*{Modeling the Entities}
+% %For our motivating example, we divide the population into three types, two of which belong to the covert group -- management and subordinates -- and a third representing ordinary individuals in the background population.
 We choose this number of types because we are using that many models of activity, though different degree nodes will present somewhat differently.
@@
 Fishing with different mechanics may yield a better historical fit, but not necessarily a better forecast.
-
-%\subsubsection*{Background Population}
-%Most observable action will be that of the general populace surrounding the clandestine group.
+% 
+% %\subsubsection*{Background Population}
+% %Most observable action will be that of the general populace surrounding the clandestine group.
 This population has some structural component -- {\em e.g.} family sizes, typical numbers of working members, tendency towards assortativity -- though not necessarily well-known when a given investigation begins and perhaps even assumed to be something it is not.
@@
 Therefore, assessment of any particular pair of opposed strategies should cut across multiple models of the background, each independently parametrized around what data is available.
-
-As an example background population, we have the ordinary individuals form small groups, which in turn connect into larger groups, those groups into still larger groups, and so on until the background consists of a single component.
+% 
+% As an example background population, we have the ordinary individuals form small groups, which in turn connect into larger groups, those groups into still larger groups, and so on until the background consists of a single component.
 If one were inclined to require that this description corresponded to a particular mechanism, this might loosely be interpreted as individuals forming households, households forming blocks, blocks forming neighborhoods, {\em ad nauseum}.
@@
 If demographic data for households were available, then we could plausibly parameterize the lowest level, then possibly combine that with mortality and mobility data to characterize how closely connected households remained, and so on.
-
-Independently, we establish a second set of edges with a different flavor.
+% 
+% Independently, we establish a second set of edges with a different flavor.
 The previously described edges we label ``Familial'', these we call ``Economic''.
@@
 Again, we emphasize: this choice is purely an academic fiction, where we have added this extra fiction purely to highlight the need for multiple dimensions to represent different kinds of relationships in the population.
-
-For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with allowance to handle an arbitrary total population size).  That is:\begin{enumerate}
-\item create and randomly permute a population $P_0$, with $N$ members (later results use $N=100$).
-\item\label{grouping_init} divide $P_0$ into equal groups of size $n$ (later results use $n=3$);
-\item for each group $i$, completely connect the individuals, and label that group $C_i^0$
-\item\label{grouping_end} form a population from the $C_i^0$
-\item repeat steps \ref{grouping_init} to \ref{grouping_end} with the $C_i^0$ connecting each edge between the $C_i^0$ to a uniformly drawn individual within the group, then with the $C_i^1$, etc until a single component is obtained
-\end{enumerate}
-
-\begin{code}[title=Pseudo-Code: Hierarchical Cliques]
-// form a clique from a vertex collection, col:
-def clique(col : Collection[V]) =
-    for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
-      left <~> right // form a bidirectional edge across the pair
-
-def hierarchicalClique(col: Collection[V], size:Int) = 
-   val thisLvl : Collection[Collection[V]] =
-     // make the lowest level cliques 
-     for ( subGroup <- groupBySize(col, size) ) yield {
-       clique(subGroup)
-       subGroup
-     }
-   cliqueGroups(thisLvl, size)
-
-@recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
-  val thisLvl : Collection[Collection[V]] =
-    // gather the groups being made into a higher level cliques
-    for ( subGroups <- groupBySize(col, size) ) yield {
-      // clique a group of groups
-      for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
-        leftGroup.randomMember <~> rightGroup.randomMember
-      }
-      // merge this set of subgroups into a single new group
-      merge(subGroups)
-    }
-  // repeat with the cliqued cliques, unless everything has been connected
-  if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
-\end{code}
-
-Lastly for this model, we establish a final set of edges with a third flavor: ``Religious''.
+% 
+% For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with allowance to handle an arbitrary total population size).  That is:\begin{enumerate}
+% \item create and randomly permute a population $P_0$, with $N$ members (later results use $N=100$).
+% \item\label{grouping_init} divide $P_0$ into equal groups of size $n$ (later results use $n=3$);
+% \item for each group $i$, completely connect the individuals, and label that group $C_i^0$
+% \item\label{grouping_end} form a population from the $C_i^0$
+% \item repeat steps \ref{grouping_init} to \ref{grouping_end} with the $C_i^0$ connecting each edge between the $C_i^0$ to a uniformly drawn individual within the group, then with the $C_i^1$, etc until a single component is obtained
+% \end{enumerate}
+% 
+% \begin{code}[title=Pseudo-Code: Hierarchical Cliques]
+% // form a clique from a vertex collection, col:
+% def clique(col : Collection[V]) =
+%     for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
+%       left <~> right // form a bidirectional edge across the pair
+%       
+% def hierarchicalClique(col: Collection[V], size:Int) = 
+%    val thisLvl : Collection[Collection[V]] =
+%      // make the lowest level cliques 
+%      for ( subGroup <- groupBySize(col, size) ) yield {
+%        clique(subGroup)
+%        subGroup
+%      }
+%    cliqueGroups(thisLvl, size)
+%    
+% @recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
+%   val thisLvl : Collection[Collection[V]] =
+%     // gather the groups being made into a higher level cliques
+%     for ( subGroups <- groupBySize(col, size) ) yield {
+%       // clique a group of groups
+%       for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
+%         leftGroup.randomMember <~> rightGroup.randomMember
+%       }
+%       // merge this set of subgroups into a single new group
+%       merge(subGroups)
+%     }
+%   // repeat with the cliqued cliques, unless everything has been connected
+%   if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
+% \end{code}
+% 
+% Lastly for this model, we establish a final set of edges with a third flavor: ``Religious''.
 These edges occur between members with a probability based the distance between the individuals on the ``Familial'' graph.
@@
 The detailed algorithm is:\begin{enumerate}
-\item\label{religious_init} for each individual $i$:
-\item assign $d_i=1$ and $F_i$ to the set of all their familial connections, excluding individuals already considered
-\item with probability $p^{d_i}$ connect $i$ to the members of $F_i$
-\item\label{religious_end} increment $d_i$, move all of $F_i$ to $P_i$, then add all of the familial connections of $P_i$ that are not in $P_i$ (or previously considered) to $F_i$.
-\item repeat steps \ref{religious_init} to \ref{religious_end} until $F_i$ has no members added in step \ref{religious_end} 
-\end{enumerate}
-
-<<readInCliqued,echo=FALSE>>=
-library(igraph)
-pree <- read.table("../1-EL.txt",col.names=c("sender_id","recipient_id","type"))
-pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
-pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
-edgeinfo <- t(pree)
-# hack: doesn't seem reasonable to have sim output other vertex labels
-#edgeinfo[1,which(edgeinfo[1,] == 107)] <- 103
-#edgeinfo[2,which(edgeinfo[2,] == 107)] <- 103
-#edgeinfo[1,which(edgeinfo[1,] == 112)] <- 104
-#edgeinfo[2,which(edgeinfo[2,] == 112)] <- 104
-
-gWhole <- graph(edgeinfo[-3,])
-V(gWhole)$size<-2
-V(gWhole)[c(102,103,104)]$size<-6
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWhole)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
-E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWhole)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
-E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
-E(gWhole)[which(edgeinfo[3,]=="Plot")]$color <- "red"
-#gAlt <- delete.vertices(gWhole, which(degree(gWhole) < 1)-1)
-l<-layout.auto(gWhole)
-@
-
-\begin{rnwfig}
-<<all, fig=TRUE, width=7.5, echo=FALSE>>=
-plot(gWhole, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<religion, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
-E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- NA
-plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<work, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- NA
-plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{Now showing work (green) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<family, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
-plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{Now showing family (blue) links.}
-\end{rnwfig}
-
-We also consider an alternative background: instead of cliques for ``Familial'' and ``Economic'' ties, individuals form trees.
+% \item\label{religious_init} for each individual $i$:
+% \item assign $d_i=1$ and $F_i$ to the set of all their familial connections, excluding individuals already considered
+% \item with probability $p^{d_i}$ connect $i$ to the members of $F_i$
+% \item\label{religious_end} increment $d_i$, move all of $F_i$ to $P_i$, then add all of the familial connections of $P_i$ that are not in $P_i$ (or previously considered) to $F_i$.
+% \item repeat steps \ref{religious_init} to \ref{religious_end} until $F_i$ has no members added in step \ref{religious_end} 
+% \end{enumerate}
+% 
+% <<readInCliqued,echo=FALSE>>=
+% library(igraph)
+% pree <- read.table("../1-EL.txt",col.names=c("sender_id","recipient_id","type"))
+% pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
+% pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
+% edgeinfo <- t(pree)
+% # hack: doesn't seem reasonable to have sim output other vertex labels
+% #edgeinfo[1,which(edgeinfo[1,] == 107)] <- 103
+% #edgeinfo[2,which(edgeinfo[2,] == 107)] <- 103
+% #edgeinfo[1,which(edgeinfo[1,] == 112)] <- 104
+% #edgeinfo[2,which(edgeinfo[2,] == 112)] <- 104
+% 
+% gWhole <- graph(edgeinfo[-3,])
+% V(gWhole)$size<-2
+% V(gWhole)[c(102,103,104)]$size<-6
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
+% E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
+% E(gWhole)[which(edgeinfo[3,]=="Plot")]$color <- "red"
+% #gAlt <- delete.vertices(gWhole, which(degree(gWhole) < 1)-1)
+% l<-layout.auto(gWhole)
+% @
+% 
+% \begin{rnwfig}
+% <<all, fig=TRUE, width=7.5, echo=FALSE>>=
+% plot(gWhole, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<religion, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- NA
+% plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{The same graph, showing the plot connections (red) and religious (purple) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<work, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- NA
+% plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing work (green) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<family, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
+% plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing family (blue) links.}
+% \end{rnwfig}
+% 
+% We also consider an alternative background: instead of cliques for ``Familial'' and ``Economic'' ties, individuals form trees.
 The treatment for ``Religious'' ties remains the same.
@@
 None of that supposition is particularly relevant: we picked another academic fiction to compare with our previous one.
-
-<<readInTree,echo=FALSE>>=
-pree <- read.table("../1-EL-alt.txt",col.names=c("sender_id","recipient_id","type"))
-pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
-pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
-edgeinfo <- t(pree)
-
-gWholeAlt <- graph(edgeinfo[-3,])
-V(gWholeAlt)$size<-2
-V(gWholeAlt)[c(102,103,104)]$size<-6
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
-E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
-E(gWholeAlt)[which(edgeinfo[3,]=="Plot")]$color <- "red"
-l<-layout.auto(gWholeAlt)
-@
-
-\begin{rnwfig}
-<<allAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-plot(gWholeAlt, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<religionAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- NA
-plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<workAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
-E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- NA
-plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{Now showing work (green) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<familyAlt, fig=TRUE, width=7.5, echo=FALSE>>=
-E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
-E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
-plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
-@
-\caption{Now showing family (blue) links.}
-\end{rnwfig}
-
-\subsubsection*{Covert Actors}
-Sageman, Qin, et al.
+% 
+% <<readInTree,echo=FALSE>>=
+% pree <- read.table("../1-EL-alt.txt",col.names=c("sender_id","recipient_id","type"))
+% pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
+% pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
+% edgeinfo <- t(pree)
+% 
+% gWholeAlt <- graph(edgeinfo[-3,])
+% V(gWholeAlt)$size<-2
+% V(gWholeAlt)[c(102,103,104)]$size<-6
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
+% E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Plot")]$color <- "red"
+% l<-layout.auto(gWholeAlt)
+% @
+% 
+% \begin{rnwfig}
+% <<allAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% plot(gWholeAlt, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<religionAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- NA
+% plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{The same graph, showing the plot connections (red) and religious (purple) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<workAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- NA
+% plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing work (green) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<familyAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
+% plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing family (blue) links.}
+% \end{rnwfig}
+% 
+% \subsubsection*{Covert Actors}
+% Sageman, Qin, et al.
 describe the structure of the Salafi networks as comprising a few key individuals with links to a large group of lieutenants -- the middle management of terror -- that are each connected to several tightly clustered subordinate groups -- terrorist cells -- that execute plots.
 The lieutenants typically integrate with the regular population, while the subordinate groups are largely cloistered.
-
-So we consider a covert organization consisting of two types, those directing a plot, {\em Management}, and those carrying out the day-to-day details, {\em Subordinates}.
+% 
+% So we consider a covert organization consisting of two types, those directing a plot, {\em Management}, and those carrying out the day-to-day details, {\em Subordinates}.
 For this demonstration model, we consider a ``small enough'' plot with a ``narrow enough'' schedule, such that only a single plot will occur during the simulation and that only a single manager is necessary to run that plot.
-
-This manager, $M$, exists along side the background population.
+% 
+% This manager, $M$, exists along side the background population.
 Binomially sample the $C_i^0$ for ``Familial'' (with probability $p_F$) and ``Economic'' (with probability $p_E$) graphs and add $M$ to the selected $C_i^0$.
 Then create ``Religious'' edges according to the same procedure used by ordinary individuals.
-
-The subordinates, however, are isolated from the background population.  The manager and subordinates form a clique with a new type of edge: ``Plot''.
-
-We also consider an alternative model: $M$ joins the background population as described above, but connects to the subordinate group via a hierarchical tree.
-
-\subsection*{Modeling Action \&\ Observation}
-Our proposed types have differences in their structural organization, but we also use those types to distinguish activity by those types on their related structure.
+% 
+% The subordinates, however, are isolated from the background population.  The manager and subordinates form a clique with a new type of edge: ``Plot''.
+% 
+% We also consider an alternative model: $M$ joins the background population as described above, but connects to the subordinate group via a hierarchical tree.
+% 
+% \subsection*{Modeling Action \&\ Observation}
+% Our proposed types have differences in their structural organization, but we also use those types to distinguish activity by those types on their related structure.
 In this assessment, we represent activity only as monitorable communications, and those communications have their content flattened into two categories: ``Good'' and ``Bad''.
@@
 A time and group sensitive partitioning of intercepts for variety and distribution could plausibly form a basis for such a fit; one would have to consider, however, the distinction between the open source background communications (i.e., generally known to be public) versus the intercepted communications of the clandestine group (generally assumed private).
-
-One last technical issue before proceeding to our example implementation: activity models may obviously differ in the data they generate, both structurally and semantically.
+% 
+% One last technical issue before proceeding to our example implementation: activity models may obviously differ in the data they generate, both structurally and semantically.
 The underlying masking and investigating strategies may be framed relative to a particular action model, which may then require an adaptation of the activity data to be consistent with those adjacent models.
-
-\subsubsection*{Background Action}
-Each simulation time step, individuals in the background population generate messages by binomially sampling from all of their available connections.
+% 
+% \subsubsection*{Background Action}
+% Each simulation time step, individuals in the background population generate messages by binomially sampling from all of their available connections.
 They exhibit no preference for the type of those connections (beyond the structural consequence of having different numbers of different types).
@@
 In this proposed continuous model, the aggregation for a strategy that only considers whole messages might be to average (on some specific real time scale) the total vocabulary in each direction of the conversation and then send one message from each party.
-
-As for message content, the background population sends ``Good'' messages with a higher probability than ``Bad'' messages.
-
-\subsubsection*{Covert Action}
-The clandestine group's manager, $M$, behaves much like background population relative to his non-``Plot'' edges.
+% 
+% As for message content, the background population sends ``Good'' messages with a higher probability than ``Bad'' messages.
+% 
+% \subsubsection*{Covert Action}
+% The clandestine group's manager, $M$, behaves much like background population relative to his non-``Plot'' edges.
 His tendency to send ``Bad'' messages to the background population is defined relative to the background probability.
-
-However, he additionally sends direction to the subordinate group via the ``Plot'' connections with some probability.  Presumably, he would balance execution rate with secrecy.
-
-The subordinates, however, do not interact with the background population.
+% 
+% However, he additionally sends direction to the subordinate group via the ``Plot'' connections with some probability.  Presumably, he would balance execution rate with secrecy.
+% 
+% The subordinates, however, do not interact with the background population.
 They randomly speak with each other, with an enhanced probability of sending ``Bad'' messages after having received them from another member of the clandestine group.
-
-\subsection*{Modeling Observation}
-Incomplete information is the norm in these investigations, much like any work in the non-physical sciences.
+% 
+% \subsection*{Modeling Observation}
+% Incomplete information is the norm in these investigations, much like any work in the non-physical sciences.
 However, it is typically possible to modulate what information is available (by investing more resources, by redirecting assets, etc).
@@
 The type of edge used to transmit the message is also not disclosed.
-
-\subsection*{Modeling Reaction}
-We rate investigating entities that implement a few different strategies.
+% 
+% \subsection*{Modeling Reaction}
+% We rate investigating entities that implement a few different strategies.
 One strategy is purely structural based on degrees, another purely content-oriented based on ``Bad'' frequency, and the third mixes the first two.
-
-\subsubsection*{Structural Strategy}
-Structural identification strategies range from having strong prior belief about a particular feature -- e.g., degree distribution -- and a relatively simple detection computation to belief only that there is a meaningful structural distinction (and a covert group to detect at all), thus running up against practical computational considerations trying to analyze all possible distinctions.
-
-We demonstrate a case of the former, positing that the unique structural feature is to do with degree distribution -- specifically, a relatively high degree distribution for the manager, a relatively low degree distribution for the subordinates, and connection between them.
+% 
+% \subsubsection*{Structural Strategy}
+% Structural identification strategies range from having strong prior belief about a particular feature -- e.g., degree distribution -- and a relatively simple detection computation to belief only that there is a meaningful structural distinction (and a covert group to detect at all), thus running up against practical computational considerations trying to analyze all possible distinctions.
+% 
+% We demonstrate a case of the former, positing that the unique structural feature is to do with degree distribution -- specifically, a relatively high degree distribution for the manager, a relatively low degree distribution for the subordinates, and connection between them.
 The criteria for labeling a member of the covert group is then a matter of setting what slices of the distribution to take from the top and bottom, and then testing for an observed path between the manager and the subordinate.
-
-\subsubsection*{Content Strategy}
-A pure content strategy ignores details about sender and receiver arrangement, instead focusing on the sending and receiving of different types of messages.
+% 
+% \subsubsection*{Content Strategy}
+% A pure content strategy ignores details about sender and receiver arrangement, instead focusing on the sending and receiving of different types of messages.
 For our simple model, we consider a strategy that measures relative in and outflow of messages and the frequency of ``Bad'' messages.
-
-\subsubsection*{Combination Strategy}
-For our example combination strategy, we simply require individuals pass both threshold measures.
-
-\section*{Evaluation}
-There are two basic aspects of evaluation, which correspond to the general scientific method questions of model fitting versus model selection.
+% 
+% \subsubsection*{Combination Strategy}
+% For our example combination strategy, we simply require individuals pass both threshold measures.
+% 
+% \section*{Evaluation}
+% There are two basic aspects of evaluation, which correspond to the general scientific method questions of model fitting versus model selection.
 Note that these are entirely separate from detailed application of these aspects to particular model components; those activities are certainly critical in narrow assessments, but we must again emphasize the ability of both sides to adapt their strategies, improve technological capabilities, and so on, all of which will present disruptive changes to any established model.
-
-The aspect which corresponds to questions of fit is basically identifying, for a particular model context -- specific combination of opposed strategies and background activity -- parameter surfaces for optimal performance on the chosen metrics.
-
-The question of model selection corresponds more to considering these performance surfaces across a wide breadth of combinations.
+% 
+% The aspect which corresponds to questions of fit is basically identifying, for a particular model context -- specific combination of opposed strategies and background activity -- parameter surfaces for optimal performance on the chosen metrics.
+% 
+% The question of model selection corresponds more to considering these performance surfaces across a wide breadth of combinations.
 That is, how well does a particular covert strategy perform across multiple background population behaviors and against varying investigatory capabilities?
 Vice versa for the investigatory system?
+% 
+% For our toy systems we consider a simple performance metric, the Receiver Operator Characteristic discrimination statistic, across a background parameter sweep and as a function of time duration. 
+% 
+% <<preRes,echo=FALSE>>=
+% structureFiles <- sort(list.files("..","structure_\\d+\\.txt"))
+% contentFiles <- sort(list.files("..","content_\\d+\\.txt"))
+% sandcFiles <- sort(list.files("..","sandc_\\d+\\.txt"))
+% badFiles <- sort(list.files("..","bad_\\d+\\.txt"))
+% astructureFiles <- sort(list.files("..","astructure_\\d+\\.txt"))
+% acontentFiles <- sort(list.files("..","acontent_\\d+\\.txt"))
+% asandcFiles <- sort(list.files("..","asandc_\\d+\\.txt"))
+% abadFiles <- sort(list.files("..","abad_\\d+\\.txt"))
+% popsize <- 100
+% plotsize <- 6
+% 
+% qData<-function(src) {
+%   FPR<-sapply(src,function(fname) {
+%     read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$FPR
+%   },simplify="matrix") / popsize
+%   FPRq <- apply(FPR,1,quantile)
+%   TPR<-sapply(src,function(fname) {
+%     read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$TPR
+%   },simplify="matrix") / plotsize
+%   TPRq <- apply(TPR,1,quantile)
+%   list(TPRq=TPRq,FPRq=FPRq)
+% }
+% sQData <- qData(structureFiles)
+% sFPRq <- sQData$FPRq
+% sTPRq <- sQData$TPRq
+% asQData <- qData(astructureFiles)
+% asFPRq <- asQData$FPRq
+% asTPRq <- asQData$TPRq
+% 
+% cQData <- qData(contentFiles)
+% cFPRq <- cQData$FPRq
+% cTPRq <- cQData$TPRq
+% acQData <- qData(acontentFiles)
+% acFPRq <- acQData$FPRq
+% acTPRq <- acQData$TPRq
+% 
+% sandcQData <- qData(sandcFiles)
+% sandcFPRq <- sandcQData$FPRq
+% sandcTPRq <- sandcQData$TPRq
+% asandcQData <- qData(asandcFiles)
+% asandcFPRq <- asandcQData$FPRq
+% asandcTPRq <- asandcQData$TPRq
+% 
+% badMR <- sapply(badFiles,function(fname) {
+%   read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
+% },simplify="matrix")
+% badMR <- badMR / max(badMR)
+% badMRq <- apply(badMR,1,quantile)
+% 
+% abadMR <- sapply(abadFiles,function(fname) {
+%   read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
+% },simplify="matrix")
+% abadMR <- abadMR / max(abadMR)
+% abadMRq <- apply(abadMR,1,quantile)
+% 
+% timesteps <- 100
+% samples <- 100
+% t<-1:timesteps
+% tm <- matrix(t,nrow=timesteps,ncol=samples)
+% backpoly<-function(t,x,color,lb=0) {
+%   polygon( c(min(t), t, max(t)), c( lb, x, lb), col=color, border=NA )
+% }
+% liner<-function(t,TPRq,FPRq,TPRxy,FPRxy) {
+%   lines(x=t,y=TPRq[3,],col="green3",type="l",xlab="iterate",ylab="FPR",ylim=c(0,1))
+%   lines(x=t,y=TPRq[4,],col="darkolivegreen1")
+%   lines(x=t,y=TPRq[2,],col="darkolivegreen1")
+%   lines(x=t,y=FPRq[3,],col="red")
+%   lines(x=t,y=FPRq[4,],col="coral")
+%   lines(x=t,y=FPRq[2,],col="coral")
+%   text(TPRxy[1],TPRxy[2],"TPR",col="green3")
+%   text(FPRxy[1],FPRxy[2],"FPR",col="red")
+% }
+% plotter<-function(t,badMRq,TPRq,FPRq,TPRxy,FPRxy) {
+%   plot(t,badMRq[5,],type="l",col="grey93",ylab="",xlab="",yaxt="n",xaxt="n",bty="n")
+%   backpoly(t,badMRq[5,],col="grey93")
+%   backpoly(t,badMRq[4,],col="grey85")
+%   backpoly(t,badMRq[2,],col="grey93")
+%   backpoly(t,badMRq[1,],col="white",lb=-1)
+%   axis(side=1,at=c(min(t),max(t)),labels=c(expression(t[min]),expression(t[max])),tick=F,line=-1)
+%   text(75,0.5,"MR")
+%   liner(t,TPRq,FPRq,TPRxy,FPRxy)
+% }
+% @
+% \subsection*{Structural Performance}
+% \begin{rnwfig}
+% <<structure, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,badMRq,sTPRq,sFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure-Only performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<astructure, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,abadMRq,asTPRq,asFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure-Only performance versus tree-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<content, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,badMRq,cTPRq,cFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Content-Only performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<acontent, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,abadMRq,acTPRq,acFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Content-Only performance versus tree-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<sandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,badMRq,sandcTPRq,sandcFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure and Content performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<asandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,abadMRq,asandcTPRq,asandcFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure and Content performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \section*{Discussion}
+% \todoCP{Get figures to give some direction for this.  Probably going to see the interesting stuff in the differently organized pops.}
 
-For our toy systems we consider a simple performance metric, the Receiver Operator Characteristic discrimination statistic, across a background parameter sweep and as a function of time duration.
-
-<<preRes,echo=FALSE>>=
-structureFiles <- sort(list.files("..","structure_\\d+\\.txt"))
-contentFiles <- sort(list.files("..","content_\\d+\\.txt"))
-sandcFiles <- sort(list.files("..","sandc_\\d+\\.txt"))
-badFiles <- sort(list.files("..","bad_\\d+\\.txt"))
-astructureFiles <- sort(list.files("..","astructure_\\d+\\.txt"))
-acontentFiles <- sort(list.files("..","acontent_\\d+\\.txt"))
-asandcFiles <- sort(list.files("..","asandc_\\d+\\.txt"))
-abadFiles <- sort(list.files("..","abad_\\d+\\.txt"))
-popsize <- 100
-plotsize <- 6
-
-qData<-function(src) {
-  FPR<-sapply(src,function(fname) {
-    read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$FPR
-  },simplify="matrix") / popsize
-  FPRq <- apply(FPR,1,quantile)
-  TPR<-sapply(src,function(fname) {
-    read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$TPR
-  },simplify="matrix") / plotsize
-  TPRq <- apply(TPR,1,quantile)
-  list(TPRq=TPRq,FPRq=FPRq)
-}
-sQData <- qData(structureFiles)
-sFPRq <- sQData$FPRq
-sTPRq <- sQData$TPRq
-asQData <- qData(astructureFiles)
-asFPRq <- asQData$FPRq
-asTPRq <- asQData$TPRq
-
-cQData <- qData(contentFiles)
-cFPRq <- cQData$FPRq
-cTPRq <- cQData$TPRq
-acQData <- qData(acontentFiles)
-acFPRq <- acQData$FPRq
-acTPRq <- acQData$TPRq
-
-sandcQData <- qData(sandcFiles)
-sandcFPRq <- sandcQData$FPRq
-sandcTPRq <- sandcQData$TPRq
-asandcQData <- qData(asandcFiles)
-asandcFPRq <- asandcQData$FPRq
-asandcTPRq <- asandcQData$TPRq
-
-badMR <- sapply(badFiles,function(fname) {
-  read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
-},simplify="matrix")
-badMR <- badMR / max(badMR)
-badMRq <- apply(badMR,1,quantile)
-
-abadMR <- sapply(abadFiles,function(fname) {
-  read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
-},simplify="matrix")
-abadMR <- abadMR / max(abadMR)
-abadMRq <- apply(abadMR,1,quantile)
-
-timesteps <- 100
-samples <- 100
-t<-1:timesteps
-tm <- matrix(t,nrow=timesteps,ncol=samples)
-backpoly<-function(t,x,color,lb=0) {
-  polygon( c(min(t), t, max(t)), c( lb, x, lb), col=color, border=NA )
-}
-liner<-function(t,TPRq,FPRq,TPRxy,FPRxy) {
-  lines(x=t,y=TPRq[3,],col="green3",type="l",xlab="iterate",ylab="FPR",ylim=c(0,1))
-  lines(x=t,y=TPRq[4,],col="darkolivegreen1")
-  lines(x=t,y=TPRq[2,],col="darkolivegreen1")
-  lines(x=t,y=FPRq[3,],col="red")
-  lines(x=t,y=FPRq[4,],col="coral")
-  lines(x=t,y=FPRq[2,],col="coral")
-  text(TPRxy[1],TPRxy[2],"TPR",col="green3")
-  text(FPRxy[1],FPRxy[2],"FPR",col="red")
-}
-plotter<-function(t,badMRq,TPRq,FPRq,TPRxy,FPRxy) {
-  plot(t,badMRq[5,],type="l",col="grey93",ylab="",xlab="",yaxt="n",xaxt="n",bty="n")
-  backpoly(t,badMRq[5,],col="grey93")
-  backpoly(t,badMRq[4,],col="grey85")
-  backpoly(t,badMRq[2,],col="grey93")
-  backpoly(t,badMRq[1,],col="white",lb=-1)
-  axis(side=1,at=c(min(t),max(t)),labels=c(expression(t[min]),expression(t[max])),tick=F,line=-1)
-  text(75,0.5,"MR")
-  liner(t,TPRq,FPRq,TPRxy,FPRxy)
-}
-@
-\subsection*{Structural Performance}
-\begin{rnwfig}
-<<structure, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,badMRq,sTPRq,sFPRq,c(20,0.5),c(80,.15))
-@
-\caption{Structure-Only performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<astructure, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,abadMRq,asTPRq,asFPRq,c(20,0.5),c(80,.15))
-@
-\caption{Structure-Only performance versus tree-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<content, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,badMRq,cTPRq,cFPRq,c(20,0.5),c(80,.15))
-@
-\caption{Content-Only performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<acontent, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,abadMRq,acTPRq,acFPRq,c(20,0.5),c(80,.15))
-@
-\caption{Content-Only performance versus tree-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<sandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,badMRq,sandcTPRq,sandcFPRq,c(20,0.5),c(80,.15))
-@
-\caption{Structure and Content performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-<<asandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
-plotter(t,abadMRq,asandcTPRq,asandcFPRq,c(20,0.5),c(80,.15))
-@
-\caption{Structure and Content performance versus clique-based background.}
-\end{rnwfig}
-
-\section*{Discussion}
-\todoCP{Get figures to give some direction for this.  Probably going to see the interesting stuff in the differently organized pops.}
+\bibliography{manuscript}{}
+\bibliographystyle{plain}
 

### 2014-01-22 rscripts/manuscript.tex
update, shifting direction of discussion
not yet reflected in code updates

@@
 \subsection*{The Observation Model}
+Activity is understood by the observing entities by an imperfect process.
+Depending on the situation modeled, that process might simply miss events.
+Alternatively, the error could be misunderstanding the type or content of the activity, or between what individuals and interaction is occurring.
+In general, the observing entities should be able to modify what is observed at what rates.
+Also, the observation process may itself alter the state of the world -- i.e., the covert group may have the opportunity to respond to the observation process.
 
-%To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
+\section*{A Language Matching the Simulation Requirements: Scala}
+Scala is an mixin inheritance object oriented, functional language, which is also includes an message-passing based model for parallel computing (which is transitioning from the initial implementation to that developed by the Akka project).
+Indeed, Scala is used under the hood for the network simulation framework NetLogo.
+
+Why do these traits make Scala particularly effective for the simulation requirements described above?
+
+First, the covert group members and the background population share behavior (in the real and simulation sense), but are not necessarily simple extensions, which makes a mixin inheritance object system a natural fit for the model subjects.
+
+Second, many interactions in an organization can be thought of as management directing labor to complete a task; in a functional paradigm, functions are themselves treated as objects.
+This translates natural to sending a function object from one simulated individual (in a leadership position) to another (in a subordinate position).
+Loosely adopting some of the other functional paradigms, as Scala does -- emphasis on immutability, principally -- provides practical advantages to concurrent programming, which is necessary for large scale simulations.
+
+Finally, the message-passing framework is an exact analogy to an individual / agent / actor - based model.
+Having this baked into the language, makes it a feature more like control loops or collections, than a more complicated syntax available only advanced programmers (sometimes available on via library support) rather than more phenomena and model focused scientists.
+
+\section*{Practical Application}
+The practical application of these ideas is -- for a simple simulation -- as straightforward as stated.
+
+Using the Scala actor framework, we simply define how each agent in the system responds to events, including the observing entities, and how they generate new events.
+We accomplish the first by appropriately defining {\texttt receive} and the handling of different {\texttt case}s of messages.
+We accomplish the second by having actors know about other actors, and having their internal state determine if they send events to these actors at any particular time.
+
+In considering the salafi network, we will observe a highly constrained set of activity: only communication, with all of that communication having been reduced to ``Good'' or ``Bad''.
+The communication connections between individuals will be static, therefore we may set the social network at the outset of the simulation.
+
+In the code, we could define different kinds of actors -- the covert leader, covert followers, background individuals -- as different classes of simulation object.
+However, we instead take advantage of the context switching in the newer Scala actor framework.
+By approaching the simulation this way, we lay a good foundation for observing the dynamics of individuals -- transitioning from background to covert member to covert leader -- as well as the dynamics of relationship and group formation and dissolution.
+
+One syntactical aside for the code the below sections: the Scala collections (like \texttt{Set}) define methods for adding and removing elements in terms of the natural mathematical operators (\texttt{+} and \texttt{-}) and for similar operations with groups of elements (\texttt{++} and \texttt{--}).
+Similarly, the names for message senders (\texttt{sender}), changing context (\texttt{context become ...}), et cetera are all quite clear.
+Given the informative, mixed-natural-and-mathematical-like syntax, we use actual code snippets instead of pseudo-code to illustrate.
+
+\subsection*{The Covert Group}
+First, we define the covert group context:
+
+\begin{code}
+  def plotter(
+    superiors:Set[Plotter] = Set.empty,
+    collaborators:Set[Plotter] = Set.empty,
+    subordinates:Set[Plotter] = Set.empty) : Receive = {
+      case _ => // do nothing with messages
+  }
+\end{code}
+
+We need to define how they adopt and evolve this context.
+Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though be taking this approach we have laid the ground for monitoring that activity as part of future detection scenarios.
+The simulation runner will send a message \texttt{Radicalize} to the initial cabal, containing all its members:
+
+\begin{code}
+  case class Radicalize(id:Long, leaders:Set[Plotter])
+
+  def receive = {
+    case Radicalize(_, cabal) =>
+      context become plotter(collaborators = (cabal - self))
+    case _ => // do nothing with other messages
+  }
+\end{code}
+
+These leaders then need to recruit subordinates.
+In this particular simulation, we are using exactly the network from Sageman et al., so we will have the simulation runner inform each plotter whom they are recruiting with a \texttt{Subordinates} message.
+The leads will then send \texttt{Recruit} messages to those individuals and they will adopt the plotter context.
+
+\begin{code}
+  case class Subordinates(id:Long, recruits:Set[Plotter])
+  case class Recruit(id:Long)
+  case class RecruitAck(id:Long) // acknowledge recruitment
+
+  def plotter(...) = {
+    case Subordinates(_, recruits) =>
+      recruits foreach { recruit =>
+        recruit ! Recruit(0)
+      }
+    case RecruitAck(_) =>
+      context become plotter(superiors, collaborators, subordinates + sender)
+    // ... other cases
+  }
+
+  def receive = {
+    case Recruit(id) =>
+      context become plotter(superiors = Set(sender))
+      sender ! RecruitAck(id)
+    // ... other cases
+  }
+\end{code}
+
+Some of these new recruits might be introduced to others as collaborators, so we need to need to handle collaboration messages, again assuming that the simulation runner is setting the particular covert structure.
+
+\begin{code}
+  case class Collaboration(id:Long, members:Set[Plotter])
+
+  def plotter(...) = {
+    case Collaboration(_, members) =>
+      context become plotter(superiors, collaborators ++ members, subordinates)
+    // ... other cases
+  }
+\end{code}
+
+Similarly, we can advise these recruits about other actors for them to recruit.
+
+Then we need to consider how these actors will know about each other in the simulation, how they will interact, etc.
+In this example, we consider a static group, and as such the simulation could simply initialize the group.
+However, we know that these groups do not really appear fully formed, and thus someday it will likely be desirable to include recruitment in the model.
+We can anticipate that day by building our static initialization in those terms.
+
+We will represent recruitment activity with two messages, which we add to the definition of the Leader:
+
+\begin{code}[title=Covert Leader Helper Object]
+object Leader {
+  case class Recruit(id:Long) // try to recuit a member
+  case class RecruitAck(id:Long) // join the group
+}
+\end{code}
+
+We then modify the definitions of Leader and Member to respond to these messages:
+
+\begin{code}[title=Using Messages]
+class Leader extends Actor {
+  var recruits = Set.empty[Member]
+
+  def receive = {
+    case RecruitAck(_) =>
+      recruits += sender
+    case _ =>
+  }
+}
+
+class Member extends Actor {
+  var leader;
+
+  def receive = {
+    case Recruit(id) =>
+      leader = sender
+    case _ =>
+  }
+
+}
+\end{code}
+% // form a clique from a vertex collection, col:
+% def clique(col : Collection[V]) =
+%     for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
+%       left <~> right // form a bidirectional edge across the pair
+%       
+% def hierarchicalClique(col: Collection[V], size:Int) = 
+%    val thisLvl : Collection[Collection[V]] =
+%      // make the lowest level cliques 
+%      for ( subGroup <- groupBySize(col, size) ) yield {
+%        clique(subGroup)
+%        subGroup
+%      }
+%    cliqueGroups(thisLvl, size)
+%    
+% @recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
+%   val thisLvl : Collection[Collection[V]] =
+%     // gather the groups being made into a higher level cliques
+%     for ( subGroups <- groupBySize(col, size) ) yield {
+%       // clique a group of groups
+%       for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
+%         leftGroup.randomMember <~> rightGroup.randomMember
+%       }
+%       // merge this set of subgroups into a single new group
+%       merge(subGroups)
+%     }
+%   // repeat with the cliqued cliques, unless everything has been connected
+%   if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
+% \end{code}
+% 
+% %To test strategies for either the covert group or survillience entity, one must have a model capable of representing their strategies.
 That means modeling entities that take action, modeling how that action is observed (including if it is observed correctly, or at all), and modeling how those observations are digested into reactions.
@@
 Again, we emphasize: this choice is purely an academic fiction, where we have added this extra fiction purely to highlight the need for multiple dimensions to represent different kinds of relationships in the population.
-
-For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with allowance to handle an arbitrary total population size).  That is:\begin{enumerate}
-\item create and randomly permute a population $P_0$, with $N$ members (later results use $N=100$).
-\item\label{grouping_init} divide $P_0$ into equal groups of size $n$ (later results use $n=3$);
-\item for each group $i$, completely connect the individuals, and label that group $C_i^0$
-\item\label{grouping_end} form a population from the $C_i^0$
-\item repeat steps \ref{grouping_init} to \ref{grouping_end} with the $C_i^0$ connecting each edge between the $C_i^0$ to a uniformly drawn individual within the group, then with the $C_i^1$, etc until a single component is obtained
-\end{enumerate}
-
-\begin{code}[title=Pseudo-Code: Hierarchical Cliques]
-// form a clique from a vertex collection, col:
-def clique(col : Collection[V]) =
-    for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
-      left <~> right // form a bidirectional edge across the pair
-
-def hierarchicalClique(col: Collection[V], size:Int) = 
-   val thisLvl : Collection[Collection[V]] =
-     // make the lowest level cliques 
-     for ( subGroup <- groupBySize(col, size) ) yield {
-       clique(subGroup)
-       subGroup
-     }
-   cliqueGroups(thisLvl, size)
-
-  val thisLvl : Collection[Collection[V]] =
-    // gather the groups being made into a higher level cliques
-    for ( subGroups <- groupBySize(col, size) ) yield {
-      // clique a group of groups
-      for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
-        leftGroup.randomMember <~> rightGroup.randomMember
-      }
-      // merge this set of subgroups into a single new group
-      merge(subGroups)
-    }
-  // repeat with the cliqued cliques, unless everything has been connected
-  if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
-\end{code}
-
-Lastly for this model, we establish a final set of edges with a third flavor: ``Religious''.
+% 
+% For both of these types, the ``grouping'' operation is to try to form cliques of size $n$ (with allowance to handle an arbitrary total population size).  That is:\begin{enumerate}
+% \item create and randomly permute a population $P_0$, with $N$ members (later results use $N=100$).
+% \item\label{grouping_init} divide $P_0$ into equal groups of size $n$ (later results use $n=3$);
+% \item for each group $i$, completely connect the individuals, and label that group $C_i^0$
+% \item\label{grouping_end} form a population from the $C_i^0$
+% \item repeat steps \ref{grouping_init} to \ref{grouping_end} with the $C_i^0$ connecting each edge between the $C_i^0$ to a uniformly drawn individual within the group, then with the $C_i^1$, etc until a single component is obtained
+% \end{enumerate}
+% 
+% \begin{code}[title=Pseudo-Code: Hierarchical Cliques]
+% // form a clique from a vertex collection, col:
+% def clique(col : Collection[V]) =
+%     for ( (left, right) <- undirectedPairs(col) ) // for each unique, undirected pair in col:
+%       left <~> right // form a bidirectional edge across the pair
+%       
+% def hierarchicalClique(col: Collection[V], size:Int) = 
+%    val thisLvl : Collection[Collection[V]] =
+%      // make the lowest level cliques 
+%      for ( subGroup <- groupBySize(col, size) ) yield {
+%        clique(subGroup)
+%        subGroup
+%      }
+%    cliqueGroups(thisLvl, size)
+%    
+% @recursive def cliqueGroups(col: Collection[Collection[V]], size:Int) =
+%   val thisLvl : Collection[Collection[V]] =
+%     // gather the groups being made into a higher level cliques
+%     for ( subGroups <- groupBySize(col, size) ) yield {
+%       // clique a group of groups
+%       for ( (leftGroup, rightGroup) <- undirectedPairs(subGroup) ) {
+%         leftGroup.randomMember <~> rightGroup.randomMember
+%       }
+%       // merge this set of subgroups into a single new group
+%       merge(subGroups)
+%     }
+%   // repeat with the cliqued cliques, unless everything has been connected
+%   if (thisLvl.size != 1) cliqueGroups(thisLvl, size)  
+% \end{code}
+% 
+% Lastly for this model, we establish a final set of edges with a third flavor: ``Religious''.
 These edges occur between members with a probability based the distance between the individuals on the ``Familial'' graph.
@@
 The detailed algorithm is:\begin{enumerate}
-\item\label{religious_init} for each individual $i$:
-\item assign $d_i=1$ and $F_i$ to the set of all their familial connections, excluding individuals already considered
-\item with probability $p^{d_i}$ connect $i$ to the members of $F_i$
-\item\label{religious_end} increment $d_i$, move all of $F_i$ to $P_i$, then add all of the familial connections of $P_i$ that are not in $P_i$ (or previously considered) to $F_i$.
-\item repeat steps \ref{religious_init} to \ref{religious_end} until $F_i$ has no members added in step \ref{religious_end} 
-\end{enumerate}
-
-\begin{rnwfig}
-\includegraphics{manuscript-all}
-\caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-religion}
-\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-work}
-\caption{Now showing work (green) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-family}
-\caption{Now showing family (blue) links.}
-\end{rnwfig}
-
-We also consider an alternative background: instead of cliques for ``Familial'' and ``Economic'' ties, individuals form trees.
+% \item\label{religious_init} for each individual $i$:
+% \item assign $d_i=1$ and $F_i$ to the set of all their familial connections, excluding individuals already considered
+% \item with probability $p^{d_i}$ connect $i$ to the members of $F_i$
+% \item\label{religious_end} increment $d_i$, move all of $F_i$ to $P_i$, then add all of the familial connections of $P_i$ that are not in $P_i$ (or previously considered) to $F_i$.
+% \item repeat steps \ref{religious_init} to \ref{religious_end} until $F_i$ has no members added in step \ref{religious_end} 
+% \end{enumerate}
+% 
+% <<readInCliqued,echo=FALSE>>=
+% library(igraph)
+% pree <- read.table("../1-EL.txt",col.names=c("sender_id","recipient_id","type"))
+% pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
+% pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
+% edgeinfo <- t(pree)
+% # hack: doesn't seem reasonable to have sim output other vertex labels
+% #edgeinfo[1,which(edgeinfo[1,] == 107)] <- 103
+% #edgeinfo[2,which(edgeinfo[2,] == 107)] <- 103
+% #edgeinfo[1,which(edgeinfo[1,] == 112)] <- 104
+% #edgeinfo[2,which(edgeinfo[2,] == 112)] <- 104
+% 
+% gWhole <- graph(edgeinfo[-3,])
+% V(gWhole)$size<-2
+% V(gWhole)[c(102,103,104)]$size<-6
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
+% E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
+% E(gWhole)[which(edgeinfo[3,]=="Plot")]$color <- "red"
+% #gAlt <- delete.vertices(gWhole, which(degree(gWhole) < 1)-1)
+% l<-layout.auto(gWhole)
+% @
+% 
+% \begin{rnwfig}
+% <<all, fig=TRUE, width=7.5, echo=FALSE>>=
+% plot(gWhole, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Overlay of all connections; the large vertex represents the cluster of subordinates.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<religion, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- NA
+% plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{The same graph, showing the plot connections (red) and religious (purple) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<work, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWhole)[which(edgeinfo[3,]=="Religion")]$color <- NA
+% plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing work (green) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<family, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWhole)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWhole)[which(edgeinfo[3,]=="Work")]$color <- NA
+% plot(gWhole, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing family (blue) links.}
+% \end{rnwfig}
+% 
+% We also consider an alternative background: instead of cliques for ``Familial'' and ``Economic'' ties, individuals form trees.
 The treatment for ``Religious'' ties remains the same.
@@
 None of that supposition is particularly relevant: we picked another academic fiction to compare with our previous one.
-
-\begin{rnwfig}
-\includegraphics{manuscript-allAlt}
-\caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-religionAlt}
-\caption{The same graph, showing the plot connections (red) and religious (purple) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-workAlt}
-\caption{Now showing work (green) links.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-familyAlt}
-\caption{Now showing family (blue) links.}
-\end{rnwfig}
-
-\subsubsection*{Covert Actors}
-Sageman, Qin, et al.
+% 
+% <<readInTree,echo=FALSE>>=
+% pree <- read.table("../1-EL-alt.txt",col.names=c("sender_id","recipient_id","type"))
+% pree[which(pree[,1]<0),1] <- -pree[which(pree[,1]<0),1]
+% pree[which(pree[,2]<0),2] <- -pree[which(pree[,2]<0),2]
+% edgeinfo <- t(pree)
+% 
+% gWholeAlt <- graph(edgeinfo[-3,])
+% V(gWholeAlt)$size<-2
+% V(gWholeAlt)[c(102,103,104)]$size<-6
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$weight <- 0.1
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$weight <- 0.1
+% E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- "purple"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Plot")]$color <- "red"
+% l<-layout.auto(gWholeAlt)
+% @
+% 
+% \begin{rnwfig}
+% <<allAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% plot(gWholeAlt, edge.arrow.size=0.01, edge.color="grey", vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Overlay of all connections; again, the large vertex represents the cluster of subordinates.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<religionAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- NA
+% plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{The same graph, showing the plot connections (red) and religious (purple) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<workAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- "green"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Religion")]$color <- NA
+% plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing work (green) links.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<familyAlt, fig=TRUE, width=7.5, echo=FALSE>>=
+% E(gWholeAlt)[which(edgeinfo[3,]=="Family")]$color <- "blue"
+% E(gWholeAlt)[which(edgeinfo[3,]=="Work")]$color <- NA
+% plot(gWholeAlt, edge.arrow.size=0.01, vertex.label=NA, vertex.frame.color=NA, layout=l)
+% @
+% \caption{Now showing family (blue) links.}
+% \end{rnwfig}
+% 
+% \subsubsection*{Covert Actors}
+% Sageman, Qin, et al.
 describe the structure of the Salafi networks as comprising a few key individuals with links to a large group of lieutenants -- the middle management of terror -- that are each connected to several tightly clustered subordinate groups -- terrorist cells -- that execute plots.
 The lieutenants typically integrate with the regular population, while the subordinate groups are largely cloistered.
-
-So we consider a covert organization consisting of two types, those directing a plot, {\em Management}, and those carrying out the day-to-day details, {\em Subordinates}.
+% 
+% So we consider a covert organization consisting of two types, those directing a plot, {\em Management}, and those carrying out the day-to-day details, {\em Subordinates}.
 For this demonstration model, we consider a ``small enough'' plot with a ``narrow enough'' schedule, such that only a single plot will occur during the simulation and that only a single manager is necessary to run that plot.
-
-This manager, $M$, exists along side the background population.
+% 
+% This manager, $M$, exists along side the background population.
 Binomially sample the $C_i^0$ for ``Familial'' (with probability $p_F$) and ``Economic'' (with probability $p_E$) graphs and add $M$ to the selected $C_i^0$.
 Then create ``Religious'' edges according to the same procedure used by ordinary individuals.
-
-The subordinates, however, are isolated from the background population.  The manager and subordinates form a clique with a new type of edge: ``Plot''.
-
-We also consider an alternative model: $M$ joins the background population as described above, but connects to the subordinate group via a hierarchical tree.
-
-\subsection*{Modeling Action \&\ Observation}
-Our proposed types have differences in their structural organization, but we also use those types to distinguish activity by those types on their related structure.
+% 
+% The subordinates, however, are isolated from the background population.  The manager and subordinates form a clique with a new type of edge: ``Plot''.
+% 
+% We also consider an alternative model: $M$ joins the background population as described above, but connects to the subordinate group via a hierarchical tree.
+% 
+% \subsection*{Modeling Action \&\ Observation}
+% Our proposed types have differences in their structural organization, but we also use those types to distinguish activity by those types on their related structure.
 In this assessment, we represent activity only as monitorable communications, and those communications have their content flattened into two categories: ``Good'' and ``Bad''.
@@
 The type of edge used to transmit the message is also not disclosed.
-
-\subsection*{Modeling Reaction}
-We rate investigating entities that implement a few different strategies.
+% 
+% \subsection*{Modeling Reaction}
+% We rate investigating entities that implement a few different strategies.
 One strategy is purely structural based on degrees, another purely content-oriented based on ``Bad'' frequency, and the third mixes the first two.
-
-\subsubsection*{Structural Strategy}
-Structural identification strategies range from having strong prior belief about a particular feature -- e.g., degree distribution -- and a relatively simple detection computation to belief only that there is a meaningful structural distinction (and a covert group to detect at all), thus running up against practical computational considerations trying to analyze all possible distinctions.
-
-We demonstrate a case of the former, positing that the unique structural feature is to do with degree distribution -- specifically, a relatively high degree distribution for the manager, a relatively low degree distribution for the subordinates, and connection between them.
+% 
+% \subsubsection*{Structural Strategy}
+% Structural identification strategies range from having strong prior belief about a particular feature -- e.g., degree distribution -- and a relatively simple detection computation to belief only that there is a meaningful structural distinction (and a covert group to detect at all), thus running up against practical computational considerations trying to analyze all possible distinctions.
+% 
+% We demonstrate a case of the former, positing that the unique structural feature is to do with degree distribution -- specifically, a relatively high degree distribution for the manager, a relatively low degree distribution for the subordinates, and connection between them.
 The criteria for labeling a member of the covert group is then a matter of setting what slices of the distribution to take from the top and bottom, and then testing for an observed path between the manager and the subordinate.
-
-\subsubsection*{Content Strategy}
-A pure content strategy ignores details about sender and receiver arrangement, instead focusing on the sending and receiving of different types of messages.
+% 
+% \subsubsection*{Content Strategy}
+% A pure content strategy ignores details about sender and receiver arrangement, instead focusing on the sending and receiving of different types of messages.
 For our simple model, we consider a strategy that measures relative in and outflow of messages and the frequency of ``Bad'' messages.
-
-\subsubsection*{Combination Strategy}
-For our example combination strategy, we simply require individuals pass both threshold measures.
-
-\section*{Evaluation}
-There are two basic aspects of evaluation, which correspond to the general scientific method questions of model fitting versus model selection.
+% 
+% \subsubsection*{Combination Strategy}
+% For our example combination strategy, we simply require individuals pass both threshold measures.
+% 
+% \section*{Evaluation}
+% There are two basic aspects of evaluation, which correspond to the general scientific method questions of model fitting versus model selection.
 Note that these are entirely separate from detailed application of these aspects to particular model components; those activities are certainly critical in narrow assessments, but we must again emphasize the ability of both sides to adapt their strategies, improve technological capabilities, and so on, all of which will present disruptive changes to any established model.
-
-The aspect which corresponds to questions of fit is basically identifying, for a particular model context -- specific combination of opposed strategies and background activity -- parameter surfaces for optimal performance on the chosen metrics.
-
-The question of model selection corresponds more to considering these performance surfaces across a wide breadth of combinations.
+% 
+% The aspect which corresponds to questions of fit is basically identifying, for a particular model context -- specific combination of opposed strategies and background activity -- parameter surfaces for optimal performance on the chosen metrics.
+% 
+% The question of model selection corresponds more to considering these performance surfaces across a wide breadth of combinations.
 That is, how well does a particular covert strategy perform across multiple background population behaviors and against varying investigatory capabilities?
 Vice versa for the investigatory system?
+% 
+% For our toy systems we consider a simple performance metric, the Receiver Operator Characteristic discrimination statistic, across a background parameter sweep and as a function of time duration. 
+% 
+% <<preRes,echo=FALSE>>=
+% structureFiles <- sort(list.files("..","structure_\\d+\\.txt"))
+% contentFiles <- sort(list.files("..","content_\\d+\\.txt"))
+% sandcFiles <- sort(list.files("..","sandc_\\d+\\.txt"))
+% badFiles <- sort(list.files("..","bad_\\d+\\.txt"))
+% astructureFiles <- sort(list.files("..","astructure_\\d+\\.txt"))
+% acontentFiles <- sort(list.files("..","acontent_\\d+\\.txt"))
+% asandcFiles <- sort(list.files("..","asandc_\\d+\\.txt"))
+% abadFiles <- sort(list.files("..","abad_\\d+\\.txt"))
+% popsize <- 100
+% plotsize <- 6
+% 
+% qData<-function(src) {
+%   FPR<-sapply(src,function(fname) {
+%     read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$FPR
+%   },simplify="matrix") / popsize
+%   FPRq <- apply(FPR,1,quantile)
+%   TPR<-sapply(src,function(fname) {
+%     read.table(paste("../",fname,sep=""),header=F,sep=" ",col.names=c("FPR","TPR"))$TPR
+%   },simplify="matrix") / plotsize
+%   TPRq <- apply(TPR,1,quantile)
+%   list(TPRq=TPRq,FPRq=FPRq)
+% }
+% sQData <- qData(structureFiles)
+% sFPRq <- sQData$FPRq
+% sTPRq <- sQData$TPRq
+% asQData <- qData(astructureFiles)
+% asFPRq <- asQData$FPRq
+% asTPRq <- asQData$TPRq
+% 
+% cQData <- qData(contentFiles)
+% cFPRq <- cQData$FPRq
+% cTPRq <- cQData$TPRq
+% acQData <- qData(acontentFiles)
+% acFPRq <- acQData$FPRq
+% acTPRq <- acQData$TPRq
+% 
+% sandcQData <- qData(sandcFiles)
+% sandcFPRq <- sandcQData$FPRq
+% sandcTPRq <- sandcQData$TPRq
+% asandcQData <- qData(asandcFiles)
+% asandcFPRq <- asandcQData$FPRq
+% asandcTPRq <- asandcQData$TPRq
+% 
+% badMR <- sapply(badFiles,function(fname) {
+%   read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
+% },simplify="matrix")
+% badMR <- badMR / max(badMR)
+% badMRq <- apply(badMR,1,quantile)
+% 
+% abadMR <- sapply(abadFiles,function(fname) {
+%   read.table(paste("../",fname,sep=""),header=F,col.names=c("badMR"))$badMR
+% },simplify="matrix")
+% abadMR <- abadMR / max(abadMR)
+% abadMRq <- apply(abadMR,1,quantile)
+% 
+% timesteps <- 100
+% samples <- 100
+% t<-1:timesteps
+% tm <- matrix(t,nrow=timesteps,ncol=samples)
+% backpoly<-function(t,x,color,lb=0) {
+%   polygon( c(min(t), t, max(t)), c( lb, x, lb), col=color, border=NA )
+% }
+% liner<-function(t,TPRq,FPRq,TPRxy,FPRxy) {
+%   lines(x=t,y=TPRq[3,],col="green3",type="l",xlab="iterate",ylab="FPR",ylim=c(0,1))
+%   lines(x=t,y=TPRq[4,],col="darkolivegreen1")
+%   lines(x=t,y=TPRq[2,],col="darkolivegreen1")
+%   lines(x=t,y=FPRq[3,],col="red")
+%   lines(x=t,y=FPRq[4,],col="coral")
+%   lines(x=t,y=FPRq[2,],col="coral")
+%   text(TPRxy[1],TPRxy[2],"TPR",col="green3")
+%   text(FPRxy[1],FPRxy[2],"FPR",col="red")
+% }
+% plotter<-function(t,badMRq,TPRq,FPRq,TPRxy,FPRxy) {
+%   plot(t,badMRq[5,],type="l",col="grey93",ylab="",xlab="",yaxt="n",xaxt="n",bty="n")
+%   backpoly(t,badMRq[5,],col="grey93")
+%   backpoly(t,badMRq[4,],col="grey85")
+%   backpoly(t,badMRq[2,],col="grey93")
+%   backpoly(t,badMRq[1,],col="white",lb=-1)
+%   axis(side=1,at=c(min(t),max(t)),labels=c(expression(t[min]),expression(t[max])),tick=F,line=-1)
+%   text(75,0.5,"MR")
+%   liner(t,TPRq,FPRq,TPRxy,FPRxy)
+% }
+% @
+% \subsection*{Structural Performance}
+% \begin{rnwfig}
+% <<structure, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,badMRq,sTPRq,sFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure-Only performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<astructure, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,abadMRq,asTPRq,asFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure-Only performance versus tree-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<content, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,badMRq,cTPRq,cFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Content-Only performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<acontent, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,abadMRq,acTPRq,acFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Content-Only performance versus tree-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<sandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,badMRq,sandcTPRq,sandcFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure and Content performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \begin{rnwfig}
+% <<asandcontent, fig=TRUE, width=7.5, echo=FALSE>>=
+% plotter(t,abadMRq,asandcTPRq,asandcFPRq,c(20,0.5),c(80,.15))
+% @
+% \caption{Structure and Content performance versus clique-based background.}
+% \end{rnwfig}
+% 
+% \section*{Discussion}
+% \todoCP{Get figures to give some direction for this.  Probably going to see the interesting stuff in the differently organized pops.}
 
-For our toy systems we consider a simple performance metric, the Receiver Operator Characteristic discrimination statistic, across a background parameter sweep and as a function of time duration.
-
-\subsection*{Structural Performance}
-\begin{rnwfig}
-\includegraphics{manuscript-structure}
-\caption{Structure-Only performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-astructure}
-\caption{Structure-Only performance versus tree-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-content}
-\caption{Content-Only performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-acontent}
-\caption{Content-Only performance versus tree-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-sandcontent}
-\caption{Structure and Content performance versus clique-based background.}
-\end{rnwfig}
-
-\begin{rnwfig}
-\includegraphics{manuscript-asandcontent}
-\caption{Structure and Content performance versus clique-based background.}
-\end{rnwfig}
-
-\section*{Discussion}
-\todoCP{Get figures to give some direction for this.  Probably going to see the interesting stuff in the differently organized pops.}
+\bibliography{manuscript}{}
+\bibliographystyle{plain}
 

### 2014-01-23 rscripts/manuscript.Rnw
assorted updates

@@
 
-% PLOS One TEMPLATE
-\usepackage{amsmath,amssymb,graphicx,cite,setspace}
+\usepackage{amsmath,amssymb,graphicx,cite,setspace} % PLOS One TEMPLATE
+\usepackage{listings, url, color, appendix} % things I usually use
+% items to make todonotes useful
 \usepackage[textwidth=1.75in]{todonotes} 
 \usepackage[total={6in,9.5in},top=0.5in,left=0.5in,includefoot]{geometry}
-\usepackage{listings}
-\usepackage{url}
-\usepackage{color}
+\newcommand{\wantFig}[2]{
+  \begin{figure}[h]\centering
+    \missingfigure{}
+    \caption{#2}
+    \figlabel{#1}
+  \end{figure}
+}
+% alt layout formatting
+%\doublespacing
+
+% Text layout
+%\topmargin 0.0cm
+%\oddsidemargin 0.5cm
+%\evensidemargin 0.5cm
+%\textwidth 16cm 
+%\textheight 21cm
 
@@
 
-%\renewcommand{\lstlistingname}{Scala Code}
-
-%\doublespacing
-
-% Text layout
-%\topmargin 0.0cm
-%\oddsidemargin 0.5cm
-%\evensidemargin 0.5cm
-%\textwidth 16cm 
-%\textheight 21cm
-
 \usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
@@
 \newcommand{\todoCP}[1]{\todo{CP, #1}}
+
+\newcommand{\figlabel}[1]{\label{fig_#1}}
+\newcommand{\figref}[1]{fig. \ref{fig_#1}}
+\newcommand{\Figref}[1]{Fig. \ref{fig_#1}}
+
+\newcommand{\applabel}[1]{\label{app_#1}}
+\newcommand{\appref}[1]{appendix \ref{app_#1}}
+\newcommand{\Appref}[1]{Appendix \ref{fig_#1}}
 
@@
 \\
-$\ast$ E-mail: cap10@ufl.edu
+$\ast$ E-mail: cap10 \{at\} ufl \{dot\} edu
 \end{flushleft}
@@
 Moreover, increasing active observation would be expected to perturb these social systems.
-However, these groups never have perfect cover, their secrecy may make them conspicuously absent from the everyday activity of the background population, and finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc.
+However, these groups never have perfect cover.
+Their secrecy may make them conspicuously absent from the everyday activity of the background population.
+Finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc.
 -- that might predict their actions.
 
-This situation -- strong constraints on passively and actively obtained direct empirical data, overwhelming indirect data, and competing mechanical explanations -- recommends simulation.
-However, because the of the censoring issues in the data, that simulation requires particular attention to characterizing uncertainty and avoiding overfitting.
+The structure of covert networks is highly variable from one context to another, and they change composition and interaction patterns over time.
+With any covert network, there are strong constraints on passively and actively obtainable direct empirical data, overwhelming amounts of indirect data, and competing mechanical explanations for what is observed.
+Studying the detailed history of any one network, or set of them, will not necessarily provide insight about how to effectively detect such networks in the future.
+This situation recommends simulation of a variety of scenarios with accompanying detection strategies.
+Because of the censoring issues in the data, simulation requires particular attention to characterizing uncertainty and avoiding overfitting.
 
 Herein, we discuss these issues in terms of network models.
-Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) finally some philosophy relative to implementation.
+Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) some philosophy relative to implementation.
 
@@
 Receiver Operator Characteristic).
-Finally, we also demonstrate a small extension.
+Finally, we also demonstrate a small model extensionand show how it affects the (unextended) detection performance.
 
@@
 In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.
-Of course that revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
+Of course the revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
 Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.
@@
 The underlying drive for these opposed efforts is the implacably expanding byte trail.
-Once, the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including morbidity -- but now information era tools produce a near endless supply or 1s and 0s.
-In some cases, this production is a side effect of organized commercial instinct -- monetizing search through targetted advertisement, e.g.
+Once, the recorded information of an entire life might have amounted only to a few bytes -- e.g., parish records on births and deaths, perhaps including morbidity.
+Now information era tools produce a near endless supply of 1s and 0s.
+In some cases, this production is a side effect of organized commercial instinct -- Google monetizing search through targeted advertisement, for example.
 In other cases, individuals are voluntarily advancing their own brand -- Instagram ``selfies'', e.g.
 However, in many cases this bitwise production is an engineering consequence of the associated technology.
-E.g., cellular phones transmit constant location data, financial systems are increasingly just data-based, and of course any use of the internet produces veritable bit contrails.
+E.g., cellular phones transmit constant location data, financial systems are increasing just digital ledgers with almost no physical exchange, and of course any use of the internet produces veritable bit contrails.
 This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.
@@
 We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.
-Like ``algorithmic'' bubble prediction, we tend to side more with Fama than Shiller.
+Like ``algorithmic'' bubble prediction, we side more with Fama than Shiller.
+
+Many such filters are based on social network descriptions.
+These social networks can be a compact way to represent measured relations in a population.
+However, these approaches all boil down to measuring properties of mathematical representation, with those properties having a tenuous connection to a mechanistic explanation of those phenomena.
 
@@
 Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' ranges from incomplete to deceptive, and experiments could have dangerous side effects.
-Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior and technological innovation (without even further abstraction), even assuming the historical data are more than victor's embellishment.
+Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior and technological innovation (without an additional layer of abstraction that would create an even more complex empirical challenge), even assuming the historical data are more than victor's embellishment.
 
@@
 Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable agreement for a particular time and situation -- there would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.
+There have been other attempts at such an approach\cite{carley2003dynamic}.
 
@@
 Relative to the covert group, the simulation must model how the group members act and interact.
-Typically, that further entails a modeling few moving pieces: (i) the members internal state (including what other members they know and communicate with), (ii) how that internal state evolves relative to external forces or states, and (iii) what actions those members undertake based on their internal state (possibly in response to some specific external event).
+Typically, that further entails modeling a few moving pieces: (i) the members internal state (including what other members they know and communicate with), (ii) how that internal state evolves relative to external forces or states, and (iii) what actions those members undertake based on their internal state (possibly in response to some specific external event).
 
@@
 
-An alternative representation for the background would be to model it as a continuous entity vice individuals in a network.
+An alternative representation for the background would be to model it as a continuous entity instead of individuals in a network.
 This could be used to reduce the simulation size.
@@
 \section*{A Language Matching the Simulation Requirements: Scala}
-Scala is an mixin inheritance object oriented, functional language, which is also includes an message-passing based model for parallel computing (which is transitioning from the initial implementation to that developed by the Akka project).
+Scala\cite{odersky2004overview} is an mixin-inheritance\cite{bracha1990mixin}, object-oriented, functional language, which also includes an message-passing based model for parallel computing\footnote{The actor model is transitioning from the initial implementation to that developed by the Akka project\cite{akka}.
+We base our discussion on the Akka model.
+The most important difference is the notion of contexts and context switching: with context switching, an actor alters how it responds to events.
+This is a useful abstraction for representing evolving individual state.}.
 Indeed, Scala is used under the hood for the network simulation framework NetLogo.
@@
 Loosely adopting some of the other functional paradigms, as Scala does -- emphasis on immutability, principally -- provides practical advantages to concurrent programming, which is necessary for large scale simulations.
+Additionally, the functional emphasis bakes in other programming constructs to do with collection processing (e.g., iteration, transformation, filtering) that are less convenient in earlier languages.
 
 Finally, the message-passing framework is an exact analogy to an individual / agent / actor - based model.
-Having this baked into the language, makes it a feature more like control loops or collections, than a more complicated syntax available only advanced programmers (sometimes available on via library support) rather than more phenomena and model focused scientists.
+Having this baked into the language makes it a feature more like control branches (e.g., \texttt{if}, \texttt{else}, \texttt{for}) or collections, than a more complicated syntax available only to advanced programmers (and often only via library support) rather than more phenomena- and model-focused scientists.
 
@@
 
-In considering the salafi network, we will observe a highly constrained set of activity: only communication, with all of that communication having been reduced to ``Good'' or ``Bad''.
+In considering the salafi network (see \figref{salafi_network}), we will observe a highly constrained set of activity: only communication, with all of that communication having been reduced to \texttt{Good} or \texttt{Bad}.
 The communication connections between individuals will be static, therefore we may set the social network at the outset of the simulation.
+
+\wantFig{salafi_network}{the salafi network}
 
@@
 \subsection*{The Covert Group}
-First, we define the covert group context:
+First, we define the covert group \texttt{Receive} context:
 
@@
   def plotter(
+    // a plotter has superiors, collaborators, and subordinates
+    // these groups are initially empty
     superiors:Set[Plotter] = Set.empty,
@@
     subordinates:Set[Plotter] = Set.empty) : Receive = {
-      case _ => // do nothing with messages
+      // ... to-be-defined behavior
+      case _ => // ignore unmatched messages
   }
@@
 We need to define how they adopt and evolve this context.
-Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though be taking this approach we have laid the ground for monitoring that activity as part of future detection scenarios.
-The simulation runner will send a message \texttt{Radicalize} to the initial cabal, containing all its members:
+Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though this approach means we can add monitoring that activity as part of future detection scenarios with minimal coding.
+The simulation runner will send a \texttt{Radicalize} message to the initial cabal, then a \texttt{Collaboration} message to the same group:
 
 \begin{code}
-  case class Radicalize(id:Long, leaders:Set[Plotter])
+  case class Radicalize(id:Long)
+  case class Collaboration(id:Long, group:Set[Plotter])
 
   def receive = {
-    case Radicalize(_, cabal) =>
-      context become plotter(collaborators = (cabal - self))
-    case _ => // do nothing with other messages
+    case Radicalize(_) =>
+      context become plotter()
+    case _ => // ignore other messages
+  }
+
+  def plotter(...) => {
+    case Collaboration(_, group) =>
+      context become plotter(collaborators = collaborators ++ (group - self))
+    // ...
   }
@@
 
-Some of these new recruits might be introduced to others as collaborators, so we need to need to handle collaboration messages, again assuming that the simulation runner is setting the particular covert structure.
+At this point, we have all the necessary messages to construct the given network.
+Those messages, based on the labels in Sageman et al.
+are covered in \Appref{sage_network}.
+However, we still need to model how the plotters actually plot.
+We will assume that all plotting (i) originates with leadership, (ii) when intermediates receive direction, they work with their \texttt{collaborators} as well as delegating work, (iii) that directions occasionally result in feedback to higher-ups, and finally (iv) all of the ``work'' and direction is captured as passing \texttt{Bad} messages.
 
-\begin{code}
-  case class Collaboration(id:Long, members:Set[Plotter])
-
-  def plotter(...) = {
-    case Collaboration(_, members) =>
-      context become plotter(superiors, collaborators ++ members, subordinates)
-    // ... other cases
-  }
-\end{code}
-
-Similarly, we can advise these recruits about other actors for them to recruit.
-
-Then we need to consider how these actors will know about each other in the simulation, how they will interact, etc.
-In this example, we consider a static group, and as such the simulation could simply initialize the group.
-However, we know that these groups do not really appear fully formed, and thus someday it will likely be desirable to include recruitment in the model.
-We can anticipate that day by building our static initialization in those terms.
-
-We will represent recruitment activity with two messages, which we add to the definition of the Leader:
-
-\begin{code}[title=Covert Leader Helper Object]
-object Leader {
-  case class Recruit(id:Long) // try to recuit a member
-  case class RecruitAck(id:Long) // join the group
-}
-\end{code}
-
-We then modify the definitions of Leader and Member to respond to these messages:
-
-\begin{code}[title=Using Messages]
-class Leader extends Actor {
-  var recruits = Set.empty[Member]
-
-  def receive = {
-    case RecruitAck(_) =>
-      recruits += sender
-    case _ =>
-  }
-}
-
-class Member extends Actor {
-  var leader;
-
-  def receive = {
-    case Recruit(id) =>
-      leader = sender
-    case _ =>
-  }
-
-}
-\end{code}
 % // form a clique from a vertex collection, col:
@@
 
+\begin{appendices}
+\section{Messages to Construct Salafi Network}\applabel{sage_network}
+\begin{code}
+  // ...TODO figure out these messages
+  val cabal = Set(``A'',``B'',``C'');
+  cabal foreach { _ ! Radicalize(...) }
+  cabal foreach { _ ! Collaboration(..., cabal) } 
+\end{code}
+
+\end{appendices}
+
 \end{document}

### 2014-01-23 rscripts/manuscript.Rnw
tweaks

@@
 
-One syntactical aside for the code the below sections: the Scala collections (like \texttt{Set}) define methods for adding and removing elements in terms of the natural mathematical operators (\texttt{+} and \texttt{-}) and for similar operations with groups of elements (\texttt{++} and \texttt{--}).
+One syntactical aside for the code the below sections: the Scala collections (like \texttt{Set}) define methods for adding and removing elements in terms of the natural mathematical operators (\texttt{+}, \texttt{-}) and for similar operations with groups of elements (\texttt{++}, \texttt{--}).
 Similarly, the names for message senders (\texttt{sender}), changing context (\texttt{context become ...}), et cetera are all quite clear.
-Given the informative, mixed-natural-and-mathematical-like syntax, we use actual code snippets instead of pseudo-code to illustrate.
+Given the informative, mixed-natural-and-mathematical-like syntax, we prefer actual code snippets to pseudo-code to illustrate simulation steps.
 

### 2014-01-24 rscripts/manuscript.Rnw
add discussion of covert comms model

@@
 We will assume that all plotting (i) originates with leadership, (ii) when intermediates receive direction, they work with their \texttt{collaborators} as well as delegating work, (iii) that directions occasionally result in feedback to higher-ups, and finally (iv) all of the ``work'' and direction is captured as passing \texttt{Bad} messages.
+
+The simulation runner sends \texttt{Time(t:Long)} messages to all actors in the simulation to indicate the march of time.
+In general, during each time step a leader may initiate a new \texttt{Bad} message.
+Other members of the covert group circulate \texttt{Bad} messages subsequent to receiving a \texttt{Bad} message themselves, with different probabilities for sending messages to superiors, collaborators, and subordinates.
+Those probabilities do not depend on the origin of the received message (though this could be a useful mechanical tweak to the model).
+
+Though those probabilities could be coded into the actors
 

### 2014-01-24 rscripts/manuscript.Rnw
changes + compiled diff

@@
 
-The simulation runner sends \texttt{Time(t:Long)} messages to all actors in the simulation to indicate the march of time.
+The simulation runner sends \texttt{Time(t:Long)} messages to all actors in the simulation to indicate the march of time (waiting for them to complete all their business for that step and reply with \texttt{TimeAck(t)}).
 In general, during each time step a leader may initiate a new \texttt{Bad} message.
@@
 
-Though those probabilities could be coded into the actors
+Though those probabilities could be coded into the actors as part of their initialization, if they are set and adjusted by events then the actors (i) more accurrately reflect real behavior and (ii) the evolution of individuals occurs via a consistent simulation mechanism.
+The former provides a lever for expert insight and empirical study.
+The latter makes it easier to produce real-alike data -- i.e., simulation data that has the form of real observations.
+
+In our simplified example, we do not consider how ``real'' events would translate into adjusting a covert member's behavior.
+Instead, the simulation runner sends a message, giving us a wedge for later adjustment without frontloading too much model complexity.
+This represents the final step (in our simulation) for how the plotters evolve in state, so the following repeats the previous snippets with some condensing modifications to the code
+
+\begin{code}
+object SimActor {
+// the companion object, where we define common elements for SimActors
+  case class Radicalize(id:Long)  // become a plotter
+  case class Collaboration(id:Long, group:Set[Plotter])
+                                  // link existing plotters
+  case class Subordinates(id:Long, recruits:Set[Plotter])
+                                  // tell a plotter to recruit subordinates
+  case class Recruit(id:Long)     // tell an actor to join plotters
+  case class RecruitAck(id:Long)  // acknowledge recruitment
+  type Plotter = SimActor
+  type Probability = Double
+}
+
+class SimActor extends Actor {
+
+  def plotter(
+    superiors:Set[Plotter] = Set.empty,
+    collaborators:Set[Plotter] = Set.empty,
+    subordinates:Set[Plotter] = Set.empty,
+    initiate:Probability = 0.0,     // prob. of starting plot comms
+    report:Probability = 0.0,       // prob. of reporting to superiors
+    collaborate:Probability = 0.0,  // prob. of working w/ peers
+    delegate:Probability = 0.0      // prob. of delegating work
+    ) = {
+      def update(
+        newSuperiors:Set[Plotter] = superiors,
+        newCollaborators:Set[Plotter] = collaborators,
+        newSubordinates:Set[Plotter] = subordinates,
+        newInitiate:Probability = initiate,
+        newReport:Probability = report,
+        newCollaborate:Probability = collaborate,
+        newDelegate:Probability = delegate
+        ) = {
+          context become plotter(newSuperiors, newCollaborators, newSubordinates,
+            newInitiate, newReport, newCollaborate, newDelegate)
+      }
+
+      {
+        case Subordinates(_, recruits) =>
+          recruits foreach { recruit =>
+            recruit ! Recruit(0)
+          }
+        case RecruitAck(_) => update(newSubordinates = subordinates + sender)
+        // ... other cases
+      }
+  }
+}
+\end{code}
 

### 2014-01-28 rscripts/manuscript.Rnw
more updates

--- 2014-01-24 rscripts/manuscript.Rnw
@@
 % define Scala
-\lstdefinelanguage{scala}{morekeywords={class,object,trait,extends,with,new,if,while,for,def,val,var,this,case,->,=>},
+\lstdefinelanguage{scala}{morekeywords={package,class,object,trait,extends,with,new,if,while,for,def,val,var,this,case,import,type,->,=>},
 %otherkeywords={->,=>},
@@
 In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.
-Of course the revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
+The revealation of these activities has sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
 Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.
@@
 Given the informative, mixed-natural-and-mathematical-like syntax, we prefer actual code snippets to pseudo-code to illustrate simulation steps.
+However, we have elided some declarations that enable the more informative syntax from the initial appearance in code.
 
@@
     // these groups are initially empty
-    superiors:Set[Plotter] = Set.empty,
-    collaborators:Set[Plotter] = Set.empty,
-    subordinates:Set[Plotter] = Set.empty) : Receive = {
+    superiors:Plotters = empty,
+    collaborators:Plotters = empty,
+    subordinates:Plotters = empty
+    /*, ... other args */) : Receive = {
       // ... to-be-defined behavior
@@
 Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though this approach means we can add monitoring that activity as part of future detection scenarios with minimal coding.
-The simulation runner will send a \texttt{Radicalize} message to the initial cabal, then a \texttt{Collaboration} message to the same group:
+The simulation runner will send a \texttt{Radicalize} message to the initial cabal, then a \texttt{Collaborate} message to the same group:
 
 \begin{code}
-  case class Radicalize(id:Long)
-  case class Collaboration(id:Long, group:Set[Plotter])
+  object Radicalize
+  case class Collaborate(group:Plotters)
 
   def receive = {
-    case Radicalize(_) =>
-      context become plotter()
+    case Radicalize =>
+      context become plotter
     case _ => // ignore other messages
@@
 \begin{code}
-  case class Subordinates(id:Long, recruits:Set[Plotter])
-  case class Recruit(id:Long)
-  case class RecruitAck(id:Long) // acknowledge recruitment
+  case class Subordinates(recruits:Plotters)
+  object Recruit
 
@@
       recruits foreach { recruit =>
-        recruit ! Recruit(0)
+        recruit ! Recruit
       }
-    case RecruitAck(_) =>
-      context become plotter(superiors, collaborators, subordinates + sender)
     // ... other cases
@@
   def receive = {
-    case Recruit(id) =>
+    case Recruit =>
       context become plotter(superiors = Set(sender))
-      sender ! RecruitAck(id)
     // ... other cases
@@
 
-Though those probabilities could be coded into the actors
+Though those probabilities could be coded into the actors as part of their initialization, if they are set and adjusted by events then the actors (i) more accurrately reflect real behavior and (ii) the evolution of individuals occurs via a consistent simulation mechanism.
+The former provides a lever for expert insight and empirical study.
+The latter makes it easier to produce real-alike data -- i.e., simulation data that has the form of real observations.
+
+In our simplified example, we do not consider how ``real'' events would translate into adjusting a covert member's behavior.
+Instead, the simulation runner sends a message, giving us a wedge for later adjustment without frontloading too much model complexity.
+This represents the final step for how the plotters evolve in this simulation, so the following repeats the previous snippets with some condensing modifications to the code.
+The most notable one is making use of a \texttt{case class} to represent all of the parameters that form a plotter's state.
+
+\begin{code}
+package object simactor { // defs to make implementation more natural
+  type Plotters = Set[ActorRef]
+  def empty : Plotters = Set.empty
+  def group(list:ActorRef*) : Plotters = list.toSet
+  type Probability = Double
+}
+
+object SimActor { // defines common elements for SimActors
+  // MESSAGES ABOUT INTERNAL STATE
+  object Radicalize                           // become a plotter
+  object Recruit                              // work for a plotter
+  case class Collaborate(group:Plotters)      // work with other plotters
+  case class Subordinates(recruits:Plotters)  // recruit subordinates
+
+  object PType extends Enumeration { // plotter behavior probabilites
+    type PType = Value
+    val Initiate, // prob. of starting plot comms
+        Report,   // prob. of reporting to superiors
+    	  Collab,	  // prob. of working w/ peers
+    	  Delegate  // prob. of delegating work
+    		  = Value
+  }
+  case class UpdateProb(changes:Map[PType.Value,Probability])
+
+  // MESSAGES ABOUT EXTERNAL ACTIVITY
+  object Bad
+  object Good
+
+  case class PlotterState( // the plotter state description
+    superiors:Plotters = empty,
+    collaborators:Plotters = empty,
+    subordinates:Plotters = empty,
+    probs:Map[PType.Value,Probability] = PType.values.map { (_, 0.0) } toMap
+    // default each plotting probability 0.0
+  )
+}
+
+class SimActor extends Actor {
+  import SimActor._
+
+  def receive = {
+    case Radicalize =>
+      context become plotter()
+    case Recruit =>
+      context become plotter(PlotterState(superiors = group(sender)))
+    case _ => // ignore other messages
+  }
+
+  def plotter(ps:PlotterState = PlotterState()) : Receive = {
+    import ps._
+    def update(newPs:PlotterState) = context become plotter(newPs)
+
+    {
+       case Collaborate(group) =>
+         update(copy(collaborators = collaborators ++ (group - self)))
+       case Subordinates(recruits) =>
+         recruits foreach { recruit => recruit ! Recruit }
+         update(copy(subordinates = subordinates ++ recruits))
+       case Recruit =>
+         update(copy(superiors = superiors + sender))
+       case UpdateProb(changes) =>
+         update(copy(probs = probs ++ changes))
+       case _ => // ignore other messages
+    }
+  }
+}
+\end{code}
+
+Now we address how the plotters plot: if they receive a \texttt{Bad} message from within the covert group, they have some probability of continuing plot activity with their peers, some probability of delegating some work to subordinates (if they have any), and some probability of reporting to their supervisors (again if they have any).
+We will use this as the probability of {\em any} \texttt{Bad} interaction with those groups, and that in general their interaction with these groups is binomially distributed.
+Therefore, if a particular plotter has $k$ collaborators (or supervisors, or subordinates), the probability of interacting with an individual in that group, $p_i$, is related to the probability of interacting with the group, $p_g$ -- which is the probability in the simulation state -- by:
+
+$$
+(1-p_i)^k = 1 - p_g \rightarrow p_i = 1-\sqrt[k]{1-p_g}
+$$
 

### 2014-01-28 rscripts/manuscript.tex
more updates

@@
 
-% PLOS One TEMPLATE
-\usepackage{amsmath,amssymb,graphicx,cite,setspace}
+\usepackage{amsmath,amssymb,graphicx,cite,setspace} % PLOS One TEMPLATE
+\usepackage{listings, url, color, appendix} % things I usually use
+% items to make todonotes useful
 \usepackage[textwidth=1.75in]{todonotes} 
 \usepackage[total={6in,9.5in},top=0.5in,left=0.5in,includefoot]{geometry}
-\usepackage{listings}
-\usepackage{url}
-\usepackage{color}
+\newcommand{\wantFig}[2]{
+  \begin{figure}[h]\centering
+    \missingfigure{}
+    \caption{#2}
+    \figlabel{#1}
+  \end{figure}
+}
+% alt layout formatting
+%\doublespacing
+
+% Text layout
+%\topmargin 0.0cm
+%\oddsidemargin 0.5cm
+%\evensidemargin 0.5cm
+%\textwidth 16cm 
+%\textheight 21cm
 
 % define Scala
-\lstdefinelanguage{scala}{morekeywords={class,object,trait,extends,with,new,if,while,for,def,val,var,this,case,->,=>},
+\lstdefinelanguage{scala}{morekeywords={package,class,object,trait,extends,with,new,if,while,for,def,val,var,this,case,import,type,->,=>},
 %otherkeywords={->,=>},
@@
 
-One syntactical aside for the code the below sections: the Scala collections (like \texttt{Set}) define methods for adding and removing elements in terms of the natural mathematical operators (\texttt{+} and \texttt{-}) and for similar operations with groups of elements (\texttt{++} and \texttt{--}).
+One syntactical aside for the code the below sections: the Scala collections (like \texttt{Set}) define methods for adding and removing elements in terms of the natural mathematical operators (\texttt{+}, \texttt{-}) and for similar operations with groups of elements (\texttt{++}, \texttt{--}).
 Similarly, the names for message senders (\texttt{sender}), changing context (\texttt{context become ...}), et cetera are all quite clear.
-Given the informative, mixed-natural-and-mathematical-like syntax, we use actual code snippets instead of pseudo-code to illustrate.
+Given the informative, mixed-natural-and-mathematical-like syntax, we prefer actual code snippets to pseudo-code to illustrate simulation steps.
+However, we have elided some declarations that enable the more informative syntax from the initial appearance in code.
 
 \subsection*{The Covert Group}
-First, we define the covert group context:
+First, we define the covert group \texttt{Receive} context:
 
@@
   def plotter(
-    superiors:Set[Plotter] = Set.empty,
-    collaborators:Set[Plotter] = Set.empty,
-    subordinates:Set[Plotter] = Set.empty) : Receive = {
-      case _ => // do nothing with messages
+    // a plotter has superiors, collaborators, and subordinates
+    // these groups are initially empty
+    superiors:Plotters = empty,
+    collaborators:Plotters = empty,
+    subordinates:Plotters = empty
+    /*, ... other args */) : Receive = {
+      // ... to-be-defined behavior
+      case _ => // ignore unmatched messages
   }
@@
 We need to define how they adopt and evolve this context.
-Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though be taking this approach we have laid the ground for monitoring that activity as part of future detection scenarios.
-The simulation runner will send a message \texttt{Radicalize} to the initial cabal, containing all its members:
+Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though this approach means we can add monitoring that activity as part of future detection scenarios with minimal coding.
+The simulation runner will send a \texttt{Radicalize} message to the initial cabal, then a \texttt{Collaborate} message to the same group:
 
 \begin{code}
-  case class Radicalize(id:Long, leaders:Set[Plotter])
+  object Radicalize
+  case class Collaborate(group:Plotters)
 
   def receive = {
-    case Radicalize(_, cabal) =>
-      context become plotter(collaborators = (cabal - self))
-    case _ => // do nothing with other messages
+    case Radicalize =>
+      context become plotter
+    case _ => // ignore other messages
+  }
+
+  def plotter(...) => {
+    case Collaboration(_, group) =>
+      context become plotter(collaborators = collaborators ++ (group - self))
+    // ...
   }
@@
 
-Some of these new recruits might be introduced to others as collaborators, so we need to need to handle collaboration messages, again assuming that the simulation runner is setting the particular covert structure.
+At this point, we have all the necessary messages to construct the given network.
+Those messages, based on the labels in Sageman et al.
+are covered in \Appref{sage_network}.
+However, we still need to model how the plotters actually plot.
+We will assume that all plotting (i) originates with leadership, (ii) when intermediates receive direction, they work with their \texttt{collaborators} as well as delegating work, (iii) that directions occasionally result in feedback to higher-ups, and finally (iv) all of the ``work'' and direction is captured as passing \texttt{Bad} messages.
+
+The simulation runner sends \texttt{Time(t:Long)} messages to all actors in the simulation to indicate the march of time (waiting for them to complete all their business for that step and reply with \texttt{TimeAck(t)}).
+In general, during each time step a leader may initiate a new \texttt{Bad} message.
+Other members of the covert group circulate \texttt{Bad} messages subsequent to receiving a \texttt{Bad} message themselves, with different probabilities for sending messages to superiors, collaborators, and subordinates.
+Those probabilities do not depend on the origin of the received message (though this could be a useful mechanical tweak to the model).
+
+Though those probabilities could be coded into the actors as part of their initialization, if they are set and adjusted by events then the actors (i) more accurrately reflect real behavior and (ii) the evolution of individuals occurs via a consistent simulation mechanism.
+The former provides a lever for expert insight and empirical study.
+The latter makes it easier to produce real-alike data -- i.e., simulation data that has the form of real observations.
+
+In our simplified example, we do not consider how ``real'' events would translate into adjusting a covert member's behavior.
+Instead, the simulation runner sends a message, giving us a wedge for later adjustment without frontloading too much model complexity.
+This represents the final step for how the plotters evolve in this simulation, so the following repeats the previous snippets with some condensing modifications to the code.
+The most notable one is making use of a \texttt{case class} to represent all of the parameters that form a plotter's state.
 
 \begin{code}
-  case class Collaboration(id:Long, members:Set[Plotter])
+package object simactor { // defs to make implementation more natural
+  type Plotters = Set[ActorRef]
+  def empty : Plotters = Set.empty
+  def group(list:ActorRef*) : Plotters = list.toSet
+  type Probability = Double
+}
 
-  def plotter(...) = {
-    case Collaboration(_, members) =>
-      context become plotter(superiors, collaborators ++ members, subordinates)
-    // ... other cases
+object SimActor { // defines common elements for SimActors
+  // MESSAGES ABOUT INTERNAL STATE
+  object Radicalize                           // become a plotter
+  object Recruit                              // work for a plotter
+  case class Collaborate(group:Plotters)      // work with other plotters
+  case class Subordinates(recruits:Plotters)  // recruit subordinates
+
+  object PType extends Enumeration { // plotter behavior probabilites
+    type PType = Value
+    val Initiate, // prob. of starting plot comms
+        Report,   // prob. of reporting to superiors
+    	  Collab,	  // prob. of working w/ peers
+    	  Delegate  // prob. of delegating work
+    		  = Value
+  }
+  case class UpdateProb(changes:Map[PType.Value,Probability])
+
+  // MESSAGES ABOUT EXTERNAL ACTIVITY
+  object Bad
+  object Good
+
+  case class PlotterState( // the plotter state description
+    superiors:Plotters = empty,
+    collaborators:Plotters = empty,
+    subordinates:Plotters = empty,
+    probs:Map[PType.Value,Probability] = PType.values.map { (_, 0.0) } toMap
+    // default each plotting probability 0.0
+  )
+}
+
+class SimActor extends Actor {
+  import SimActor._
+
+  def receive = {
+    case Radicalize =>
+      context become plotter()
+    case Recruit =>
+      context become plotter(PlotterState(superiors = group(sender)))
+    case _ => // ignore other messages
+  }
+
+  def plotter(ps:PlotterState = PlotterState()) : Receive = {
+    import ps._
+    def update(newPs:PlotterState) = context become plotter(newPs)
+
+    {
+       case Collaborate(group) =>
+         update(copy(collaborators = collaborators ++ (group - self)))
+       case Subordinates(recruits) =>
+         recruits foreach { recruit => recruit ! Recruit }
+         update(copy(subordinates = subordinates ++ recruits))
+       case Recruit =>
+         update(copy(superiors = superiors + sender))
+       case UpdateProb(changes) =>
+         update(copy(probs = probs ++ changes))
+       case _ => // ignore other messages
+    }
+  }
 }
@@
 
-Similarly, we can advise these recruits about other actors for them to recruit.
+Now we address how the plotters plot: if they receive a \texttt{Bad} message from within the covert group, they have some probability of continuing plot activity with their peers, some probability of delegating some work to subordinates (if they have any), and some probability of reporting to their supervisors (again if they have any).
+We will use this as the probability of {\em any} \texttt{Bad} interaction with those groups, and that in general their interaction with these groups is binomially distributed.
+Therefore, if a particular plotter has $k$ collaborators (or supervisors, or subordinates), the probability of interacting with an individual in that group, $p_i$, is related to the probability of interacting with the group, $p_g$ -- which is the probability in the simulation state -- by:
 
-Then we need to consider how these actors will know about each other in the simulation, how they will interact, etc.
-In this example, we consider a static group, and as such the simulation could simply initialize the group.
-However, we know that these groups do not really appear fully formed, and thus someday it will likely be desirable to include recruitment in the model.
-We can anticipate that day by building our static initialization in those terms.
+$$
+(1-p_i)^k = 1 - p_g \rightarrow p_i = 1-\sqrt[k]{1-p_g}
+$$
 
-We will represent recruitment activity with two messages, which we add to the definition of the Leader:
-
-\begin{code}[title=Covert Leader Helper Object]
-object Leader {
-  case class Recruit(id:Long) // try to recuit a member
-  case class RecruitAck(id:Long) // join the group
-}
-\end{code}
-
-We then modify the definitions of Leader and Member to respond to these messages:
-
-\begin{code}[title=Using Messages]
-class Leader extends Actor {
-  var recruits = Set.empty[Member]
-
-  def receive = {
-    case RecruitAck(_) =>
-      recruits += sender
-    case _ =>
-  }
-}
-
-class Member extends Actor {
-  var leader;
-
-  def receive = {
-    case Recruit(id) =>
-      leader = sender
-    case _ =>
-  }
-
-}
-\end{code}
 % // form a clique from a vertex collection, col:

### 2014-01-29 src/edu/cap10/simactor/SimActor.scala
moving code over from other project

@@ -1,0 +1,90 @@
+package edu.cap10.simactor
+
+import akka.actor.Actor
+import akka.actor.Props
+import akka.actor.Actor
+import akka.actor.ActorRef
+
+object SimActor { // defines common elements for SimActors
+  // MESSAGES ABOUT INTERNAL STATE
+  object Radicalize                           // become a plotter
+  object Recruit                              // work for a plotter
+  case class Collaborate(group:Plotters)      // work with other plotters
+  case class Subordinates(recruits:Plotters)  // recruit subordinates
+  
+  object PType extends Enumeration { // plotter behavior probabilites
+    type PType = Value
+    val	Initiate, // prob. of starting plot comms
+        Report,   // prob. of reporting to superiors
+    	Collab,	  // prob. of working w/ peers
+    	Delegate  // prob. of delegating work
+    	  = Value
+  }
+  case class UpdateProb(changes:Map[PType.Value,Probability])
+
+  // MESSAGES ABOUT EXTERNAL ACTIVITY
+  object Bad
+  object Good
+
+  case class PlotterState( // the plotter state description
+    superiors:Plotters = empty,
+    collaborators:Plotters = empty,
+    subordinates:Plotters = empty,
+    probs:Map[PType.Value,Probability] = PType.values.map { (_, 0.0) } toMap,
+    heardBad:Boolean = false
+    // default each communication type to probability 0.0
+  )
+
+}
+
+import SimRunner._
+
+class SimActor extends Actor with BaseSimActor {
+  import SimActor._
+  import context._
+  
+  def receive = {
+    case Radicalize =>
+      context become plotter()
+    case Recruit =>
+      context become plotter(PlotterState(superiors = group(sender)))
+    case _ => // ignore other messages
+  }
+  
+  def plotter(ps:PlotterState = PlotterState(), time:Long=0) : Receive = {
+    import ps._
+    def update(newPs:PlotterState) = context become plotter(newPs, time)
+    def step = context become plotter(ps,time+1)
+    
+    val base : Receive = {
+       case Collaborate(group) =>
+         update(copy(collaborators = collaborators ++ (group - self)))
+       case Subordinates(recruits) =>
+         recruits foreach { recruit => recruit ! Recruit }
+         update(copy(subordinates = subordinates ++ recruits))
+       case Recruit =>
+         update(copy(superiors = superiors + sender))
+       case UpdateProb(changes) =>
+         update(copy(probs = probs ++ changes))
[…]

### 2014-01-29 src/edu/cap10/simactor/SimRunner.scala

@@ -1,0 +1,73 @@
+package edu.cap10.simactor
+
+import akka.actor.Actor
+import akka.actor.ActorRef
+import akka.actor.Props
+import akka.actor.ActorContext
+
+object SimRunner {
+  case class Time(t:Long)
+  case class TimeAck(t:Long)
+}
+
+class SimRunner extends Actor {
+  import SimRunner._
+  
+  
+  def receive = {
+    case TimeAck(t) =>
+  }
+  
+}
+
+object AckHelp {
+  case class ExpectFrom(senders:Set[ActorRef])
+  object DoneAck
+  object Ack
+  def props(): Props = Props(classOf[AckHelp])
+}
+
+class AckHelp extends Actor {
+  import AckHelp._
+  import context._
+  def receive = {
+    case ExpectFrom(senders) => become( expecting(senders) )
+    case _ => // do nothing
+  }
+  
+  def expecting(senders:Set[ActorRef]) : Receive = {
+    case Ack if (senders(sender)) => 
+      senders - sender match {
+        case leftovers if leftovers.isEmpty => 
+          parent ! DoneAck
+          unbecome
+        case leftovers => become ( expecting(leftovers) )
+      }
+      
+    case ExpectFrom(moreSenders) =>
+      become ( expecting(senders ++ moreSenders) )
+  }
+  
+}
+
+trait BaseSimActor {
+  implicit val context : ActorContext
+  import context._
+  import AckHelp._
+  import Actor.Receive
+  import SimRunner._
+  
+  val ackHelper = actorOf(AckHelp.props)
+  
+  def simBase(step: =>Unit, time: =>Long) : Receive = {
+    case a @ Ack => ackHelper forward a
+    // ackHelper manages whether all my messages have been received
+  	case DoneAck =>
+  	// when it tells me they are, notify the simulation runner (parent)
+      parent ! TimeAck(time)
+      step
+  }
+  
+  def awaiting(who:Set[ActorRef]) = ackHelper ! ExpectFrom(who)
+  def ack = sender ! Ack
+}

### 2014-01-29 src/edu/cap10/simactor/package.scala

@@ -1,0 +1,11 @@
+
+import akka.actor.ActorRef
+package object simactor {
+  type Plotters = Set[ActorRef]
+  type People = Set[ActorRef]
+  type Sim = ActorRef
+  def empty[A] = Set.empty[A]
+  def group(list:ActorRef*) : Plotters = list.toSet
+  type Probability = Double
+  def random = Math.random
+}

### 2014-01-29 rscripts/manuscript.Rnw
updating manuscript

@@
 We base our discussion on the Akka model.
-The most important difference is the notion of contexts and context switching: with context switching, an actor alters how it responds to events.
+The most important difference is the notion of contexts and their switching and stacking: with a switching and stacking, an actor alters how it responds to events.
 This is a useful abstraction for representing evolving individual state.}.

### 2014-01-30 src/edu/cap10/simactor/package.scala
compile fix

@@ -1,5 +1,6 @@
-
+package edu.cap10
 import akka.actor.ActorRef
+
 package object simactor {
   type Plotters = Set[ActorRef]
   type People = Set[ActorRef]

### 2014-01-30 src/edu/cap10/simactor/SimActor.scala
remove unnecessary reference to context

@@ -44,17 +44,15 @@ class SimActor extends Actor with BaseSimActor {
   import context._
   
   def receive = {
-    case Radicalize =>
-      context become plotter()
-    case Recruit =>
-      context become plotter(PlotterState(superiors = group(sender)))
+    case Radicalize => become( plotter() )
+    case Recruit => become( plotter(PlotterState(superiors = group(sender))) )
     case _ => // ignore other messages
   }
   
   def plotter(ps:PlotterState = PlotterState(), time:Long=0) : Receive = {
     import ps._
-    def update(newPs:PlotterState) = context become plotter(newPs, time)
-    def step = context become plotter(ps,time+1)
+    def update(newPs:PlotterState) = become( plotter(newPs, time) )
+    def step = become( plotter(ps,time+1) )
     
     val base : Receive = {
        case Collaborate(group) =>

### 2014-01-31 rscripts/manuscript.Rnw
some changes

@@
 
-The simulation runner sends \texttt{Time(t:Long)} messages to all actors in the simulation to indicate the march of time (waiting for them to complete all their business for that step and reply with \texttt{TimeAck(t)}).
-In general, during each time step a leader may initiate a new \texttt{Bad} message.
+The simulation runs as a series of iterations, and during each time step a leader may initiate a new \texttt{Bad} message.
 Other members of the covert group circulate \texttt{Bad} messages subsequent to receiving a \texttt{Bad} message themselves, with different probabilities for sending messages to superiors, collaborators, and subordinates.
@@
 In our simplified example, we do not consider how ``real'' events would translate into adjusting a covert member's behavior.
-Instead, the simulation runner sends a message, giving us a wedge for later adjustment without frontloading too much model complexity.
+Instead, the simulation runner sends a message, giving us an entry point for later adjustment without frontloading too much model complexity.
 This represents the final step for how the plotters evolve in this simulation, so the following repeats the previous snippets with some condensing modifications to the code.
@@
   def plotter(ps:PlotterState = PlotterState()) : Receive = {
-    import ps._
-    def update(newPs:PlotterState) = context become plotter(newPs)
+    import ps.{copy => change, _}
+    def evolve(newPs:PlotterState) = context become plotter(newPs)
 
@@
        case Collaborate(group) =>
-         update(copy(collaborators = collaborators ++ (group - self)))
+         evolve(change(collaborators = collaborators ++ (group - self)))
        case Subordinates(recruits) =>
          recruits foreach { recruit => recruit ! Recruit }
-         update(copy(subordinates = subordinates ++ recruits))
+         evolve(change(subordinates = subordinates ++ recruits))
        case Recruit =>
-         update(copy(superiors = superiors + sender))
+         evolve(change(superiors = superiors + sender))
        case UpdateProb(changes) =>
-         update(copy(probs = probs ++ changes))
+         evolve(change(probs = probs ++ changes))
        case _ => // ignore other messages
@@
 Now we address how the plotters plot: if they receive a \texttt{Bad} message from within the covert group, they have some probability of continuing plot activity with their peers, some probability of delegating some work to subordinates (if they have any), and some probability of reporting to their supervisors (again if they have any).
+We will treat these activities as (i) independent, and (ii) the interactions with a particular group as a series of independent binomial trials.
+We are not aware of any mechanical explanation justifying this model of interaction; indeed, it implies that, in a finite slice of time for an individual with finite resources, interactions that consume time and resources never interfere.
+For a large interval of time, that might be reasonable -- but large intervals of time conflicts with assumption that actions stimulate responses in subsequent intervals.
+Alternatively, we might infer that there is substantial variability in events (in terms of their duration and resource requirements) such that many interactions means small events, but this undermines the model assumption that these activities can be lumped together into a single category.
+
+The upside of this ``ideal gas'' model of social activity is ease of implementation.
+However, we can circumscribe the appearance of this assumption in our simulation, so that some future Van der Waals can easily clean up our mess later with a more detailed interaction model.
+
 We will use this as the probability of {\em any} \texttt{Bad} interaction with those groups, and that in general their interaction with these groups is binomially distributed.

### 2014-01-31 src/edu/cap10/simactor/SimActor.scala

@@ -6,26 +6,34 @@ import akka.actor.Actor
 import akka.actor.ActorRef
 
 object SimActor { // defines common elements for SimActors
-  // MESSAGES ABOUT INTERNAL STATE
-  object Radicalize                           // become a plotter
-  object Recruit                              // work for a plotter
-  case class Collaborate(group:Plotters)      // work with other plotters
-  case class Subordinates(recruits:Plotters)  // recruit subordinates
+  object Radicalize // become a plotter
   
-  object PType extends Enumeration { // plotter behavior probabilites
-    type PType = Value
-    val	Initiate, // prob. of starting plot comms
-        Report,   // prob. of reporting to superiors
-    	Collab,	  // prob. of working w/ peers
-    	Delegate  // prob. of delegating work
-    	  = Value
-  }
-  case class UpdateProb(changes:Map[PType.Value,Probability])
-
   // MESSAGES ABOUT EXTERNAL ACTIVITY
-  object Bad
-  object Good
+  object SimEvents extends Enumeration { 
+    val Bad, Good = Value
+  }
+  import SimEvents.{Value => SimEvent, _}
 
+  object Relationships extends Enumeration {
+    val Supervisor, Peer, Subordinate, All = Value
+  }
+  import Relationships.{Value => Relationship, _}
+    
+  object SocialContexts extends Enumeration {
+    val Familial, Economic, Religious, Covert = Value
+  }
+  import SocialContexts.{Value => SocialContext, _}
+  
+  type RelationshipMap = Map[SocialContext,Map[Relationship,People]]
+  type BehaviorMap = Map[SocialContext,Map[Relationship,Probability]]
+  
+  // CREATE RELATIONSHIPS IN PARTICULAR SOCIAL CONTEXTS                   
+  case class Recruit(sc:SocialContext) // add a Supervisor relationship w/ sender
+  case class Peer(peers:People, sc:SocialContext) // add Peer relationship w/ peers
+  case class Supervise(recruits:People, sc:SocialContext) // recruit subordinates
+
+  case class UpdateBehavior(changes:Map[Relationship,Probability], sc:SocialContext)
+  
   case class PlotterState( // the plotter state description
     superiors:Plotters = empty,
     collaborators:Plotters = empty,
@@ -45,43 +53,50 @@ class SimActor extends Actor with BaseSimActor {
   
   def receive = {
     case Radicalize => become( plotter() )
[…]

### 2014-01-31 src/edu/cap10/simactor/SimActor.scala
the root update cases
next need to define some objects to create more convenient messages

@@ -27,6 +27,9 @@ object SimActor { // defines common elements for SimActors
   type RelationshipMap = Map[SocialContext,Map[Relationship,People]]
   type BehaviorMap = Map[SocialContext,Map[Relationship,Probability]]
   
+  case class UpdateRelationships(add:RelationshipMap = Map.empty,remove:RelationshipMap = Map.empty)
+  case class UpdateBehaviors(add:BehaviorMap = Map.empty, remove:BehaviorMap = Map.empty)
+  
   // CREATE RELATIONSHIPS IN PARTICULAR SOCIAL CONTEXTS                   
   case class Recruit(sc:SocialContext) // add a Supervisor relationship w/ sender
   case class Peer(peers:People, sc:SocialContext) // add Peer relationship w/ peers

### 2014-02-04 src/edu/cap10/simactor/SimActor.scala
updating

@@ -28,12 +28,9 @@ object SimActor { // defines common elements for SimActors
   type BehaviorMap = Map[ (SocialContext, Relationship), Map[SimEvent, Probability]]
   
   case class UpdateRelationships(sc:SocialContext, rel:Relationship, add:People = empty, remove:People = empty)
-  case class UpdateBehavior(sc:SocialContext, rel:Relationship, event:SimEvent, p:Probability)
+  case class UpdateBehaviors(sc:SocialContext, rel:Relationship, event:SimEvent, p:Probability)
   
-  // CREATE RELATIONSHIPS IN PARTICULAR SOCIAL CONTEXTS                   
-  case class Recruit(sc:SocialContext) // add a Supervisor relationship w/ sender
-  case class Peer(peers:People, sc:SocialContext) // add Peer relationship w/ peers
-  case class Supervise(recruits:People, sc:SocialContext) // recruit subordinates
+  // CREATE RELATIONSHIPS IN PARTICULAR SOCIAL CONTEXTS
   
   case class PersonState( // the plotter state description
     relationships : RelationshipMap = Map.empty,
@@ -46,55 +43,74 @@ import SimRunner._
 
 class SimActor extends Actor with BaseSimActor {
   import SimActor._
-  import context._
   
-  def receive = {
-    case Radicalize => become( plotter() )
-    case Recruit => become( plotter(state = PlotterState(superiors = group(sender))) )
-    case _ => // ignore other messages
+  def evolve : Receive => Unit = context become _
+  
+  def connections(ps:PersonState = PersonState())(implicit other : Receive) : Receive = {
+    import ps.{copy => change, _}  
+	other orElse {
+	  case UpdateRelationships(sc, rel, add, remove) =>
+	    this evolve connections(change(relationships = relationships + ( (sc,rel) -> (relationships((sc,rel)) ++ add -- remove) )))
+	  case UpdateBehaviors(sc, rel, event, p) =>
+	    this evolve connections(change(behaviors = behaviors + ( (sc,rel) -> (behaviors((sc,rel)) + (event -> p)) )))
+	  case Radicalize =>
+	    this evolve connections(ps)(plotter)
+	}
   }
   
-  def basePlot(evolve: (PlotterState)=>Unit, state:PlotterState ) : Receive = {
-    import state.{copy => change, _}
-    {  case Collaborate(group) =>
-         evolve(change(collaborators = collaborators ++ (group - self))) 
-       
[…]

### 2014-02-04 src/edu/cap10/simactor/SimActor.scala
mostly there on runner

@@ -13,16 +13,19 @@ object SimActor { // defines common elements for SimActors
     val Bad, Good = Value
   }
   import SimEvents.{Value => SimEvent, _}
-
+  def simEvents = SimEvents.values
+  
   object Relationships extends Enumeration {
     val Supervisor, Peer, Subordinate, All = Value
   }
   import Relationships.{Value => Relationship, _}
-    
+  def relationshipTypes = Relationships.values 
+  
   object SocialContexts extends Enumeration {
     val Familial, Economic, Religious, Covert = Value
   }
   import SocialContexts.{Value => SocialContext, _}
+  def socialContexts = SocialContexts.values
   
   type RelationshipMap = Map[ (SocialContext, Relationship), People]
   type BehaviorMap = Map[ (SocialContext, Relationship), Map[SimEvent, Probability]]
@@ -34,8 +37,14 @@ object SimActor { // defines common elements for SimActors
   
   case class PersonState( // the plotter state description
     relationships : RelationshipMap = Map.empty,
-    behaviors : BehaviorMap = Map.empty
-  )
+    behaviors : BehaviorMap = Map.empty,
+    time:Long = 0
+  ) {
+    def wholeSocialContextGroup(sc:SocialContext) : People =
+      relationshipTypes.foldLeft(empty[ActorRef])((group,rel) => relationships((Covert,rel)) ++ group)
+    def behaviorMapForSocialContext(sc:SocialContext) : Map[Relationship,Map[SimEvent, Probability]] =
+      relationshipTypes.foldLeft(Map.empty[Relationship,Map[SimEvent,Probability]])( (res, rel) => res + (rel -> behaviors((sc,rel)) ) )
+  }
 
 }
 
@@ -46,6 +55,9 @@ class SimActor extends Actor with BaseSimActor {
   
   def evolve : Receive => Unit = context become _
   
+  import SimActor.SimEvents.{Value => SimEvent, _}
+  import SimActor.SocialContexts.{Value => SocialContext, _}
+  
   def connections(ps:PersonState = PersonState())(implicit other : Receive) : Receive = {
     import ps.{copy => change, _}  
 	other orElse {
@@ -54,14 +66,36 @@ class SimActor extends Actor with BaseSimActor {
 	  case UpdateBehaviors(sc, rel, event, p) =>
 	    this evolve connections(change(behaviors = behaviors + ( (sc,rel) -> (behaviors((sc,rel)) + (event -> p)) )))
 	  case Radicalize =>
[…]

### 2014-02-04 src/edu/cap10/simactor/SimRunner.scala

@@ -4,20 +4,78 @@ import akka.actor.Actor
 import akka.actor.ActorRef
 import akka.actor.Props
 import akka.actor.ActorContext
+import akka.actor.ActorSystem
+
+object Simulation extends App {
+  val system = ActorSystem()
+  // TODO parse args
+  val runner = system.actorOf[SimRunner] 
+}
 
 object SimRunner {
+  case class Start(duration:Long)
+  case class Clock(left:Long,now:Long=0) {
+    def unary_+ = Clock(left-1,now+1)
+    lazy val done = left == 0
+  }
+  
   case class Time(t:Long)
-  case class TimeAck(t:Long)
+  case class TimeAck(t:Long) {
+    lazy val next = Time(t+1)
+  }
+  case class Create(props:Props, initMsgs:Seq[Any])
+  case class CreateBlock(blocks:Map[Props,(Int,Seq[Any])])
+  def props(): Props = Props(classOf[SimRunner])
 }
 
 class SimRunner extends Actor {
   import SimRunner._
+  import context.{ become, actorOf => create }
+  // 
   
-  
-  def receive = {
-    case TimeAck(t) =>
+  def init(c:Any) = {
+    become(initial())
+    self ! c
   }
   
+  def receive = {
+    case m => init(m)
+  }
+  
+  def initial(members : Set[ActorRef] = Set.empty) : Receive = {
+    case Create(props, msgs) =>
+      val newMember = create(props)
+      msgs foreach { newMember ! _}
+    case CreateBlock(blocks) =>
+      blocks foreach {
+        case (props,(count,msgs)) =>
+          val send = Create(props,msgs)
+          for (i <- 0 until count) self ! send
+      }
+    case Start(duration) => 
+      become(running(members)( members,Clock(duration)) )
+      val go = Time(0)
+      members foreach { _ ! go }
+  }
+  
+  def running(awaiting:Set[ActorRef] = Set.empty)(implicit members:Set[ActorRef], clock:Clock) : Receive = {
+	import clock._
+    
+    {
+      case time @ TimeAck(t) if t == now => 
+        awaiting - sender match {
+          case done if done.isEmpty =>
+            become(running()(members,+clock))
+            members foreach { _ ! time.next }
+          case left => become(running(left))
+        }
+      case Create(props, msgs) => 
+        val newMember = create(props)
+        msgs foreach { newMember ! _ }
+        become( running(awaiting)(members + newMember,clock) )
+    }
+  }
+
 }
 
 object AckHelp {

### 2014-04-21 rscripts/manuscript.Rnw
mostly there on runner
lots of overhaul

@@
 {\Large
-\textbf{Modeling Requirements for Simulating Covert Social Groups}
+\textbf{Agent-Based Modeling of Covert Social Groups, Applying Software Engineering Patterns and Planning for Cluster Computation}
 }
@@
 \section*{Abstract}
+We discuss covert social network analysis (SNA) in the larger scientific context, highlight pitfalls, and propose what we view as necessary analysis that seems absent.
+Because SNA typically occurs on aggregated observations (e.g., ties are across types and times), and furthermore covert SNA involves at least some (and often severe) censoring of those observations, it is critical that the implications of that aggregation and observation process be considered.
+
+We demonstrate an implementation framework for performing that kind of analysis, based on a distributed, agent-based model.
+We also apply past software engineering insights to show how they can make for more natural (and therefore comprehensible and compartmentally testable) agent-based models in this framework.
+
+We close with discussion of how this framework might be used in a practical application, and how it is might be pieced together with other analytical tools.
+
+\newpage
+
 Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.
-Moreover, increasing active observation would be expected to perturb these social systems.
-However, these groups never have perfect cover.
-Their secrecy may make them conspicuously absent from the everyday activity of the background population.
+They watch the watchers as well, and increasing effort to reveal them should be expected to encourage adaptation.
+Yet, as impossible as it is for outsiders to perfectly see the covert group, it is equally impossible for these groups to perfectly hide and accomplish anything of note.
+
+% Their secrecy may make them conspicuously absent from the everyday activity of the background population.
 Finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc.
 -- that might predict their actions.
-
-The structure of covert networks is highly variable from one context to another, and they change composition and interaction patterns over time.
+% 
+% Often these mechanically characterizations are framed in terms of social networks.
+Individuals and their ties are thought to have certain graph theoretic features based on their importance, and the observation of these features is then used to focus attention accordingly.
+Yet, often these networks are little more than diverse shadows on the wall.
+They are not the mechanisms or the phenomena, they are not even observation: they are mere representation.
+% 
+% The structure of covert networks is highly variable from one context to another, and they change composition and interaction patterns over time.
 With any covert network, there are strong constraints on passively and actively obtainable direct empirical data, overwhelming amounts of indirect data, and competing mechanical explanations for what is observed.
@@
 Because of the censoring issues in the data, simulation requires particular attention to characterizing uncertainty and avoiding overfitting.
-
-Herein, we discuss these issues in terms of network models.
+% 
+% Herein, we discuss these issues in terms of network models.
 Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) some philosophy relative to implementation.
-
-As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.
+% 
+% As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.
 We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g.
@@
 
-%\todoCP{{\em I think not for this round, but maybe:} Finally, we consider the implications of {\em forged} messages.
-In the basic model, we consider incomplete information about the communications, but the available information is always accurate.
-In this extension, we allow the Observer and the clandestine group to forge messages.
-We again measure various Observer performance traits relative to properties of the observed network.}
-
-\newpage
-
 \section*{Introduction}
-For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior groups operating at the edge of the observable is a principle concern.
+For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior of groups operating at the edge of the observable is a principle concern.
 In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.
@@
 However, in many cases this bitwise production is an engineering consequence of the associated technology.
-E.g., cellular phones transmit constant location data, financial systems are increasing just digital ledgers with almost no physical exchange, and of course any use of the internet produces veritable bit contrails.
+E.g., cellular phones transmit constant location data, financial systems are increasing just digital ledgers substituting the certainty of physical exchange with sophisticated book-keeping to ensure accounts are square, and of course any use of the internet produces veritable bit contrails.
 This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.
@@
 We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.
-Like ``algorithmic'' bubble prediction, we side more with Fama than Shiller.
+Like ``algorithmic'' economic bubble prediction, we side more with Fama than Shiller.
 
 Many such filters are based on social network descriptions.
-These social networks can be a compact way to represent measured relations in a population.
-However, these approaches all boil down to measuring properties of mathematical representation, with those properties having a tenuous connection to a mechanistic explanation of those phenomena.
+Social networks are a compact way to represent measured relations in a population.
+That compression, however, often involves irreversible aggregations: in strength -- e.g., co-employment as measured by working at the same university versus the same department versus in same the laboratory; in type -- e.g., engaging in commerce versus attending the same church versus having married relations; and in time -- e.g., co-appearance at a coffeeshop months ago versus attending classes together on a weekly schedule versus living together.
+
+The mathematics may provide a formally guaranteed result from some particular input, but the more aggregation that occurs translating observations -- themselves imperfect snapshots of actual phenomena -- to representation, the more convincing evidence that must be provided before we should accept that those representations (and subsequent results from manipulating them) have meaning.
 
@@
 It allows for comparison across both detection and masking strategies, consideration of multiple background contexts, forecasting of risks and tradeoffs in a way that allows uncertainties, and in general providing a framework for imaginative assessment.
-Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable agreement for a particular time and situation -- there would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.
+Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable agreement for a particular time and situation -- there would be a natural tendency to ``adjust'' scenario parameters until they indicated the success of strategy.
+Simple models are equally problematic, as they replace the parameter fudge-factor with the complementary assumptions fudge-factors.
+Thus, we must always be skeptical - of hordes of purely fitted parameters that do not have an attached mechanism we can validate the resulting numbers against, and of the assumptions buried simple models that may in fact be driving model outcomes.
+
+A messy business all around, and we have not even gotten to the technically hard details.
+
+\section*{Skeptical Simulation}
+If, as we believe, generating synthetic data for the purpose of testing covert SNA methods is the only pragmatic option, how should one proceed in light of all the obstacles?
+
+As the point is to assess the conclusions of those methods against what is ``actually'' happening ({\em in silico}), on one end the process must obviously produce output that can be ultimately represented as a social network.
+Since any the subject SNA method should include the process of converting real observations to the represention, the output should be in the form of those observations.
+That part is conceptually straightforward, if not necessarily technically so.
+
+The other end -- what is ``actually'' happening, at least in simulation -- is less conceptually straightforward, the problem being that we do not know all of what is actually happening.
+We propose here that researchers adopt a modeling philosophy that makes them open to criticism.
+That is, the model meets the following criteria:
+\begin{itemize}
+\item it is clear which parts are explicit (subject to fitting -- e.g., rate constants) versus implicit (assumptions -- e.g., interaction forms)
+\item it is clear what those explicit and implicit processes are doing
+\item it is clear where parts interact
+\item it is clear how the observation process interacts with the modeled system
+\end{itemize}
+
+Consider the covert SNA described by Bright, et al.
+of the social networks created by judge's sentencing comments indicating personal association in methamphetamine production enterprises.
+They provide the key components for the proposed modeling philosophy:
+\begin{itemize}
+\item a description of the subject process - e.g., acquisition of precursor, production in clandestine laboratory, distribution, as well as necessity of specialists
+\item a description of how that process is observed - e.g., law enforcement investigation and subsequent prosecution
+\item how the observations are flattened into a network representation - translation of judge's sentencing comments
+\item the specific SNA methods applied and desired inferences - e.g., various centralities and applications to intervention
+\end{itemize}
+
+With the exception of the actual SNA, however, these are ``talking'' models.
+We assert that to genuinely assess if the SNA based on judicial sentencing notes, we must explicitly understand how gang activity ultimately percolates into that ink.
+That it is necessary to turn the descriptive models into computational ones, while preserving (and hopefully enhancing) our ability to argue with each piece of that story and all the between-the-lines implications.
+
+\section*{Features of Good Computational Model}
+What makes a good computational model?
+Modellers in the traditional hard sciences (physics, chemistry, biology, and assorted sub disciplines) have historically emphasized purely computational concerns, which is quite sensible for problems that have simpler conceptual models (i.e., those that are mathematically formalizable) but relatively intense numerical tasks.
+On the other hand, many non-academic software products deal with a quite sophisticated conceptual model with relatively simple numerical tasks, and that demand tends to push their development more towards programming and engineering concerns.
+We assert this domain -- software engineering -- is the better reference for social science modeling, and that the lessons of that discipline are more keenly relevant (though they are certainly relevant to the natural sciences, especially as the biological sciences deal with more of the heterogeniety intrinsics to that domain).
+
+We will emphasize a few key lessons, some ``ancient'' wisdom (two plus decades old) and some a bit more recent.
+Good modeling is quite like good (scientific) story telling, with the exception that the story must be told by the code (though we include in ``code'' more than just what runs a particular simulation).
+
+The overall goal is to make implementation choices that make the model itself communicate with other researchers.
+This means the right level of dividing up the model.
+Too little, and the rest of us have to comprehend the whole thing to comprehend a single atom of the story.
+Too much, and instead we have to go to quarks to understand an atom of the story.
+
+This begins with choosing a setup (language, libraries, frameworks, etc.) that make this possible.
+
+\begin{itemize}
+\item literate programming
+\item design patterns - particularly emphasis on composition, decoration, adaptation over inheritance + achieving that with chain of responsibility 
+\end{itemize}
+
+They provide a descriptive model of drugs production that appears nearly identical to most industrial enterprises: acquisition of materials, production, retailing, wholesaling, and regulatory ``compliance'' (bribing officials).
+Notably, the detail of this description and its quite reasonable analogy to regular enterprise.
+
 There have been other attempts at such an approach\cite{carley2003dynamic}.
@@
 \begin{appendices}
+\section{Code for Base Simulation Actor}
+
 \section{Messages to Construct Salafi Network}\applabel{sage_network}

### 2014-04-22 rscripts/manuscript.Rnw
eh, commenting out a thing

@@
 
-Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.
+% Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.
 They watch the watchers as well, and increasing effort to reveal them should be expected to encourage adaptation.

### 2014-04-22 rscripts/manuscript.tex
eh, commenting out a thing

@@
 
-%\todoCP{{\em I think not for this round, but maybe:} Finally, we consider the implications of {\em forged} messages.
-In the basic model, we consider incomplete information about the communications, but the available information is always accurate.
-In this extension, we allow the Observer and the clandestine group to forge messages.
-We again measure various Observer performance traits relative to properties of the observed network.}
-
-\newpage
-
 \section*{Introduction}
-For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior groups operating at the edge of the observable is a principle concern.
+For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior of groups operating at the edge of the observable is a principle concern.
 In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.
-Of course the revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
+The revealation of these activities has sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.
 Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.

### 2015-04-03 rscripts/data_review.Rmd
updates, convert to manuscript

---
output: pdf_document
---
## Analysis and plots of montreal data.

```{r, echo=FALSE, include=FALSE}
req.packages <- c("data.table", "ggplot2", "reshape2", "grid")
sapply(req.packages, require, character.only = T)
```

### Required packages: `r req.packages`.

Source data:

```{r, echo=FALSE}
src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ")
setnames(src.dt, c("user_id", "location_id", "login", "logout"))
setkey(src.dt, login, logout, user_id, location_id)
src.dt
```

```{r}
max_hours <- 24
min_logins <- 2
```

Trim data by:

1. sessions exceeding `r max_hours` hours
2. eliminate users that have fewer than `r min_logins` logins
3. eliminate locations that have fewer than `r min_logins` logins
4. recursively apply steps 2-3 until no users / locations eliminated

```{r, echo=FALSE}
censor.dt <- src.dt[(logout - login) <= max_hours*60*60, ]
invalid.users <- censor.dt[,.N, by=user_id ][N < min_logins, user_id]
while(length(invalid.users) > 0) {
  censor.dt <- censor.dt[!user_id %in% invalid.users]
  invalid.locs <- censor.dt[,.N, by=location_id ][N < min_logins, location_id]
  censor.dt <- censor.dt[!location_id %in% invalid.locs]
  invalid.users <- censor.dt[,.N, by=user_id ][N < min_logins, user_id]
}
```

```{r, echo=FALSE}
# src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ")
# setnames(src.dt, c("user_id", "location_id", "login", "logout"))
# setkey(src.dt, login, logout, user_id, location_id)
# 
# max_real_duration <- 60*60*24 # at most, continuous login of a day
# 
# bad.list <- with(src.dt[(logout - login) > max_real_duration,],
#   list(bad_users = unique(user_id), bad_locs = unique(location_id), bad_entries = length(user_id) )
# )
# 
# duration.censored.dt <- src.dt[(logout - login) <= max_real_duration,]
# 
# eliminated.users <- setdiff(bad.list$bad_users, unique(duration.censored.dt$user_id))
# eliminated.locs <- setdiff(bad.list$bad_locs, unique(duration.censored.dt$location_id))
# warning("eliminated ", length(eliminated.users)," users with only >24 hour sessions: ", paste(eliminated.users, collapse=" "))
# 
# user.location.count.dt <- duration.censored.dt[,
# 	list(location_count = length(unique(location_id))),
# 	by="user_id"
# ]
# valid.users <- user.location.count.dt[location_count > 1,user_id]
# 
# user_and_duration.censored.dt <- duration.censored.dt[user_id %in% valid.users,]
# 
# locs.dt <- user_and_duration.censored.dt[,
# 	list(
# 		unique_users = length(unique(user_id)),
# 		total_login_time = sum(logout - login)),
# 	by="location_id"
# ] # unique users per location
# 
# min_total_login <- 60*60 # 1 hour
# 
# invalid.locs <- locs.dt[(unique_users == 1) | (total_login_time < min_total_login), location_id] # locations with only one user, or total log in time < 1 hour
# 
# loc_user_duration.censored.dt <- user_and_duration.censored.dt[!(location_id %in% invalid.locs),]
# loc_user_duration.censored.dt[,user_id := .GRP, by=user_id]
# loc_user_duration.censored.dt[,location_id := .GRP, by=location_id]
# 
# loc_view.dt <- loc_user_duration.censored.dt[,
#   list(
#     unique_users = length(unique(user_id)),
#     total_login_time = sum(logout - login),
#     mean_login = mean(logout - login),
#     sd_login = sd(logout - login)
#   ),
#   by="location_id"]
# 
# user_view.dt <- loc_user_duration.censored.dt[,
# 	list(
# 		unique_locs = length(unique(location_id)),
# 		total_login_time = sum(logout - login),
# 		mean_login = mean(logout - login),
# 		sd_login = sd(logout - login),
#     first = min(login),
#     last = max(logout)
# 	),
# 	by="user_id"
# ]
# 
# user_view.dt[, login_proportion := total_login_time / (last-first) ]
# 
# ggplot(user_view.dt) + theme_bw() +
#   aes(x=user_id, ymin=first, ymax=last, alpha=login_proportion) + geom_linerange() + coord_flip()
# 
# 
# break_categorization <- function(dt, target, output) {
# 	bks <- hist(dt[[target]], plot = F)$breaks
# 	lbls <- paste(head(bks, -1), tail(bks, -1), sep="-")
# 	dt[[output]] <- sapply(dt[[target]], function(n) factor(lbls[which.max(n <= bks)-1], levels=lbls))
# 	dt
# }
# 
# #loc_count_breaks <- hist(user_view.dt$unique_locs, plot = F)$breaks
# 
# #loc_count_cats <- paste(head(loc_count_breaks, -1), tail(loc_count_breaks, -1), sep="-")
# 
# #user_view.dt[,loc_count_cat := sapply(unique_locs, function(n) which.max(n <= loc_count_breaks)-1)]
# #user_view.dt$loc_count_cat <- factor(loc_count_cats[user_view.dt$loc_count_cat], levels=loc_count_cats)
# 
# user_view.dt <- break_categorization(user_view.dt, "unique_locs", "loc_count_cat")
# 
# # sum(loc_user_duration.censored.dt[,length(unique(location_id)),by="user_id"]$V1 == 1) == 0 # => no more user ids to censor
# 
# total_login_duration <- sum(as.numeric(loc_view.dt$total_login_time))
# loc_view.dt[,proportion := total_login_time/total_login_duration]
# 
# nets.dt <- loc_user_duration.censored.dt[, list(first=min(login),last=max(logout)), by=c("user_id","location_id")]
# 
# melt_nets.dt <- melt(nets.dt, id.vars = c("user_id", "location_id"))
# melt_nets.dt$inc <- ifelse(melt_nets.dt$variable == "first", 1, -1)
# setkey(melt_nets.dt, value, user_id, location_id)
# 
# melt_nets.dt[,net_users := cumsum(inc), by="location_id"]
# melt_nets.dt[,net_locs  := cumsum(inc), by="user_id"]
# 
# acc_loc_time <- melt_nets.dt[,
# 	list(ave_concurrent_locs = 
# 	  sum(diff(value)*head(net_locs,-1)) /
# 	  sum(diff(value)*(head(net_locs,-1)>0)),
# 	  unique_locs = sum(abs(inc))/2
# 	), # exclude time intervals w/ no presence
# 	by="user_id"]
# acc_loc_time <- break_categorization(acc_loc_time, "ave_concurrent_locs","con_locs_cat")
# acc_loc_time <- break_categorization(acc_loc_time, "unique_locs","unique_locs_cat")
# 
# ggplot(acc_loc_time[ave_concurrent_locs >= 2,]) + theme_bw() +
# 	aes(x=ave_concurrent_locs, fill=unique_locs_cat) + geom_bar() +
# 	scale_x_log10() #+ scale_y_log10()
# #sample(loc_view.dt$location_id, 5, F, loc_view.dt$proportion)
# 
# p.loc <- ggplot(loc_view.dt) + theme_bw()
# p.user <- ggplot(user_view.dt) + theme_bw()
# 
# p.user + theme(panel.margin = unit(0.5, "lines")) +
# 	aes(fill=loc_count_cat, x=total_login_time/60) +
# 	facet_grid(loc_count_cat ~ ., scales="free_y", shrink=T, drop=F) +
# 	geom_bar() + scale_x_log10(name="total login time per individual, in minutes") +
# 	ylab("number of individuals") +
# 	scale_fill_discrete(name="locations per individual", drop=F) +
# 	scale_y_continuous(breaks=function(lims) c(head(lims,1), tail(lims, 1)), expand=c(0,0))
# 
# p + aes(x=(logout - login)) + geom_bar() + geom_vline(x=150, color="red") + scale_x_log10()
# 
# #####
# p + aes(x=unique_users) + geom_bar() + scale_x_log10()
# p + aes(x=total_login_time) + geom_bar() + scale_x_log10()
# 
# src.dt[,length(unique(user_id)),by="location_id"] # unique users per location
# src.dt[,length(unique(location_id)),by="user_id"] # unique locations by user
# 
# regular_users <- src.dt[,list(regular = length(location_id) > 1), by="user_id"] # users w/ more than one login
# regular_users <- regular_users[regular == T, user_id, keyby="user_id"]$user_id
```

### 2015-04-03 rscripts/data_review.R

@@ -3,12 +3,20 @@
 require(data.table); require(ggplot2); require(grid); require(reshape2)
 
 src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ")
-setnames(src.dt, c("user_id","location_id","login","logout"))
+setnames(src.dt, c("user_id", "location_id", "login", "logout"))
 setkey(src.dt, login, logout, user_id, location_id)
 
 max_real_duration <- 60*60*24 # at most, continuous login of a day
 
+bad.list <- with(src.dt[(logout - login) > max_real_duration,],
+  list(bad_users = unique(user_id), bad_locs = unique(location_id), bad_entries = length(user_id) )
+)
+
 duration.censored.dt <- src.dt[(logout - login) <= max_real_duration,]
+
+eliminated.users <- setdiff(bad.list$bad_users, unique(duration.censored.dt$user_id))
+eliminated.locs <- setdiff(bad.list$bad_locs, unique(duration.censored.dt$location_id))
+warning("eliminated ", length(eliminated.users)," users with only >24 hour sessions: ", paste(eliminated.users, collapse=" "))
 
 user.location.count.dt <- duration.censored.dt[,
 	list(location_count = length(unique(location_id))),
@@ -33,10 +41,6 @@ loc_user_duration.censored.dt <- user_and_duration.censored.dt[!(location_id %in
 loc_user_duration.censored.dt[,user_id := .GRP, by=user_id]
 loc_user_duration.censored.dt[,location_id := .GRP, by=location_id]
 
-
-ggplot(loc_user_duration.censored.dt[,list(first=min(login), last=max(logout)),by=user_id]) + theme_bw() +
-  aes(ymin=user_id+1-0.5, ymax=user_id+1+0.5, xmin = first, xmax=last) + geom_rect()
-
 loc_view.dt <- loc_user_duration.censored.dt[,
   list(
     unique_users = length(unique(user_id)),
@@ -51,10 +55,18 @@ user_view.dt <- loc_user_duration.censored.dt[,
 		unique_locs = length(unique(location_id)),
 		total_login_time = sum(logout - login),
 		mean_login = mean(logout - login),
-		sd_login = sd(logout - login)
+		sd_login = sd(logout - login),
+    first = min(login),
+    last = max(logout)
 	),
 	by="user_id"
 ]
+
+user_view.dt[, login_proportion := total_login_time / (last-first) ]
+
[…]

### 2015-04-07 rscripts/data_review.Rmd
implementing bus method one

@@
 ```{r, echo=FALSE}
-src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ")
+src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ", colClasses = c("integer","integer","integer","integer"))
 setnames(src.dt, c("user_id", "location_id", "login", "logout"))
@@
 }
+censor.dt
+```
+
+Relabel source data:
+
+1. re-assign user ids from 1, ordered by first appearance
+2. re-assign location ids from 1, ordered by first appearance
+3. TODO split entries that cross days (~ 3% of data; of crossing entries, most look like late PM->early AM sessions, aka bar / club visits)
+4. convert login / logout times to `log(in|out)_day` and `log(in|out)_time`
+
+```{r, echo=FALSE}
+censor.dt[,
+  user_id := .GRP, by=user_id
+][,
+  location_id := .GRP, by=location_id
+][,
+  login_day := login %/% (24*60*60)
+][,
+  logout_day := logout %/% (24*60*60)
+][,
+  login_time := login - login_day*24*60*60
+][,
+  logout_time := logout - logout_day*24*60*60
+]
+setkey(censor.dt, login, logout, user_id, location_id)
+save(censor.dt, file = "~/Dropbox/montreal/censored.Rdata")
+# censor.dt[,list(
+#   login_day = login %/% (24*60*60),
+#   logout_day = logout %/% (24*60*60),
+#   login_time = login %% (24*60*60)/(60*60),
+#   logout_time = logout %% (24*60*60)/(60*60)
+# ) ]
 ```
@@
 ```{r, echo=FALSE}
-# src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ")
-# setnames(src.dt, c("user_id", "location_id", "login", "logout"))
-# setkey(src.dt, login, logout, user_id, location_id)
-# 
-# max_real_duration <- 60*60*24 # at most, continuous login of a day
-# 
 # bad.list <- with(src.dt[(logout - login) > max_real_duration,],

### 2015-04-07 rscripts/detection_outline.R

@@ -1,0 +1,94 @@
+## detection schemes
+
+load("~/Dropbox/montreal/censored.Rdata")
+censor.dt[, run_id := 0][,sample_id := 0][, target := FALSE]
+setkeyv(censor.dt, c("run_id","sample_id", key(censor.dt)))
+samples.dt <- fread("../test-1.csv", colClasses = "integer")
+setnames(samples.dt, c("run_id", "sample_id", "user_id", "location_id", "login", "logout"))
+setkeyv(samples.dt, key(censor.dt))
+samples.dt[,
+  user_id := user_id + max(censor.dt$user_id)+1L
+][,
+  login_day := login %/% (24*60*60)
+][,
+  logout_day := logout %/% (24*60*60)
+][,
+  login_time := login - login_day*24*60*60
+][,
+  logout_time := logout - logout_day*24*60*60
+][,
+  login_day := login_day + min(censor.dt$login_day)
+][,
+  logout_day := logout_day + min(censor.dt$login_day)
+][,
+  target := TRUE
+]
+
+meld.dt <- rbind(censor.dt, samples.dt, use.names = TRUE)
+
+synthesize <- function(rid, sid, combo.dt = meld.dt) {
+  res.dt <- combo.dt[
+    ((run_id == 0) & (sample_id == 0)) | ((run_id == rid) & (sample_id == sid)),
+    list(user_id, location_id, login_day, login_time, logout_day, logout_time, target)
+  ]
+  setkey(res.dt, login_day, login_time, logout_day, logout_time, user_id, location_id)
+  
+  res.dt[,
+    user_id := .GRP, by=user_id
+  ][,
+    location_id := .GRP, by=location_id
+  ]
+}
+
+view <- function(synthesized.dt) {
+  synthesized.dt[,list(user_id, location_id, login_day, login_time, logout_day, logout_time)]
+}
+
+asynth.dt <- synthesize(1, 1)
+aview.dt <- view(asynth.dt)
+
+method_1_slice <- function(dt, start, width) {
+  dt[(start <= login_day) & (login_day < start + width)]
+}
+
+method_1_target <- function(slice.dt)
+  slice.dt[,
+    list(login_count = .N),
+    by=list(user_id, login_day)
+  ][,
+    list(only_one = all(login_count == 1)),
+    by=user_id
+  ][only_one == T,
+    user_id
+  ]
+
[…]

### 2015-04-07 src/main/scala/edu/cap10/actormodels/demo/DataParams.scala

@@ -13,3 +13,7 @@ case class DataParams(
 )
 
 object MontrealProps extends DataParams(345, 356*5, 0.5, 5*60, 4)
+
+class LocationParams
+
+class AdvancedDataParams

### 2015-04-08 rscripts/detection_outline.R
fiddling a bit w/ bus method 1 approach
stub out 1a as well

@@ -62,14 +62,24 @@ method_1_target <- function(slice.dt)
     user_id
   ]
 
-method_1 <- function(dt, width, start = min(dt$login_day)+365, max_increments = 6, target_pop = 50) {
+method_1 <- function(dt, width, start = min(dt$login_day)+4*365, max_increments = 6, target_pop = 25) {
   i <- 0
   target_users <- method_1_target(method_1_slice(dt, start+i*width, width))
   while((i < max_increments) & (length(target_users) > target_pop)) {
     i <- i+1
     target_users <- intersect(target_users, method_1_target(method_1_slice(dt, start+i*width, width)))
   }
-  target_users
+  list(targets=target_users, inc=i)
+}
+
+method_1a <- function(dt, width, start = min(dt$login_day)+4*365, max_increments = 6, target_pop = 25) {
+  i <- 0
+  target_users <- method_1_target(method_1_slice(dt, start+i*width, width))
+  while((i < max_increments) & (length(target_users) > target_pop)) {
+    i <- i+1
+    target_users <- intersect(target_users, method_1_target(method_1_slice(dt, start+i*width, width)))
+  }
+  list(targets=target_users, inc=i)
 }
 
 for (rid in meld.dt[run_id != 0, unique(run_id)]) for (sid in meld.dt[run_id == rid, unique(sample_id)]) {
@@ -77,14 +87,16 @@ for (rid in meld.dt[run_id != 0, unique(run_id)]) for (sid in meld.dt[run_id == 
   targets <- synth.dt[target == T,]
   view.dt <- view(synth.dt)
   target_14 <- method_1(view.dt, 14)
-  target_21 <- method_1(view.dt, 21)
-  target_28 <- method_1(view.dt, 28)
+  covert_targets <- target_14$targets
+#  target_21 <- method_1(view.dt, 21)
+#  target_28 <- method_1(view.dt, 28)
   misses_ref <- targets[,length(unique(user_id))]
-  hits <- targets[user_id %in% target_14, length(unique(user_id))]
+  hits <- targets[user_id %in% covert_targets, length(unique(user_id))]
   misses <- misses_ref - hits
-  collateral <- length(target_14)-hits
+  collateral <- length(covert_targets)-hits
   print("14:")
   print(c(
+    steps = target_14$inc,
     run_id = rid,
     sample_id = sid,
     hits=hits,

### 2015-04-09 rscripts/data_review.Rmd
re-arrange project, introduce arg parser

@@
 ```{r, echo=FALSE}
-src.dt <- fread("~/Dropbox/montreal/merged.o", header = F, sep=" ", colClasses = c("integer","integer","integer","integer"))
+src.dt <- fread("../input/merged.o", header = F, sep=" ", colClasses = c("integer","integer","integer","integer"))
 setnames(src.dt, c("user_id", "location_id", "login", "logout"))
@@
 
-Trim data by:
+Trim data:
 
-1. sessions exceeding `r max_hours` hours
+1. remove sessions exceeding `r max_hours` hours
 2. eliminate users that have fewer than `r min_logins` logins
@@
 
+Removed `r dim(src.dt)[1] - dim(censor.dt)[1]` rows, or `r (dim(src.dt)[1] - dim(censor.dt)[1])/dim(src.dt)[1]` reduction.
+
 Relabel source data:
@@
 4. convert login / logout times to `log(in|out)_day` and `log(in|out)_time`
+5. mark data as the empirical source: set `run_id = 0`, `sample_id = 0`, `target = FALSE`.
 
@@
   logout_time := logout - logout_day*24*60*60
+][,
+  run_id := 0
+][,
+  sample_id := 0
+][,
+  target := FALSE
 ]
-setkey(censor.dt, login, logout, user_id, location_id)
-save(censor.dt, file = "~/Dropbox/montreal/censored.Rdata")
-# censor.dt[,list(
-#   login_day = login %/% (24*60*60),
-#   logout_day = logout %/% (24*60*60),
-#   login_time = login %% (24*60*60)/(60*60),
-#   logout_time = logout %% (24*60*60)/(60*60)
-# ) ]
+setkey(censor.dt, run_id, sample_id, login_day, login_time, logout_day, logout_time, user_id, location_id)
+save(censor.dt, file = "../input/censored.Rdata")
 ```

### 2015-04-09 rscripts/detection_outline.R

@@ -1,8 +1,8 @@
 ## detection schemes
+require(argparser)
 
-load("~/Dropbox/montreal/censored.Rdata")
-censor.dt[, run_id := 0][,sample_id := 0][, target := FALSE]
-setkeyv(censor.dt, c("run_id","sample_id", key(censor.dt)))
+
+load("../input/censored.Rdata")
 samples.dt <- fread("../test-1.csv", colClasses = "integer")
 setnames(samples.dt, c("run_id", "sample_id", "user_id", "location_id", "login", "logout"))
 setkeyv(samples.dt, key(censor.dt))

### 2015-04-09 rscripts/data_review.Rmd
a little more DRY, reorg

@@
 ```{r, echo=FALSE}
-censor.dt[,
+breakoutDays <- function(dt) {
+  dt[,
     user_id := .GRP, by=user_id
@@
     logout_time := logout - logout_day*24*60*60
+  ]
+}
+breakoutDays(censor.dt[,
+  user_id := .GRP, by=user_id
+][,
+  location_id := .GRP, by=location_id
 ][,
@@
   target := FALSE
-]
+])
 setkey(censor.dt, run_id, sample_id, login_day, login_time, logout_day, logout_time, user_id, location_id)
-save(censor.dt, file = "../input/censored.Rdata")
+save(censor.dt, breakoutDays, file = "../input/censored.Rdata")
 ```

### 2015-04-09 rscripts/detection_outline.R

@@ -1,28 +1,20 @@
 ## detection schemes
 require(argparser)
+a.parser <- arg.parser("Run Detection approaches against synthetic data", "detector")
+a.parser <- add.argument(a.parser, "--src", "source data", default = "../input/censored.Rdata")
+a.parser <- add.argument(a.parser, "target", "target synthetic data", type="character")
+a.parser <- add.argument(a.parser, "--out", "the output file; will default to out-$tar", type="character")
+argv <- parse.args(a.parser, c("../output/test-1.csv"))
 
-
-load("../input/censored.Rdata")
-samples.dt <- fread("../test-1.csv", colClasses = "integer")
+load(argv$src)
+samples.dt <- fread(argv$target, colClasses = "integer")
 setnames(samples.dt, c("run_id", "sample_id", "user_id", "location_id", "login", "logout"))
-setkeyv(samples.dt, key(censor.dt))
-samples.dt[,
+breakoutDays(samples.dt[,
   user_id := user_id + max(censor.dt$user_id)+1L
 ][,
-  login_day := login %/% (24*60*60)
-][,
-  logout_day := logout %/% (24*60*60)
-][,
-  login_time := login - login_day*24*60*60
-][,
-  logout_time := logout - logout_day*24*60*60
-][,
-  login_day := login_day + min(censor.dt$login_day)
-][,
-  logout_day := logout_day + min(censor.dt$login_day)
-][,
   target := TRUE
-]
+])
+setkeyv(samples.dt, key(censor.dt))
 
 meld.dt <- rbind(censor.dt, samples.dt, use.names = TRUE)
 

### 2015-04-19 rscripts/data_review.Rmd
extract day function and trimming into separate script

--- 2015-04-09 rscripts/data_review.Rmd
@@
 ```{r, echo=FALSE}
-censor.dt[,
+source("munging.R")
+breakoutDays(censor.dt[,
   user_id := .GRP, by=user_id
@@
   location_id := .GRP, by=location_id
-][,
-  login_day := login %/% (24*60*60)
-][,
-  logout_day := logout %/% (24*60*60)
-][,
-  login_time := login - login_day*24*60*60
-][,
-  logout_time := logout - logout_day*24*60*60
 ][,
@@
   target := FALSE
-]
+])
 setkey(censor.dt, run_id, sample_id, login_day, login_time, logout_day, logout_time, user_id, location_id)
-save(censor.dt, file = "../input/censored.Rdata")
+
+```
+
+When generating our synthetic data, we assume locations exist before and after they join the service.
+So the synthetic population may make visits to these locations outside of time when the service is operating.
+We trim that activity from the synthetic data according the time that locations are active in the observed data.
+
+```{r, echo=FALSE}
+limits.dt <- censor.dt[,list(first=min(login), last=max(logout)), keyby=location_id]
+limits.dt
+save(censor.dt, limits.dt, file = "../input/censored.Rdata")
 ```

### 2015-04-19 rscripts/detection_outline.R

@@ -1,6 +1,7 @@
 ## detection schemes
 req.packages <- c("data.table", "argparser", "ggplot2")
 sapply(req.packages, require, character.only = T)
+source("munging.R")
 
 a.parser <- arg.parser("Run Detection approaches against synthetic data", "detector")
 a.parser <- add.argument(a.parser, "--src", "source data", default = "../input/censored.Rdata")
@@ -15,18 +16,14 @@ breakoutDays(samples.dt[,
   user_id := user_id + max(censor.dt$user_id)+1L
 ][,
   target := TRUE
-])[, login_day := login_day + min(censor.dt$login_day)
-  ][,
+])[, login_day  := login_day  + min(censor.dt$login_day)
+][,
      logout_day := logout_day + min(censor.dt$login_day)
 ]
 setkeyv(samples.dt, key(censor.dt))
 
-limits <- censor.dt[,list(first=min(login), last=max(logout)), keyby=location_id]
-samp.dt <- merge(samples.dt, limits, by = "location_id")
-samp.dt <- samp.dt[(login >= first) & (logout <= last),]
-samp.dt$first <- samp.dt$last <- NULL
+samp.dt <- trimLimits(samples.dt, limits.dt)
 samp.dt[, user_id := user_id + max(censor.dt$user_id)]
-setkeyv(samp.dt, key(samples.dt))
 
 stopifnot(
   samples.dt[,
@@ -43,6 +40,16 @@ method_1_eval <- function(dt, start, width)
     list(atmost_one = all(login_count <= 1), seen = any(login_count > 0)),
     by=user_id
   ]
+
+method_1a_eval <- function(dt, start, width)
+  dt[(start <= login_day) & (login_day < start + width),
+     list(login_count = .N),
+     by=list(user_id)
+  ][,
+    list(exactly_one = (login_count == width)),
+    by=user_id
+  ]
+
 
 method_1 <- function(dt, width = 14, start = 4*365, max_increments = 6) {
   init.dt <- dt[,list(atmost_one = T, seen = F), by=user_id]
@@ -62,24 +69,56 @@ method_1 <- function(dt, width = 14, start = 4*365, max_increments = 6) {
   res.dt[, inc := .I-1]
 }
 
-# method_1a <- function(dt, width, start = min(dt$login_day)+4*365, max_increments = 6, target_pop = 25) {
-#   i <- 0
[…]

### 2015-04-19 rscripts/munging.R

@@ -1,0 +1,24 @@
+## data munging utils
+require(data.table)
+breakoutDays <- function(dt) {
+  dt[,
+    user_id     := .GRP, by=user_id
+  ][,
+    location_id := .GRP, by=location_id
+  ][,
+    login_day   := login %/% (24*60*60)
+  ][,
+    logout_day  := logout %/% (24*60*60)
+  ][,
+    login_time  := login - login_day*24*60*60
+  ][,
+    logout_time := logout - logout_day*24*60*60
+  ]
+}
+
+trimLimits <- function(tar.dt, lim.dt) {
+  res.dt <- merge(tar.dt, lim.dt, by = "location_id")[(login >= first) & (logout <= last),]
+  res.dt$first <- res.dt$last <- NULL
+  setkeyv(res.dt, key(tar.dt))
+  res.dt
+}

### 2015-04-30 src/main/scala/edu/cap10/util/NaturalNumber.scala
tweak utilities

@@ -11,7 +11,7 @@ object NaturalInt {
   
   implicit def convertItoN(i:Int) = this(i)
   implicit def convertNtoI(n:NaturalInt) = n.underlying
-  implicit def ordering : Ordering[NaturalInt] = Ordering[Int].on(n => n.underlying)
+  implicit def ordering[NI <: NaturalInt] : Ordering[NI] = Ordering[Int].on(n => n.underlying)
   
 }
 

### 2015-04-30 src/main/scala/edu/cap10/util/TimeStamp.scala

@@ -3,8 +3,8 @@ package edu.cap10.util
 case class TimeStamp(hour:Int, min:Int, sec:Int) {
 
   require(0 <= hour && hour < 24, f"hour $hour must be [0,24).")
-  require(0 <= min && min < 60, f"hour $hour must be [0,60).")
-  require(0 <= sec && sec < 60, f"hour $hour must be [0,60).")
+  require(0 <= min && min < 60, f"min $min must be [0,60).")
+  require(0 <= sec && sec < 60, f"sec $sec must be [0,60).")
   
   override val toString = f"$hour%02d:$min%02d:$sec%02d"
  

### 2015-04-30 src/main/scala/edu/cap10/actormodels/demo/Locations.scala
draft locations object
eventually need these to support different kinds of locations

@@ -1,0 +1,5 @@
+package edu.cap10.actormodels.demo
+
+import edu.cap10.util.TimeStamp
+
+case class Locations(id:LocationID, open:TimeStamp = TimeStamp(8,0,0), close:TimeStamp = TimeStamp(17,0,0))

### 2015-04-30 src/main/scala/edu/cap10/actormodels/demo/HauntGenerators.scala
revisions to haunt gen

@@ -1,10 +1,12 @@
 package edu.cap10.actormodels.demo
 
 import scala.util.Random
+import edu.cap10.util.NaturalInt
+import scala.languageFeature.implicitConversions
 
 case class HauntGenerator(rng : Random) {
 
-  def uniform(locations : Seq[Int]) : (Int) => Seq[Int] = {
+  def uniform(locations : Seq[LocationID]) : (Int) => Seq[LocationID] = {
     require(locations.size > 0, "Cannot sample from non-existent locations.")
     (count : Int) => {
       require(count >= 0, "Cannot request a negative location count.")
@@ -13,6 +15,15 @@ case class HauntGenerator(rng : Random) {
     } ensuring(_.size == count)
   }
  
-  def distributed(locations: Seq[Int])(mean:Double) : Seq[Int] = ???
+  def distributed(locations: Seq[LocationID], min : Int = 1) : (Double) => Seq[LocationID] = {
+    require(locations.size > 0, "Cannot sample from non-existent locations.")
+    (mean : Double) => {
+      require(mean >= min, "Cannot request a mean number of locations less than the min "+min)
+      require(mean < locations.size, "The expected number of locations can be the total number of locations.")
+      val p = (mean - min) / (locations.size - min)
+      val (keep, possible) = rng.shuffle(locations).splitAt(min)
+      keep ++ possible.filter { _ => rng.nextDouble < p }
+    } ensuring(_.size >= min)
+  }
   
 }

### 2015-04-30 src/main/scala/edu/cap10/actormodels/demo/SimpleAgent.scala
update SimpleX for other code changes

@@ -2,6 +2,22 @@ package edu.cap10.actormodels.demo
 
 import edu.cap10.util.Probability
 import edu.cap10.util.LocalRNG
+import akka.actor.{ TypedActor, TypedActorFactory, TypedProps}
+
+object SimpleAgent {
+  
+  def props(
+    id : AgentID,
+    haunts : Seq[LocationID],
+    runConfig : SimpleParams,
+    globalConfig : DataParams,
+    seed : Long
+  ) = TypedProps(
+    classOf[Dispatchable[TravelEvent]],
+    new SimpleAgent(id, haunts, globalConfig.dailyVisitProb, 60*60, seed)
+  )
+  
+}
 
 /*An agent that
  * - can be dispatched, and does exactly as dispatched
@@ -23,7 +39,8 @@ class SimpleAgent (
   override def _tick(when:Int) =
     if (_dispatched || !makeVisit)
       super._tick(when)
-    else 
+    else {
       TravelEvent.random(id, haunts, meanVisitDuration.toInt) +: super._tick(when)
+    }
 
 }

### 2015-04-30 src/main/scala/edu/cap10/actormodels/demo/SimpleMain.scala

@@ -23,12 +23,13 @@ object SimpleMain extends App {
     case Some(config) => {    
       val results = for (res <- SimpleSystem(config, MontrealProps).run()) yield { 
         res map { case (day, events) => events map { _ + day } }
+      } map { 
+        dailyEvents =>
+          stdout.println(dailyEvents mkString System.lineSeparator)
+          stdout.flush
       }
       results.onComplete { _ match {
-        case Success(output) => {
-          stdout.println((output.flatten) mkString System.lineSeparator)
-          stdout.flush()
-        }
+        case Success(output) => stdout.flush
         case Failure(e) => stderr.println(e)
       } }
       Await.result(results, Duration(30, SECONDS))

### 2015-04-30 src/main/scala/edu/cap10/actormodels/demo/SimpleUniverse.scala

@@ -33,16 +33,24 @@ object SimpleUniverse {
       )
   
   def createAgents(
-    seeds:Seq[Long],
+    count:NaturalInt,
     runConfig: SimpleParams,
     globalConfig : DataParams
-  )(implicit rng : Random) : Seq[Dispatchable[TravelEvent]] = {
-    implicit val sys = TypedActor.get(TypedActor.context)
-    import runConfig._
-    import globalConfig._
-    seeds.zipWithIndex map { case (agentSeed, id) => 
-      val haunts = rng.shuffle(1 to uniqueLocations).take(meanVisitedLocations.toInt)
-      createAgent(id, haunts, dailyVisitProb, meanVisitDuration, seed)
+  )(
+    implicit rng : Random,
+    sys : TypedActorFactory = TypedActor.get(TypedActor.context)
+  ) : Seq[Dispatchable[TravelEvent]] = {
+    import runConfig._, globalConfig._
+    val hauntGen = HauntGenerator(rng).uniform(1 to uniqueLocations)
+    (0 until count) map { id =>
+      val haunts = hauntGen(meanVisitedLocations.toInt)
+      val agentSeed = rng.nextLong
+      sys.typedActorOf(
+        SimpleAgent.props(
+            id, haunts, runConfig, globalConfig, agentSeed 
+        ),
+        "agent"+id
+      )
     }
   }
 }
@@ -56,18 +64,14 @@ class SimpleUniverse(
   import runConfig.{seed => _, _}, globalConfig._, SimpleUniverse._
 
   val meetingGenerator = PoissonGenerator(meanMeetingPeriod)
-  import rng.{ shuffle, nextLong => newSeed }
-  import meetingGenerator.{ next => daysToNextMeeting }
+  import rng.shuffle, meetingGenerator.{ next => daysToNextMeeting }
   
-  val meetingLocations = shuffle((0 to (uniqueLocations-1))).take(locationCount)
-  val agents = createAgents(
-    replicate(agentCount)( newSeed ),
-    runConfig, globalConfig
-  )
+  val meetingLocations = HauntGenerator(rng).uniform(1 to uniqueLocations)(locationCount)
+  val agents = createAgents(agentCount, runConfig, globalConfig)
   
   var timeToNextMeeting : Int = daysToNextMeeting
   def timeToMeet : Boolean = timeToNextMeeting == 0
-
+  
   override def tick(when:Int) = {
     var dispatch : Future[Seq[Unit]] = Future.successful(Seq())
     

### 2015-04-30 src/main/scala/edu/cap10/actormodels/demo/LingeringAgent.scala
stub advanced agent behavior

@@ -1,0 +1,46 @@
+package edu.cap10.actormodels.demo
+
+import edu.cap10.util.Probability
+import edu.cap10.util.LocalRNG
+import akka.actor.{ TypedActor, TypedActorFactory, TypedProps}
+
+object LingeringAgent {
+  
+  def props(
+    id : AgentID,
+    haunts : Seq[LocationID],
+    runConfig : SimpleParams,
+    globalConfig : DataParams,
+    seed : Long
+  ) = TypedProps(
+    classOf[Dispatchable[TravelEvent]],
+    new LingeringAgent(id, haunts, globalConfig.dailyVisitProb, 60*60, seed)
+  )
+  
+}
+
+/*An agent that
+ * - can be dispatched, and does exactly as dispatched
+ * - if not dispatched, may visit somewhere on its own, with uniform random behavior
+ * */
+class LingeringAgent (
+    id:AgentID,
+    haunts:Seq[LocationID],
+    dailyVisitProbability:Probability,
+    meanVisitDuration:Double, // seconds
+    override val seed:Long
+) extends Dispatchable[TravelEvent] with LocalRNG { 
+  
+  private def makeVisit = rng.nextDouble < dailyVisitProbability
+  
+  override def _dispatch(te:TravelEvent) = 
+    super._dispatch(te.copy(agentID = id))
+
+  override def _tick(when:Int) =
+    if (_dispatched || !makeVisit)
+      super._tick(when)
+    else {
+      TravelEvent.random(id, haunts, meanVisitDuration.toInt) +: super._tick(when)
+    }
+
+}

### 2015-04-30 rscripts/data_review.Rmd
revising data review

@@
   location_id := .GRP, by=location_id
-][,
-  run_id := 0
-][,
-  sample_id := 0
-][,
-  target := FALSE
 ])
-setkey(censor.dt, run_id, sample_id, login_day, login_time, logout_day, logout_time, user_id, location_id)
-
+setkey(censor.dt, login_day, login_time, logout_day, logout_time, user_id, location_id)
+censor.dt
 ```
@@
 ```{r, echo=FALSE}
+## TODO switch to log(in|out) day
 limits.dt <- censor.dt[,list(first=min(login), last=max(logout)), keyby=location_id]
@@
 
+The locations have different patterns of activity, reflecting their different hours of operation.
+
 ```{r, echo=FALSE}
+hours.dt <- censor.dt[,
+  list(login_day, login_time, logout_day, logout_time, logout_time_adj=0),
+  keyby=list(location_id, user_id)
+]
+hours.dt[ login_day != logout_day,
+   logout_time_adj := logout_time + 24*60*60
+]
+hours.dt <- hours.dt[,
+  list(open=min(login_time), close=max(logout_time, logout_time_adj)),
+  by=list(login_day, location_id)
+]
+p <- ggplot(hours.dt[,
+  list(med_open = median(open), med_close=median(close)), by=location_id
+][, open_id:=.GRP, by=list(med_open, location_id)])+aes(x=open_id, y=med_open/3600) + geom_point() + geom_hline(yintercept=1:23, color="red") + theme_bw()
 # bad.list <- with(src.dt[(logout - login) > max_real_duration,],

### 2015-04-30 rscripts/detection_outline.R

@@ -10,6 +10,9 @@ a.parser <- add.argument(a.parser, "--out", "the output file; will default to ou
 argv <- parse.args(a.parser, c("../output/runs-1.csv"))
 
 load(argv$src)
+
+censor.dt[, run_id := 0 ][, sample_id := 0 ][, target := FALSE ]
+
 samples.dt <- fread(argv$target, colClasses = "integer")
 setnames(samples.dt, c("run_id", "sample_id", "user_id", "location_id", "login", "logout"))
 breakoutDays(samples.dt[,

### 2015-05-21 src/main/scala/edu/cap10/actormodels/demo/CovertAgent.scala
grammar

@@ -5,7 +5,7 @@ import edu.cap10.util.Probability
 import edu.cap10.util.Probability._
 import edu.cap10.util.TimeStamp
 import edu.cap10.util.TimeStamp._
-import scala.languageFeature.implicitConversions
+import scala.language.implicitConversions
 
 trait CovertAgent extends Dispatchable[TravelEvent] with LocalRNG {
 

### 2015-05-21 src/test/scala/edu/cap10/actormodels/demo/TimeStampTests.scala
give up on property checks for now

@@ -1,96 +1,27 @@
 package edu.cap10.actormodels.demo
 
-import org.scalatest.{
-  FunSuite,
-  BeforeAndAfter
-}
-import org.scalatest.prop.PropertyChecks
-
-import scala.util.Random
-import scala.language.implicitConversions
+import org.scalatest.FunSuite
 
 import edu.cap10.util.TimeStamp
 import edu.cap10.util.TimeStamp._
 
-class TimeStampTests extends FunSuite with BeforeAndAfter with PropertyChecks {
+class TimeStampTests extends FunSuite {
  
-  val seed : Long = 10
-  var rng : Seq[Random] = _
+  // TODO get scalacheck working 
   
-  test("TimeStamp Hour should be between 0 and 23") { 
-    forAll { (n:Int) => 
-      whenever(n < 0 || n > 23) {
-        intercept[IllegalArgumentException] { Hour(n) }
-      }
-    }
+  test("timestamp hours must be between 0 and 23") {
+    intercept[IllegalArgumentException] { Hour(-1) }
+    intercept[IllegalArgumentException] { Hour(24) }
   }
   
-  test("TimeStamp Minute should be between 0 and 59") { 
-    forAll { (n:Int) => 
-      whenever(n < 0 || n > 59) {
-        intercept[IllegalArgumentException] { Minute(n) }
-      }
-    }
+  test("timestamp minutes must be between 0 and 59") {
+    intercept[IllegalArgumentException] { Minute(-1) }
+    intercept[IllegalArgumentException] { Minute(60) }
   }
   
-  test("TimeStamp Second should be between 0 and 59") { 
-    forAll { (n:Int) => 
-      whenever(n < 0 || n > 59) {
-        intercept[IllegalArgumentException] { Second(n) }
-      }
-    }
+  test("timestamp seconds must be between 0 and 59") {
+    intercept[IllegalArgumentException] { Second(-1) }
+    intercept[IllegalArgumentException] { Second(60) }
   }
-//  val locations : IndexedSeq[Int] = 0 until 100
-//  var HG : Seq[HauntGenerator] = _
-//  
-//  before {
-//    rng = List(seed, seed, seed*2) map { new Random(_) }
-//    HG = rng map { HauntGenerator(_, locations) }
-//  }
-//  
-//  test("properly set up random number generators") {
-//    assert(rng(0) !== rng(1), "rng(0) vs rng(1) refs are incorrectly equal.")
[…]

### 2015-05-21 rscripts/data_review.Rmd
exploring population more

@@
 ```{r, echo=FALSE}
-src.dt <- fread("../input/merged.o", header = F, sep=" ", colClasses = c("integer","integer","integer","integer"))
-setnames(src.dt, c("user_id", "location_id", "login", "logout"))
-setkey(src.dt, login, logout, user_id, location_id)
+src.dt <- 
+  setkey(
+    setnames(
+      fread("../input/merged.o", header = F, sep=" ", colClasses = list(integer=1.4)),
+      c("user_id", "location_id", "login", "logout")
+    ),
+    login, logout, user_id, location_id
+  )
 src.dt
@@
 1. remove sessions exceeding `r max_hours` hours
-2. eliminate users that have fewer than `r min_logins` logins
-3. eliminate locations that have fewer than `r min_logins` logins
-4. recursively apply steps 2-3 until no users / locations eliminated
+2. remove sessions with no time (these correspond to rows in the source data with no logout time)
+3. eliminate users that have fewer than `r min_logins` logins
+4. eliminate locations that have fewer than `r min_logins` logins
+5. recursively apply steps 2-3 until no users / locations eliminated
 
 ```{r, echo=FALSE}
-censor.dt <- src.dt[(logout - login) <= max_hours*60*60, ]
+censor.dt <- src.dt[(logout != login) & ((logout - login) <= max_hours*60*60), ]
 invalid.users <- censor.dt[,.N, by=user_id ][N < min_logins, user_id]
@@
 source("munging.R")
-breakoutDays(censor.dt[,
+setkey(breakoutDays(censor.dt[,
   user_id := .GRP, by=user_id
@@
   location_id := .GRP, by=location_id
-])
-setkey(censor.dt, login_day, login_time, logout_day, logout_time, user_id, location_id)
+]), login_day, login_time, logout_day, logout_time, user_id, location_id)
 censor.dt
@@
 save(censor.dt, limits.dt, file = "../input/censored.Rdata")
+```
+
+Our synthetic population comprises long term users of the system.  When synthezing their behavior, we should be thinking relative to other long term users of the system.
+
+```{r, echo=FALSE}
+reference.dt <- merge(censor.dt, censor.dt[,list(tot = (max(logout) - min(login))/(365*24*60*60)), by=user_id][tot > 1, list(user_id = user_id)], by="user_id")
+setkeyv(reference.dt, key(censor.dt))
+ref.dt <- reference.dt[,
+  list(
+    overnight_logins = sum(login_day != logout_day),
+    non_overnight_logins = sum(login_day == logout_day),
+    mean_session_min = mean(logout-login)/60,
+    total_session_min = sum(logout-login)/60,
+    user_lifetime_days = max(logout_day) - min(login_day)
+), by=user_id][,total_logins := overnight_logins + non_overnight_logins]
+hist(ref.dt$overnight_logins/c(ref.dt$overnight_logins + ref.dt$non_overnight_logins))
+hist(ref.dt$mean_session_min)
+fitting <- function(runs, p_continue) sum(rnbinom(n=runs, size=1, prob=1-p_continue))/runs
+with(ref.dt, {
+  plot(total_logins, mean_session_min)
+  points(total_logins, sapply(total_logins, fitting, p_continue = 0.99), pch=".", col = "red")
+})
+
+```
+
+```{r, echo=FALSE}
+
+plot()
 ```

### 2015-05-22 rscripts/data_review.Rmd
sorting out the view of reference individuals

@@
 ), by=user_id][,total_logins := overnight_logins + non_overnight_logins]
-hist(ref.dt$overnight_logins/c(ref.dt$overnight_logins + ref.dt$non_overnight_logins))
-hist(ref.dt$mean_session_min)
-fitting <- function(runs, p_continue) sum(rnbinom(n=runs, size=1, prob=1-p_continue))/runs
-with(ref.dt, {
-  plot(total_logins, mean_session_min)
-  points(total_logins, sapply(total_logins, fitting, p_continue = 0.99), pch=".", col = "red")
-})
+setkey(ref.dt, total_logins)
+```
 
+```{r,echo=FALSE}
+sess_scale_min <- log10(c(10, 60, 2*60, 24*60, 7*24*60, 30*24*60, 6*30*24*60, 365*24*60))
+sess_labels <- c("10 min", "1 hr", "2 hr", "1 day", "1 week", "1 month", "6 months", "1 year")
+
+use_period_scale_day <- log10((1:5)*365)
+use_period_labels <- c(1,2,3,4,5)
+
+ggplot(ref.dt) + theme_bw() + 
+  aes(x=log10(user_lifetime_days), y=log10(total_session_min), color=log10(total_logins)) + 
+  geom_point() +
+  scale_y_continuous("Total Session Time", breaks=sess_scale_min, labels=sess_labels) +
+  scale_x_continuous("Usage Period (Years)", breaks=use_period_scale_day, labels=use_period_labels) +
+  scale_color_continuous("Unique Logins", breaks=log10(c(10,100,1000)), labels=c(10,100,1000))
 ```
@@
 ```{r, echo=FALSE}
+#with(ref.dt, {
 
-plot()
+#  hist(ref.dt$overnight_logins/c(ref.dt$overnight_logins + ref.dt$non_overnight_logins))
+#hist(ref.dt$mean_session_min)
+#fitting <- function(runs, p_continue) sum(rnbinom(n=runs, size=1, prob=1-p_continue))/runs
+
+#  plot(user_lifetime_days, total_logins, pch=".", alpha=0.5)
+#  plot(total_logins, mean_session_min)
+#  points(total_logins, sapply(total_logins, fitting, p_continue = 0.99), pch=".", col = "red")
+#})
 ```

### 2015-06-10 rscripts/data_review.Rmd
updates to data_review

--- 2015-05-21 rscripts/data_review.Rmd
@@
 
-Our synthetic population comprises long term users of the system.  When synthezing their behavior, we should be thinking relative to other long term users of the system.
+Our synthetic population comprises long term users of the system.
+When synthezing their behavior, we should be thinking relative to other long term users of the system.
+We define *long term* as having greater than six months between first and last observed use of the system by a particular user id.
 
-```{r, echo=FALSE}
-reference.dt <- merge(censor.dt, censor.dt[,list(tot = (max(logout) - min(login))/(365*24*60*60)), by=user_id][tot > 1, list(user_id = user_id)], by="user_id")
-setkeyv(reference.dt, key(censor.dt))
-ref.dt <- reference.dt[,
+For long term users, we are interested in a few first-order behaviors: how much of their time are they using the system, and how is the time spent using the system broken up into sessions?
+
+```{r}
+reference.dt <-
+  setkey(censor.dt[,
     list(
@@
       non_overnight_logins = sum(login_day == logout_day),
-    mean_session_min = mean(logout-login)/60,
       total_session_min = sum(logout-login)/60,
       user_lifetime_days = max(logout_day) - min(login_day)
-), by=user_id][,total_logins := overnight_logins + non_overnight_logins]
-hist(ref.dt$overnight_logins/c(ref.dt$overnight_logins + ref.dt$non_overnight_logins))
-hist(ref.dt$mean_session_min)
-fitting <- function(runs, p_continue) sum(rnbinom(n=runs, size=1, prob=1-p_continue))/runs
-with(ref.dt, {
-  plot(total_logins, mean_session_min)
-  points(total_logins, sapply(total_logins, fitting, p_continue = 0.99), pch=".", col = "red")
-})
-
+    ), by=user_id
+  ][
+    user_lifetime_days > 365/2
+  ][,
+    total_logins := overnight_logins + non_overnight_logins
+  ], total_logins)
 ```
 
+First, we survey the relationship between total time spent on the system, period of use, and total unique logins:
+
+```{r,echo=FALSE}
+sess_scale_min <- log10(c(10, 60, 2*60, 24*60, 7*24*60, 30*24*60, 6*30*24*60, 365*24*60))
+sess_labels <- c("10 min", "1 hr", "2 hr", "1 day", "1 week", "1 month", "6 months", "1 year")
+
+use_period_scale_day <- log10(c(0.5,1:5)*365)
+use_period_labels <- c(0.5,1:5)
+
+base.plot <- ggplot(reference.dt) + theme_bw() + 
+  aes(
+    color=log10(user_lifetime_days),
+    y=log10(total_session_min),
+    x=log10(total_logins)
+  ) + 
+  scale_y_continuous(
+    "Cumulative Usage",
+    breaks=sess_scale_min,
+    labels=sess_labels
+  ) +
+  scale_color_continuous(
+    "Usage Period (Years)",
+    breaks=use_period_scale_day,
+    labels=use_period_labels
+  ) +
+  scale_x_continuous(
+    "Unique Logins",
+    breaks=log10(c(10,100,1000)),
+    labels=c(10,100,1000)
+  ) +
+  geom_point() +
+  stat_smooth(method=lm, color="red")
+base.plot
+```
+
+Visual inspection indicates that there is a log-linear dependency of total session time, $\sum{t_i}$, on login counts, $n_l$, and the period of activity seems to explain little.
+We can check this with a linear regression:
+
 ```{r, echo=FALSE}
+user.model <- lm(
+  log10(total_session_min) ~ log10(user_lifetime_days) + log10(total_logins),
+  reference.dt
+)
+summary(user.model)
+```
 
-plot()
+This confirms that usage period is not substantively driving total session time - unique logins are.
+However, is usage period driving the total lifetime logins?
+Certainly that would seem reasonable - the longer someone uses the system, the more login opportunities they have.
+
+```{r, echo=FALSE}
+logins.plot <- ggplot(reference.dt) + theme_bw() + 
+  aes(
+    x=log10(user_lifetime_days),
+    y=log10(total_logins)
+  ) + 
+  scale_x_continuous(
+    "Usage Period (Years)",
+    breaks=use_period_scale_day,
+    labels=use_period_labels
+  ) +
+  scale_y_continuous(
+    "Unique Logins",
+    breaks=log10(c(10,100,1000)),
+    labels=c(10,100,1000)
+  ) +
+  geom_point(alpha=0.3, size=3) +
+  stat_smooth(method=lm, color="red")
+logins.plot
 ```
+
+There seems to be a steady trend:
+
+```{r, echo=FALSE}
+logins.model <- lm(
+  log10(total_logins) ~ log10(user_lifetime_days),
+  reference.dt
+)
+summary(logins.model)
+```
+
+However, little of the variation is explained.
+By visual inspection, the variation seems to remain roughly constant with the increasing trend.
+Failing to have predictive terms for that variance is not a problem, as long as we are able to replicate that amount of variation in our synthetic population.
+
+Before we try some models to reproduce that, let's consider our user behavior a bit more closely, since the system itself is growing as time proceeds.
+Particularly, we want to consider whether the growth in number of available login sites influences the number of unique logins.
+
+Consider the number of available sites:
+
+```{r,echo=FALSE}
+site.dt <- setkey(melt(
+    censor.dt[,list(end=max(logout), start=min(login)), by=location_id],
+    id.var="location_id",
+    value.name="time",
+    variable.name = "event"),
+  "time")
+site.dt[,increment := ifelse(event=="end",-1,1)]
+site.dt[,count := cumsum(increment)]
+mintime <- site.dt$time[1]
+site.dt[, time := time - mintime]
+locations.plot <- ggplot(site.dt) + theme_bw() + aes(x=time/(365*24*60*60), y=count) +
+  geom_step() +
+  scale_x_continuous("year", limits=c(0,NA))
+locations.plot
+```
+
+From this view, we see a division of eras: a period of steady growth, a period of stable turnover, then another growth period terminating in the end of the data set (where locations that likely continue to exist are censored by the end of the data set).
+While we could do a more rigorous interpretation, this is a practical approximation, and we can use straightforward linear regression to fit three trend lines, assuming that the middle line should have zero slope.
+To curtail the terminal region in the data, we will simply identify where the trend is monotonically declining.
+
+```{r, echo=FALSE}
+grow1 <- function(break1, dt) sum(lm(count ~ time, dt[time < break1])$residuals^2)
+stable <- function(break1, break2, dt) sum(dt[(break1 <= time) & (time < break2), count - mean(count)]^2)
+grow2 <- function(break2, break3, dt) sum(lm(count ~ time, dt[(break2 <= time) & (time < break3)])$residuals^2)
+decline <- function(break3, dt) sum(lm(count ~ time, dt[break3 <= time])$residuals^2)
+partition <- function(x, maxt) {
+  ## x is logit of proportions remaining in the series
+  es <- exp(x)
+  bs <- es/(1+es)
+  break1 <- bs[1]*maxt
+  break2 <- (maxt-break1)*bs[2] + break1
+  break3 <- (maxt-break2)*bs[3] + break2
+  grow1(break1) + stable(break1, break2) + grow2(break2, break3) + decline(break3)
+}
+guessp <- c(0.5, 0.5, 0.9)
+res <- with(optim(log(guessp/(1-guessp)), partition, maxt=max(site.dt$time), dt=site.dt), {
+  es <- exp(par)
+  ps <- es/(1+es)
+  maxt <- max(site.dt$time)
+  break1 <- ps[1]*maxt
+  break2 <- (maxt-break1)*ps[2] + break1
+  break3 <- (maxt-break2)*ps[3] + break2
+  c(b1=break1, b2=break2, b3=break3)  
+})
+locations.plot + geom_vline(x=res/(365*24*60*60), color="red")
+```
+
+This means fitting three values: the slope in the growth region, the transition point, and the end point for stable region.
+
+If we propose some distributions of unique logins around an increasing mean number of logins, then we can determine via maximum likelihood tests (1) what shape parameters best capture the data, and (2) which distribution models the data best.
+There are a few features of the distribution to capture:
+
+ - it should be left censored.
+The outcomes are bound by $\log_{10}(2)=`r log10(2)`$ as the minimum number of logins (required to appear at least long enough to measure a lifetime in the dataset greater than a 6 month duration).
+
+First, let's consider the Poisson distribution, with the shape parameter $\lambda$ as the rate of new sessions per available location.
+That is, users of the system have an activity rate per location, so more locations.
+
+This implies that we may be able to create synthetic activity informed by the whole dataset, independently of how long a period (within the bounds of the actual empirical data) we are simulating.
+
+TODO: some simulation approaches, show those simulated people on that plot.
 

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/Entity.scala
initial agent spec

@@ -1,0 +1,31 @@
+package edu.cap10.actormodels.tb
+
+import edu.cap10.util.LocalRNG
+
+trait Entity extends LocalRNG {
+
+  // has state
+  // has (time until, state transition)
+  // receives ticks, possibly with infectious interactions
+  // replies with state transitions, infections generated
+  
+}
+
+//trait CovertAgent extends Dispatchable[TravelEvent] with LocalRNG {
+//
+//  protected def id : AgentID 
+//  
+//  final override def _dispatch(te:TravelEvent) = 
+//    super._dispatch( dispatchModifier(te).copy(agentID = id) )
+//
+//  final override def _tick(when:Int) =
+//    timesGenerator(locationSelector) ++: super._tick(when)
+//    
+//  protected def timesGenerator(locations: Seq[LocationID]) : Seq[TravelEvent]
+//  
+//  protected def dispatchModifier(te:TravelEvent) : TravelEvent
+//  
+//  protected def locationSelector : Seq[LocationID]
+//  
+//  
+//}

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/States.scala
create system states

@@ -1,0 +1,15 @@
+sealed class TBStrain
+case object Resistant extends TBStrain
+case object Sensitive extends TBStrain
+
+sealed class HostTBState
+case object Susceptible extends HostTBState
+case class Resistant(strain:TBStrain) extends HostTBState
+case class Exposed(strain:TBStrain) extends HostTBState
+case class Infectious(strain:TBStrain) extends HostTBState
+case class Treated(strain:TBStrain) extends HostTBState
+case class Chronic(strain:TBStrain) extends HostTBState
+
+sealed class HostHIVState
+case object Negative extends HostHIVState
+case class Positive(indicator:Double) extends HostHIVState

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala
first test failing

@@ -1,0 +1,12 @@
+package edu.cap10.actormodels.tb
+
+import TBStrain._
+
+case class InfectionModel(some:Double)
+  extends Function2[HostTBState, Seq[TBStrain], Option[HostTBState]]
+{
+  def apply(
+    hostTBstate : HostTBState,
+    infectiousContacts:Seq[TBStrain]
+  ) : Option[HostTBState] = ???
+}

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/States.scala

@@ -1,6 +1,11 @@
-sealed class TBStrain
-case object Resistant extends TBStrain
-case object Sensitive extends TBStrain
+package edu.cap10.actormodels.tb
+
+object TBStrain extends Enumeration {
+  type TBStrain = Value
+  val Resistant, Sensitive = Value
+}
+
+import TBStrain._
 
 sealed class HostTBState
 case object Susceptible extends HostTBState

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/package.scala

@@ -1,0 +1,5 @@
+package edu.cap10.actormodels
+
+package object tb {
+  type EntityID = Int
+}

### 2015-06-10 src/test/scala/edu/cap10/actormodels/tb/InfectionModelTests.scala

@@ -1,0 +1,27 @@
+package edu.cap10.actormodels.tb
+
+import org.scalatest.FunSuite
+
+import TBStrain.{values => strains, _}
+
+class InfectionModelTests extends FunSuite {
+
+  val allHostStates : Seq[HostTBState] = Seq(Susceptible) ++ (
+    strains map Resistant
+  ) ++ (
+    strains map Exposed
+  ) ++ (
+    strains map Infectious
+  ) ++ (
+    strains map Treated
+  ) ++ (
+    strains map Chronic
+  )
+
+  test("exposing hosts to no TB returns Option == None") {
+    val infModel = InfectionModel(0.5)
+    val res = allHostStates map { hs => infModel(hs, Seq.empty) }
+    assert(res.forall { _.isEmpty } === true)
+  }
+
+}

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala
handle non-infections properly

@@ -8,5 +8,8 @@ case class InfectionModel(some:Double)
   def apply(
     hostTBstate : HostTBState,
     infectiousContacts:Seq[TBStrain]
-  ) : Option[HostTBState] = ???
+  ) : Option[HostTBState] = infectiousContacts match {
+    case Seq() => None
+    case _ => None
+  }
 }

### 2015-06-10 src/test/scala/edu/cap10/actormodels/tb/InfectionModelTests.scala
(fake) pass handling exposing already infected hosts

@@ -6,7 +6,7 @@ import TBStrain.{values => strains, _}
 
 class InfectionModelTests extends FunSuite {
 
-  val allHostStates : Seq[HostTBState] = Seq(Susceptible) ++ (
+  val nonSusceptibleHosts = Seq() ++ (
     strains map Resistant
   ) ++ (
     strains map Exposed
@@ -14,13 +14,22 @@ class InfectionModelTests extends FunSuite {
     strains map Infectious
   ) ++ (
     strains map Treated
-  ) ++ (
+  )
+  
+  val allHostStates : Seq[HostTBState] = Seq(Susceptible) ++
+    nonSusceptibleHosts ++ (
     strains map Chronic
   )
-
+  
   test("exposing hosts to no TB returns Option == None") {
     val infModel = InfectionModel(0.5)
     val res = allHostStates map { hs => infModel(hs, Seq.empty) }
+    assert(res.forall { _.isEmpty } === true)
+  }
+  
+  test("exposing hosts that are Exposed, Resistant, or Infectious (with any strain) returns Option == None") {
+    val infModel = InfectionModel(0.5)
+    val res = nonSusceptibleHosts map { hs => infModel(hs, strains.toSeq) }
     assert(res.forall { _.isEmpty } === true)
   }
 

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/Entity.scala
infection right for 1.0 infectivity, draft entity

@@ -4,6 +4,8 @@ import edu.cap10.util.LocalRNG
 
 trait Entity extends LocalRNG {
 
+  protected def id : EntityID
+  
   // has state
   // has (time until, state transition)
   // receives ticks, possibly with infectious interactions
@@ -13,10 +15,8 @@ trait Entity extends LocalRNG {
 
 //trait CovertAgent extends Dispatchable[TravelEvent] with LocalRNG {
 //
-//  protected def id : AgentID 
+ 
 //  
-//  final override def _dispatch(te:TravelEvent) = 
-//    super._dispatch( dispatchModifier(te).copy(agentID = id) )
 //
 //  final override def _tick(when:Int) =
 //    timesGenerator(locationSelector) ++: super._tick(when)

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala

@@ -10,6 +10,9 @@ case class InfectionModel(some:Double)
     infectiousContacts:Seq[TBStrain]
   ) : Option[HostTBState] = infectiousContacts match {
     case Seq() => None
-    case _ => None
+    case _ => hostTBstate match {
+      case Susceptible => Some(Exposed(infectiousContacts.head))
+      case _ => None
+    }
   }
 }

### 2015-06-10 src/test/scala/edu/cap10/actormodels/tb/InfectionModelTests.scala

@@ -32,5 +32,12 @@ class InfectionModelTests extends FunSuite {
     val res = nonSusceptibleHosts map { hs => infModel(hs, strains.toSeq) }
     assert(res.forall { _.isEmpty } === true)
   }
+  
+  test("exposing Susceptible hosts to one infectious contact of a particular strain, returns an Exposed of the appropriate type, when the infection probability is 1.0") {
+    val infModel = InfectionModel(1.0)
+    strains.foreach { strain =>
+      assert(infModel(Susceptible, Seq(strain)) === Option(Exposed(strain))) 
+    }
+  }
 
 }

### 2015-06-10 src/main/scala/edu/cap10/util/LocalRNG.scala
let LocalRNGs be lazy

@@ -5,6 +5,6 @@ import scala.util.Random
 trait LocalRNG {
 
   def seed : Long
-  implicit protected val rng = new Random(seed)
+  implicit protected lazy val rng = new Random(seed)
   
 }

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala
correct infection of susceptibles

@@ -1,9 +1,12 @@
 package edu.cap10.actormodels.tb
 
 import TBStrain._
+import edu.cap10.util.LocalRNG
 
-case class InfectionModel(some:Double)
+
+case class InfectionModel(infectionProb:Double, val seed:Long)
   extends Function2[HostTBState, Seq[TBStrain], Option[HostTBState]]
+  with LocalRNG
 {
   def apply(
     hostTBstate : HostTBState,
@@ -11,7 +14,7 @@ case class InfectionModel(some:Double)
   ) : Option[HostTBState] = infectiousContacts match {
     case Seq() => None
     case _ => hostTBstate match {
-      case Susceptible => Some(Exposed(infectiousContacts.head))
+      case Susceptible => if (rng.nextDouble() < infectionProb) Some(Exposed(infectiousContacts.head)) else None
       case _ => None
     }
   }

### 2015-06-10 src/test/scala/edu/cap10/actormodels/tb/InfectionModelTests.scala

@@ -1,10 +1,11 @@
 package edu.cap10.actormodels.tb
 
-import org.scalatest.FunSuite
+import org.scalatest.{FunSuite, BeforeAndAfter}
+import scala.util.Random
 
 import TBStrain.{values => strains, _}
 
-class InfectionModelTests extends FunSuite {
+class InfectionModelTests extends FunSuite with BeforeAndAfter {
 
   val nonSusceptibleHosts = Seq() ++ (
     strains map Resistant
@@ -21,23 +22,41 @@ class InfectionModelTests extends FunSuite {
     strains map Chronic
   )
   
+  val seed : Long = 1
+  
+  var rng : Random = _
+  
+  before {
+    rng = new Random(seed)
+  }
+  
   test("exposing hosts to no TB returns Option == None") {
-    val infModel = InfectionModel(0.5)
+    val infModel = InfectionModel(0.5, seed)
     val res = allHostStates map { hs => infModel(hs, Seq.empty) }
     assert(res.forall { _.isEmpty } === true)
   }
   
   test("exposing hosts that are Exposed, Resistant, or Infectious (with any strain) returns Option == None") {
-    val infModel = InfectionModel(0.5)
+    val infModel = InfectionModel(0.5, seed)
     val res = nonSusceptibleHosts map { hs => infModel(hs, strains.toSeq) }
     assert(res.forall { _.isEmpty } === true)
   }
   
   test("exposing Susceptible hosts to one infectious contact of a particular strain, returns an Exposed of the appropriate type, when the infection probability is 1.0") {
-    val infModel = InfectionModel(1.0)
+    val infModel = InfectionModel(1.0, seed)
     strains.foreach { strain =>
       assert(infModel(Susceptible, Seq(strain)) === Option(Exposed(strain))) 
     }
   }
 
+  test("exposing Susceptible hosts to one infectious contact of a particular strain, returns an Exposed of the appropriate type, at the appropriate rate, when the infection probability is not 1.0") {
+    val infProb = 0.3
+    val infModel = InfectionModel(infProb, seed) 
+    strains foreach { strain =>
+      val draws : Seq[Option[HostTBState]] = Seq.fill(10)( rng.nextDouble() < infProb ) map { if(_) Some(Exposed(strain)) else None }
+      val res = Seq.fill(10)( infModel(Susceptible, Seq(strain)) )
+      assert(res === draws)
+    } 
+  }
+  
 }

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala
type inference fix

@@ -1,16 +1,15 @@
 package edu.cap10.actormodels.tb
 
-import TBStrain._
+import TBStrain.Value
 import edu.cap10.util.LocalRNG
 
-
 case class InfectionModel(infectionProb:Double, val seed:Long)
-  extends Function2[HostTBState, Seq[TBStrain], Option[HostTBState]]
+  extends Function2[HostTBState, Seq[TBStrain.Value], Option[HostTBState]]
   with LocalRNG
 {
   def apply(
     hostTBstate : HostTBState,
-    infectiousContacts:Seq[TBStrain]
+    infectiousContacts:Seq[TBStrain.Value]
   ) : Option[HostTBState] = infectiousContacts match {
     case Seq() => None
     case _ => hostTBstate match {

### 2015-06-10 src/test/scala/edu/cap10/actormodels/tb/InfectionModelTests.scala
test chronic to infected failing

@@ -59,4 +59,15 @@ class InfectionModelTests extends FunSuite with BeforeAndAfter {
     } 
   }
   
+  test("exposing Chronic (of either type) hosts to one infectious contact of a particular strain, returns an Infected of the appropriate type, at the appropriate rate, when the infection probability is not 1.0") {
+    val infProb = 0.3
+    val chronicEnhance = 2
+    val infModel = InfectionModel(infProb, seed) 
+    strains foreach { strain =>
+      val draws : Seq[Option[HostTBState]] = Seq.fill(10)( rng.nextDouble() < infProb*chronicEnhance ) map { if(_) Some(Infectious(strain)) else None }
+      val res = Seq.fill(10)( infModel(Chronic(TBStrain.Resistant), Seq(strain)) )
+      assert(res === draws)
+    } 
+  }
+  
 }

### 2015-06-10 src/main/scala/edu/cap10/util/Probability.scala
add complement to Probability for convenience

@@ -27,7 +27,8 @@ object Probability {
 class Probability private (val underlying: Double) extends AnyVal {
   def |(p:Probability) = {
     require(underlying + p.underlying <= 1, "this and p must be exclusive events.")
-    Probability(underlying + p.underlying)
+    new Probability(underlying + p.underlying)
   }
-  def &(p:Probability) = Probability(underlying * p.underlying)
+  def &(p:Probability) = new Probability(underlying * p.underlying)
+  def complement = new Probability(1-underlying)
 }

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala
introduce ModelParams data class, pass Chronic infection tests

@@ -3,17 +3,27 @@ package edu.cap10.actormodels.tb
 import TBStrain.Value
 import edu.cap10.util.LocalRNG
 
-case class InfectionModel(infectionProb:Double, val seed:Long)
+case class InfectionModel(mp:ModelParams, val seed:Long)
   extends Function2[HostTBState, Seq[TBStrain.Value], Option[HostTBState]]
   with LocalRNG
 {
+  import mp._
   def apply(
     hostTBstate : HostTBState,
     infectiousContacts:Seq[TBStrain.Value]
   ) : Option[HostTBState] = infectiousContacts match {
     case Seq() => None
     case _ => hostTBstate match {
-      case Susceptible => if (rng.nextDouble() < infectionProb) Some(Exposed(infectiousContacts.head)) else None
+      case Susceptible =>
+        if (rng.nextDouble() < infection)
+          Some(Exposed(infectiousContacts.head))
+        else
+          None
+      case Chronic(_) => 
+        if (rng.nextDouble() < (1-Math.pow(infection.complement, chronicEnhancement)))
+          Some(Infectious(infectiousContacts.head))
+        else
+          None
       case _ => None
     }
   }

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/ModelParams.scala

@@ -1,0 +1,7 @@
+package edu.cap10.actormodels.tb
+
+import edu.cap10.util.Probability
+
+case class ModelParams(infection:Probability, chronicEnhancement:Double) {
+
+}

### 2015-06-10 src/test/scala/edu/cap10/actormodels/tb/InfectionModelTests.scala

@@ -2,8 +2,11 @@ package edu.cap10.actormodels.tb
 
 import org.scalatest.{FunSuite, BeforeAndAfter}
 import scala.util.Random
+import scala.language.implicitConversions
 
 import TBStrain.{values => strains, _}
+
+import edu.cap10.util.Probability._
 
 class InfectionModelTests extends FunSuite with BeforeAndAfter {
 
@@ -24,6 +27,9 @@ class InfectionModelTests extends FunSuite with BeforeAndAfter {
   
   val seed : Long = 1
   
+  val nullModel = ModelParams(1.0, 1)
+  val refModel = ModelParams(0.3, 2)
+  
   var rng : Random = _
   
   before {
@@ -31,40 +37,38 @@ class InfectionModelTests extends FunSuite with BeforeAndAfter {
   }
   
   test("exposing hosts to no TB returns Option == None") {
-    val infModel = InfectionModel(0.5, seed)
+    val infModel = InfectionModel(nullModel, seed)
     val res = allHostStates map { hs => infModel(hs, Seq.empty) }
     assert(res.forall { _.isEmpty } === true)
   }
   
   test("exposing hosts that are Exposed, Resistant, or Infectious (with any strain) returns Option == None") {
-    val infModel = InfectionModel(0.5, seed)
+    val infModel = InfectionModel(nullModel, seed)
     val res = nonSusceptibleHosts map { hs => infModel(hs, strains.toSeq) }
     assert(res.forall { _.isEmpty } === true)
   }
   
   test("exposing Susceptible hosts to one infectious contact of a particular strain, returns an Exposed of the appropriate type, when the infection probability is 1.0") {
-    val infModel = InfectionModel(1.0, seed)
+    val infModel = InfectionModel(nullModel, seed)
     strains.foreach { strain =>
       assert(infModel(Susceptible, Seq(strain)) === Option(Exposed(strain))) 
     }
   }
 
   test("exposing Susceptible hosts to one infectious contact of a particular strain, returns an Exposed of the appropriate type, at the appropriate rate, when the infection probability is not 1.0") {
-    val infProb = 0.3
-    val infModel = InfectionModel(infProb, seed) 
+    val infModel = InfectionModel(refModel, seed) 
     strains foreach { strain =>
-      val draws : Seq[Option[HostTBState]] = Seq.fill(10)( rng.nextDouble() < infProb ) map { if(_) Some(Exposed(strain)) else None }
[…]

### 2015-06-10 src/main/scala/edu/cap10/actormodels/tb/InfectionModel.scala
refactor infection model to handle multi-exposure

@@ -2,6 +2,7 @@ package edu.cap10.actormodels.tb
 
 import TBStrain.Value
 import edu.cap10.util.LocalRNG
+import edu.cap10.util.Probability
 
 case class InfectionModel(mp:ModelParams, val seed:Long)
   extends Function2[HostTBState, Seq[TBStrain.Value], Option[HostTBState]]
@@ -15,16 +16,14 @@ case class InfectionModel(mp:ModelParams, val seed:Long)
     case Seq() => None
     case _ => hostTBstate match {
       case Susceptible =>
-        if (rng.nextDouble() < infection)
-          Some(Exposed(infectiousContacts.head))
-        else
-          None
-      case Chronic(_) => 
-        if (rng.nextDouble() < (1-Math.pow(infection.complement, chronicEnhancement)))
-          Some(Infectious(infectiousContacts.head))
-        else
-          None
+        anyInfection(infectiousContacts, infection) map { strain => Exposed(strain) }
+      case Chronic(_) =>
+        anyInfection(infectiousContacts, (1-Math.pow(infection.complement, chronicEnhancement))).map { strain => Infectious(strain) }
       case _ => None
     }
   }
+  
+  def anyInfection(contacts:Seq[TBStrain.Value], success:Probability) : Option[TBStrain.Value] = 
+    contacts.find(_ => rng.nextDouble() < success)
+  
 }

### 2015-06-24 src/main/scala/edu/cap10/wifi/DispatchModel.scala
stub to try something different

@@ -1,0 +1,10 @@
+package edu.cap10.wifi
+
+/**
+ * @author cap10
+ */
+trait DispatchModel {
+  
+  def apply(orders:Orders)
+  
+}

### 2015-06-24 src/main/scala/edu/cap10/wifi/package.scala

@@ -1,0 +1,12 @@
+package edu.cap10
+
+/**
+ * @author cap10
+ */
+package object wifi {
+  type AgentID = Int
+  type HotspotID = Int
+  type Orders = Seq[(HotspotID, edu.cap10.util.TimeStamp)]
+  type Events = Seq[Int]
+  
+}

### 2015-07-07 src/main/scala/edu/cap10/actormodels/covert/HotSpot.scala
more agent modeling ideas

@@ -11,23 +11,21 @@ case object Active extends HotSpotState
 case object NoLongerActive extends HotSpotState
 
 case class HotSpot(
-  id:HotSpotID, activation:Day, shutdown:Day
+  id:HotSpotID, activation:Day, shutdown:Day,
+  startHourDistro: CDF = Array.empty
 ) {
   
   private var _clock : Day = 0
   private var _state : HotSpotState = if (activation == 0) Active else NotYetActive
   
   def clock : Day = _clock
+  def state : HotSpotState = _state
   
   def tick(time:Day) : HotSpotState = {
     _clock += 1
     if (_clock == activation)   _setState(Active)
     else if (_clock > shutdown) _setState(NoLongerActive)
     _state 
-  }
-
-  private def _record(userID:Long, timeIn:Long, timeOut:Long) : Boolean = {
-    true
   }
   
   private def _setState(s : HotSpotState) = {
@@ -44,14 +42,14 @@ case class HotSpot(
   } else {
     false
   }
-  
+    
   private val _record : MList[AccessRecord] = MList()
   
   def logout(userID:Long, time:Long) = _cache remove userID match {
     case Some(loginTime) => if (loginTime <= time) {
-      _record += ((_clock, userID, loginTime, time))
-      true
-    } else false
+        _record += ((_clock, userID, loginTime, time))
+        true
+      } else false
     case None => false
   }
   

### 2015-07-07 src/main/scala/edu/cap10/actormodels/covert/HotSpotUseStrategy.scala

@@ -1,0 +1,27 @@
+package edu.cap10.actormodels.covert
+
+import scala.util.Random.nextInt
+
+trait HotSpotUseStrategy {
+  def plan(day:Day, options:Set[HotSpot]) : Seq[AccessPlan]
+}
+
+trait StartTimeStrategy {
+  def start(day:Day, hs:HotSpot, otherEvents:List[AccessPlan]) : Time
+}
+
+trait EndTimeStrategy {
+  def end(day:Day, hs:HotSpot, start:Time, otherEvents:List[AccessPlan]) : Time
+}
+
+case class OnePerDay(s:StartTimeStrategy, e:EndTimeStrategy) extends HotSpotUseStrategy {
+  import s.start
+  import e.end
+  override def plan(day:Day, options:Set[HotSpot]) = {
+    val which = nextInt(options.size)
+    val hs = options.drop(which).head
+    val sTime = start(day, hs, List.empty)
+    val eTime = end(day, hs, sTime, List.empty)
+    Seq((hs, sTime, eTime))
+  }
+}

### 2015-07-07 src/main/scala/edu/cap10/actormodels/covert/Provocateur.scala

@@ -1,6 +1,9 @@
 package edu.cap10.actormodels.covert
 
-case class Provocateur(id:UserID) {
+import scala.util.Random.nextDouble
+import scala.collection.mutable.{Set => MSet}
+
+case class Provocateur(id:UserID, acceptNewHotSpot:Double) {
 
   private var _clock : Day = 0
   
@@ -10,5 +13,15 @@ case class Provocateur(id:UserID) {
     _clock += 1
     true
   }
+  
+  private val _hotspots = MSet[HotSpot]()
+  
+  def aware(hotspot:HotSpot) =
+    if (nextDouble() < acceptNewHotSpot)
+      _hotspots add hotspot
+    else
+      false
+  
+  def hotspots : Set[HotSpot] = _hotspots.toSet
   
 }

### 2015-07-07 src/main/scala/edu/cap10/actormodels/covert/package.scala

@@ -8,5 +8,8 @@ package object covert {
   type Time = Long
   
   type AccessRecord = (Day, UserID, Time, Time)
+  type AccessPlan = (HotSpot, Time, Time)
+  
+  type CDF = Array[Double]
   
 }

### 2015-07-07 src/test/scala/edu/cap10/actormodels/covert/ProvocateurTests.scala

@@ -6,13 +6,28 @@ import scala.language.implicitConversions
 
 class ProvocateurTests extends FunSuite {
     
-  test("sending ticks to a HotSpot advances its clock") {
-    val P1 = Provocateur(1)
+  test("sending ticks to a Provo advances its clock") {
+    val P1 = Provocateur(1, 1)
     assert(P1.clock === 0)
     for (i <- 1 to 5) {
       P1.tick(i)
       assert(P1.clock === i)
     }
+  }
+  
+  test("Provos can add HotSpots to their use list") {
+    val pID = 1
+    val P1 = Provocateur(pID, 1)
+    val HS1 = HotSpot(1, 0, 10)
+    val HS2 = HotSpot(2, 0, 10)
+    assert(P1.aware(HS1), "P1 did not accept HS1, despite a 100% probability")
+    assert(P1.aware(HS2), "P1 did not accept HS2, despite a 100% probability")
+    assert(P1.hotspots.contains(HS1), "after accepting HS1, it didn't appear in its list")
+    assert(P1.hotspots.contains(HS2), "after accepting HS2, it didn't appear in its list")
+  }
+  
+  test("Provos") {
+    
   }
   
 }

### 2015-07-07 rscripts/data_review.Rmd
add plots of location usage

@@
 ```
+
+Now we must try to understand the remaining data.
+
+First, what happens with our hotspot locations?
+We need to understand how people appear connected to them, in order to properly model how our covert agents might use them.
+To do that, we need to speculate about how these hotspots are behaving, *e.g.* are they new wifi access locations?
+are they substituting for existing services?
+
+We can start by looking at all of locations, and their first and last access times.
+
+```{r, echo=FALSE}
+ggplot(censor.dt[,list(t_min=min(login),t_max=max(logout)), by=location_id]) +
+  theme_bw() +
+  aes(x=location_id, ymin=t_min, ymax=t_max) +
+  geom_linerange() +
+  coord_flip() + labs(y="seconds since start", x="hotspot ordered by start time")
+```
+
+The fairly consistent slope of site start times indicates a reliable creation rate.
+We can investigate this further by considering the average creation rate for the whole time period, and the rolling average.
+
+We can determine the overall average creation rate straightforwardly, by taking the number of hotspots created over the whole duration of the data set.
+We should also compare that average to an aggregation (*e.g.*, the number created in each week) to get a feel for how creation varies around that.
+
+```{r, echo=FALSE}
+weekly_ave <- censor.dt[,(length(unique(location_id))-2)/(max(login) - min(login))*60*60*24*7]
+weekly_count <- censor.dt[,list(creation_week=(min(login_day) %/% 7)), by=location_id][,list(count=.N),by=creation_week]
+ggplot(weekly_count) + theme_bw() + aes(x=creation_week, y=count) +
+  geom_bar(stat="identity", width=1) +
+  geom_hline(y = weekly_ave, color="red")
+```
+
+We can also be a bit more statistical, fitting a least squares regression model (here, creations by day instead of by week):
+
+```{r, echo=FALSE}
+creations <- setkey(censor.dt[,list(creation_day=min(login_day)), by=location_id][,list(count=.N), by=creation_day][,list(creation_day = creation_day - min(creation_day), count)], "creation_day")
+zeros <- data.table(creation_day = 0:max(creations$creation_day), count=0, key = "creation_day")
+creations <- merge(creations, zeros, all=TRUE)[,list(count=max(count.x, count.y, na.rm=T)),keyby=creation_day]
+ggplot(creations) + theme_bw() + aes(x=creation_day, y=count) +
+  stat_smooth(method="lm") + geom_point()
+```
+
+This leaves a bit to be desired - clearly, there is a mostly constant level with possibly a very small trend, but least squares is not the best fitting approach since the underlying data is restricted to be positive.
+We can use maximum likelihood to better describe the number of new hotspots created per day.
+We can posit some processes, like creation of new businesses, tendency to accummulate requests and complete them all on one day, *etc.* to inspire our distribution.
+
+TODO: zero-inflated poisson?  zero-inflated negative binomial?
+
+Unlike creation, elimination of hotspot IDs seems to be much more variable.  So they are steadily created, but the lifetime after creation is variable.
+
+```{r,echo=FALSE}
+# dt <- censor.dt[,list(duration=(max(logout) - min(login))/60/60/24/7), by=location_id]
+# h <- hist(dt$duration, 0:ceiling(max(dt$duration)), plot=F)
+# ggplot() +
+#   theme_bw() +
+#   aes(x=duration/60/60/24/7) +
+#   geom_bar(binwidth=1) + labs(x="weeks of hotspot life", y="hotspots with that duration")
+```
+
+The usage profile of hostspots also varies.  We can use this to categorize "types" of locations using a *k-means* approach.  First let's look at grouping by typical *login* times:
+
+```{r, echo=FALSE}
+shft <- 4
+orig <- 0:23
+hr_shf <- (orig - shft) %% 24
+hr_lab <- (orig + shft) %% 24
+base <- censor.dt[,list(login_hour=floor(login_time/60/60)), by=location_id][,list(count=.N),keyby=list(location_id, login_hour)][,list(login_hour=hr_shf[login_hour+1],count=count/max(count)),keyby=location_id]
+# by max vs by sum
+byid <- acast(base, location_id ~ login_hour, value.var = "count", fill = 0)
+kmplotter <- function(n) {
+  km <- kmeans(byid, n, nstart = 5)
+  ggplot(base[,list(location_id,login_hour,count,cluster=factor(km$cluster[location_id]))]) + theme_bw() +
+    aes(x = factor(location_id), y = login_hour, fill = count) +
+    facet_grid(. ~ cluster, scales = "free") + scale_x_discrete("location", labels="") +
+    geom_raster() + scale_y_discrete(name="login hour", labels=hr_lab, breaks=0:23) +
+    labs(fill="proportion max")
+}
+print(kmplotter(4))
+print(kmplotter(5))
+print(kmplotter(6))
+print(kmplotter(7))
+print(kmplotter(8))
+print(kmplotter(9))
+print(kmplotter(10))
+```
+
+We can also look at logout times:
+
+```{r, echo=FALSE}
+shft <- 4
+orig <- 0:23
+hr_shf <- (orig - shft) %% 24
+hr_lab <- (orig + shft) %% 24
+base <- censor.dt[,list(logout_hour=floor(logout_time/60/60)), by=location_id][,list(count=.N),keyby=list(location_id, logout_hour)][,list(logout_hour=hr_shf[logout_hour+1],count=count/max(count)),keyby=location_id]
+# by max vs by sum
+byid <- acast(base, location_id ~ logout_hour, value.var = "count", fill = 0)
+kmplotter <- function(n) {
+  km <- kmeans(byid, n, nstart = 5)
+  ggplot(base[,list(location_id,logout_hour,count,cluster=factor(km$cluster[location_id]))]) + theme_bw() +
+    aes(x = factor(location_id), y = logout_hour, fill = count) +
+    facet_grid(. ~ cluster, scales = "free") + scale_x_discrete("location", labels="") +
+    geom_raster() + scale_y_discrete(name="logout hour", labels=hr_lab, breaks=0:23) +
+    labs(fill="proportion max")
+}
+print(kmplotter(4))
+print(kmplotter(5))
+print(kmplotter(6))
+print(kmplotter(7))
+print(kmplotter(8))
+print(kmplotter(9))
+print(kmplotter(10))
+```
+
+* * *
 
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, eval=FALSE}
 grow1 <- function(break1, dt) sum(lm(count ~ time, dt[time < break1])$residuals^2)

### 2015-07-11 rscripts/data_review.Rmd
review refinement

@@
 
-```{r}
+```{r, echo=FALSE}
 max_hours <- 24
@@
 ```{r, echo=FALSE}
-ggplot(censor.dt[,list(t_min=min(login),t_max=max(logout)), by=location_id]) +
+minmax.dt <- censor.dt[,list(t_min=min(login),t_max=max(logout)), keyby=location_id]
+ggplot(minmax.dt) +
   theme_bw() +
-  aes(x=location_id, ymin=t_min, ymax=t_max) +
+  aes(x=location_id, ymin=t_min/60/60/24/7, ymax=t_max/60/60/24/7) +
   geom_linerange() +
-  coord_flip() + labs(y="seconds since start", x="hotspot ordered by start time")
+  coord_flip() + labs(y="weeks since start", x="hotspot")
 ```
@@
 ```{r, echo=FALSE}
-weekly_ave <- censor.dt[,(length(unique(location_id))-2)/(max(login) - min(login))*60*60*24*7]
-weekly_count <- censor.dt[,list(creation_week=(min(login_day) %/% 7)), by=location_id][,list(count=.N),by=creation_week]
-ggplot(weekly_count) + theme_bw() + aes(x=creation_week, y=count) +
+weekly_ave <- minmax.dt[,
+  (length(unique(location_id))-2)/(max(t_max) - min(t_min))*60*60*24*7
+]
+weekly.dt <- minmax.dt[,
+  list(
+    creation = (t_min/60/60/24)%/%7,
+    shutdown = (t_max/60/60/24)%/%7
+  ),
+  by = location_id
+]
+count.dt <- weekly.dt[,
+  .N, by = creation
+]
+ggplot(count.dt) + theme_bw() + 
+  aes(x=creation, y=N) +
+  labs(x="week", y="hotspots created", color="ave") +
   geom_bar(stat="identity", width=1) +
@@
 ```{r, echo=FALSE}
-creations <- setkey(censor.dt[,list(creation_day=min(login_day)), by=location_id][,list(count=.N), by=creation_day][,list(creation_day = creation_day - min(creation_day), count)], "creation_day")
-zeros <- data.table(creation_day = 0:max(creations$creation_day), count=0, key = "creation_day")
-creations <- merge(creations, zeros, all=TRUE)[,list(count=max(count.x, count.y, na.rm=T)),keyby=creation_day]
-ggplot(creations) + theme_bw() + aes(x=creation_day, y=count) +
-  stat_smooth(method="lm") + geom_point()
+creations <- setkey(censor.dt[,
+  list(creation_day=min(login_day)),
+  by=location_id
+][,
+  .N,
+  by=creation_day
+][,
+  list(creation_day = creation_day - min(creation_day), N)
+], creation_day)
+zeros <- data.table(
+  creation_day = 0:max(creations$creation_day),
+  N = 0,
+  key = "creation_day"
+)
+creations <- merge(creations, zeros, all=TRUE)[,
+  list(N = max(N.x, N.y, na.rm=T)),
+  keyby = creation_day
+]
+ggplot(creations) + theme_bw() +
+  aes(x=creation_day, y=N) +
+  stat_smooth(method="lm") + geom_point() +
+  labs(x="creation day", y="hotspots created")
 ```
@@
 
-Unlike creation, elimination of hotspot IDs seems to be much more variable.  So they are steadily created, but the lifetime after creation is variable.
+Unlike creation, elimination of hotspot IDs seems to be much more variable.
+While they are steadily created, their lifetime after creation has a long tailed distribution.
+This is further complicated by the termination of the data, left censoring many entries.
+
+TODO: break out entries that start near the end of the dataset and highlight their contribution to the spike in low weeks.
 
 ```{r,echo=FALSE}
-# dt <- censor.dt[,list(duration=(max(logout) - min(login))/60/60/24/7), by=location_id]
+dt <- minmax.dt[,
+  list(duration = (t_max-t_min)/60/60/24/7),
+  by=location_id
+]
 # h <- hist(dt$duration, 0:ceiling(max(dt$duration)), plot=F)
-# ggplot() +
-#   theme_bw() +
-#   aes(x=duration/60/60/24/7) +
-#   geom_bar(binwidth=1) + labs(x="weeks of hotspot life", y="hotspots with that duration")
+ggplot(dt) +
+  theme_bw() +
+  aes(x=duration) +
+  geom_bar(binwidth=1) + labs(x="weeks of hotspot life", y="hotspots with that duration")
 ```
 
-The usage profile of hostspots also varies.  We can use this to categorize "types" of locations using a *k-means* approach.  First let's look at grouping by typical *login* times:
+The usage profile of hostspots also varies.  We can use this to categorize "types" of locations using a *k-means* approach.  First, let's look at grouping by typical *login* and *logout* times:
 
@@
 hr_lab <- (orig + shft) %% 24
-base <- censor.dt[,list(login_hour=floor(login_time/60/60)), by=location_id][,list(count=.N),keyby=list(location_id, login_hour)][,list(login_hour=hr_shf[login_hour+1],count=count/max(count)),keyby=location_id]
-# by max vs by sum
-byid <- acast(base, location_id ~ login_hour, value.var = "count", fill = 0)
-kmplotter <- function(n) {
-  km <- kmeans(byid, n, nstart = 5)
-  ggplot(base[,list(location_id,login_hour,count,cluster=factor(km$cluster[location_id]))]) + theme_bw() +
-    aes(x = factor(location_id), y = login_hour, fill = count) +
-    facet_grid(. ~ cluster, scales = "free") + scale_x_discrete("location", labels="") +
-    geom_raster() + scale_y_discrete(name="login hour", labels=hr_lab, breaks=0:23) +
-    labs(fill="proportion max")
+base <- censor.dt[,
+  list(
+    login_hour  = floor(login_time/60/60),
+    logout_hour = floor(logout_time/60/60)
+  ),
+  by = location_id
+]
+both <- rbind(base[,
+  list(login_count=.N),
+  keyby=list(location_id, login_hour)
+][,
+  list(
+    hour=hr_shf[login_hour+1],
+    count=login_count/max(login_count),
+    event=factor("login", levels=c("logout","login"))
+  ),
+  keyby=location_id
+],
+base[,
+  list(logout_count=.N),
+  keyby=list(location_id, logout_hour)
+][,
+  list(
+    hour=hr_shf[logout_hour+1],
+    count=logout_count/max(logout_count),
+    event=factor("logout", levels=c("logout","login"))
+  ),
+  keyby=location_id
+])
+setkey(both, location_id, event, hour)
+# by max vs by sum?
+byid <- acast(both, location_id ~ hour + event, value.var = "count", fill = 0)
+
+kmplotter <- function(n, castdata, src,
+  facet = facet_grid(event ~ cluster, scales = "free"),
+  refevent = "login"
+) {
+    km <- kmeans(castdata, n, nstart = 5)
+    cluster.dt <- data.table(
+      location_id = 1:src[,max(location_id)],
+      cluster = km$cluster, 
+      key="location_id"
+    )
+    peakhours <- src[cluster.dt][,
+      list(peak=sum(hour*count)/sum(count)),
+      by=list(event, cluster)
+    ]
+    setkey(peakhours, event, peak)
+    ## get the order from login
+    ## relabel
+    reorder <- peakhours[event == refevent, order(cluster)]
+    peakhours[, cluster := reorder[cluster]]
+    cluster.dt[,cluster := reorder[cluster]]
+  # reorder cluster ids
+    ggplot(src[cluster.dt]) +
+      theme_bw() +
+      aes(x = factor(location_id), y = hour, fill = count) + facet +
+      geom_raster() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
+      scale_x_discrete("location",  labels="") +
+      scale_y_discrete(name="hour", labels=hr_lab, limits=0:23, expand = c(0,0)) +
+      labs(fill="prop. max count")
 }
-print(kmplotter(4))
-print(kmplotter(5))
-print(kmplotter(6))
-print(kmplotter(7))
-print(kmplotter(8))
-print(kmplotter(9))
-print(kmplotter(10))
+# print(kmplotter(3))
+# print(kmplotter(4))
+print(kmplotter(5, byid, both)) 
+# print(kmplotter(6))
+# print(kmplotter(7))
+# print(kmplotter(8))
+# print(kmplotter(9))
+# print(kmplotter(10))
+# TODO: sort out clustering - either different clustering algorithm, or principled selection of k?
 ```
 
-We can also look at logout times:
+An alternative view would be to consider something like total users present in each hour.
+
+TODO: could weight a user contribution to an hour by how much time they are present during it, but seems reasonable to just do counts at the moment.
 
 ```{r, echo=FALSE}
-shft <- 4
-orig <- 0:23
-hr_shf <- (orig - shft) %% 24
-hr_lab <- (orig + shft) %% 24
-base <- censor.dt[,list(logout_hour=floor(logout_time/60/60)), by=location_id][,list(count=.N),keyby=list(location_id, logout_hour)][,list(logout_hour=hr_shf[logout_hour+1],count=count/max(count)),keyby=location_id]
-# by max vs by sum
-byid <- acast(base, location_id ~ logout_hour, value.var = "count", fill = 0)
-kmplotter <- function(n) {
-  km <- kmeans(byid, n, nstart = 5)
-  ggplot(base[,list(location_id,logout_hour,count,cluster=factor(km$cluster[location_id]))]) + theme_bw() +
-    aes(x = factor(location_id), y = logout_hour, fill = count) +
-    facet_grid(. ~ cluster, scales = "free") + scale_x_discrete("location", labels="") +
-    geom_raster() + scale_y_discrete(name="logout hour", labels=hr_lab, breaks=0:23) +
-    labs(fill="proportion max")
+if (file.access("extend.RData") != 0) {
+  proto <- array(0L, dim = c(1,24), dimnames = list(drop=c(),hours=0:23))
+  extend <- cbind(censor.dt, proto)
+  extend[, login_hour := floor(login_time/60/60) ][, logout_hour:=floor(logout_time/60/60)]
+  extend[, "0"  := ifelse((0  <= logout_hour) & (login_hour <= 0),  1L, 0L)]
+  extend[, "1"  := ifelse((1  <= logout_hour) & (login_hour <= 1),  1L, 0L)]
+  extend[, "2"  := ifelse((2  <= logout_hour) & (login_hour <= 2),  1L, 0L)]
+  extend[, "3"  := ifelse((3  <= logout_hour) & (login_hour <= 3),  1L, 0L)]
+  extend[, "4"  := ifelse((4  <= logout_hour) & (login_hour <= 4),  1L, 0L)]
+  extend[, "5"  := ifelse((5  <= logout_hour) & (login_hour <= 5),  1L, 0L)]
+  extend[, "6"  := ifelse((6  <= logout_hour) & (login_hour <= 6),  1L, 0L)]
+  extend[, "7"  := ifelse((7  <= logout_hour) & (login_hour <= 7),  1L, 0L)]
+  extend[, "8"  := ifelse((8  <= logout_hour) & (login_hour <= 8),  1L, 0L)]
+  extend[, "9"  := ifelse((9  <= logout_hour) & (login_hour <= 9),  1L, 0L)]
+  extend[, "10" := ifelse((10 <= logout_hour) & (login_hour <= 10), 1L, 0L)]
+  extend[, "11" := ifelse((11 <= logout_hour) & (login_hour <= 11), 1L, 0L)]
+  extend[, "12" := ifelse((12 <= logout_hour) & (login_hour <= 12), 1L, 0L)]
+  extend[, "13" := ifelse((13 <= logout_hour) & (login_hour <= 13), 1L, 0L)]
+  extend[, "14" := ifelse((14 <= logout_hour) & (login_hour <= 14), 1L, 0L)]
+  extend[, "15" := ifelse((15 <= logout_hour) & (login_hour <= 15), 1L, 0L)]
+  extend[, "16" := ifelse((16 <= logout_hour) & (login_hour <= 16), 1L, 0L)]
+  extend[, "17" := ifelse((17 <= logout_hour) & (login_hour <= 17), 1L, 0L)]
+  extend[, "18" := ifelse((18 <= logout_hour) & (login_hour <= 18), 1L, 0L)]
+  extend[, "19" := ifelse((19 <= logout_hour) & (login_hour <= 19), 1L, 0L)]
+  extend[, "20" := ifelse((20 <= logout_hour) & (login_hour <= 20), 1L, 0L)]
+  extend[, "21" := ifelse((21 <= logout_hour) & (login_hour <= 21), 1L, 0L)]
+  extend[, "22" := ifelse((22 <= logout_hour) & (login_hour <= 22), 1L, 0L)]
+  extend[, "23" := ifelse((23 <= logout_hour) & (login_hour <= 23), 1L, 0L)]
+} else {
+  load("extend.RData")
 }
-print(kmplotter(4))
-print(kmplotter(5))
-print(kmplotter(6))
-print(kmplotter(7))
-print(kmplotter(8))
-print(kmplotter(9))
-print(kmplotter(10))
+if (file.access("comp.RData") != 0) {
+  comp <- extend[,list(
+    `0` =sum(`0`),
+    `1` =sum(`1`),
+    `2` =sum(`2`),
+    `3` =sum(`3`),
+    `4` =sum(`4`),
+    `5` =sum(`5`),
+    `6` =sum(`6`),
+    `7` =sum(`7`),
+    `8` =sum(`8`),
+    `9` =sum(`9`),
+    `10`=sum(`10`),
+    `11`=sum(`11`),
+    `12`=sum(`12`),
+    `13`=sum(`13`),
+    `14`=sum(`14`),
+    `15`=sum(`15`),
+    `16`=sum(`16`),
+    `17`=sum(`17`),
+    `18`=sum(`18`),
+    `19`=sum(`19`),
+    `20`=sum(`20`),
+    `21`=sum(`21`),
+    `22`=sum(`22`),
+    `23`=sum(`23`)
+  ),by=location_id]
+  save(comp, file="comp.RData")
+} else {
+  load("comp.RData")
+}
+
+comp.short <- melt(comp, id.var="location_id", variable.name = "hour", value.name="count")[count != 0]
+comp.short$count <- as.double(comp.short$count)
+comp.short$hour <- as.integer(comp.short$hour)
+setkey(comp.short, location_id, hour)
+comp.short[, count := count/max(count), by=location_id][,event := "usage" ]
+compbyID <- acast(comp.short, location_id ~ hour, value.var = "count", fill = 0)
+
+kmplotter(5, compbyID, comp.short, facet = facet_grid(. ~ cluster, scale="free"), refevent="usage")
 ```
+
+These analyses identify similar kinds of locations based on their aggegrated usage profile.
+There is not an obviously preferable approach to the signature we used for clustering -- *i.e.*, login and logout densities *vs.* concurrent usage density -- so we choose the usage-based clusters.
+
+TODO: there probably could be, if we did some statistics to determine best *k*, and then did some information criterion between the two models?  Additionally, compare overlap in clusters?
+
+If we revisit our view of the presence of locations over time in the dataset, what do we see?
+
+```{r,echo=FALSE}
+usagekm <- kmeans(compbyID, 8, nstart = 5)$cluster
+km.dt <- data.table(location_id=1:length(usagekm), cluster=factor(usagekm), key=c("location_id"))
+ggplot(minmax.dt[km.dt]) +
+  theme_bw() +
+  aes(x=location_id, ymin=t_min, ymax=t_max, color=cluster) +
+  geom_linerange() +
+  coord_flip() + labs(y="seconds since start", x="hotspot ordered by start time") +
+  scale_color_discrete()
+```
+
+There are perhaps some trends like-location creation, indicated by the solid color bands.
+However, this does not clarify what sort of locations are available throughout the whole duration of the data set.
+What if we look at the availability of a type cluster at anytime?
+
+```{r, echo=FALSE}
+week.melt <- setkey(melt(weekly.dt, id.var = "location_id", variable.name = "event", value.name = "week"), location_id)[km.dt]
+week.melt[, inc := ifelse(event == "creation", 1, -1)]
+week.melt <- week.melt[,list(inc = sum(inc)), keyby=list(cluster, week) ]
+# ggplot(week.totals[cluster==1]) + aes(x = week, y = agg, fill=factor(cluster), group=cluster) + geom_area(stat="hline")
+week.melt.fill <- setkey(Reduce(function(left, right) {
+  ul <- left[!(week %in% right[,unique(week)]), unique(week)]
+  ur <- right[!(week %in% left[,unique(week)]), unique(week)]
+  ulc <- left[,unique(cluster)]
+  rbind(left, right, 
+    data.table(week=ul, inc=0, cluster=right[,unique(cluster)]),
+    data.table(week=rep(ur, times=length(ulc)), inc=0, cluster=rep(ulc, each=length(ur)))
+  )
+}, Map(function(clus) week.melt[cluster == clus], week.melt[,unique(cluster)])), cluster, week)
+week.totals <- week.melt.fill[,list(agg = cumsum(inc), week), keyby=cluster]
+clusorder <- week.totals[,list(mx=max(agg)),keyby=cluster][,cluster,keyby=mx]$cluster
+week.totals$cluster <- factor(week.totals$cluster, levels=clusorder)
+ggplot(week.totals) + aes(x = week, y = agg, fill=cluster, order=cluster) + geom_area(position="fill")
+```
+
+This suggests that, at least to a reasonable approximation, the prevalance of cluster types is constant through the range we are considering for the population synthesis.
+
+Now, how do real users use these locations?
 

### 2015-07-13 rscripts/data_review.Rmd
de-ugly houring box; identified error w/ binning across day

@@
 min_logins <- 2
+min_lifetime <- 30 # days
 ```
@@
 3. eliminate users that have fewer than `r min_logins` logins
+4. eliminate users that have a *lifetime* `(max(logout) - min(login))` less than `r min_lifetime` days
 4. eliminate locations that have fewer than `r min_logins` logins
@@
 censor.dt <- src.dt[(logout != login) & ((logout - login) <= max_hours*60*60), ]
-invalid.users <- censor.dt[,.N, by=user_id ][N < min_logins, user_id]
+invalid.users <- censor.dt[,list(.N, lifetime = (max(logout) - min(login))/60/60/24), by=user_id ][N < min_logins | lifetime < min_lifetime, user_id]
 while(length(invalid.users) > 0) {
@@
   censor.dt     <- censor.dt[!location_id %in% invalid.locs]
-  invalid.users <- censor.dt[,.N, by=user_id ][N < min_logins, user_id]
+  invalid.users <- censor.dt[,list(.N, lifetime = (max(logout) - min(login))/60/60/24), by=user_id ][N < min_logins | lifetime < min_lifetime, user_id]
 }
@@
     hour=hr_shf[login_hour+1],
-    count=login_count/max(login_count),
+    norm=login_count/max(login_count),
     event=factor("login", levels=c("logout","login"))
@@
     hour=hr_shf[logout_hour+1],
-    count=logout_count/max(logout_count),
+    norm=logout_count/max(logout_count),
     event=factor("logout", levels=c("logout","login"))
@@
 # by max vs by sum?
-byid <- acast(both, location_id ~ hour + event, value.var = "count", fill = 0)
+byid <- acast(both, location_id ~ hour + event, value.var = "norm", fill = 0)
 
@@
     peakhours <- src[cluster.dt][,
-      list(peak=sum(hour*count)/sum(count)),
+      list(peak=sum(hour*norm)/sum(norm)),
       by=list(event, cluster)
@@
       theme_bw() +
-      aes(x = factor(location_id), y = hour, fill = count) + facet +
+      aes(x = factor(location_id), y = hour, fill = norm) + facet +
       geom_raster() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
@@
 ```{r, echo=FALSE}
-if (file.access("extend.RData") != 0) {
-  proto <- array(0L, dim = c(1,24), dimnames = list(drop=c(),hours=0:23))
-  extend <- cbind(censor.dt, proto)
-  extend[, login_hour := floor(login_time/60/60) ][, logout_hour:=floor(logout_time/60/60)]
-  extend[, "0"  := ifelse((0  <= logout_hour) & (login_hour <= 0),  1L, 0L)]
-  extend[, "1"  := ifelse((1  <= logout_hour) & (login_hour <= 1),  1L, 0L)]
-  extend[, "2"  := ifelse((2  <= logout_hour) & (login_hour <= 2),  1L, 0L)]
-  extend[, "3"  := ifelse((3  <= logout_hour) & (login_hour <= 3),  1L, 0L)]
-  extend[, "4"  := ifelse((4  <= logout_hour) & (login_hour <= 4),  1L, 0L)]
-  extend[, "5"  := ifelse((5  <= logout_hour) & (login_hour <= 5),  1L, 0L)]
-  extend[, "6"  := ifelse((6  <= logout_hour) & (login_hour <= 6),  1L, 0L)]
-  extend[, "7"  := ifelse((7  <= logout_hour) & (login_hour <= 7),  1L, 0L)]
-  extend[, "8"  := ifelse((8  <= logout_hour) & (login_hour <= 8),  1L, 0L)]
-  extend[, "9"  := ifelse((9  <= logout_hour) & (login_hour <= 9),  1L, 0L)]
-  extend[, "10" := ifelse((10 <= logout_hour) & (login_hour <= 10), 1L, 0L)]
-  extend[, "11" := ifelse((11 <= logout_hour) & (login_hour <= 11), 1L, 0L)]
-  extend[, "12" := ifelse((12 <= logout_hour) & (login_hour <= 12), 1L, 0L)]
-  extend[, "13" := ifelse((13 <= logout_hour) & (login_hour <= 13), 1L, 0L)]
-  extend[, "14" := ifelse((14 <= logout_hour) & (login_hour <= 14), 1L, 0L)]
-  extend[, "15" := ifelse((15 <= logout_hour) & (login_hour <= 15), 1L, 0L)]
-  extend[, "16" := ifelse((16 <= logout_hour) & (login_hour <= 16), 1L, 0L)]
-  extend[, "17" := ifelse((17 <= logout_hour) & (login_hour <= 17), 1L, 0L)]
-  extend[, "18" := ifelse((18 <= logout_hour) & (login_hour <= 18), 1L, 0L)]
-  extend[, "19" := ifelse((19 <= logout_hour) & (login_hour <= 19), 1L, 0L)]
-  extend[, "20" := ifelse((20 <= logout_hour) & (login_hour <= 20), 1L, 0L)]
-  extend[, "21" := ifelse((21 <= logout_hour) & (login_hour <= 21), 1L, 0L)]
-  extend[, "22" := ifelse((22 <= logout_hour) & (login_hour <= 22), 1L, 0L)]
-  extend[, "23" := ifelse((23 <= logout_hour) & (login_hour <= 23), 1L, 0L)]
-} else {
-  load("extend.RData")
-}
-if (file.access("comp.RData") != 0) {
-  comp <- extend[,list(
-    `0` =sum(`0`),
-    `1` =sum(`1`),
-    `2` =sum(`2`),
-    `3` =sum(`3`),
-    `4` =sum(`4`),
-    `5` =sum(`5`),
-    `6` =sum(`6`),
-    `7` =sum(`7`),
-    `8` =sum(`8`),
-    `9` =sum(`9`),
-    `10`=sum(`10`),
-    `11`=sum(`11`),
-    `12`=sum(`12`),
-    `13`=sum(`13`),
-    `14`=sum(`14`),
-    `15`=sum(`15`),
-    `16`=sum(`16`),
-    `17`=sum(`17`),
-    `18`=sum(`18`),
-    `19`=sum(`19`),
-    `20`=sum(`20`),
-    `21`=sum(`21`),
-    `22`=sum(`22`),
-    `23`=sum(`23`)
-  ),by=location_id]
-  save(comp, file="comp.RData")
-} else {
-  load("comp.RData")
-}
-
-comp.short <- melt(comp, id.var="location_id", variable.name = "hour", value.name="count")[count != 0]
-comp.short$count <- as.double(comp.short$count)
-comp.short$hour <- as.integer(comp.short$hour)
-setkey(comp.short, location_id, hour)
-comp.short[, count := count/max(count), by=location_id][,event := "usage" ]
-compbyID <- acast(comp.short, location_id ~ hour, value.var = "count", fill = 0)
-
+comp.short <- setkey(rbindlist(
+  Map(function(hour) censor.dt[
+    (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600)), 
+    list(count=.N, hour=hour, event="usage"),
+    by=location_id
+  ], 0:23)), location_id, hour
+)
+comp.short[,norm := count/max(count), by=location_id]
+compbyID <- acast(comp.short, location_id ~ hour, value.var = "norm", fill = 0)
 kmplotter(5, compbyID, comp.short, facet = facet_grid(. ~ cluster, scale="free"), refevent="usage")
@@
 
-* * *
+Have user, cluster id, time in, time out
 
-When generating our synthetic data, we assume locations exist before and after they join the service.
-So the synthetic population may make visits to these locations outside of time when the service is operating.
-We trim that activity from the synthetic data according the time that locations are active in the observed data.
+For long term users, what is the distribution of their life time cluster usage?  do they tend to have a fixed number of items from a clusters they use?  do they replace locations in a cluster?
+
+on last question: how would we know this from data?
+ - looking at a usage time series, non-overlapping / non-interleaving use of different locations from the same cluster.
+replacement == first location no longer available (other users also stop using).
+displacement == first location still available, user simply stops going there
+
+ - for each user, for each cluster:
+  - does a location use stop before another starts?  relaxation: does it overlap for only a short period of time, w/ declining user-time while the other increases in user time?
+  - so this looks like a combinations test: for each pair of locations?
+
+Alt: try to decompose users into time series.  How to compare users w/ different lifetimes?  Do the same as locations, and give them hourly signatures.
+Then sort by (user) cluster, appearance in time.  display video of users, showing usage time-bars by cluster.  preset room for max number of streams that appear in a cluster?
 
 ```{r, echo=FALSE}
-## TODO switch to log(in|out) day
-limits.dt <- censor.dt[,list(first=min(login), last=max(logout)), keyby=location_id]
-limits.dt
-save(censor.dt, limits.dt, file = "../input/censored.Rdata")
+setkey(censor.dt, location_id)
+number_of_rows <- censor.dt[km.dt][,
+  list(ul = length(unique(location_id))),
+  by=list(user_id, cluster)
+][,
+  list(nrows = sum(max(ul))),
+  by=user_id
+][,
+  max(nrows)
+]
+
+## TODO: handle day splits!
+
+comp.user.short <- setkey(rbindlist(
+  Map(function(hour) censor.dt[
+    (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600)), 
+    list(count=.N, hour=hour, event="usage"),
+    by=user_id
+  ], 0:23)), user_id, hour
+)
+comp.user.short[,norm := count/max(count), by=user_id]
+compuserbyID <- acast(comp.user.short, user_id ~ hour, value.var = "norm", fill = 0)
+
+kmUserPlotter <- function(n, castdata, src,
+  facet = facet_grid(. ~ cluster, scales = "free"),
+  refevent = "usage"
+) {
+    km <- kmeans(castdata, n, nstart = 5)
+    cluster.dt <- data.table(
+      user_id = 1:src[,max(user_id)],
+      cluster = km$cluster, 
+      key="user_id"
+    )
+    peakhours <- src[cluster.dt][,
+      list(peak=sum(hour*norm)/sum(norm)),
+      by=list(event, cluster)
+    ]
+    setkey(peakhours, event, peak)
+    ## get the order from login
+    ## relabel
+    reorder <- peakhours[event == refevent, order(cluster)]
+    peakhours[, cluster := reorder[cluster]]
+    cluster.dt[,cluster := reorder[cluster]]
+  # reorder cluster ids
+    ggplot(src[cluster.dt]) +
+      theme_bw() +
+      aes(x = factor(user_id), y = hour, fill = norm) + facet +
+      geom_raster() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
+      scale_x_discrete("user",  labels="") +
+      scale_y_discrete(name="hour", labels=hr_lab, limits=0:23, expand = c(0,0)) +
+      labs(fill="prop. max count")
+}
+
+kmUserPlotter(5, compuserbyID, comp.user.short)
+
+#thing <- censor.dt[km.dt][,list(ul = length(unique(location_id))),by=list(user_id, cluster)][,max(ul),by=cluster]
+#test_users <- censor.dt[,.N,by=user_id][which(N==median(N)), user_id]
+#setkey(censor.dt, location_id)
+#typical_users <- censor.dt[km.dt][user_id %in% test_users, list(cluster=unique(cluster), count=.N, first_login=min(login), last_logout=max(logout)), keyby=list(user_id, location_id)]
+#typical_users[,.N,by=list(user_id,cluster)][N>1]
+
 ```
-
-Our synthetic population comprises long term users of the system.
-When synthezing their behavior, we should be thinking relative to other long term users of the system.
-We define *long term* as having greater than six months between first and last observed use of the system by a particular user id.
-
-For long term users, we are interested in a few first-order behaviors: how much of their time are they using the system, and how is the time spent using the system broken up into sessions?
-
-```{r}
-reference.dt <-
-  setkey(censor.dt[,
-    list(
-      overnight_logins = sum(login_day != logout_day),
-      non_overnight_logins = sum(login_day == logout_day),
-      total_session_min = sum(logout-login)/60,
-      user_lifetime_days = max(logout_day) - min(login_day)
-    ), by=user_id
-  ][
-    user_lifetime_days > 365/2
-  ][,
-    total_logins := overnight_logins + non_overnight_logins
-  ], total_logins)
-```
-
-First, we survey the relationship between total time spent on the system, period of use, and total unique logins:
-
-```{r,echo=FALSE}
-sess_scale_min <- log10(c(10, 60, 2*60, 24*60, 7*24*60, 30*24*60, 6*30*24*60, 365*24*60))
-sess_labels <- c("10 min", "1 hr", "2 hr", "1 day", "1 week", "1 month", "6 months", "1 year")
-
-use_period_scale_day <- log10(c(0.5,1:5)*365)
-use_period_labels <- c(0.5,1:5)
-
-base.plot <- ggplot(reference.dt) + theme_bw() + 
-  aes(
-    color=log10(user_lifetime_days),
-    y=log10(total_session_min),
-    x=log10(total_logins)
-  ) + 
-  scale_y_continuous(
-    "Cumulative Usage",
-    breaks=sess_scale_min,
-    labels=sess_labels
-  ) +
-  scale_color_continuous(
-    "Usage Period (Years)",
-    breaks=use_period_scale_day,
-    labels=use_period_labels
-  ) +
-  scale_x_continuous(
-    "Unique Logins",
-    breaks=log10(c(10,100,1000)),
-    labels=c(10,100,1000)
-  ) +
-  geom_point() +
-  stat_smooth(method=lm, color="red")
-base.plot
-```
-
-Visual inspection indicates that there is a log-linear dependency of total session time, $\sum{t_i}$, on login counts, $n_l$, and the period of activity seems to explain little.
-We can check this with a linear regression:
-
-```{r, echo=FALSE}
-user.model <- lm(
-  log10(total_session_min) ~ log10(user_lifetime_days) + log10(total_logins),
-  reference.dt
-)
-summary(user.model)
-```
-
-This confirms that usage period is not substantively driving total session time - unique logins are.
-However, is usage period driving the total lifetime logins?
-Certainly that would seem reasonable - the longer someone uses the system, the more login opportunities they have.
-
-```{r, echo=FALSE}
-logins.plot <- ggplot(reference.dt) + theme_bw() + 
-  aes(
-    x=log10(user_lifetime_days),
-    y=log10(total_logins)
-  ) + 
-  scale_x_continuous(
-    "Usage Period (Years)",
-    breaks=use_period_scale_day,
-    labels=use_period_labels
-  ) +
-  scale_y_continuous(
-    "Unique Logins",
-    breaks=log10(c(10,100,1000)),
-    labels=c(10,100,1000)
-  ) +
-  geom_point(alpha=0.3, size=3) +
-  stat_smooth(method=lm, color="red")
-logins.plot
-```
-
-There seems to be a steady trend:
-
-```{r, echo=FALSE}
-logins.model <- lm(
-  log10(total_logins) ~ log10(user_lifetime_days),
-  reference.dt
-)
-summary(logins.model)
-```
-
-However, little of the variation is explained.
-By visual inspection, the variation seems to remain roughly constant with the increasing trend.
-Failing to have predictive terms for that variance is not a problem, as long as we are able to replicate that amount of variation in our synthetic population.
-
-Before we try some models to reproduce that, let's consider our user behavior a bit more closely, since the system itself is growing as time proceeds.
-Particularly, we want to consider whether the growth in number of available login sites influences the number of unique logins.
-
-Consider the number of available sites:
-
-```{r,echo=FALSE}
-site.dt <- setkey(melt(
-    censor.dt[,list(end=max(logout), start=min(login)), by=location_id],
-    id.var="location_id",
-    value.name="time",
-    variable.name = "event"),
-  "time")
-site.dt[,increment := ifelse(event=="end",-1,1)]
-site.dt[,count := cumsum(increment)]
-mintime <- site.dt$time[1]
-site.dt[, time := time - mintime]
-locations.plot <- ggplot(site.dt) + theme_bw() + aes(x=time/(365*24*60*60), y=count) +
-  geom_step() +
-  scale_x_continuous("year", limits=c(0,NA))
-locations.plot
-```
-
-From this view, we see a division of eras: a period of steady growth, a period of stable turnover, then another growth period terminating in the end of the data set (where locations that likely continue to exist are censored by the end of the data set).
-While we could do a more rigorous interpretation, this is a practical approximation, and we can use straightforward linear regression to fit three trend lines, assuming that the middle line should have zero slope.
-To curtail the terminal region in the data, we will simply identify where the trend is monotonically declining.
-
-```{r, echo=FALSE, eval=FALSE}
-grow1 <- function(break1, dt) sum(lm(count ~ time, dt[time < break1])$residuals^2)
-stable <- function(break1, break2, dt) sum(dt[(break1 <= time) & (time < break2), count - mean(count)]^2)
-grow2 <- function(break2, break3, dt) sum(lm(count ~ time, dt[(break2 <= time) & (time < break3)])$residuals^2)
-decline <- function(break3, dt) sum(lm(count ~ time, dt[break3 <= time])$residuals^2)
-partition <- function(x, maxt) {
-  ## x is logit of proportions remaining in the series
-  es <- exp(x)
-  bs <- es/(1+es)
-  break1 <- bs[1]*maxt
-  break2 <- (maxt-break1)*bs[2] + break1
-  break3 <- (maxt-break2)*bs[3] + break2
-  grow1(break1) + stable(break1, break2) + grow2(break2, break3) + decline(break3)
-}
-guessp <- c(0.5, 0.5, 0.9)
-res <- with(optim(log(guessp/(1-guessp)), partition, maxt=max(site.dt$time), dt=site.dt), {
-  es <- exp(par)
-  ps <- es/(1+es)
-  maxt <- max(site.dt$time)
-  break1 <- ps[1]*maxt
-  break2 <- (maxt-break1)*ps[2] + break1
-  break3 <- (maxt-break2)*ps[3] + break2
-  c(b1=break1, b2=break2, b3=break3)  
-})
-locations.plot + geom_vline(x=res/(365*24*60*60), color="red")
-```
-
-This means fitting three values: the slope in the growth region, the transition point, and the end point for stable region.
-
-If we propose some distributions of unique logins around an increasing mean number of logins, then we can determine via maximum likelihood tests (1) what shape parameters best capture the data, and (2) which distribution models the data best.
-There are a few features of the distribution to capture:
-
- - it should be left censored.
-The outcomes are bound by $\log_{10}(2)=`r log10(2)`$ as the minimum number of logins (required to appear at least long enough to measure a lifetime in the dataset greater than a 6 month duration).
-
-First, let's consider the Poisson distribution, with the shape parameter $\lambda$ as the rate of new sessions per available location.
-That is, users of the system have an activity rate per location, so more locations.
-
-This implies that we may be able to create synthetic activity informed by the whole dataset, independently of how long a period (within the bounds of the actual empirical data) we are simulating.
-
-TODO: some simulation approaches, show those simulated people on that plot.
-
-The locations have different patterns of activity, reflecting their different hours of operation.
-
-```{r, echo=FALSE}
-hours.dt <- censor.dt[,
-  list(login_day, login_time, logout_day, logout_time, logout_time_adj=0),
-  keyby=list(location_id, user_id)
-]
-hours.dt[ login_day != logout_day,
-   logout_time_adj := logout_time + 24*60*60
-]
-hours.dt <- hours.dt[,
-  list(open=min(login_time), close=max(logout_time, logout_time_adj)),
-  by=list(login_day, location_id)
-]
-p <- ggplot(hours.dt[,
-  list(med_open = median(open), med_close=median(close)), by=location_id
-][, open_id:=.GRP, by=list(med_open, location_id)])+aes(x=open_id, y=med_open/3600) + geom_point() + geom_hline(yintercept=1:23, color="red") + theme_bw()
-# bad.list <- with(src.dt[(logout - login) > max_real_duration,],
-#   list(bad_users = unique(user_id), bad_locs = unique(location_id), bad_entries = length(user_id) )
-# )
-# 
-# duration.censored.dt <- src.dt[(logout - login) <= max_real_duration,]
-# 
-# eliminated.users <- setdiff(bad.list$bad_users, unique(duration.censored.dt$user_id))
-# eliminated.locs <- setdiff(bad.list$bad_locs, unique(duration.censored.dt$location_id))
-# warning("eliminated ", length(eliminated.users)," users with only >24 hour sessions: ", paste(eliminated.users, collapse=" "))
-# 
-# user.location.count.dt <- duration.censored.dt[,
-# 	list(location_count = length(unique(location_id))),
-# 	by="user_id"
-# ]
-# valid.users <- user.location.count.dt[location_count > 1,user_id]
-# 
-# user_and_duration.censored.dt <- duration.censored.dt[user_id %in% valid.users,]
-# 
-# locs.dt <- user_and_duration.censored.dt[,
-# 	list(
-# 		unique_users = length(unique(user_id)),
-# 		total_login_time = sum(logout - login)),
-# 	by="location_id"
-# ] # unique users per location
-# 
-# min_total_login <- 60*60 # 1 hour
-# 
-# invalid.locs <- locs.dt[(unique_users == 1) | (total_login_time < min_total_login), location_id] # locations with only one user, or total log in time < 1 hour
-# 
-# loc_user_duration.censored.dt <- user_and_duration.censored.dt[!(location_id %in% invalid.locs),]
-# loc_user_duration.censored.dt[,user_id := .GRP, by=user_id]
-# loc_user_duration.censored.dt[,location_id := .GRP, by=location_id]
-# 
-# loc_view.dt <- loc_user_duration.censored.dt[,
-#   list(
-#     unique_users = length(unique(user_id)),
-#     total_login_time = sum(logout - login),
-#     mean_login = mean(logout - login),
-#     sd_login = sd(logout - login)
-#   ),
-#   by="location_id"]
-# 
-# user_view.dt <- loc_user_duration.censored.dt[,
-# 	list(
-# 		unique_locs = length(unique(location_id)),
-# 		total_login_time = sum(logout - login),
-# 		mean_login = mean(logout - login),
-# 		sd_login = sd(logout - login),
-#     first = min(login),
-#     last = max(logout)
-# 	),
-# 	by="user_id"
-# ]
-# 
-# user_view.dt[, login_proportion := total_login_time / (last-first) ]
-# 
-# ggplot(user_view.dt) + theme_bw() +
-#   aes(x=user_id, ymin=first, ymax=last, alpha=login_proportion) + geom_linerange() + coord_flip()
-# 
-# 
-# break_categorization <- function(dt, target, output) {
-# 	bks <- hist(dt[[target]], plot = F)$breaks
-# 	lbls <- paste(head(bks, -1), tail(bks, -1), sep="-")
-# 	dt[[output]] <- sapply(dt[[target]], function(n) factor(lbls[which.max(n <= bks)-1], levels=lbls))
-# 	dt
-# }
-# 
-# #loc_count_breaks <- hist(user_view.dt$unique_locs, plot = F)$breaks
-# 
-# #loc_count_cats <- paste(head(loc_count_breaks, -1), tail(loc_count_breaks, -1), sep="-")
-# 
-# #user_view.dt[,loc_count_cat := sapply(unique_locs, function(n) which.max(n <= loc_count_breaks)-1)]
-# #user_view.dt$loc_count_cat <- factor(loc_count_cats[user_view.dt$loc_count_cat], levels=loc_count_cats)
-# 
-# user_view.dt <- break_categorization(user_view.dt, "unique_locs", "loc_count_cat")
-# 
-# # sum(loc_user_duration.censored.dt[,length(unique(location_id)),by="user_id"]$V1 == 1) == 0 # => no more user ids to censor
-# 
-# total_login_duration <- sum(as.numeric(loc_view.dt$total_login_time))
-# loc_view.dt[,proportion := total_login_time/total_login_duration]
-# 
-# nets.dt <- loc_user_duration.censored.dt[, list(first=min(login),last=max(logout)), by=c("user_id","location_id")]
-# 
-# melt_nets.dt <- melt(nets.dt, id.vars = c("user_id", "location_id"))
-# melt_nets.dt$inc <- ifelse(melt_nets.dt$variable == "first", 1, -1)
-# setkey(melt_nets.dt, value, user_id, location_id)
-# 
-# melt_nets.dt[,net_users := cumsum(inc), by="location_id"]
-# melt_nets.dt[,net_locs  := cumsum(inc), by="user_id"]
-# 
-# acc_loc_time <- melt_nets.dt[,
-# 	list(ave_concurrent_locs = 
-# 	  sum(diff(value)*head(net_locs,-1)) /
-# 	  sum(diff(value)*(head(net_locs,-1)>0)),
-# 	  unique_locs = sum(abs(inc))/2
-# 	), # exclude time intervals w/ no presence
-# 	by="user_id"]
-# acc_loc_time <- break_categorization(acc_loc_time, "ave_concurrent_locs","con_locs_cat")
-# acc_loc_time <- break_categorization(acc_loc_time, "unique_locs","unique_locs_cat")
-# 
-# ggplot(acc_loc_time[ave_concurrent_locs >= 2,]) + theme_bw() +
-# 	aes(x=ave_concurrent_locs, fill=unique_locs_cat) + geom_bar() +
-# 	scale_x_log10() #+ scale_y_log10()
-# #sample(loc_view.dt$location_id, 5, F, loc_view.dt$proportion)
-# 
-# p.loc <- ggplot(loc_view.dt) + theme_bw()
-# p.user <- ggplot(user_view.dt) + theme_bw()
-# 
-# p.user + theme(panel.margin = unit(0.5, "lines")) +
-# 	aes(fill=loc_count_cat, x=total_login_time/60) +
-# 	facet_grid(loc_count_cat ~ ., scales="free_y", shrink=T, drop=F) +
-# 	geom_bar() + scale_x_log10(name="total login time per individual, in minutes") +
-# 	ylab("number of individuals") +
-# 	scale_fill_discrete(name="locations per individual", drop=F) +
-# 	scale_y_continuous(breaks=function(lims) c(head(lims,1), tail(lims, 1)), expand=c(0,0))
-# 
-# p + aes(x=(logout - login)) + geom_bar() + geom_vline(x=150, color="red") + scale_x_log10()
-# 
-# #####
-# p + aes(x=unique_users) + geom_bar() + scale_x_log10()
-# p + aes(x=total_login_time) + geom_bar() + scale_x_log10()
-# 
-# src.dt[,length(unique(user_id)),by="location_id"] # unique users per location
-# src.dt[,length(unique(location_id)),by="user_id"] # unique locations by user
-# 
-# regular_users <- src.dt[,list(regular = length(location_id) > 1), by="user_id"] # users w/ more than one login
-# regular_users <- regular_users[regular == T, user_id, keyby="user_id"]$user_id
-```

### 2015-07-14 rscripts/data_review.Rmd
fix day split issue in kmeans clustering

@@
 ```{r, echo=FALSE, include=FALSE}
-req.packages <- c("data.table", "ggplot2", "reshape2", "grid")
+req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales")
 sapply(req.packages, require, character.only = T)
@@
 ```{r, echo=FALSE}
-shft <- 4
-orig <- 0:23
-hr_shf <- (orig - shft) %% 24
-hr_lab <- (orig + shft) %% 24
+# shifthrs_trans <- function(hrs) {
+#   ihrs <- as.integer(hrs)
+#   scales::trans_new(
+#     paste0("shift",ihrs,"hrs"),
+#     function(y) {
+#       if (!is.numeric(y)) return(y)
+#       (floor(y) - ihrs) %% 24L
+#     },
+#     function(yinv) {
+#       if (!is.numeric(yinv)) return(yinv)
+#       (floor(yinv) + ihrs) %% 24L
+#     },
+#   )
+# }
+# shft <- 4
+# orig <- 0:23
+# hr_shf <- (orig - shft) %% 24
+# hr_lab <- (orig + shft) %% 24
 base <- censor.dt[,
@@
   list(
-    hour=hr_shf[login_hour+1],
+    hour=as.integer(login_hour),
     norm=login_count/max(login_count),
@@
   list(
-    hour=hr_shf[logout_hour+1],
+    hour=as.integer(logout_hour),
     norm=logout_count/max(logout_count),
@@
     ggplot(src[cluster.dt]) +
-      theme_bw() +
-      aes(x = factor(location_id), y = hour, fill = norm) + facet +
-      geom_raster() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
+      theme_bw() + theme(panel.grid=element_blank(), panel.grid.major.y = element_line()) +
+      aes(x = factor(location_id), y = factor(hour), fill = norm) + facet +
+      geom_tile() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
       scale_x_discrete("location",  labels="") +
-      scale_y_discrete(name="hour", labels=hr_lab, limits=0:23, expand = c(0,0)) +
+      scale_y_discrete(name="hour") +
+#      coord_trans(ytrans = shifthrs_trans(4)) +
       labs(fill="prop. max count")
@@
   Map(function(hour) censor.dt[
-    (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600)), 
+    ((login_day == logout_day) & (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600))) |
+    ((login_day != logout_day) & ((floor(login_time/3600) <= hour) | (hour  <= floor(logout_time/3600))) ),
     list(count=.N, hour=hour, event="usage"),
@@
 
-## TODO: handle day splits!
+## handle day splits:
+## if login_day == logout_day
+##  (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600))
+## else - n.b., already precluded logout_time > login_time, since enforced < 24hr sessions
+##  hour >= login | hour <= logout
 
@@
   Map(function(hour) censor.dt[
-    (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600)), 
+    ((login_day == logout_day) & (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600))) |
+    ((login_day != logout_day) & ((floor(login_time/3600) <= hour) | (hour  <= floor(logout_time/3600))) ), 
     list(count=.N, hour=hour, event="usage"),
@@
     ggplot(src[cluster.dt]) +
-      theme_bw() +
-      aes(x = factor(user_id), y = hour, fill = norm) + facet +
+      theme_bw() + theme(panel.grid=element_blank()) +
+      aes(x = factor(user_id), y = factor(hour), fill = norm) + facet +
       geom_raster() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
       scale_x_discrete("user",  labels="") +
-      scale_y_discrete(name="hour", labels=hr_lab, limits=0:23, expand = c(0,0)) +
+      scale_y_discrete(name="hour") +
       labs(fill="prop. max count")

### 2015-07-14 rscripts/data_review.Rmd
reorganize clustering functions

@@
 ```{r, echo=FALSE}
-# shifthrs_trans <- function(hrs) {
-#   ihrs <- as.integer(hrs)
-#   scales::trans_new(
-#     paste0("shift",ihrs,"hrs"),
-#     function(y) {
-#       if (!is.numeric(y)) return(y)
-#       (floor(y) - ihrs) %% 24L
-#     },
-#     function(yinv) {
-#       if (!is.numeric(yinv)) return(yinv)
-#       (floor(yinv) + ihrs) %% 24L
-#     },
-#   )
-# }
-# shft <- 4
-# orig <- 0:23
-# hr_shf <- (orig - shft) %% 24
-# hr_lab <- (orig + shft) %% 24
 base <- censor.dt[,
@@
 
+# animation::saveVideo({
+#   
+# })
 #thing <- censor.dt[km.dt][,list(ul = length(unique(location_id))),by=list(user_id, cluster)][,max(ul),by=cluster]

### 2015-07-14 rscripts/munging.R

@@ -22,3 +22,19 @@ trimLimits <- function(tar.dt, lim.dt) {
   setkeyv(res.dt, key(tar.dt))
   res.dt
 }
+
+kmParse <- function(src, k, nstart = k, key, refevent = "login") {
+  castdata <- reshape2::acast(src, eval(parse(text=paste0(key,"~hour+event"))), value.var = "norm", fill = 0)
+  km <- stats::kmeans(castdata, k, nstart = nstart)
+  cluster.dt <- data.table(cluster = km$cluster)
+  cluster.dt[[key]] = as.integer(names(km$cluster))
+  setkeyv(cluster.dt, key)
+  peakhours <- setkey(src[cluster.dt][,
+    list(peak=sum(hour*norm)/sum(norm)),
+    by=list(event, cluster)
+  ], event, peak)
+  reorder <- peakhours[event == refevent, order(cluster)]
+  peakhours[, cluster := reorder[cluster]]
+  cluster.dt[,cluster := reorder[cluster]]
+  return(list(clustering = cluster.dt, peaks = peakhours))
+}

### 2015-07-15 rscripts/data_review.Rmd
incipient user cluster pictures

@@
 
-## handle day splits:
-## if login_day == logout_day
-##  (floor(login_time/3600) <= hour) & (hour  <= floor(logout_time/3600))
-## else - n.b., already precluded logout_time > login_time, since enforced < 24hr sessions
-##  hour >= login | hour <= logout
-
 comp.user.short <- setkey(rbindlist(
@@
 )
+
 comp.user.short[,norm := count/max(count), by=user_id]
@@
 
-kmUserPlotter(5, compuserbyID, comp.user.short)
+## let's figure out user clusters
+## then join censor.dt to user clusters
+## then track those clusters as:
+##  panel col, color for each cluster (all in same row, since y axis shared)
+##  yaxis - cluster size / total user base at that time?
+##  line for current total in cluster
+##  width to represent turnover?  like ymax = count + additions, ymin = count - losses
+##  x axis - week; day will be too noisy w/ turnover / count?
+
+user.clusters <- kmParse(comp.user.short, 5, key="user_id", refevent="usage")
+censor.dt[user.clusters$clustering][,
+  list(leave=max(logout_time), join=min(login_time)), by=list(user_id,cluster)
+]
 

### 2015-07-16 rscripts/data_review.Rmd
user clusters displayed similarly to location clusters

@@
 req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales")
-sapply(req.packages, require, character.only = T)
+need.pkgs <- req.packages[!sapply(req.packages, require, character.only = T)]
+if (length(need.pkgs) != 0) install.packages(need.pkgs, dependencies = T)
 ```
@@
   ulc <- left[,unique(cluster)]
-  rbind(left, right, 
-    data.table(week=ul, inc=0, cluster=right[,unique(cluster)]),
-    data.table(week=rep(ur, times=length(ulc)), inc=0, cluster=rep(ulc, each=length(ur)))
-  )
+  temp <- rbind(left, right)
+  if (length(ul) != 0) temp <- rbind(temp, data.table(week=ul, inc=0, cluster=right[,unique(cluster)]))
+  if (length(ur) != 0) temp <- rbind(temp, data.table(week=rep(ur, times=length(ulc)), inc=0, cluster=rep(ulc, each=length(ur))))
+  temp
 }, Map(function(clus) week.melt[cluster == clus], week.melt[,unique(cluster)])), cluster, week)
@@
 user.clusters <- kmParse(comp.user.short, 5, key="user_id", refevent="usage")
-censor.dt[user.clusters$clustering][,
-  list(leave=max(logout_time), join=min(login_time)), by=list(user_id,cluster)
+setkey(censor.dt, user_id)
+user.melt <- melt(censor.dt[user.clusters$clustering][,
+  list(leave=max(logout), join=min(login)), by=list(user_id, cluster)
+], id.vars = c("user_id","cluster"), value.name = "time", variable.name = "event")
+
+agg.dt <- user.melt[, 
+  week := floor(time/60/60/24/7)
+][,
+  inc := ifelse(event == "leave",-1,1)
+][,list(inc = sum(inc)),keyby=list(week, cluster)]
+
+agg.fill <- setkey(Reduce(function(left, right) {
+  ul <- left[!(week %in% right[,unique(week)]), unique(week)]
+  ur <- right[!(week %in% left[,unique(week)]), unique(week)]
+  ulc <- left[,unique(cluster)]
+  temp <- rbind(left, right)
+  if (length(ul) != 0) temp <- rbind(temp, data.table(week=ul, inc=0, cluster=right[,unique(cluster)]))
+  if (length(ur) != 0) temp <- rbind(temp, data.table(week=rep(ur, times=length(ulc)), inc=0, cluster=rep(ulc, each=length(ur))))
+  temp
+}, Map(function(clus) agg.dt[cluster == clus], agg.dt[,unique(cluster)])), cluster, week)
+
+agg.final <- agg.fill[,
+  list(net = sum(inc), gain = sum(inc[inc > 0]), loss = sum(inc[inc < 0])),
+  by=list(week, cluster)
+][,
+  list(net, gain, loss, cum=cumsum(net), week),
+  keyby=cluster
 ]
+
+ggplot(agg.final) + theme_bw() +
+  aes(fill = factor(cluster), x = week, y = cum) + geom_area(position="fill") +
+  ylab("prop. users by cluster") + labs(fill="cluster") +
+  coord_cartesian(xlim=c(34, 300), ylim=c(0,1))
 

### 2015-07-20 rscripts/data_review.Rmd
notes about 3x3 plot
progress on panel figures

@@
 source("munging.R")
-setkey(breakoutDays(censor.dt[,
+zeroize(breakoutDays(censor.dt[,
   user_id := .GRP, by=user_id
@@
   location_id := .GRP, by=location_id
-]), login_day, login_time, logout_day, logout_time, user_id, location_id)
+], login_day, login_time, logout_day, logout_time, user_id, location_id))
 censor.dt
@@
   ylab("prop. users by cluster") + labs(fill="cluster") +
-  coord_cartesian(xlim=c(34, 300), ylim=c(0,1))
+  coord_cartesian(ylim=c(0,1))
 
@@
 ```
+
+clustering locations:
+
+ try categories of low, med, high usage rates
+ vs short, med, long login times
+
+ color vs y for membership
+
+ locations that disappeared between intervals - add holes?
+
+ rolling aggregration windows
+
+ highlight changes by size (+alpha)?
+
+```{r, echo=F}
+duration_breaks <- censor.dt[,quantile(logout-login, probs=(1:2)/3)]
+daily_login_breaks <- censor.dt[,.N,by=list(location_id, login_day)][,quantile(N,probs=(1:2)/3)] ## TODO zeros?
+# for 90 day window
+#  for extant locations
+#   - what percent of active days fall in each login category?
+#   - what percent of logins fall in each login category?
+```

### 2015-07-20 rscripts/munging.R

@@ -1,7 +1,7 @@
 ## data munging utils
 require(data.table)
-breakoutDays <- function(dt) {
-  dt[,
+breakoutDays <- function(dt, ...) {
+  setkey(dt[,
     user_id     := .GRP, by=user_id
   ][,
     location_id := .GRP, by=location_id
@@ -13,8 +13,18 @@ breakoutDays <- function(dt) {
     login_time  := login - login_day*24*60*60
   ][,
     logout_time := logout - logout_day*24*60*60
-  ]
+  ], ...)
 }
+
+zeroize <- function(dt, z = dt[1, login-login_time], dz = dt[1, login_day], ks=key(dt)) setkeyv(dt[,
+  login := login - z
+][,
+  logout := logout - z
+][,
+  login_day := login_day - dz
+][,
+  logout_day := logout_day - dz
+], ks)
 
 trimLimits <- function(tar.dt, lim.dt) {
   res.dt <- merge(tar.dt, lim.dt, by = "location_id")[(login >= first) & (logout <= last),]

### 2015-07-21 rscripts/data_review.Rmd
draft 9x9 plot

@@
 daily_login_breaks <- censor.dt[,.N,by=list(location_id, login_day)][,quantile(N,probs=(1:2)/3)] ## TODO zeros?
+duration_group.dt <- censor.dt[90 <= login_day & login_day < 180, 
+  list( # initial pass to get temp variables
+    low = sum(logout - login < duration_breaks[1]),
+    mid = sum(logout - login < duration_breaks[2]),
+    high = .N
+  ), 
+  keyby=location_id
+][,
+  list( # use temp vars to calc actuals
+    low = low/high,
+    mid = (mid-low)/high,
+    high = (high-mid)/high
+  ),
+  keyby=location_id
+]
+
+login_group.dt <- censor.dt[90 <= login_day & login_day < 180, 
+  list(count = .N), 
+  by=list(login_day, location_id)
+][,
+  list( # initial pass to get temp variables
+    low = sum(count < daily_login_breaks[1]),
+    mid = sum(count < daily_login_breaks[2]),
+    high = .N
+  ), 
+  keyby=location_id
+][,
+  list( # use temp vars to calc actuals
+    low = low/high,
+    mid = (mid-low)/high,
+    high = (high-mid)/high
+  ),
+  keyby=location_id
+]
+
+cross_data.dt <- merge(
+  melt(
+    duration_group.dt, id.vars = "location_id",
+    value.name="membership", variable.name="group"
+  ),
+  melt(
+    login_group.dt, id.vars = "location_id",
+    value.name="membership", variable.name="group"
+  ),
+  by = "location_id",
+  allow.cartesian = T,
+  suffixes = c(".duration",".login")
+)
+
+ggplot(cross_data.dt) + theme_bw() +
+  facet_grid(group.duration ~ group.login, labeller = function(var, val) {
+    paste0(val,ifelse(var == "group.duration"," duration"," logins"))
+  }) +
+  aes(x=location_id, y = membership.duration*membership.login) +
+  geom_point()
 # for 90 day window

### 2015-07-21 rscripts/data_review.Rmd
loop for animation

--- 2015-07-20 rscripts/data_review.Rmd
@@
 daily_login_breaks <- censor.dt[,.N,by=list(location_id, login_day)][,quantile(N,probs=(1:2)/3)] ## TODO zeros?
-# for 90 day window
-#  for extant locations
-#   - what percent of active days fall in each login category?
-#   - what percent of logins fall in each login category?
+
+offset <- window <- 90; slide <- 30 ## TODO: go by actual months
+
+divy <- function(dt) dt[,
+  list( # use temp vars to calc actuals
+    low = low/high,
+    mid = (mid-low)/high,
+    high = (high-mid)/high
+  ),
+  keyby=location_id
+]
+
+cross_group <- function(timeslice, duration_breaks, daily_login_breaks) {
+  duration_group.dt <- divy(timeslice[,
+    list( # initial pass to get temp variables
+      low = sum((logout - login) <= duration_breaks[1]),
+      mid = sum((logout - login) <= duration_breaks[2]),
+      high = .N
+    ), 
+    keyby=location_id
+  ])
+  login_group.dt <- divy(timeslice[, 
+    list(count = .N), 
+    by=list(login_day, location_id)
+  ][,
+    list( # initial pass to get temp variables
+      low = sum(count <= daily_login_breaks[1]),
+      mid = sum(count <= daily_login_breaks[2]),
+      high = .N
+    ), 
+    keyby=location_id
+  ])
+  cross_data.dt <- merge(
+    melt(
+      duration_group.dt, id.vars = "location_id",
+      value.name="membership", variable.name="group"
+    ),
+    melt(
+      login_group.dt, id.vars = "location_id",
+      value.name="membership", variable.name="group"
+    ),
+    by = "location_id",
+    allow.cartesian = T,
+    suffixes = c(".duration",".login")
+  )
+}
+# censor.dt[,max(login_day)]
+
+prev.dt <- cross_group(censor.dt[offset <= login_day & login_day < (offset+window)], duration_breaks, daily_login_breaks)
+for (start in seq(offset+slide, 600-window, by=slide)) {
+  timeslice <- censor.dt[start <= login_day & login_day < (start+window)]
+  cross_data.dt <- cross_group(timeslice, duration_breaks, daily_login_breaks)
+  plotting.dt <- merge(cross_data.dt, prev.dt, by=c("location_id", "group.duration", "group.login"), all=T, suffixes = c("",".prev"))
+  print(ggplot(plotting.dt[!is.na(membership.duration.prev)]) + theme_bw() +
+    facet_grid(group.duration ~ group.login, labeller = function(var, val) {
+      paste0(val,ifelse(var == "group.duration"," duration"," logins"))
+    }) +
+    aes(x=location_id, y = membership.duration*membership.login,
+        color = membership.duration*membership.login - membership.duration.prev*membership.login.prev) +
+    geom_point(na.rm = TRUE) +
+    geom_point(data = plotting.dt[is.na(membership.duration.prev)]) +
+    geom_point(
+      data=plotting.dt[
+        is.na(membership.duration),
+        list(
+          membership.duration=membership.duration.prev,
+          membership.login=membership.login.prev
+        ),
+        by=list(group.duration, group.login, location_id)
+      ], color="black", shape = 1) +
+    scale_color_gradient("change in membership", limits = c(-1,1), low="red", high="blue", na.value = "black")
+  )
+  prev.dt <- cross_data.dt
+}
 ```

### 2015-07-21 rscripts/data_review.Rmd
gif movie

--- 2015-07-20 rscripts/data_review.Rmd
@@
 ```{r, echo=FALSE, include=FALSE}
-req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales")
+req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation")
 need.pkgs <- req.packages[!sapply(req.packages, require, character.only = T)]
@@
 daily_login_breaks <- censor.dt[,.N,by=list(location_id, login_day)][,quantile(N,probs=(1:2)/3)] ## TODO zeros?
-# for 90 day window
-#  for extant locations
-#   - what percent of active days fall in each login category?
-#   - what percent of logins fall in each login category?
+
+offset <- window <- 90; slide <- 30 ## TODO: go by actual months
+
+divy <- function(dt) dt[,
+  list( # use temp vars to calc actuals
+    low = low/high,
+    mid = (mid-low)/high,
+    high = (high-mid)/high
+  ),
+  keyby=location_id
+]
+
+cross_group <- function(timeslice, duration_breaks, daily_login_breaks) {
+  duration_group.dt <- divy(timeslice[,
+    list( # initial pass to get temp variables
+      low = sum((logout - login) <= duration_breaks[1]),
+      mid = sum((logout - login) <= duration_breaks[2]),
+      high = .N
+    ), 
+    keyby=location_id
+  ])
+  login_group.dt <- divy(timeslice[, 
+    list(count = .N), 
+    by=list(login_day, location_id)
+  ][,
+    list( # initial pass to get temp variables
+      low = sum(count <= daily_login_breaks[1]),
+      mid = sum(count <= daily_login_breaks[2]),
+      high = .N
+    ), 
+    keyby=location_id
+  ])
+  cross_data.dt <- merge(
+    melt(
+      duration_group.dt, id.vars = "location_id",
+      value.name="membership", variable.name="group"
+    ),
+    melt(
+      login_group.dt, id.vars = "location_id",
+      value.name="membership", variable.name="group"
+    ),
+    by = "location_id",
+    allow.cartesian = T,
+    suffixes = c(".duration",".login")
+  )
+}
+# censor.dt[,max(login_day)]
+
+base_scale_x <- scale_x_discrete("location",
+  limits = 1:censor.dt[,max(unique(location_id))])
+base_scale_y <- scale_y_continuous("membership", limits=c(0,1))
+
+base_plot <- ggplot() + theme_bw() +
+    facet_grid(group.duration ~ group.login, labeller = function(var, val) {
+      paste0(val,ifelse(var == "group.duration"," duration"," logins"))
+    }) +
+    aes(x=location_id, y=membership.duration*membership.login,
+        color = membership.duration*membership.login - membership.duration.prev*membership.login.prev) +
+    base_scale_x + base_scale_y + theme(
+      axis.text.x = element_blank(),
+      panel.grid.major.x = element_blank(),
+      panel.grid.minor.x = element_blank()
+    ) +
+    scale_color_gradient("change in membership", limits = c(-1,1), low="red", high="blue", na.value = "black")
+
+prev.dt <- cross_group(censor.dt[offset <= login_day & login_day < (offset+window)], duration_breaks, daily_login_breaks)
+saveGIF(for (start in seq(offset+slide, censor.dt[,max(login_day)]-window, by=slide)) {
+  timeslice <- censor.dt[start <= login_day & login_day < (start+window)]
+  cross_data.dt <- cross_group(timeslice, duration_breaks, daily_login_breaks)
+  plotting.dt <- merge(cross_data.dt, prev.dt, by=c("location_id", "group.duration", "group.login"), all=T, suffixes = c("",".prev"))
+  print(base_plot +
+    geom_point(data = plotting.dt[!is.na(membership.duration.prev)], na.rm = TRUE) +
+    geom_point(data = plotting.dt[is.na(membership.duration.prev)]) +
+    geom_point(
+      data=plotting.dt[
+        is.na(membership.duration),
+        list(
+          membership.duration=membership.duration.prev,
+          membership.login=membership.login.prev
+        ),
+        by=list(group.duration, group.login, location_id)
+      ], color="black", shape = 1)
+  )
+  prev.dt <- cross_data.dt
+})
 ```

### 2015-08-10 rscripts/data_review.Rmd
updates

@@
 ```
+
+Looking at day-of-week, week-of-year patterns:
+
+```{r, echo=F}
+byweekday.dt <- censor.dt[,
+  .N,
+  keyby=list(location_id, weekday)
+]
+loctotals <- byweekday.dt[,list(total=sum(N)), keyby=location_id]
+weekday.kmeans.src <- byweekday.dt[loctotals][,list(distribution=N/total), keyby=list(location_id, weekday)]
+kmByWeekDayData <- kmParse2(weekday.kmeans.src, 10, 5, "location_id", "weekday")
+ggplot(weekday.kmeans.src[kmByWeekDayData$clustering]) +
+  theme_bw() + facet_grid(. ~ cluster, scales="free") +
+  aes(x=factor(location_id), y=weekday, fill=log(distribution)) +
+  geom_raster() +
+  scale_x_discrete("location_id", labels="")
+
+byweek.dt <- censor.dt[,
+  .N,
+  keyby=list(location_id, week)
+]
+week.kmeans.src <- byweek.dt[
+  byweek.dt[,list(total=sum(N)), keyby=location_id]
+][,
+  list(distribution=N/total), keyby=list(location_id, week)
+]
+kmByWeekData <- kmParse2(week.kmeans.src, 10, 5, "location_id", "week")
+ggplot(week.kmeans.src[kmByWeekData$clustering]) +
+  theme_bw() + facet_grid(. ~ cluster, scales="free") +
+  aes(x=factor(location_id), y=week, fill=log(distribution)) +
+  geom_raster() +
+  scale_x_discrete("location_id", labels="")
+
+bymonth.dt <- censor.dt[,
+  .N,
+  keyby=list(location_id, month)
+]
+month.kmeans.src <- bymonth.dt[
+  bymonth.dt[,list(total=sum(N)), keyby=location_id]
+][,
+  list(distribution=N/total), keyby=list(location_id, month)
+]
+kmBymonthData <- kmParse2(month.kmeans.src, 10, 5, "location_id", "month")
+ggplot(month.kmeans.src[kmBymonthData$clustering]) +
+  theme_bw() + facet_grid(. ~ cluster, scales="free") +
+  aes(x=factor(location_id), y=month, fill=log(distribution)) +
+  geom_raster() +
+  scale_x_discrete("location_id", labels="")
+
+```

### 2015-08-10 rscripts/munging.R

@@ -1,5 +1,5 @@
 ## data munging utils
-require(data.table)
+require(data.table); require(lubridate)
 breakoutDays <- function(dt, ...) {
   setkey(dt[,
     user_id     := .GRP, by=user_id
@@ -16,15 +16,26 @@ breakoutDays <- function(dt, ...) {
   ], ...)
 }
 
-zeroize <- function(dt, z = dt[1, login-login_time], dz = dt[1, login_day], ks=key(dt)) setkeyv(dt[,
-  login := login - z
-][,
-  logout := logout - z
-][,
-  login_day := login_day - dz
-][,
-  logout_day := logout_day - dz
-], ks)
+zeroize <- function(dt, z = dt[1, login-login_time], dz = dt[1, login_day], ks=key(dt), ref.yr=2005) {
+  b <- ymd(paste(ref.yr, 01, 01))
+  mul <- 24*3600
+  dt[,
+    weekday := wday(b+mul*(login_day-1), label = T)
+  ][,
+    week    := week(b+mul*(login_day-1))
+  ][,
+    month   := month(b+mul*(login_day-1), label = T)
+  ]
+  setkeyv(dt[,
+    login := login - z
+  ][,
+    logout := logout - z
+  ][,
+    login_day := login_day - dz
+  ][,
+    logout_day := logout_day - dz
+  ], ks)
+}
 
 trimLimits <- function(tar.dt, lim.dt) {
   res.dt <- merge(tar.dt, lim.dt, by = "location_id")[(login >= first) & (logout <= last),]
@@ -47,4 +58,13 @@ kmParse <- function(src, k, nstart = k, key, refevent = "login") {
   peakhours[, cluster := reorder[cluster]]
   cluster.dt[,cluster := reorder[cluster]]
   return(list(clustering = cluster.dt, peaks = peakhours))
+}
+
+kmParse2 <- function(src, k, nstart = k, key, category) {
+  castdata <- reshape2::acast(src, eval(parse(text=paste0(key,"~",category))), value.var = "distribution", fill = 0)
+  km <- stats::kmeans(castdata, k, nstart = nstart)
+  cluster.dt <- data.table(cluster = km$cluster)
+  cluster.dt[[key]] = as.integer(names(km$cluster))
+  setkeyv(cluster.dt, key)
+  return(list(clustering = cluster.dt))
 }

### 2015-08-10 rscripts/data_review.Rmd
minor tweaks, mostly reading atm

@@
 ```{r, echo=FALSE, include=FALSE}
-req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation")
+req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation", "randomForest")
 need.pkgs <- req.packages[!sapply(req.packages, require, character.only = T)]
@@
 prev.dt <- cross_group(censor.dt[offset <= login_day & login_day < (offset+window)], duration_breaks, daily_login_breaks)
-saveGIF(for (start in seq(offset+slide, censor.dt[,max(login_day)]-window, by=slide)) {
-  timeslice <- censor.dt[start <= login_day & login_day < (start+window)]
-  cross_data.dt <- cross_group(timeslice, duration_breaks, daily_login_breaks)
-  plotting.dt <- merge(cross_data.dt, prev.dt, by=c("location_id", "group.duration", "group.login"), all=T, suffixes = c("",".prev"))
-  print(base_plot +
-    geom_point(data = plotting.dt[!is.na(membership.duration.prev)], na.rm = TRUE) +
-    geom_point(data = plotting.dt[is.na(membership.duration.prev)]) +
-    geom_point(
-      data=plotting.dt[
-        is.na(membership.duration),
-        list(
-          membership.duration=membership.duration.prev,
-          membership.login=membership.login.prev
-        ),
-        by=list(group.duration, group.login, location_id)
-      ], color="black", shape = 1)
-  )
-  prev.dt <- cross_data.dt
-})
+# saveGIF(for (start in seq(offset+slide, censor.dt[,max(login_day)]-window, by=slide)) {
+#   timeslice <- censor.dt[start <= login_day & login_day < (start+window)]
+#   cross_data.dt <- cross_group(timeslice, duration_breaks, daily_login_breaks)
+#   plotting.dt <- merge(cross_data.dt, prev.dt, by=c("location_id", "group.duration", "group.login"), all=T, suffixes = c("",".prev"))
+#   print(base_plot +
+#     geom_point(data = plotting.dt[!is.na(membership.duration.prev)], na.rm = TRUE) +
+#     geom_point(data = plotting.dt[is.na(membership.duration.prev)]) +
+#     geom_point(
+#       data=plotting.dt[
+#         is.na(membership.duration),
+#         list(
+#           membership.duration=membership.duration.prev,
+#           membership.login=membership.login.prev
+#         ),
+#         by=list(group.duration, group.login, location_id)
+#       ], color="black", shape = 1)
+#   )
+#   prev.dt <- cross_data.dt
+# })
 ```

### 2015-08-10 rscripts/munging.R

@@ -16,7 +16,7 @@ breakoutDays <- function(dt, ...) {
   ], ...)
 }
 
-zeroize <- function(dt, z = dt[1, login-login_time], dz = dt[1, login_day], ks=key(dt), ref.yr=2005) {
+zeroize <- function(dt, z = dt[1, login-login_time], dz = dt[1, login_day], ks=key(dt), ref.yr=2004) {
   b <- ymd(paste(ref.yr, 01, 01))
   mul <- 24*3600
   dt[,

### 2015-08-11 rscripts/data_review.Rmd
prepping for PCA

@@
 ```
+
+1. consider subset - subset of records? subset of users?  subset of locations?  want predictions about location "class", so subset of locations?  later, when locations classified, subset of users.
+2. using subset and some predictor measurements
+
+```{r,echo=F}
+## principle components analysis
+predictors <- function(SD) with(SD, {
+  duration <- logout - login
+  total_login_time <- sum(duration)
+  SD[,sum(logout - login), by=weekday]
+  list(
+    lifetime = max(logout) - min(login),
+    ave_duration = mean(duration),
+    med_duration = median(duration),
+    sd_duration = sd(duration),
+    max_duration = max(duration),
+    total_user_time = total_login_time
+  )
+})
+
+censor.dt[,predictors(.SD), by=location_id]
+```
+
+* * *
 

### 2015-08-11 rscripts/munging.R

@@ -1,18 +1,23 @@
 ## data munging utils
 require(data.table); require(lubridate)
 breakoutDays <- function(dt, ...) {
+  secs_per_day <- 24*60*60
   setkey(dt[,
     user_id     := .GRP, by=user_id
   ][,
     location_id := .GRP, by=location_id
   ][,
-    login_day   := login %/% (24*60*60)
+    login_day   := login %/% secs_per_day
   ][,
-    logout_day  := logout %/% (24*60*60)
+    logout_day  := logout %/% secs_per_day
   ][,
-    login_time  := login - login_day*24*60*60
+    login_time  := login - login_day*secs_per_day
   ][,
-    logout_time := logout - logout_day*24*60*60
+    logout_time := logout - logout_day*secs_per_day
+  ][,
+    login_day_secs := ifelse(login_day == logout_day, logout_time - login_time, secs_per_day - login_time)
+  ][,
+    logout_day_secs := ifelse(login_day == logout_day, 0, logout_time)
   ], ...)
 }
 

### 2015-08-13 rscripts/data_review.Rmd
add pca

--- 2015-08-10 rscripts/data_review.Rmd
@@
 ```
+
+1. consider subset - subset of records? subset of users?  subset of locations?  want predictions about location "class", so subset of locations?  later, when locations classified, subset of users.
+2. using subset and some predictor measurements
+
+```{r,echo=F}
+## principle components analysis
+weekday_zeroes <- data.table(weekday=factor(c("Sun", "Mon", "Tues", "Wed","Thurs","Fri","Sat"), ordered = T), usage = 0, key = "weekday")
+predictors <- function(SD) with(SD, {
+  duration <- logout - login
+  total_login_time <- sum(duration)
+  usage_by_weekday <- SD[, list(usage = sum(login_day_secs)), keyby=weekday][weekday_zeroes][, list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=weekday]
+  temp <- SD[, sum(logout_day_secs), keyby=weekday][weekday_zeroes][, list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=weekday]
+  usage_by_weekday$usage <- usage_by_weekday$usage + c(temp$usage[7], temp$usage[-7])
+  store <- as.list(usage_by_weekday$usage / total_login_time)
+  names(store) <- usage_by_weekday$weekday
+  reduced_store <- list(weekend=store$Fri+store$Sat, midweek=store$Tues+store$Wed+store$Thurs, weekbegin = store$Sun+store$Mon)
+  user_count <- length(unique(user_id))
+  life <- max(logout) - min(login)
+  return(c(list(
+    log10_lifetime = log10(life),
+    ave_duration = mean(duration),
+    med_duration = median(duration),
+    sd_duration = sd(duration),
+    max_duration = max(duration),
+    total_user_time = total_login_time,
+    log10_user_per_life = log10(user_count/life),
+    log10_time_per_user = log10(total_login_time/user_count),
+    log10_time_per_life = log10(total_login_time/life),
+    log10_unique_users = log10(user_count)
+  ), reduced_store))
+})
+unique_locs <- unique(censor.dt$location_id)
+slice <- sort(sample(unique_locs, size = length(unique_locs)/5, replace = F))
+pca.dt <- censor.dt[, predictors(.SD), by=location_id]
+centered <- pca.dt[,lapply(.SD[,-1,with=F],function(col) col-mean(col)),]
+scaled <- centered[,lapply(.SD, function(col) col/sd(col))]
+prediction.dt <- scaled[slice]
+validation.dt <- scaled[!slice]
+res <- prcomp(prediction.dt[, -1, with=F], scale. = F)
+res$var <- res$sdev^2
+res$pve <- res$var/sum(res$var)
+keep <- res$pve > 0.01 # 1/length(res$pve)
+orig.p <- par(mfrow=c(1,sum(keep)-1))
+for (x in 1:(sum(keep)-1)) biplot(res, choices=c(x,x+1), scale = 0)
+par(orig.p)
+with(res, {
+  plot(pve, xlab="Principal Component", ylab="Proportion of Variance Explained", type="b", log="y")
+  abline(h=1/(length(pve)))
+  # abline(h=tail(pve, 1)*2)
+  plot(cumsum(pve), xlab="Principal Component", ylab="Cumulative Proportion of Variance Explained", ylim=c(0,1), type="b")
+  abline(h=0.9)
+  plot(log10(1-cumsum(pve)), xlab="Principal Component", ylab="Remaining Variance", type="b")
+  abline(a=0, b=log10(1-pve[1]))
+  #plot(pve/(1-c(0,head(cumsum(pve),-1))), xlab="Principal Component", ylab="Proportion of Remaining Variance Explained", ylim=c(0,1), type="b")
+})
+par(orig.p)
+res
+res$x[,]
+```
+
+* * *
 

### 2015-08-13 rscripts/pca_graph.R

@@ -1,0 +1,39 @@
+## PCA plots in ggplot
+
+require(ggplot2)
+
+pcaplot <- function(prcompres, limit=length(prcompres$sdev), validation.dt) {
+  fcts <- paste0("PC", 1:limit)
+  ref <- factor(fcts, levels=fcts, ordered=T)
+  cols <- ref[seq(from=1, to=limit, by=2)]
+  rows <- ref[seq(from=2, to=limit, by=2)]
+  pc <- dim(prcompres$x)[1]
+  (if(limit %% 2 == 0) {
+    if (limit == 2) {
+      data.table(col = cols, row = rows)
+    } else data.table(col = cols[c(1, rep(2:length(cols), each=2))], row = rows[c(rep(1:(length(rows)-1), each=2), length(rows))])
+  } else {
+    if (limit == 3) {
+      data.table(col = cols, row = rows)
+    } else data.table(col = cols[c(1, rep(2:(length(cols)-1), each=2), length(cols))], row = rep(rows, each=2))
+  }) -> template
+  
+  kcluster <- kmeans(prcompres$x[,1:limit], 5, nstart = 20)
+
+  pred.src <- template[,list(x=prcompres$x[,col], y=prcompres$x[,row], cluster=factor(kcluster$cluster)), by=list(col, row)]
+  ## on pca scale validation.dt[,-1,with=F][,(.SD-as.list(res$center))/as.list(res$scale)]
+  valid.x <- as.matrix(validation.dt) %*% res$rotation
+  pred.cluster <- apply(valid.x[,1:limit],1,function(row) {
+    which.min(colSums(row - t(kcluster$centers))^2)
+  })
+  valid.src <- template[,list(x=valid.x[,col], y=valid.x[,row], cluster=factor(pred.cluster)), by=list(col, row)]
+  ggplot(pred.src) + theme_bw() + facet_grid(row ~ col) + aes(x=x, y=y, color=cluster) + geom_point(aes(size="training", alpha="training")) + geom_point(aes(size="validation", alpha="validation"), data = valid.src) +
+    scale_size_manual(limits=c("training","validation"), values=c(10,2)) + scale_alpha_manual(limits=c("training","validation"), values=c(0.2,1))
+  
+  ## first, long format, rows odd components, cols even components
+}
+
+prcompres <- prcomp(prediction.dt, scale. = F)
+p <- pcaplot(prcompres, limit=6, validation.dt)
+print(p)
+

### 2015-09-02 rscripts/data_review.Rmd
ugh, too many changes

--- 2015-08-10 rscripts/data_review.Rmd
@@
 ```{r, echo=FALSE, include=FALSE}
-req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation", "randomForest")
+req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation", "randomForest","lubridate")
 need.pkgs <- req.packages[!sapply(req.packages, require, character.only = T)]
 if (length(need.pkgs) != 0) install.packages(need.pkgs, dependencies = T)
+source("munging.R")
 ```
@@
 ```{r, echo=FALSE}
-src.dt <- 
-  setkey(
-    setnames(
-      fread("../input/merged.o", header = F, sep=" ", colClasses = list(integer=1.4)),
-      c("user_id", "location_id", "login", "logout")
-    ),
-    login, logout, user_id, location_id
-  )
+src.dt <- loadRaw()
 src.dt
@@
 ```{r, echo=FALSE}
-source("munging.R")
 zeroize(breakoutDays(censor.dt[,
@@
 ```
+
+1. consider subset - subset of records? subset of users?  subset of locations?  want predictions about location "class", so subset of locations?  later, when locations classified, subset of users.
+2. using subset and some predictor measurements
+
+```{r,echo=F}
+## principle components analysis
+# weekday_zeroes <- data.table(weekday=factor(c("Sun", "Mon", "Tues", "Wed","Thurs","Fri","Sat"), ordered = T), usage = 0, key = "weekday")
+# predictors <- function(SD) with(SD, {
+#   duration <- logout - login
+#   total_login_time <- sum(duration)
+#   usage_by_weekday <- SD[, list(usage = sum(login_day_secs)), keyby=weekday][weekday_zeroes][, list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=weekday]
+#   temp <- SD[, sum(logout_day_secs), keyby=weekday][weekday_zeroes][, list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=weekday]
+#   usage_by_weekday$usage <- usage_by_weekday$usage + c(temp$usage[7], temp$usage[-7])
+#   store <- as.list(usage_by_weekday$usage / total_login_time)
+#   names(store) <- usage_by_weekday$weekday
+#   reduced_store <- list(weekend=store$Fri+store$Sat, midweek=store$Tues+store$Wed+store$Thurs, weekbegin = store$Sun+store$Mon)
+#   user_count <- length(unique(user_id))
+#   life <- max(logout) - min(login)
+#   return(c(list(
+#     log10_lifetime = log10(life),
+#     ave_duration = mean(duration),
+#     med_duration = median(duration),
+#     sd_duration = sd(duration),
+#     max_duration = max(duration),
+#     total_user_time = total_login_time,
+#     log10_user_per_life = log10(user_count/life),
+#     log10_time_per_user = log10(total_login_time/user_count),
+#     log10_time_per_life = log10(total_login_time/life),
+#     log10_unique_users = log10(user_count)
+#   ), reduced_store))
+# })
+# unique_locs <- unique(censor.dt$location_id)
+# slice <- sort(sample(unique_locs, size = length(unique_locs)/5, replace = F))
+# pca.dt <- censor.dt[, predictors(.SD), by=location_id]
+# centered <- pca.dt[,lapply(.SD[,-1,with=F],function(col) col-mean(col)),]
+# scaled <- centered[,lapply(.SD, function(col) col/sd(col))]
+# prediction.dt <- scaled[slice]
+# validation.dt <- scaled[!slice]
+# res <- prcomp(prediction.dt[, -1, with=F], scale. = F)
+# res$var <- res$sdev^2
+# res$pve <- res$var/sum(res$var)
+# keep <- res$pve > 0.01 # 1/length(res$pve)
+# orig.p <- par(mfrow=c(1,sum(keep)-1))
+# for (x in 1:(sum(keep)-1)) biplot(res, choices=c(x,x+1), scale = 0)
+# par(orig.p)
+# with(res, {
+#   plot(pve, xlab="Principal Component", ylab="Proportion of Variance Explained", type="b", log="y")
+#   abline(h=1/(length(pve)))
+#   # abline(h=tail(pve, 1)*2)
+#   plot(cumsum(pve), xlab="Principal Component", ylab="Cumulative Proportion of Variance Explained", ylim=c(0,1), type="b")
+#   abline(h=0.9)
+#   plot(log10(1-cumsum(pve)), xlab="Principal Component", ylab="Remaining Variance", type="b")
+#   abline(a=0, b=log10(1-pve[1]))
+#   #plot(pve/(1-c(0,head(cumsum(pve),-1))), xlab="Principal Component", ylab="Proportion of Remaining Variance Explained", ylim=c(0,1), type="b")
+# })
+# par(orig.p)
+# res
+# res$x[,]
+```
+
+* * *
 

### 2015-09-02 rscripts/location_clusters.Rmd
ugh, too many changes

---
output:
    pdf_document:
        includes:
            in_header: header.tex
        fig_caption: yes
---
## Analysis and plots of montreal data.

```{r, echo=FALSE, include=FALSE}
req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation", "randomForest","lubridate", "movMF")
need.pkgs <- req.packages[!sapply(req.packages, require, character.only = T)]
if (length(need.pkgs) != 0) install.packages(need.pkgs, dependencies = T)
source("munging.R")
source("pca_plots.R")
```

### Required packages: `r req.packages`.

Source data:

```{r}
max_hours <- 24
min_logins <- 2;    min_users <- 6 # 6 user min remove abberant location
min_lifetime <- 30; min_loc_lifetime <- 30 # days
censor.dt <- loadCensored(
  raw.dt = loadRaw(),
  max_hours = max_hours, min_logins = min_logins, min_lifetime = min_lifetime,
  min_users = min_users, min_loc_lifetime = min_loc_lifetime
)
censor.dt
set.seed(0); training_portion <- .8
unique_locations <- censor.dt[,unique(location_id)]
training_locations <- sort(sample(
  unique_locations,
  size = length(unique_locations)*training_portion, replace = F
))
# ggplot(censor.dt[,list(usage=sum(logout-login)/60/60), by=list(location_id, year, month)]) + aes(x=month, y=usage, group=location_id) + facet_grid(year ~ .) + geom_line(alpha=0.5)
# users that qualify as regular:
# censor.dt[,list(lifetime = max(logout)-min(login)), by=list(location_id, user_id)][lifetime > 60*60*24*365, unique(user_id)]
```

If want to identify locations which are "similar", so that we can inform how the covert group makes use
of locations, but we have no particular information about those locations (*e.g.*, what businesses they are, where they are geographically), what are we to do?

This question falls neatly into the sort of problems described as "unsupervised learning", so we will combine approaches from that field with "analyst discretion".  Having to rely on domain expert input here is largely unavoidable, but we can avoid some of the traps associated with that (more sophosticated, perhaps, but essentially 'p-hacking') by cross-validating the outcomes at different stages.

TODO: what stages? how? definitely need at detection stage: can we get "no detect" when making non-covert synthetic people?

Initially, we are going to trawl the data, sieving out a great variety of measurements for locations, and progressively identify which to maintain, discard, or merge, attempting to find a compelling argument as to why.  While these specific choices are unlikely to apply to another data source, we think the process will be illustrative as to what an analyst in the field should do.

Our initial measures are:

| measure | reasoning | transform? |
|--------:|-----------|-----------:|
| total lifetime | other parameters are likely normalized by this | log |
| total usage | low vs. high use locations is obvious distinction | log |
| number of unique users | few vs. many unique visitors indicates location "population" | log |
| quartiles for login durations | typically short vs. long logins indicates turnover | none |
| mean, sd of login durations | ibid. | none |
| % of usage by week day | weekend vs working vs continuous availability | none |
| % of usage by month | seasonal vs non | none |
| % of log(in\|out), usage by hour | working day vs evening use | none |

This is 101 (!) measures, so we definitely need to reduce the dimensionality.

```{r, echo=F}
process_hour_rle <- function(prefix) {
  hrsind <- 0:23
  nms <- sapply(hrsind, function(h) sprintf("%s_hr_%.2d", prefix, h))
  function(hrrle) with(hrrle, {
    missing <- hrsind[!(hrsind %in% values)]
    res <- c(lengths, rep.int(0, length(missing)))[order(c(values, missing))] / sum(lengths)
    names(res) <- nms
    res
  })
}

day.abb <- c("Sun", "Mon", "Tues", "Wed","Thurs","Fri","Sat")

weekday_zeroes <- data.table(
  weekday=factor(day.abb, levels = day.abb, ordered = T),
  usage = 0, key = "weekday"
)

monthly_zeroes <- data.table(
  month=factor(month.abb, levels = month.abb, ordered = T),
  usage = 0, key = "month"
)

hourly_zeros <- data.table(
  hour = 0:23,
  usage = 0, key = "hour"
)

login_parser <- process_hour_rle("login")
logout_parser <- process_hour_rle("logout")

with_usage_zeros <- function(orig.usage, zeros) orig.usage[zeros][,
  list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=key(zeros)
]

refpredictors <- function(SD) with(SD, {
  duration <- logout - login
  total_login_time <- sum(duration)
  usage_by_weekday <- with_usage_zeros(SD[,
    list(usage = sum(login_day_secs)), keyby=weekday
  ], weekday_zeroes)
  usage_on_logout_day <- with_usage_zeros(SD[,
    list(usage = sum(logout_day_secs)), keyby=weekday
  ], weekday_zeroes)$usage

  weekday_percent <- with(usage_by_weekday, {
    ## logout usage "on" Sunday belongs to Monday, on M to Tues, etc
    usage <- (usage + c(usage_on_logout_day[7], usage_on_logout_day[-7])) / total_login_time
    names(usage) <- weekday
    usage
  })

  usage_by_month <- with_usage_zeros(SD[,
    list(usage = sum(logout - login) / total_login_time), keyby=month
  ], monthly_zeroes)

  monthly_percent <- with(usage_by_month, {
    names(usage) <- month
    usage
  })

  user_count <- length(unique(user_id))
  life <- max(logout) - min(login)

  dur_quantiles <- quantile(duration)
  login_hours <- login_parser(rle(sort(login_hour)))
  logout_hours <- logout_parser(rle(sort(logout_hour)))

  short <- (login_hour == logout_hour) & (login_day == logout_day)

  login_usage_by_hour <- data.table(usage = 
      ifelse(short, logout_time - login_time, (login_hour + 1)*3600 - login_time), hour = login_hour)
  logout_usage_by_hour <- data.table(usage = 
      logout_time[!short] - logout_hour[!short]*3600, hour = logout_hour[!short])

  long <- (login_day != logout_day) & !(login_hour == 23 & logout_hour == 0)
  cross_day <- rbindlist(with(SD[long, list(login_hour, logout_hour)],
    mapply(function(in_hr, out_hr) {
      data.table(usage = rep.int(3600, out_hr-in_hr+23), hour = ((in_hr+1):(out_hr+23))%%24)
    }, in_hr = login_hour, out_hr = logout_hour, SIMPLIFY = F)
  ))
  # what's left?
  mid <- logout_hour > (login_hour + 1)
  inday <- rbindlist(with(SD[mid, list(login_hour, logout_hour)],
    mapply(function(in_hr, out_hr) {
      data.table(usage = rep.int(3600, out_hr-in_hr-1), hour = (in_hr+1):(out_hr-1))
    }, in_hr = login_hour, out_hr = logout_hour, SIMPLIFY = F)
  ))
  usage_by_hour <- with_usage_zeros(rbind(login_usage_by_hour, logout_usage_by_hour, cross_day, inday)[, list(usage = sum(usage)/total_login_time), keyby=hour], hourly_zeros)

  hourly_percent <- with(usage_by_hour, {
    names(usage) <- paste0("hr", hour)
    usage
  })

  return(c(list(
      log10_lifetime = log10(life),
      log10_total_duration = log10(total_login_time),
      log10_unique_users = log10(user_count),
      ave_duration = mean(duration), sd_duration = sd(duration)
    ),
    weekday_percent, monthly_percent, hourly_percent,
    as.list(dur_quantiles), as.list(login_hours), as.list(logout_hours)
  ))
})
digestor <- function(dt, preds, training, cache = "../input/initPred.RData") {
  { if (!file.exists(cache)) {
    d <- dt[, preds(.SD), by=location_id]
    saveRDS(d, cache)
    d
  } else {
    readRDS(cache)
  } } -> ref

  centered <- ref[,lapply(.SD[,-1,with=F],function(col) col-mean(col))]
  scaled <- cbind(location_id = ref$location_id, centered[,lapply(.SD, function(col) col/sd(col))])
  list(raw = ref, training = scaled[training], validation = scaled[!training])
}
redigest <- function(ref, training) {
  centered <- ref[,lapply(.SD[,-1,with=F],function(col) col-mean(col))]
  scaled <- cbind(location_id = ref$location_id, centered[,lapply(.SD, function(col) col/sd(col))])
  list(raw = ref, training = scaled[training], validation = scaled[!training])
}
dopca <- function(digest) with(digest, {
  pca <- prcomp(training[,-1,with=F], center=F, scale. = F)
  val <- predict(pca, validation[,-1,with=F])
  return(list(pca=pca, val=val))
})
initial_digest <- digestor(censor.dt, refpredictors, training_locations)
```

Let's have a look at the time signature information first:

```{r, echo=F}
temporal_data_slice <- melt(initial_digest$raw[, .SD, .SDcols = c(1, grep("hr", names(initial_digest$raw)))], id.var = "location_id")
temporal_data_slice[grepl("log", variable), measure := sub("_hr_.+","", variable)]
temporal_data_slice[!grepl("log", variable), measure := "usage"]
temporal_data_slice[, hour:= as.integer(gsub("[^\\d]","", variable, perl=T))]
setkey(temporal_data_slice, location_id, measure, hour)
```

```{r, echo=F, fig.cap="Raw proportions by hour"}
ggplot(temporal_data_slice) + theme_bw() + aes(group = location_id, x = hour, y = value) + facet_grid(measure ~ .) + geom_line(alpha=0.5) + scale_x_discrete(breaks=0:23, expand=c(0,0)) +
  scale_y_continuous("proportion", expand=c(0,0))
```

```{r, echo=F, fig.cap="Measure Peak distribution"}
peaks <- temporal_data_slice[, list(peak=which.max(value)-1), by=list(location_id, measure)]
ggplot(peaks) + theme_bw() + aes(x = peak) + facet_grid(measure ~ .) + geom_bar(binwidth=1)
```

```{r, echo=F, fig.cap="Frequency Power Signatures"}
freq_power <- function(ref, measure) {
  mn <- mean(as.matrix(ref[,-1,with=F]))
  fftref <- mvfft(t(ref[,-1,with=F])-mn)
  meaningful <- fftref[1+1:12,]
  pwr <- Re(meaningful*Conj(meaningful))
  res <- t(pwr)[,c(1,2,3,4,6,8,12)]
  dimnames(res)[[2]] <- paste0(as.integer(gsub("[^\\d]","",dimnames(res)[[2]], perl=T)), "_perday")
  cbind(melt(data.table(cbind(location_id = ref[,location_id], res)), id.var = "location_id"), measure = measure)
}

ref_usage <- initial_digest$raw[,.SD,.SDcols=c(1,grep("^hr", names(initial_digest$raw)))]
ref_login <- initial_digest$raw[,.SD,.SDcols=c(1,grep("^login", names(initial_digest$raw)))]
ref_logout <- initial_digest$raw[,.SD,.SDcols=c(1,grep("^logout", names(initial_digest$raw)))]
pwr <- rbind(
  freq_power(ref_usage, "usage"),
  freq_power(ref_login, "login"),
  freq_power(ref_logout, "logout")
)
ggplot(pwr[,
  list(power=value, frequency=as.numeric(sub("_perday","", variable))), by=list(location_id, measure)
]) + theme_bw() + aes(y = power, x=frequency, group=location_id) + facet_grid(measure ~ .) + geom_line(alpha=0.5) + scale_x_continuous("per day frequency", breaks=c(1,2,3,4,6,8,12))
```

Some clear patterns there: distinct peak locations, distinct periodicity.  Let's develop categories:
peak time characterization (for this work, just on usage - possible to have circular statistics grouping on all three peaks) and single principle mode of usage (i.e., majority of usage power in the daily frequency) vs. multi-mode.  We use a mixture von Mises model.

Similar for weekly signatures - curve, peak day grouping.

Drop monthly data - would have to de-trend?

We will initially approach the measures with Principle Components Analysis (PCA).  PCA should discover the best bases to describe the locations - orthogonal, linear combinations of measurements about those locations - in a least squares sense.  The way PCA works implies two important criteria for inputs: (1) those measures need to be approximately normal and (2) we should ensure that factors that we expect to have non-linear combinations are accommodated.  Concern (1) is principally about measures that have a one-sided, long-tailed distribution.  Concern (2) is about measures that should really be combined multiplicatively - *e.g.* total usage (one measure) per unique user (another).

Replacing measures with their logarithms is a coarse way to address these both concerns.  Taking the log compresses long-tailed distributions into something more Gaussian, and linear combinations of the logs is the same as combining the underlying metrics multiplicatively.  This approach is not perfect, but absent some principled reason to make particular transformations (or not), we anticipate it is better than not.

A typically loadings plot (showing vectors for the original dimensions in the new principle component basis) will be a dense cloud of lines at this stage, so we need a different view to identify patterns for the first round of parameter culling.  We will look at just the first two principle components on a colored bar plot - with bar height for the first component, and color for the second.  If we see patterns in heights and colors that tell a story that makes sense in the domain, then we can reduce the input measures by combination or elimination.
TODO: plot for resulting approx normality of measures?

```{r, echo=F}
initial_digest <- digestor(censor.dt, refpredictors, training_locations)
initial_pca <- dopca(initial_digest)
init_bars <- make_bars(initial_pca$pca)
```

\newpage
\blandscape
```{r, echo=F, fig.width=11.25, fig.height=7 }
plot_pca_bars(init_bars)
```
\elandscape
\newpage

There seems to be something consistent going on with the hourly measures.  Upon closer examination:

```{r, echo=F}
zoom_plot <- function(pca, target = 1:2, brs = make_bars(pca, target)) {
  pcs <- paste0("PC", target)
  zoom_bars <- setnames(
    brs[ grepl("hr", measure) ][, c(list(
      type=ifelse(grepl("login", measure),"login",ifelse(grepl("logout", measure), "logout", "usage")),
      hr = as.integer(sub("^[^\\d]+(\\d+)$","\\1", measure, perl=T))
    ), y=.SD[,2,with=F], fill=.SD[,3,with=F]), .SDcols=c("measure", pcs)],
    paste(c("y","fill"), pcs, sep='.'),
    c("y","fill")
  )
  limY <- max(abs(zoom_bars[,3,with=F]))
  limC <- max(abs(zoom_bars[,4,with=F]))
  ggplot(zoom_bars) + facet_grid(type ~ .) + aes(x = hr, y = y, fill = fill) +
    theme_bw() + geom_bar(stat="identity", position = "identity") +
    labs(y=pcs[1]) + ylim(-limY,limY) +
    scale_fill_gradient2(pcs[2], limits=c(-limC, limC), low="red", mid = "grey", high="blue")
}
zoom_plot(brs = init_bars)
```

The trends in between login, usage, and logout seem very similar in principle components 1 and 2, so it seems reasonable to just pick one of them.  Similarly, they seem to be obviously driven on a daily period, so perhaps we can just use the peak amplitude to distinguish locations.  Before we do so, let's check that trends persist for some of the higher order principle components.

```{r, echo=F}
zoom_plot(initial_pca$pca, 3:4)
```

```{r, echo=F}
zoom_plot(initial_pca$pca, 5:6)
```

So those trends seem to be preserved, but that higher frequency content is necessary as well.  We will compare that frequency to our expectations, and see how much is worthy of keeping, and also check the distribution of peak usage:
<!---
with a two caveats: morning logouts (hour 8) are consistently different from usage, as are late night logins (peaks at hours 20 and 23).  This has a pretty compelling story: the differences between locations are generally about usage, though we can additionally differentiate places with variable morning turnover (excess of logouts) and variable evening (spiking with the dinner and "party" hours) starts (excess of logins).

If we focus on just usage, the plots between the components are pretty clearly periodic in both dimensions - like rotating polarized light.  Perhaps the right usage measure is instead (1) how peaky is the distribution across hours (what is the amplitude of the wave), (2) what is the period of usage peaks (per day or multiples of per day), and (3) when are the peaks located?

delta T = 1 hr
1/(2 hr) = nyquist freq
1/day, 2/day, 3/day, 4/day, 6/day, 8/day, 12/day <- valid signals
--->
```{r, echo=F}
# http://stackoverflow.com/questions/18788748/fit-a-mixture-of-von-mises-distributions-in-r
usage_peak_hr <- initial_digest$raw[, apply(.SD, 1, which.max), .SDcols=grep("^hr", names(initial_digest$raw))]
ang <- usage_peak_hr/12*pi
unit_circ <- cbind(cos(ang), sin(ang))
#login_peak_hr <- initial_digest$raw[, apply(.SD, 1, which.max), .SDcols=grep("^login", names(initial_digest$raw))]
#logout_peak_hr <- initial_digest$raw[, apply(.SD, 1, which.max), .SDcols=grep("^logout", names(initial_digest$raw))]
#underlying <- cbind(usage_peak_hr, login_peak_hr, logout_peak_hr)/24
thing <- movMF(unit_circ, 3, list(maxiter=10000, nruns=20))
predthing <- predict(thing, unit_circ)
ggplot(data.frame(hour = usage_peak_hr-1, clz=factor(predthing))) + theme_bw() + aes(x=hour, fill=clz, group=clz) +
  geom_bar(binwidth=1) +
  scale_y_continuous(expand=c(0,0)) +
  scale_x_discrete(breaks=1:24)
```

superpose three normal distributions, on periodic boundary condition, assign usage based on most probably membership to peak 1, 2, 3

```{r, echo=F}
peak_freq_power <- {
  ref <- initial_digest$raw[,.SD,.SDcols=c(1,grep("^hr", names(initial_digest$raw)))]
  mn <- mean(as.matrix(ref[,-1,with=F]))
  ref.mlt <- melt(ref, id.var = "location_id")
#  ggplot(ref.mlt) + theme_bw() + aes(x=variable, y=value-mn, group=location_id) + geom_line(alpha=0.2)
  fftref <- mvfft(t(ref[,-1,with=F])-mn)
  meaningful <- fftref[1+1:12,] # trim average signal, freqs above Nyquist
  pwr <- Re(meaningful*Conj(meaningful))
#   thing <- melt(data.table(cbind(location_id = 1:297, t(pwr))), id.var = "location_id")
#   thing[, period := 24/as.numeric(sub("hr","", variable)) ]
#   ggplot(thing[variable %in% paste0("hr",c(1,2,3,4,6,8,12))]) + theme_bw() + aes(y = value, group = location_id, x = period) + geom_line(alpha=0.2) + scale_x_continuous(breaks=c(2,3,4,6,8,12,24))
#   ggplot(thing[variable %in% paste0("hr",c(1,2,3,4,6,8,12))]) + theme_bw() + aes(x = log(value)) + facet_grid(period ~ .) + geom_bar()
  res <- t(pwr)[,c(1,2,3,4,6,8,12)]
  dimnames(res)[[2]] <- paste0("usage_period_", 24/as.integer(sub("hr","",dimnames(res)[[2]])))
  res
}
ref <- copy(initial_digest$raw)
setnames(ref, c("0%","100%"), c("min","max"))
ref <- cbind(ref[,.SD, .SDcols=grep("(hr|%)", names(ref), invert=T)], usage_peak = usage_peak_hr, peak_freq_power)
#freqs <- melt(ps, id.var="location_id")
#freqs[, period := 24/as.numeric(sub("V","", variable))]

# net <- cbind(location_id = initial_digest$raw$location_id, initial_digest$raw[, .SD[,1:24,with=F]-.SD[,1:24+24,with=F], .SDcols=grep("log(in|out)", names(initial_digest$raw))])[, {
#   z  <- fft(t(.SD[,-1,with=F]))[1:12] # Nyquist limit is 24 samples / 24 hrs / 2 = 12 / 24 hrs = 1 / 2 hrs
#   ps <- Re(z*Conj(z))
#   as.list(ps[1:12])
# }, by=location_id]
# 
# ggplot(freqs[variable %in% paste0("V",c(1,2,3,4,6,8,12))]) + theme_bw() + aes(y = value, group = location_id, x = period) + geom_line(alpha=0.2) + scale_x_continuous(breaks=c(2,3,4,6,8,12,24))
```

so pick an indicator for the full day, divided into comparable blocks, then have indicators for wonky periods:

 usage, 2-5, 10-12, 13-15, 19-21

There seems to be pretty clear relationships between hours 9 to 15 - pretty standard working day - across all measures, but hour 16 seems perhaps to differentiate on places that close earlier versus later.  Logins, logouts, and usage appear to have a similar trend from 0 to 7 - definite sleeping or early morning hours.  There appear to be some interesting outliers in hour 20 and 23 for logins - late night hotspots perhaps?

Let's take these metrics down their weights between 0 to 7 (morning), 8 (opening), 9 to 15 (day), 16 (closing), 17-19 (evening), and then 20-23 (late night).

```{r, echo=F, include=F}
# decision free version? sort(kmeans(initial_pca$pca$rotation, 10, 10)$cluster)
# ref <- copy(initial_digest$raw)
# hr_items <- grep('hr', names(ref))
# ref[,usage_morning := apply(.SD[, hr_items[1:8],with=F],1,sum)]
# ref[,login_morning := apply(.SD[, hr_items[1:8+24],with=F],1,sum)]
# ref[,logout_morning := apply(.SD[, hr_items[1:8+48],with=F],1,sum) ]
# ref[,usage_opening := apply(.SD[, hr_items[9],with=F],1,sum)]
# ref[,login_opening := apply(.SD[, hr_items[9+24],with=F],1,sum)]
# ref[,logout_opening := apply(.SD[, hr_items[9+48],with=F],1,sum) ]
# ref[,usage_day := apply(.SD[, hr_items[10:16],with=F],1,sum)]
# ref[,login_day := apply(.SD[, hr_items[10:16+24],with=F],1,sum)]
# ref[,logout_day := apply(.SD[, hr_items[10:16+48],with=F],1,sum) ]
# ref[,usage_closing := apply(.SD[, hr_items[17],with=F],1,sum)]
# ref[,login_closing := apply(.SD[, hr_items[17+24],with=F],1,sum)]
# ref[,logout_closing := apply(.SD[, hr_items[17+48],with=F],1,sum) ]
# ref[,usage_evening := apply(.SD[, hr_items[18:20],with=F],1,sum)]
# ref[,login_evening := apply(.SD[, hr_items[18:20+24],with=F],1,sum)]
# ref[,logout_evening := apply(.SD[, hr_items[18:20+48],with=F],1,sum) ]
# ref[,usage_late := apply(.SD[, hr_items[21:24],with=F],1,sum)]
# ref[,login_late := apply(.SD[, hr_items[21:24+24],with=F],1,sum)]
# ref[,logout_late := apply(.SD[, hr_items[21:24+48],with=F],1,sum) ]
# ref <- ref[,-hr_items,with=F]
```

With the consolidated measures:

\newpage
\blandscape
```{r, echo=F, fig.width=11.25, fig.height=7 }
second_digest <- redigest(ref, training_locations)
second_pca <- dopca(second_digest)
second_bars <- make_bars(second_pca$pca)
plot_pca_bars(second_bars)
```
\elandscape
\newpage

There is pretty compellingly no information between the logins, logouts, and usage during the day.  This seems a reasonable conclusion for evening and morning as well.  We will use only login information for those categories, since that seems have the most weight in components 1 and 2.
```{r, echo=F}
# ref <- copy(second_digest$raw)
# drop <- grep("(logout|usage)_(morning|day|evening)", names(ref))
```

There also seems to be fairly similar information between the middle duration quantiles, average, and standard deviations.  The max and min seem informative, but we will remove the inter-quartiles and median.
```{r, echo=F}
# ref <- copy(second_digest$raw)
# drop <- c(drop, grep("(25|50|75)%", names(ref)))
```

Finally, Tuesday and Wednesday - early working week - seem very similar.  Thursday seems somewhat similar as well, but not enough.  We will try a Tue-Wed category - usage outside of the start of the working week (Monday), the days when people regularly go out in the evenings (Thurs-Fri), and the weekends.

\newpage
\blandscape
```{r, echo=F}
# decision free version? sort(kmeans(initial_pca$pca$rotation, 10, 10)$cluster)
# ref[,TueWed := Tues + Wed]
# drop <- c(drop, grep("^(Tues|Wed)$", names(ref)))
# ref <- ref[,-drop, with=F]
# setcolorder(ref,
#   c("location_id", "log10_lifetime", "log10_total_duration", "log10_unique_users",
#   "ave_duration", "sd_duration", "0%", "100%",
#   "Sun", "Mon", "TueWed", "Thurs", "Fri", "Sat",
# "login_morning", "usage_opening", "login_opening", "logout_opening", "login_day", "usage_closing", "login_closing", "logout_closing", "login_evening", "usage_late", "login_late", "logout_late",
#   month.abb
# ))
# third_digest <- redigest(ref, training_locations)
# third_pca <- dopca(third_digest)
# third_bars <- make_bars(third_pca$pca)
# plot_pca_bars(third_bars)
```
\elandscape
\newpage

Now if we look at the months, there seems to be a Winter group (Jan-Feb), a Summer group (Jun-Jul), and a Fall group (Oct-Nov)

\newpage
\blandscape
```{r, echo=F}
# decision free version? sort(kmeans(initial_pca$pca$rotation, 10, 10)$cluster)
# ref <- copy(third_digest$raw)
# ref[, JanFeb := Jan+Feb ]
# ref[, JunJul := Jun+Jul ]
# ref[, OctNov := Oct+Nov ]
# drop <- grep("^(Jan|Feb|Jun|Jul|Oct|Nov)$", names(ref))
# ref <- ref[,-drop, with=F]
# setcolorder(ref,
#   c("location_id", "log10_lifetime", "log10_total_duration", "log10_unique_users",
#   "ave_duration", "sd_duration", "0%", "100%",
#   "Sun", "Mon", "TueWed", "Thurs", "Fri", "Sat",
# "login_morning", "usage_opening", "login_opening", "logout_opening", "login_day", "usage_closing", "login_closing", "logout_closing", "login_evening", "usage_late", "login_late", "logout_late",
#   "JanFeb","Mar","Apr","May","JunJul","Aug","Sep","OctNov","Dec"
# ))
# fourth_digest <- redigest(ref, training_locations)
# fourth_pca <- dopca(fourth_digest)
# fourth_bars <- make_bars(fourth_pca$pca)
# plot_pca_bars(fourth_bars)
```
\elandscape
\newpage

At this point, it seems like login proportions are sufficient to consider, and that usage and logout measures can be ignored.  Though usage and logout seem to have more weight in some periods, picking just one seems like a reasonable choice.

TODO: more downselect

TODO: vector/point cloud loadings plots w/ reduced measures

TODO: re-run kmeans on reduced set to get clusters

TODO: plots of measurement distributions, color categorized by cluster

### 2015-09-02 rscripts/munging.R

@@ -1,5 +1,51 @@
 ## data munging utils
 require(data.table); require(lubridate)
+
+loadRaw <- function(raw.src = "../input/merged.o", cache.src = "../input/raw.RData") {
+  if (!file.exists(cache.src) || (file.mtime(raw.src) > file.mtime(cache.src))) {
+    res <- setkey(
+      setnames(
+        fread(raw.src, header = F, sep=" ", colClasses = list(integer=1.4)),
+        c("user_id", "location_id", "login", "logout")
+      ),
+      login, logout, user_id, location_id
+    )
+    saveRDS(res, cache.src)
+    res
+  } else readRDS(cache.src)
+}
+
+processRaw <- function(raw.dt,
+  max_hours, min_logins, min_lifetime,
+  min_users, min_loc_lifetime) {
+  censor.dt <- raw.dt[(logout != login) & ((logout - login) <= max_hours*60*60), ]
+  invalid.users <- censor.dt[,list(.N, lifetime = (max(logout) - min(login))/60/60/24), by=user_id ][N < min_logins | lifetime < min_lifetime, user_id]
+  while(length(invalid.users) > 0) {
+    censor.dt     <- censor.dt[!user_id %in% invalid.users]
+    invalid.locs  <- censor.dt[,list(.N, uc=length(unique(user_id)), lifetime=(max(logout)-min(login))/60/60/24), by=location_id ][
+      N < min_logins | uc < min_users | lifetime < min_loc_lifetime,
+      location_id
+    ]
+    censor.dt     <- censor.dt[!location_id %in% invalid.locs]
+    invalid.users <- censor.dt[,list(.N, lifetime = (max(logout) - min(login))/60/60/24), by=user_id ][N < min_logins | lifetime < min_lifetime, user_id]
+  }
+  censor.dt
+}
+
+loadCensored <- function(
+  cache.src = "../input/raw.RData",
+  cache.censor = "../input/censor.RData", ...) {
+  if (!file.exists(cache.censor) || (file.mtime(cache.src) > file.mtime(cache.censor))) {
+    res <- zeroize(breakoutDays(processRaw(...)[,
+      user_id := .GRP, by=user_id
+    ][,
+      location_id := .GRP, by=location_id
+    ], login_day, login_time, logout_day, logout_time, user_id, location_id))
+    saveRDS(res, cache.censor)
+    res
+  } else readRDS(cache.censor)
+}
+
[…]

### 2015-09-02 rscripts/pca_graph.R

@@ -1,13 +1,127 @@
 ## PCA plots in ggplot
 
-require(ggplot2)
+require(ggplot2); require(data.table); require(reshape2)
 
-pcaplot <- function(prcompres, limit=length(prcompres$sdev), validation.dt) {
+process_hour_rle <- function(prefix) {
+  hrsind <- 0:23
+  nms <- sapply(hrsind, function(h) sprintf("%s_hr_%.2d", prefix, h))
+  function(hrrle) with(hrrle, {
+    missing <- hrsind[!(hrsind %in% values)]
+    res <- c(lengths, rep.int(0, length(missing)))[order(c(values, missing))] / sum(lengths)
+    names(res) <- nms
+    res
+  })
+}
+
+refpredictors <- {
+  weekday_zeroes <- data.table(
+    weekday=factor(c("Sun", "Mon", "Tues", "Wed","Thurs","Fri","Sat"), ordered = T),
+    usage = 0, key = "weekday"
+  )
+  login_parser <- process_hour_rle("login")
+  logout_parser <- process_hour_rle("logout")
+  function(SD) with(SD, {
+    duration <- logout - login
+    total_login_time <- sum(duration)
+    usage_by_weekday <- SD[, list(usage = sum(login_day_secs)), keyby=weekday][weekday_zeroes][, list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=weekday]
+    temp <- SD[, sum(logout_day_secs), keyby=weekday][weekday_zeroes][, list(usage = ifelse(is.na(usage), i.usage, usage)), keyby=weekday]
+    usage_by_weekday$usage <- usage_by_weekday$usage + c(temp$usage[7], temp$usage[-7])
+    store <- as.list(usage_by_weekday$usage / total_login_time)
+    names(store) <- usage_by_weekday$weekday
+    user_count <- length(unique(user_id))
+    life <- max(logout) - min(login)
+    login_hour <- as.integer(login_time / 60 / 60)
+    login_hours <- login_parser(rle(sort(login_hour)))
+    logout_hour <- as.integer(logout_time / 60 / 60)
+    logout_hours <- logout_parser(rle(sort(logout_hour)))
+#     ifelse(login_hour == logout_hour,ifelse(login_day == logout_day,{
+#       
+#     },{
+#       
+#     }),{
+#       
+#     })
+    # breakdown usage proportion by hour
+    #  - foreach entry
+    #   - find end hour, begin hour
[…]

### 2015-09-02 rscripts/pca_plots.R

@@ -1,0 +1,18 @@
+## PCA plots in ggplot
+
+require(ggplot2); require(data.table); require(reshape2)
+
+make_bars <- function(pca, target = 1:2) with(pca, {
+  setnames(data.table(dcast(melt(rotation[, target]), Var1 ~ Var2)), "Var1", "measure")
+})
+
+plot_pca_bars <- function(pca_bars) {
+  limY <- max(abs(pca_bars[,PC1]))
+  limC <- max(abs(pca_bars[,PC2]))
+  
+  ggplot(pca_bars) + theme_bw() + theme(
+    axis.text.x = element_text(angle=90, size = rel(0.75))
+  ) + aes(x=measure, y=PC1, fill=PC2) + geom_bar(stat="identity", position = "identity") +
+  labs(x="Predictor", y="PC 1 Loading", fill="PC 2 Loading") + ylim(-limY, limY) +
+  scale_fill_gradient2(limits=c(-limC, limC), low="red", mid = "grey", high="blue")
+}

### 2015-09-15 rscripts/location_clusters.Rmd
breaking plots

@@
 ```{r, echo=FALSE, include=FALSE}
-req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "animation", "randomForest","lubridate", "movMF")
+req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "lubridate", "movMF")
 need.pkgs <- req.packages[!sapply(req.packages, require, character.only = T)]
@@
 
-```{r, echo=F}
+```{r, echo=F, include=F}
 temporal_data_slice <- melt(initial_digest$raw[, .SD, .SDcols = c(1, grep("hr", names(initial_digest$raw)))], id.var = "location_id")
@@
 ```{r, echo=F, fig.cap="Raw proportions by hour"}
-ggplot(temporal_data_slice) + theme_bw() + aes(group = location_id, x = hour, y = value) + facet_grid(measure ~ .) + geom_line(alpha=0.5) + scale_x_discrete(breaks=0:23, expand=c(0,0)) +
+ggplot(temporal_data_slice) + theme_bw() + aes(group = location_id, x = hour, y = value) + facet_grid(measure ~ .) + geom_line(alpha=0.2) + scale_x_discrete(breaks=0:23, expand=c(0,0)) +
   scale_y_continuous("proportion", expand=c(0,0))
@@
   fftref <- mvfft(t(ref[,-1,with=F])-mn)
-  meaningful <- fftref[1+1:12,]
-  pwr <- Re(meaningful*Conj(meaningful))
-  res <- t(pwr)[,c(1,2,3,4,6,8,12)]
+  pwr <- Re(fftref*Conj(fftref))[1+1:12,]
+  res <- t(pwr)
   dimnames(res)[[2]] <- paste0(as.integer(gsub("[^\\d]","",dimnames(res)[[2]], perl=T)), "_perday")
@@
 )
+
 ggplot(pwr[,
   list(power=value, frequency=as.numeric(sub("_perday","", variable))), by=list(location_id, measure)
-]) + theme_bw() + aes(y = power, x=frequency, group=location_id) + facet_grid(measure ~ .) + geom_line(alpha=0.5) + scale_x_continuous("per day frequency", breaks=c(1,2,3,4,6,8,12))
+]) + theme_bw() + aes(y = power, x=frequency, group=location_id) + facet_grid(measure ~ .) + geom_line(alpha=0.2) + scale_x_continuous("per day frequency", breaks=c(1,2,3,4,6,8,12))
+# pwr[, list(peak_freq = variable[which.max(value)]), by=list(location_id, measure)][,.N,by=list(measure, peak_freq)]
 ```
 
-Some clear patterns there: distinct peak locations, distinct periodicity.  Let's develop categories:
-peak time characterization (for this work, just on usage - possible to have circular statistics grouping on all three peaks) and single principle mode of usage (i.e., majority of usage power in the daily frequency) vs.
-multi-mode.
-We use a mixture von Mises model.
+By-eye conclusions:
 
-Similar for weekly signatures - curve, peak day grouping.
+ - peak login time does not separate into clean groups.  So, separate into early, middle, and late thirds.
+ - peak logouts and usage, however, seem to identify three lumps, so we will use mixture von Mises models to identify clusters.
+ - some locations have substantially different power (less than .1 vs greater than 1) in the frequencies below the Nyquist limit (fixed 24 samples per day, so limited to detecting frequences 12 per day)
+ - most locations have peak power on a daily cycle, but a few do not
 
-Drop monthly data - would have to de-trend?
+Categorizing on these features:
 
-We will initially approach the measures with Principle Components Analysis (PCA).
-PCA should discover the best bases to describe the locations - orthogonal, linear combinations of measurements about those locations - in a least squares sense.
-The way PCA works implies two important criteria for inputs: (1) those measures need to be approximately normal and (2) we should ensure that factors that we expect to have non-linear combinations are accommodated.
-Concern (1) is principally about measures that have a one-sided, long-tailed distribution.
-Concern (2) is about measures that should really be combined multiplicatively - *e.g.* total usage (one measure) per unique user (another).
+```{r,echo=F,include=F}
+# http://stackoverflow.com/questions/18788748/fit-a-mixture-of-von-mises-distributions-in-r
+psi <- peaks[measure == "usage", peak/24*pi] # usage angle
+tht <- peaks[measure == "login", peak/24*pi] # login angle
+phi <- peaks[measure == "logout", peak/24*pi] # logout angle
+x0 <- cos(psi)
+x1 <- sin(psi)*cos(tht)
+x2 <- sin(psi)*sin(tht)*cos(phi)
+x3 <- sin(psi)*sin(tht)*sin(phi)
+pos <- cbind(x0,x1,x2,x3)
+temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(movMF(pos, 3, list(maxiter=10000, nruns=20)), pos), key="location_id")
 
-Replacing measures with their logarithms is a coarse way to address these both concerns.
-Taking the log compresses long-tailed distributions into something more Gaussian, and linear combinations of the logs is the same as combining the underlying metrics multiplicatively.
-This approach is not perfect, but absent some principled reason to make particular transformations (or not), we anticipate it is better than not.
+# usage_peak_hr <- peaks[measure == "usage", peak]
+# ang <- usage_peak_hr/12*pi
+# unit_circ <- cbind(cos(ang), sin(ang))
+# predusage <- predict(movMF(unit_circ, 3, list(maxiter=10000, nruns=20)), unit_circ)
+# peaks[measure == "usage", cluster := paste0("usage_", predusage[location_id])]
+# 
+# logout_peak_hr <- peaks[measure == "logout", peak]
+# ang <- logout_peak_hr/12*pi
+# unit_circ <- cbind(cos(ang), sin(ang))
+# predlogout <- predict(movMF(unit_circ, 3, list(maxiter=10000, nruns=20)), unit_circ)
+# peaks[measure == "logout", cluster := paste0("logout_", predlogout[location_id])]
+# 
+# login_peak_hr <- peaks[measure == "login", peak]
+# splits <- quantile(login_peak_hr, probs = c(1,2)/3)
+# login_clusters <- ifelse(login_peak_hr <= splits[1], "low",ifelse(login_peak_hr <= splits[2],"mid", "high"))
+# peaks[measure == "login", cluster := login_clusters[location_id]]
+```
 
-A typically loadings plot (showing vectors for the original dimensions in the new principle component basis) will be a dense cloud of lines at this stage, so we need a different view to identify patterns for the first round of parameter culling.
-We will look at just the first two principle components on a colored bar plot - with bar height for the first component, and color for the second.
-If we see patterns in heights and colors that tell a story that makes sense in the domain, then we can reduce the input measures by combination or elimination.
-TODO: plot for resulting approx normality of measures?
+```{r, echo=F, fig.cap="Login, Logout, Usage Peak Clusters"}
+ggplot(peaks[temporal_cluster]) + theme_bw() + aes(x = peak, fill=factor(vMFcluster)) + facet_grid(measure ~ .) + geom_bar(binwidth=1) + labs(fill="von Mises-Fisher\n3-Sphere\nCluster")
+```
+
+Let's turn to the powers:
+
+```{r, echo=F, fig.cap="Frequency Power Clusters"}
+setkey(pwr, measure, location_id, variable)
+peak_freq_cat <- pwr[,list(peak_freq = ifelse(variable[which.max(value)] == "1_perday","daily","multimode")), keyby=list(measure, location_id)]
+tot_pwr <- pwr[,list(tot_pwr = sum(value)), keyby=list(measure, location_id)]
+pwr_divisions <- tot_pwr[,{
+  res <- as.list(quantile(tot_pwr, probs = c(1,2)/3))
+  names(res) <- c("low","hi")
+  res
+}, keyby=measure]
+pwr_cluster <- tot_pwr[pwr_divisions][peak_freq_cat][,
+  list(peak_freq, pwr_cat = ifelse(tot_pwr < low,"low",ifelse(tot_pwr < hi, "med","hi"))),
+  keyby=list(measure, location_id)
+]
+ggplot(pwr[pwr_cluster][,
+  list(power=value, frequency=as.numeric(sub("_perday","", variable)), peak_freq, pwr_cat),
+  by=list(location_id, measure)
+]) + theme_bw() + aes(y = power, x=frequency, group=location_id, alpha=peak_freq, color = pwr_cat) + facet_grid(measure ~ .) + geom_line() + scale_x_continuous("per day frequency", breaks=c(1,2,3,4,6,8,12))
+```
+
+Showing the power cluster breakdown
+```{r}
+pre_pwr_clust <- data.table(dcast(pwr_cluster, location_id ~ measure, value.var = "pwr_cat"))
+pwr_multi_clusters <- setorder(pre_pwr_clust[,.N,by=list(login, logout, usage)], -N)
+pwr_multi_clusters
+reduced_pwr_clust <- pre_pwr_clust[, list(pwr_clust = ifelse((login == logout) || (login == usage), login, logout)), keyby=location_id]
+```
+Probably worth having all-(low|med|high) clusters; the other cases all (with one exception) are two the same, the third off by one.
+
+And now the peak frequency clusters:
+```{r}
+freq_multi_clusters <- setorder(data.table(dcast(pwr_cluster, location_id ~ measure, value.var = "peak_freq"))[,.N,by=list(login, logout, usage)], -N)
+freq_multi_clusters
+```
+All daily is the vast majority.
+Not worth attempting to categorize by this dimension, except *possibly* as all daily vs.
+not.
+Then again, maybe makes sense to remove daily signal, and then look at what remains?
+
+note: 1 yr ~= log10(lifetime) == 7.5
+
+Finally, we can look at the relative usage per user:
+\newpage\blandscape
+```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="Locations clustered excess usage per user."}
+agg.dt <- initial_digest$raw[,list(log10_total_duration,log10_unique_users,log10_lifetime), keyby=location_id]
+lg10lm <- lm(log10_total_duration-log10_unique_users ~ log10_lifetime, agg.dt)
+agg.dt <- cbind(agg.dt, predict(lg10lm, agg.dt, interval="prediction", level=1/3))[,
+  lifetime_cat := ifelse(log10_total_duration-log10_unique_users < lwr, "low",
+                  ifelse(log10_total_duration-log10_unique_users < upr, "mid", "high"))
+][reduced_pwr_clust][temporal_cluster]
+p <- ggplot(agg.dt) + theme_bw() +
+  aes(y = log10_total_duration-log10_unique_users, x = log10_lifetime, color = factor(vMFcluster), size = pwr_clust, alpha = pwr_clust) +
+  with(agg.dt, annotate("ribbon", x=log10_lifetime, ymax=upr, ymin=lwr, fill="lightgrey")) +
+  geom_vline(xintercept=log10(365*24*60*60), color="grey") +
+  annotate("text", x=log10(365*24*60*60)+c(-0.02,0.02), y=5.75, label=c("< 1 year","> 1 year"), angle=90, color="grey") +
+  geom_point() +
+#  geom_ribbon(aes(ymax=upr, ymin=lwr, y=NULL, color=NULL, size=NULL), alpha=0.2) +
+  labs(
+    x = expression(log[10]("lifetime")),
+    y = expression(log[10]*bgroup("(",frac("total usage","# unique users"),")")),
+    color = "von Mises-Fisher\nPeak Time\nCluster"
+  ) + scale_size_manual("power\ncluster", limits=c("low","med","hi"), values=c(2,5,8)) +
+  scale_alpha_manual("power\ncluster", limits=c("low","med","hi"), values=(2^0.5)/c(2,5,8)^0.5)
+p
+```
+\elandscape\newpage
+
+\newpage\blandscape
+```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="Raw proportions by hour"}
+setkey(pwr_cluster, location_id, measure)
+ggplot(temporal_data_slice[temporal_cluster][pwr_cluster]) + theme_bw() +
+  aes(group = location_id, x = hour, y = value, color=pwr_cat) +
+  facet_grid(measure ~ vMFcluster) + geom_line(alpha=0.3) +
+  scale_x_discrete(breaks=0:23, expand=c(0,0)) +
+  scale_y_continuous("proportion", expand=c(0,0))
+```
+\elandscape\newpage
+
+\newpage\blandscape
+```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="random sample of usage paths"}
+## select some users w/ 5+ locations
+travelers <- sample(censor.dt[,list(ul=length(unique(location_id))),by=user_id][ul >= 5, user_id], 5)
+trips <- setkey(setnames(
+  censor.dt[user_id %in% travelers, location_id, by=user_id][,
+    rle(location_id), by=user_id
+  ], "values","location_id"), user_id)
+locations <- agg.dt[,
+  list(x=log10_lifetime, y=log10_total_duration-log10_unique_users),
+  by=location_id
+]
+paths <- trips[,cbind(locations[location_id, list(x, y)], lengths, agg.dt[location_id, list(lifetime_cat, pwr_clust, vMFcluster)], user=.GRP), by=user_id]
+p + geom_path(aes(x=x, y=y, linetype=factor(user), color=NULL, alpha=NULL, size=NULL), paths) +
+  with(paths[lengths > 1],
+    annotate("point", x=x, y=y, shape=c(0,1,2,5,6)[user], size=10*sqrt(lengths/max(lengths)))
+  ) +
+  guides(color="none", alpha="none", size="none") + labs(linetype="user")
+```
+\elandscape\newpage
+
+Now if take these clusters and characterize user visits in terms of cluster visits, instead of particular location visits, we can consider their usage of cluster "types".
+There are two main questions: how much (relatively) do individuals use a type?
+and how often do they rotate between types?
 
 ```{r, echo=F}
-initial_digest <- digestor(censor.dt, refpredictors, training_locations)
-initial_pca <- dopca(initial_digest)
-init_bars <- make_bars(initial_pca$pca)
-```
+cluster_rle_src <- merge(censor.dt, agg.dt, by="location_id")
+setkey(cluster_rle_src, login)
+ref <- censor.dt[,
+  list(
+    ul=length(unique(location_id)),
+    life=max(logout)-min(login),
+    usage=sum(logout-login)),
+  keyby=user_id
+]
 
-\newpage
-\blandscape
-```{r, echo=F, fig.width=11.25, fig.height=7 }
-plot_pca_bars(init_bars)
-```
-\elandscape
-\newpage
+categorizer_help <- function(opts) function(SD) SD[,
+  c(list(tot_visits = sum(lengths), mn_reps=mean(lengths), switches=.N),
+    lapply(opts, function(v) SD[values == v, sum(lengths)])
+  )
+]
 
-There seems to be something consistent going on with the hourly measures.  Upon closer examination:
+categorizer <- function(src.dt, cat, ref) {
+  rle_expr <- paste0("rle(",cat,")")
+  key_expr <- c("user_id",cat)
+  mx_expr <- parse(text=paste0("list(main=",cat,"[which.max(prop)],share=max(prop))"))
+  rle_dt <- src.dt[, rle(eval(parse(text=cat))), keyby=user_id]
+  prop_ref <- src.dt[,list(prop=sum(logout-login)), keyby=key_expr][ref][,list(prop=prop/usage), keyby=key_expr]
+  prop_lab <- prop_ref[, eval(mx_expr), by=user_id]
+  opts <- sort(unique(rle_dt$values))
+  names(opts) <- paste0("cat_", opts)
+  fun <- categorizer_help(opts)
+  ## cast data, maybe?
+  digest <- rle_dt[, list(tot_visits = sum(lengths), mn_reps=mean(lengths), switches=.N), keyby=user_id][ref][prop_lab]
+  list(rle_dt=rle_dt, prop_ref=prop_ref, prop_lab=prop_lab, digest=digest)
+}
 
-```{r, echo=F}
-zoom_plot <- function(pca, target = 1:2, brs = make_bars(pca, target)) {
-  pcs <- paste0("PC", target)
-  zoom_bars <- setnames(
-    brs[ grepl("hr", measure) ][, c(list(
-      type=ifelse(grepl("login", measure),"login",ifelse(grepl("logout", measure), "logout", "usage")),
-      hr = as.integer(sub("^[^\\d]+(\\d+)$","\\1", measure, perl=T))
-    ), y=.SD[,2,with=F], fill=.SD[,3,with=F]), .SDcols=c("measure", pcs)],
-    paste(c("y","fill"), pcs, sep='.'),
-    c("y","fill")
-  )
-  limY <- max(abs(zoom_bars[,3,with=F]))
-  limC <- max(abs(zoom_bars[,4,with=F]))
-  ggplot(zoom_bars) + facet_grid(type ~ .) + aes(x = hr, y = y, fill = fill) +
-    theme_bw() + geom_bar(stat="identity", position = "identity") +
-    labs(y=pcs[1]) + ylim(-limY,limY) +
-    scale_fill_gradient2(pcs[2], limits=c(-limC, limC), low="red", mid = "grey", high="blue")
-}
-zoom_plot(brs = init_bars)
-```
+#lifetime_dts <- categorizer(cluster_rle_src, "lifetime_cat", ref)
+#pwr_dts <- categorizer(cluster_rle_src, "pwr_clust", ref)
+#vMF_dts <- categorizer(cluster_rle_src, "vMFcluster", ref)
 
-The trends in between login, usage, and logout seem very similar in principle components 1 and 2, so it seems reasonable to just pick one of them.
-Similarly, they seem to be obviously driven on a daily period, so perhaps we can just use the peak amplitude to distinguish locations.
-Before we do so, let's check that trends persist for some of the higher order principle components.
+# lifetime_rle <- cluster_rle_src[,rle(lifetime_cat),keyby=user_id]
+# lifetime_prop <- cluster_rle_src[,list(prop=sum(logout-login)), keyby=list(user_id, lifetime_cat)][ref][,list(prop=prop/usage), keyby=list(user_id, lifetime_cat)]
+# lifetime_cat_dt <- lifetime_prop[, list(main=lifetime_cat[which.max(prop)], majority=max(prop)>0.5), by=user_id]
+# 
+# pwr_rle <- cluster_rle_src[,rle(pwr_clust),keyby=user_id]
+# pwr_prop <- cluster_rle_src[,list(prop=sum(logout-login)), keyby=list(user_id, pwr_clust)][ref][,list(prop=prop/usage), keyby=list(user_id, pwr_clust)]
+# pwr_prop_dt <- pwr_prop[, list(main=pwr_clust[which.max(prop)], majority=max(prop)>0.5), by=user_id]
+# 
+# 
+# vMF_rle <- cluster_rle_src[,rle(vMFcluster), keyby=user_id]
+# vMF_prop <- cluster_rle_src[,list(prop=sum(logout-login)), keyby=list(user_id, vMFcluster)][ref][,list(prop=prop/usage), keyby=list(user_id, vMFcluster)]
+# vMF_prop_dt <- vMF_prop[, list(main=vMFcluster[which.max(prop)], majority=max(prop)>0.5), by=user_id]
+# 
+# 
+# digest_lifetime_rle <- lifetime_rle[,list(
+#     tot_visits = sum(lengths),
+#     lows=sum((values=="low")*lengths),
+#     mids=sum((values=="mid")*lengths),
+#     highs=sum((values=="high")*lengths),
+#     mn_reps=mean(lengths),
+#     switches = .N
+#   ), keyby=user_id][ref][lifetime_cat_dt]
+# digest_pwr_rle <- pwr_rle[,list(
+#     tot_visits = sum(lengths),
+#     lows=sum((values=="low")*lengths),
+#     mids=sum((values=="mid")*lengths),
+#     highs=sum((values=="high")*lengths),
+#     mn_reps=mean(lengths),
+#     switches = .N
+#   ), keyby=user_id][ref][pwr_prop_dt]
+# digest_vMF_rle <- vMF_rle[,list(
+#     tot_visits = sum(lengths),
+#     As=sum((values==1)*lengths),
+#     Bs=sum((values==2)*lengths),
+#     Cs=sum((values==3)*lengths),
+#     mn_reps=mean(lengths),
+#     switches = .N
+#   ), keyby=user_id][ref][vMF_prop_dt]
 
-```{r, echo=F}
-zoom_plot(initial_pca$pca, 3:4)
+# path_by_cluster <- paths[, list(user_id, lengths, loc_cluster = interaction(lifetime_cat, pwr_clust, vMFcluster))]
 ```
@@
 ```{r, echo=F}
-zoom_plot(initial_pca$pca, 5:6)
+#ggplot(lifetime_dts$digest) + theme_bw() + aes(x=log10(switches/usage), fill=main, alpha=majority) + geom_bar() + labs(title="By Lifetime Categories") + scale_alpha_manual(values=c(0.6, 1))
+#ggplot(pwr_dts$digest) + theme_bw() + aes(x=log10(switches/usage), fill=main, alpha=majority) + geom_bar() + labs(title="By Frequency Power Categories") + scale_alpha_manual(values=c(0.6, 1))
+#ggplot(vMF_dts$digest) + theme_bw() + aes(x=log10(switches/usage), fill=factor(main), alpha=majority) + geom_bar() + labs(title="By von Mises-Fisher Categories") + scale_alpha_manual(values=c(0.6, 1))
 ```
-
-So those trends seem to be preserved, but that higher frequency content is necessary as well.
-We will compare that frequency to our expectations, and see how much is worthy of keeping, and also check the distribution of peak usage:
-<!---
-with a two caveats: morning logouts (hour 8) are consistently different from usage, as are late night logins (peaks at hours 20 and 23).
-This has a pretty compelling story: the differences between locations are generally about usage, though we can additionally differentiate places with variable morning turnover (excess of logouts) and variable evening (spiking with the dinner and "party" hours) starts (excess of logins).
-
-If we focus on just usage, the plots between the components are pretty clearly periodic in both dimensions - like rotating polarized light.
-Perhaps the right usage measure is instead (1) how peaky is the distribution across hours (what is the amplitude of the wave), (2) what is the period of usage peaks (per day or multiples of per day), and (3) when are the peaks located?
-
-delta T = 1 hr
-1/(2 hr) = nyquist freq
-1/day, 2/day, 3/day, 4/day, 6/day, 8/day, 12/day <- valid signals
---->
-```{r, echo=F}
-# http://stackoverflow.com/questions/18788748/fit-a-mixture-of-von-mises-distributions-in-r
-usage_peak_hr <- initial_digest$raw[, apply(.SD, 1, which.max), .SDcols=grep("^hr", names(initial_digest$raw))]
-ang <- usage_peak_hr/12*pi
-unit_circ <- cbind(cos(ang), sin(ang))
-#login_peak_hr <- initial_digest$raw[, apply(.SD, 1, which.max), .SDcols=grep("^login", names(initial_digest$raw))]
-#logout_peak_hr <- initial_digest$raw[, apply(.SD, 1, which.max), .SDcols=grep("^logout", names(initial_digest$raw))]
-#underlying <- cbind(usage_peak_hr, login_peak_hr, logout_peak_hr)/24
-thing <- movMF(unit_circ, 3, list(maxiter=10000, nruns=20))
-predthing <- predict(thing, unit_circ)
-ggplot(data.frame(hour = usage_peak_hr-1, clz=factor(predthing))) + theme_bw() + aes(x=hour, fill=clz, group=clz) +
-  geom_bar(binwidth=1) +
-  scale_y_continuous(expand=c(0,0)) +
-  scale_x_discrete(breaks=1:24)
-```
-
-superpose three normal distributions, on periodic boundary condition, assign usage based on most probably membership to peak 1, 2, 3
-
-```{r, echo=F}
-peak_freq_power <- {
-  ref <- initial_digest$raw[,.SD,.SDcols=c(1,grep("^hr", names(initial_digest$raw)))]
-  mn <- mean(as.matrix(ref[,-1,with=F]))
-  ref.mlt <- melt(ref, id.var = "location_id")
-#  ggplot(ref.mlt) + theme_bw() + aes(x=variable, y=value-mn, group=location_id) + geom_line(alpha=0.2)
-  fftref <- mvfft(t(ref[,-1,with=F])-mn)
-  meaningful <- fftref[1+1:12,] # trim average signal, freqs above Nyquist
-  pwr <- Re(meaningful*Conj(meaningful))
-#   thing <- melt(data.table(cbind(location_id = 1:297, t(pwr))), id.var = "location_id")
-#   thing[, period := 24/as.numeric(sub("hr","", variable)) ]
-#   ggplot(thing[variable %in% paste0("hr",c(1,2,3,4,6,8,12))]) + theme_bw() + aes(y = value, group = location_id, x = period) + geom_line(alpha=0.2) + scale_x_continuous(breaks=c(2,3,4,6,8,12,24))
-#   ggplot(thing[variable %in% paste0("hr",c(1,2,3,4,6,8,12))]) + theme_bw() + aes(x = log(value)) + facet_grid(period ~ .) + geom_bar()
-  res <- t(pwr)[,c(1,2,3,4,6,8,12)]
-  dimnames(res)[[2]] <- paste0("usage_period_", 24/as.integer(sub("hr","",dimnames(res)[[2]])))
-  res
-}
-ref <- copy(initial_digest$raw)
-setnames(ref, c("0%","100%"), c("min","max"))
-ref <- cbind(ref[,.SD, .SDcols=grep("(hr|%)", names(ref), invert=T)], usage_peak = usage_peak_hr, peak_freq_power)
-#freqs <- melt(ps, id.var="location_id")
-#freqs[, period := 24/as.numeric(sub("V","", variable))]
-
-# net <- cbind(location_id = initial_digest$raw$location_id, initial_digest$raw[, .SD[,1:24,with=F]-.SD[,1:24+24,with=F], .SDcols=grep("log(in|out)", names(initial_digest$raw))])[, {
-#   z  <- fft(t(.SD[,-1,with=F]))[1:12] # Nyquist limit is 24 samples / 24 hrs / 2 = 12 / 24 hrs = 1 / 2 hrs
-#   ps <- Re(z*Conj(z))
-#   as.list(ps[1:12])
-# }, by=location_id]
-# 
-# ggplot(freqs[variable %in% paste0("V",c(1,2,3,4,6,8,12))]) + theme_bw() + aes(y = value, group = location_id, x = period) + geom_line(alpha=0.2) + scale_x_continuous(breaks=c(2,3,4,6,8,12,24))
-```
-
-so pick an indicator for the full day, divided into comparable blocks, then have indicators for wonky periods:
-
- usage, 2-5, 10-12, 13-15, 19-21
-
-There seems to be pretty clear relationships between hours 9 to 15 - pretty standard working day - across all measures, but hour 16 seems perhaps to differentiate on places that close earlier versus later.
-Logins, logouts, and usage appear to have a similar trend from 0 to 7 - definite sleeping or early morning hours.
-There appear to be some interesting outliers in hour 20 and 23 for logins - late night hotspots perhaps?
-
-Let's take these metrics down their weights between 0 to 7 (morning), 8 (opening), 9 to 15 (day), 16 (closing), 17-19 (evening), and then 20-23 (late night).
-
-```{r, echo=F, include=F}
-# decision free version? sort(kmeans(initial_pca$pca$rotation, 10, 10)$cluster)
-# ref <- copy(initial_digest$raw)
-# hr_items <- grep('hr', names(ref))
-# ref[,usage_morning := apply(.SD[, hr_items[1:8],with=F],1,sum)]
-# ref[,login_morning := apply(.SD[, hr_items[1:8+24],with=F],1,sum)]
-# ref[,logout_morning := apply(.SD[, hr_items[1:8+48],with=F],1,sum) ]
-# ref[,usage_opening := apply(.SD[, hr_items[9],with=F],1,sum)]
-# ref[,login_opening := apply(.SD[, hr_items[9+24],with=F],1,sum)]
-# ref[,logout_opening := apply(.SD[, hr_items[9+48],with=F],1,sum) ]
-# ref[,usage_day := apply(.SD[, hr_items[10:16],with=F],1,sum)]
-# ref[,login_day := apply(.SD[, hr_items[10:16+24],with=F],1,sum)]
-# ref[,logout_day := apply(.SD[, hr_items[10:16+48],with=F],1,sum) ]
-# ref[,usage_closing := apply(.SD[, hr_items[17],with=F],1,sum)]
-# ref[,login_closing := apply(.SD[, hr_items[17+24],with=F],1,sum)]
-# ref[,logout_closing := apply(.SD[, hr_items[17+48],with=F],1,sum) ]
-# ref[,usage_evening := apply(.SD[, hr_items[18:20],with=F],1,sum)]
-# ref[,login_evening := apply(.SD[, hr_items[18:20+24],with=F],1,sum)]
-# ref[,logout_evening := apply(.SD[, hr_items[18:20+48],with=F],1,sum) ]
-# ref[,usage_late := apply(.SD[, hr_items[21:24],with=F],1,sum)]
-# ref[,login_late := apply(.SD[, hr_items[21:24+24],with=F],1,sum)]
-# ref[,logout_late := apply(.SD[, hr_items[21:24+48],with=F],1,sum) ]
-# ref <- ref[,-hr_items,with=F]
-```
-
-With the consolidated measures:
-
-\newpage
-\blandscape
-```{r, echo=F, fig.width=11.25, fig.height=7 }
-second_digest <- redigest(ref, training_locations)
-second_pca <- dopca(second_digest)
-second_bars <- make_bars(second_pca$pca)
-plot_pca_bars(second_bars)
-```
-\elandscape
-\newpage
-
-There is pretty compellingly no information between the logins, logouts, and usage during the day.
-This seems a reasonable conclusion for evening and morning as well.
-We will use only login information for those categories, since that seems have the most weight in components 1 and 2.
-```{r, echo=F}
-# ref <- copy(second_digest$raw)
-# drop <- grep("(logout|usage)_(morning|day|evening)", names(ref))
-```
-
-There also seems to be fairly similar information between the middle duration quantiles, average, and standard deviations.
-The max and min seem informative, but we will remove the inter-quartiles and median.
-```{r, echo=F}
-# ref <- copy(second_digest$raw)
-# drop <- c(drop, grep("(25|50|75)%", names(ref)))
-```
-
-Finally, Tuesday and Wednesday - early working week - seem very similar.
-Thursday seems somewhat similar as well, but not enough.
-We will try a Tue-Wed category - usage outside of the start of the working week (Monday), the days when people regularly go out in the evenings (Thurs-Fri), and the weekends.
-
-\newpage
-\blandscape
-```{r, echo=F}
-# decision free version? sort(kmeans(initial_pca$pca$rotation, 10, 10)$cluster)
-# ref[,TueWed := Tues + Wed]
-# drop <- c(drop, grep("^(Tues|Wed)$", names(ref)))
-# ref <- ref[,-drop, with=F]
-# setcolorder(ref,
-#   c("location_id", "log10_lifetime", "log10_total_duration", "log10_unique_users",
-#   "ave_duration", "sd_duration", "0%", "100%",
-#   "Sun", "Mon", "TueWed", "Thurs", "Fri", "Sat",
-# "login_morning", "usage_opening", "login_opening", "logout_opening", "login_day", "usage_closing", "login_closing", "logout_closing", "login_evening", "usage_late", "login_late", "logout_late",
-#   month.abb
-# ))
-# third_digest <- redigest(ref, training_locations)
-# third_pca <- dopca(third_digest)
-# third_bars <- make_bars(third_pca$pca)
-# plot_pca_bars(third_bars)
-```
-\elandscape
-\newpage
-
-Now if we look at the months, there seems to be a Winter group (Jan-Feb), a Summer group (Jun-Jul), and a Fall group (Oct-Nov)
-
-\newpage
-\blandscape
-```{r, echo=F}
-# decision free version? sort(kmeans(initial_pca$pca$rotation, 10, 10)$cluster)
-# ref <- copy(third_digest$raw)
-# ref[, JanFeb := Jan+Feb ]
-# ref[, JunJul := Jun+Jul ]
-# ref[, OctNov := Oct+Nov ]
-# drop <- grep("^(Jan|Feb|Jun|Jul|Oct|Nov)$", names(ref))
-# ref <- ref[,-drop, with=F]
-# setcolorder(ref,
-#   c("location_id", "log10_lifetime", "log10_total_duration", "log10_unique_users",
-#   "ave_duration", "sd_duration", "0%", "100%",
-#   "Sun", "Mon", "TueWed", "Thurs", "Fri", "Sat",
-# "login_morning", "usage_opening", "login_opening", "logout_opening", "login_day", "usage_closing", "login_closing", "logout_closing", "login_evening", "usage_late", "login_late", "logout_late",
-#   "JanFeb","Mar","Apr","May","JunJul","Aug","Sep","OctNov","Dec"
-# ))
-# fourth_digest <- redigest(ref, training_locations)
-# fourth_pca <- dopca(fourth_digest)
-# fourth_bars <- make_bars(fourth_pca$pca)
-# plot_pca_bars(fourth_bars)
-```
-\elandscape
-\newpage
-
-At this point, it seems like login proportions are sufficient to consider, and that usage and logout measures can be ignored.
-Though usage and logout seem to have more weight in some periods, picking just one seems like a reasonable choice.
-
-TODO: more downselect
-
-TODO: vector/point cloud loadings plots w/ reduced measures
-
-TODO: re-run kmeans on reduced set to get clusters
-
-TODO: plots of measurement distributions, color categorized by cluster

### 2015-09-17 rscripts/location_clusters.Rmd
jp commit something

@@
 pos <- cbind(x0,x1,x2,x3)
+
+lls <- lapply(8:11, function(k) movMF(pos, k, list(maxiter=10000, nruns=20)))
+bics <- lapply(lls, stats::BIC)
+
 temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(movMF(pos, 3, list(maxiter=10000, nruns=20)), pos), key="location_id")
+
+alt_temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(lls[[1]], pos), key="location_id")
 
@@
 pwr_cluster <- tot_pwr[pwr_divisions][peak_freq_cat][,
-  list(peak_freq, pwr_cat = ifelse(tot_pwr < low,"low",ifelse(tot_pwr < hi, "med","hi"))),
+  list(peak_freq, pwr_cat = ifelse(tot_pwr < low,"lo",ifelse(tot_pwr < hi, "med","hi"))),
   keyby=list(measure, location_id)
@@
 
+location diversity - share of locations in each type, in each dimension.
+We know by construction that location diversity is uniform for low, medium, and high usage per lifetime categorization: we created the categories based on quantiles.
+Same is true for use variation frequency power.
+For the clusters of peak utilization characteristics, however,
+
+```{r, echo=F}
+multinomtest <- function(xs, nh=rep(1/length(xs), times=length(xs))) {
+  N <- sum(xs)
+  2*N*log(N) - 2*sum(xs*log(nh/xs))
+}
+agg.dt[,.N,by=lifetime_cat][,list(lifetime_cat, share=N/sum(N))]
+agg.dt[,.N,by=pwr_clust][,list(pwr_clust, share=N/sum(N))]
+agg.dt[,.N,by=vMFcluster][,list(vMFcluster, share=N/sum(N))]
+```
+
 Now if take these clusters and characterize user visits in terms of cluster visits, instead of particular location visits, we can consider their usage of cluster "types".
@@
   prop_lab <- prop_ref[, eval(mx_expr), by=user_id]
-  opts <- sort(unique(rle_dt$values))
-  names(opts) <- paste0("cat_", opts)
-  fun <- categorizer_help(opts)
+  chunks <- data.table(dcast(rle_dt, user_id ~ values, value.var="lengths", fun.aggregate = sum), key="user_id")
   ## cast data, maybe?
-  digest <- rle_dt[, list(tot_visits = sum(lengths), mn_reps=mean(lengths), switches=.N), keyby=user_id][ref][prop_lab]
+  digest <- rle_dt[, list(tot_visits = sum(lengths), mn_reps=mean(lengths), switches=.N), keyby=user_id][ref][prop_lab][chunks]
   list(rle_dt=rle_dt, prop_ref=prop_ref, prop_lab=prop_lab, digest=digest)
@@
 
-#lifetime_dts <- categorizer(cluster_rle_src, "lifetime_cat", ref)
-#pwr_dts <- categorizer(cluster_rle_src, "pwr_clust", ref)
-#vMF_dts <- categorizer(cluster_rle_src, "vMFcluster", ref)
-
-# lifetime_rle <- cluster_rle_src[,rle(lifetime_cat),keyby=user_id]
-# lifetime_prop <- cluster_rle_src[,list(prop=sum(logout-login)), keyby=list(user_id, lifetime_cat)][ref][,list(prop=prop/usage), keyby=list(user_id, lifetime_cat)]
-# lifetime_cat_dt <- lifetime_prop[, list(main=lifetime_cat[which.max(prop)], majority=max(prop)>0.5), by=user_id]
-# 
-# pwr_rle <- cluster_rle_src[,rle(pwr_clust),keyby=user_id]
-# pwr_prop <- cluster_rle_src[,list(prop=sum(logout-login)), keyby=list(user_id, pwr_clust)][ref][,list(prop=prop/usage), keyby=list(user_id, pwr_clust)]
-# pwr_prop_dt <- pwr_prop[, list(main=pwr_clust[which.max(prop)], majority=max(prop)>0.5), by=user_id]
-# 
-# 
-# vMF_rle <- cluster_rle_src[,rle(vMFcluster), keyby=user_id]
-# vMF_prop <- cluster_rle_src[,list(prop=sum(logout-login)), keyby=list(user_id, vMFcluster)][ref][,list(prop=prop/usage), keyby=list(user_id, vMFcluster)]
-# vMF_prop_dt <- vMF_prop[, list(main=vMFcluster[which.max(prop)], majority=max(prop)>0.5), by=user_id]
-# 
-# 
-# digest_lifetime_rle <- lifetime_rle[,list(
-#     tot_visits = sum(lengths),
-#     lows=sum((values=="low")*lengths),
-#     mids=sum((values=="mid")*lengths),
-#     highs=sum((values=="high")*lengths),
-#     mn_reps=mean(lengths),
-#     switches = .N
-#   ), keyby=user_id][ref][lifetime_cat_dt]
-# digest_pwr_rle <- pwr_rle[,list(
-#     tot_visits = sum(lengths),
-#     lows=sum((values=="low")*lengths),
-#     mids=sum((values=="mid")*lengths),
-#     highs=sum((values=="high")*lengths),
-#     mn_reps=mean(lengths),
-#     switches = .N
-#   ), keyby=user_id][ref][pwr_prop_dt]
-# digest_vMF_rle <- vMF_rle[,list(
-#     tot_visits = sum(lengths),
-#     As=sum((values==1)*lengths),
-#     Bs=sum((values==2)*lengths),
-#     Cs=sum((values==3)*lengths),
-#     mn_reps=mean(lengths),
-#     switches = .N
-#   ), keyby=user_id][ref][vMF_prop_dt]
+lifetime_dts <- categorizer(cluster_rle_src, "lifetime_cat", ref)
+pwr_dts <- categorizer(cluster_rle_src, "pwr_clust", ref)
+vMF_dts <- categorizer(cluster_rle_src, "vMFcluster", ref)
 
@@
 
-```{r, echo=F}
-#ggplot(lifetime_dts$digest) + theme_bw() + aes(x=log10(switches/usage), fill=main, alpha=majority) + geom_bar() + labs(title="By Lifetime Categories") + scale_alpha_manual(values=c(0.6, 1))
-#ggplot(pwr_dts$digest) + theme_bw() + aes(x=log10(switches/usage), fill=main, alpha=majority) + geom_bar() + labs(title="By Frequency Power Categories") + scale_alpha_manual(values=c(0.6, 1))
-#ggplot(vMF_dts$digest) + theme_bw() + aes(x=log10(switches/usage), fill=factor(main), alpha=majority) + geom_bar() + labs(title="By von Mises-Fisher Categories") + scale_alpha_manual(values=c(0.6, 1))
+\newpage\blandscape
+```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="distribution of main location type usage"}
+digests <- rbind(
+  lifetime_dts$digest[,list(switch_rate = switches/usage, main, share, type="usage")],
+  pwr_dts$digest[,list(switch_rate = switches/usage, main, share, type="power")],
+  vMF_dts$digest[,list(switch_rate = switches/usage, main, share, type="peaktime")]
+)
+ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1) + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
+ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1, position="fill") + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
 ```
+\elandscape\newpage
+
+\newpage\blandscape
+```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="user categorization, by primary location type utilization"}
+plot_ref <- ref[lifetime_dts$digest[, list(lifetime_main = main), keyby=user_id]][pwr_dts$digest[, list(pwr_main = main), keyby=user_id]][vMF_dts$digest[, list(peak_main = main), keyby=user_id]]
+#thing[, list(user_id, mains=interaction(lifetime_main, pwr_main, peak_main))][,.N,by=mains]
+ggplot(plot_ref[life > 365*24*60*60 & ul > 2 & ul < 20, list(y=log10(usage), x=log10(life), pwr_main, lifetime_main, peak_main)]) +facet_grid(pwr_main ~ lifetime_main) + aes(x=x,y=y, color=factor(peak_main)) + geom_point(alpha=0.3) + theme_bw()
+```
+\elandscape\newpage

### 2015-09-21 rscripts/location_clusters.Rmd
assorted updates

@@
 
-lls <- lapply(8:11, function(k) movMF(pos, k, list(maxiter=10000, nruns=20)))
-bics <- lapply(lls, stats::BIC)
+#lls <- lapply(8:11, function(k) movMF(pos, k, list(maxiter=10000, nruns=20)))
+#bics <- lapply(lls, stats::BIC)
+vmfmodel <- movMF(pos, 3, list(maxiter=10000, nruns=20))
+temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(vmfmodel, pos), key="location_id")
+vmfrename <- peaks[measure == "usage"][temporal_cluster][,
+  list(med=median(peak)),by=vMFcluster
+][,vMFcluster, keyby=med][,list(vMFcluster, rename=factor(c("early","middle","late"), levels=c("early","middle","late"), ordered = T))]
+temporal_cluster <- merge(temporal_cluster, vmfrename, by="vMFcluster")[,list(vMFcluster=rename), keyby=location_id]
 
-temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(movMF(pos, 3, list(maxiter=10000, nruns=20)), pos), key="location_id")
-
-alt_temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(lls[[1]], pos), key="location_id")
+#alt_temporal_cluster <- data.table(location_id = peaks[measure == "usage"]$location_id, vMFcluster = predict(lls[[1]], pos), key="location_id")
 
@@
 ```{r, echo=F, fig.cap="Login, Logout, Usage Peak Clusters"}
-ggplot(peaks[temporal_cluster]) + theme_bw() + aes(x = peak, fill=factor(vMFcluster)) + facet_grid(measure ~ .) + geom_bar(binwidth=1) + labs(fill="von Mises-Fisher\n3-Sphere\nCluster")
+ggplot(peaks[temporal_cluster]) + theme_bw() + aes(x = peak, fill=vMFcluster) + facet_grid(measure ~ .) + geom_bar(binwidth=1) + labs(fill="von Mises-Fisher\n3-Sphere\nCluster\nPeak Time")
 ```
@@
 pwr_multi_clusters
-reduced_pwr_clust <- pre_pwr_clust[, list(pwr_clust = ifelse((login == logout) || (login == usage), login, logout)), keyby=location_id]
+reduced_pwr_clust <- pre_pwr_clust[, list(pwr_clust = factor(ifelse((login == logout) || (login == usage), login, logout), levels=c("lo","med","hi"), ordered=T)), keyby=location_id]
 ```
@@
 agg.dt <- cbind(agg.dt, predict(lg10lm, agg.dt, interval="prediction", level=1/3))[,
-  lifetime_cat := ifelse(log10_total_duration-log10_unique_users < lwr, "low",
-                  ifelse(log10_total_duration-log10_unique_users < upr, "mid", "high"))
+  lifetime_cat := factor(ifelse(log10_total_duration-log10_unique_users < lwr, "low",
+                  ifelse(log10_total_duration-log10_unique_users < upr, "mid", "high")), levels=c("low","mid","high"), ordered=T)
 ][reduced_pwr_clust][temporal_cluster]
 p <- ggplot(agg.dt) + theme_bw() +
-  aes(y = log10_total_duration-log10_unique_users, x = log10_lifetime, color = factor(vMFcluster), size = pwr_clust, alpha = pwr_clust) +
+  aes(y = log10_total_duration-log10_unique_users, x = log10_lifetime, color = vMFcluster, size = pwr_clust, alpha = pwr_clust) +
   with(agg.dt, annotate("ribbon", x=log10_lifetime, ymax=upr, ymin=lwr, fill="lightgrey")) +
@@
 
-categorizer_help <- function(opts) function(SD) SD[,
-  c(list(tot_visits = sum(lengths), mn_reps=mean(lengths), switches=.N),
-    lapply(opts, function(v) SD[values == v, sum(lengths)])
-  )
-]
-
 categorizer <- function(src.dt, cat, ref) {
@@
   key_expr <- c("user_id",cat)
-  mx_expr <- parse(text=paste0("list(main=",cat,"[which.max(prop)],share=max(prop))"))
-  rle_dt <- src.dt[, rle(eval(parse(text=cat))), keyby=user_id]
+  lvls <- levels(src.dt[[cat]])
+  mx_expr <- parse(text=paste0("list(main=factor(",cat,"[which.max(prop)], ordered=T, levels=lvls),share=max(prop))"))
+  rle_dt <- src.dt[, rle(as.numeric(eval(parse(text=cat)))), keyby=user_id]
   prop_ref <- src.dt[,list(prop=sum(logout-login)), keyby=key_expr][ref][,list(prop=prop/usage), keyby=key_expr]
@@
 ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1) + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
+```
+\elandscape\newpage
+
+\newpage\blandscape
+```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="normalized distribution of main location type usage"}
 ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1, position="fill") + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
@@
 ```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="user categorization, by primary location type utilization"}
-plot_ref <- ref[lifetime_dts$digest[, list(lifetime_main = main), keyby=user_id]][pwr_dts$digest[, list(pwr_main = main), keyby=user_id]][vMF_dts$digest[, list(peak_main = main), keyby=user_id]]
+plot_ref <- ref[
+  lifetime_dts$digest[, list(lifetime_main = main, lifetime_share = share), keyby=user_id]
+][
+  pwr_dts$digest[, list(pwr_main = main, pwr_share = share), keyby=user_id]
+][
+  vMF_dts$digest[, list(peak_main = main, peak_share = share), keyby=user_id]
+]
+user_src.dt <- plot_ref[between(life, 365*24*60*60, 3*365*24*60*60) & between(ul,3,20)]
 #thing[, list(user_id, mains=interaction(lifetime_main, pwr_main, peak_main))][,.N,by=mains]
-ggplot(plot_ref[life > 365*24*60*60 & ul > 2 & ul < 20, list(y=log10(usage), x=log10(life), pwr_main, lifetime_main, peak_main)]) +facet_grid(pwr_main ~ lifetime_main) + aes(x=x,y=y, color=factor(peak_main)) + geom_point(alpha=0.3) + theme_bw()
+ggplot(user_src.dt[, list(y=log10(usage), x=log10(life), pwr_main, lifetime_main, peak_main, membership = lifetime_share * pwr_share * peak_share)]) +facet_grid(pwr_main ~ lifetime_main, labeller = function(var, val){
+  if (var == "pwr_main") {
+    ifelse(val == "lo", "smooth usage", ifelse(val=="med","some usage variability","highly variable usage"))
+  } else {
+    ifelse(val == "low", "low usage per time", ifelse(val=="mid","typical usage per time","high usage per time"))
+  }
+}) + aes(x=x,y=y, color=factor(peak_main), alpha = membership) + geom_point() + theme_bw() + labs(y="log(usage)", x="log(life)", color="usage\npeak")
 ```
 \elandscape\newpage
+
+```{r, echo=F}
+# thing <- 
+#   data.table(
+#     x=lifetime_dts$digest[, list(pos = (0.5*`2`+1*`3`)/(`1`+`2`+`3`), main, share, user_id)]$pos,
+#     y=pwr_dts$digest[, list(pos = (0.5*`2`+1*`3`)/(`1`+`2`+`3`), main, share, user_id)]$pos,
+#     z=vMF_dts$digest[, list(pos = (0.5*`2`+1*`3`)/(`1`+`2`+`3`), main, share, user_id)]$pos,
+#     key=c("x","y","z")
+#   )
+# uthing <- unique(thing)
+# lenthing <- dim(uthing)[1]
+# cloud(z ~ x*y, data=unique(thing)[sample.int(lenthing, 250)])
+# censor.dt[user_id %in% user_src.dt$user_id, .N*60*60*24*30/(max(logout)-min(login)), by=user_id][V1>1]
+```

### 2015-10-06 src/test/scala/edu/cap10/actormodels/covert/SynthUserTests.scala
update tests

@@ -1,0 +1,26 @@
+package edu.cap10.actormodels.covert
+
+import org.scalatest.FunSuite
+import org.scalactic.Tolerance._
+
+/**
+ * @author cap10
+ */
+class SynthUserTests extends FunSuite {
+  
+  val thing = Array.fill(12)(0d)
+  
+  
+  val testLocOne = Location(-1, Array.fill(24)(1d/24), Array.fill(24)(1d), Array.fill(24)(1d))
+  val testLocTwo = Location(-2,
+      (List.fill(12)(0d) ++ List.fill(12)(1d/12)).toArray,
+      Array.fill(24)(1d),
+      (List.fill(12)(0d) ++ List.fill(12)(1d)).toArray
+  )
+  
+  test("a synth user with one location just uses that location pdf") {
+    val userA = SynthUser(-1, 1, 1, Array(testLocOne), Array(1.0), .9)
+    assert(userA.pdfHour.length === testLocOne.pdf.length)
+    for (i <- 0 until 24) assert(userA.pdfHour(i) === testLocOne.pdf(i) +- 1e-8)
+  }
+}

### 2015-10-15 rscripts/location_clusters.Rmd
update tests
working covert sim

@@
 user_src.dt <- plot_ref[between(life, 365*24*60*60, 3*365*24*60*60) & between(ul,3,20)]
+saveRDS(user_src.dt, "../input/user.RData")
 #thing[, list(user_id, mains=interaction(lifetime_main, pwr_main, peak_main))][,.N,by=mains]
@@
 ```{r, echo=F}
+saveRDS(agg.dt[,list(lifetime_cat, pwr_clust, vMFcluster), by=location_id], "../input/locClusters.RData")
+intermediate <- censor.dt[user_id %in% user_src.dt$user_id, list(visits = .N), by=list(user_id, location_id)][,
+  list(location_id, pref=visits/sum(visits)), keyby=user_id
+][,list(user_id, pref), keyby=location_id]
+saveRDS(intermediate[agg.dt[location_id %in% unique(intermediate$location_id), list(lifetime_cat, pwr_clust, vMFcluster), keyby=location_id]][,
+  list(lifetime_cat, pwr_clust, vMFcluster, pref),
+  keyby=user_id
+],"../input/userPrefs.RData")
+
+#[,.N,keyby=list(user_id, lifetime_cat, pwr_clust, vMFcluster)][,list(pref = N/sum(N),keyby=list(user_id, lifetime_cat, pwr_clust, vMFcluster)]
 # thing <- 
@@
 ```
+
+show swtiching rate normalized by visit rate on a per user basis
+
+show 27 categories

### 2015-10-15 src/main/scala/edu/cap10/actormodels/covert/Locations.scala

@@ -7,14 +7,13 @@ import breeze.stats.distributions._
 
 case class UseEvent(startDaySecs:Long, endDaySecs:Long)
 
-case class Location(id: HotSpotID, pdf:Array[Double], means:Array[Double], ks:Array[Double]) {
+case class Location(id: HotSpotID, val pdf:Array[Double], means:Array[Double], ks:Array[Double]) {
   val rng = new scala.util.Random(id)
   val gen = means.zip(ks).map { 
     case (mu, k) if k != 0d => Gamma(k, mu/k)
     case _ => null
   }
-  val cdf = pdf2CDF(pdf)
-  val searchSrc = scala.collection.Searching.search(cdf)
+   
   def draw(hour:Int) = {
     assert(gen(hour) != null, hour)
     val mean = means(hour)
@@ -25,7 +24,7 @@ case class Location(id: HotSpotID, pdf:Array[Double], means:Array[Double], ks:Ar
   
   override val toString = {
     id.toString() + System.lineSeparator() +
-    "probabilities ("+cdf.size+"): "+ cdf.mkString(", ") +System.lineSeparator() + //
+    "probabilities ("+pdf.size+"): "+ pdf.mkString(", ") +System.lineSeparator() + //
     "means("+means.size+"): "+ means.mkString(", ") +System.lineSeparator() // 
   }
 }
@@ -39,11 +38,11 @@ object Locations {
   // each line is a location id, followed by info for 0, 1, 2, ..., 23 hr of day
   
   val alllocs = {
-    val stream_cdf = io.Source.fromFile(locationCDFSrc).getLines.toList
+    val stream_pdf = io.Source.fromFile(locationPDFSrc).getLines.toList
     val stream_means = io.Source.fromFile(locationMeanSrc).getLines.toList
     val stream_ks = io.Source.fromFile(locationShapeSrc).getLines.toList
     
-    val stream = (stream_cdf, stream_means, stream_ks).zipped.toList
+    val stream = (stream_pdf, stream_means, stream_ks).zipped.toList
     
     stream.map { case (pline, mline, kline) =>
       val means = strsToDoubles(mline.split(",").tail)
@@ -54,6 +53,8 @@ object Locations {
     }.toArray
   }
 
+  def get(l:Iterable[Int]) = l.map(alllocs(_)).toArray
+  
 }
 
 //object Main extends App {

### 2015-10-15 src/main/scala/edu/cap10/actormodels/covert/SynthUser.scala

@@ -2,17 +2,28 @@ package edu.cap10.actormodels.covert
 
 import breeze.stats.distributions._
 
-case class LoginEvent(user_id:UserID, location_id:HotSpotID, login:Long, logout:Long)
+case class LoginEvent(user_id:UserID, location_id:HotSpotID, login:Long, logout:Long, purpose:String = "background") {
+  override val toString = f"$user_id,$location_id,$login,$logout,$purpose"
+}
+
+import scala.math.Ordering
+
+object LoginEventOrder extends Ordering[LoginEvent] {
+  def compare(e1:LoginEvent, e2:LoginEvent) = e1.login compare e2.login
+}
 
 /**
  * @author cap10
  */
 case class SynthUser(id:Int,
-    shape:Double, mean:Double, 
-    locations:Array[Location], prefPDF:Array[Double],
-    geomP:Double) {
+    shape:Double, mean:Double, binop:Double, 
+    locations:Array[Location], prefPDF:Array[Double]
+) {
   
   // build prefs
+  
+  assert(locations.length == prefPDF.length)
+  assert(Math.abs(prefPDF.sum - 1) <= 1e-6)
   
   val pdfHour = {
     val lp = locations.zip(prefPDF)
@@ -22,7 +33,8 @@ case class SynthUser(id:Int,
     val tot = pref.sum
     pref.map(_ / tot)
   }
-  val cdfHour = pdf2CDF(pdfHour)
+
+  val maxVisits = pdfHour.filter(_ != 0d).length
   
   val pdfLocs = {
     Array.tabulate[Double](24, locations.size) {
@@ -32,44 +44,44 @@ case class SynthUser(id:Int,
       x.map(_ / tot)
     } }
   }
-  
-  val cdfLocs = pdfLocs.map { x => pdf2CDF(x) }
-  
+   
   val gen = Gamma(shape, mean/shape)
-  val geomVisits = Geometric(geomP)
-  val searchSrc = scala.collection.Searching.search(prefPDF)
+  val geomVisits = Binomial(maxVisits-1, binop*9/(maxVisits-1))
   val rng = new scala.util.Random(id)
 
   var tilEvent : Int = gen.draw.toInt
   var meeting : Option[LoginEvent] = None
   def meet(l:Location, e:UseEvent) = {
-    meeting = Some(LoginEvent(id, l.id, e.startDaySecs, e.endDaySecs))
+    meeting = Some(LoginEvent(id, l.id, e.startDaySecs, e.endDaySecs, "covert"))
   }
   
[…]

### 2015-10-15 src/main/scala/edu/cap10/actormodels/covert/package.scala

@@ -18,19 +18,17 @@ package object covert {
   def pdfFind(pdf:Array[Double], p:Double) = {
     var in = 0
     var draw = p
-    System.out.println("pdfFind draw ", draw)
-    System.out.println("cdf: ", pdf2CDF(pdf) mkString ", ")
     while (draw > pdf(in)) {
       draw -= pdf(in)
       in += 1
     }
-    System.out.println("n: ", in)
     in
   }
   
   val locationMeanSrc = "./input/loc_means.csv"
   val locationShapeSrc = "./input/loc_shapes.csv"
   val locationCDFSrc = "./input/loc_cdf.csv"
+  val locationPDFSrc = "./input/loc_probs.csv"
   
   def strsToDoubles(ss:Array[String]) = ss.map(_.trim).map(_.toDouble)
   

### 2015-10-15 src/main/scala/edu/cap10/actormodels/demo/SimpleMain.scala

@@ -15,26 +15,26 @@ case class SimpleSystem(runConfig : SimpleParams, globalConfig : DataParams)
     
 }
 
-object SimpleMain extends App {
-  
-  val (stdout, stderr) = (System.out, System.err)
-  
-  SimpleParams().parse(args.toList) match {
-    case Some(config) => {    
-      val results = for (res <- SimpleSystem(config, MontrealProps).run()) yield { 
-        res map { case (day, events) => events map { _ + day } }
-      } map { 
-        dailyEvents =>
-          stdout.println(dailyEvents mkString System.lineSeparator)
-          stdout.flush
-      }
-      results.onComplete { _ match {
-        case Success(output) => stdout.flush
-        case Failure(e) => stderr.println(e)
-      } }
-      Await.result(results, Duration(30, SECONDS))
-    }
-    case None => stderr.println(SimpleParams.usage)
-  }
-
-}
+//object SimpleMain extends App {
+//  
+//  val (stdout, stderr) = (System.out, System.err)
+//  
+//  SimpleParams().parse(args.toList) match {
+//    case Some(config) => {    
+//      val results = for (res <- SimpleSystem(config, MontrealProps).run()) yield { 
+//        res map { case (day, events) => events map { _ + day } }
+//      } map { 
+//        dailyEvents =>
+//          stdout.println(dailyEvents mkString System.lineSeparator)
+//          stdout.flush
+//      }
+//      results.onComplete { _ match {
+//        case Success(output) => stdout.flush
+//        case Failure(e) => stderr.println(e)
+//      } }
+//      Await.result(results, Duration(30, SECONDS))
+//    }
+//    case None => stderr.println(SimpleParams.usage)
+//  }
+//
+//}

### 2015-10-15 src/main/scala/edu/cap10/cora/demo/Demo.scala

@@ -25,33 +25,33 @@ object SimulationProperties {
 
 object Demo {
 
-  def main(args:Array[String]) : Unit = {
-    val as = ActorSystem("Demo")
-    val system = TypedActor(as)
-    
-    import MontrealProps._
-    import SimulationProperties._  
-        
-    for (
-      agentN <- minAgents to maxAgents by 2;
-      plotPeriod <- maxPlotPeriod to minPlotPeriod by -10;
-      meetLocs <- 1 to maxMeetLocations;
-      agentP = dailyVisitProb - 1.0/plotPeriod;
-      sample <- 1 to sampleSize;
-      fname = f"./simdata/$agentN-$plotPeriod-$meetLocs-$sample%03d"
-    ) {
-      
-        val fh = CSVLogger.makeFH(fname)
-        val universe = system.typedActorOf(Universe.props(plotPeriod, agentN, uniqueLocs, meetLocs, agentP, avgLocs, fh))
-        (startDay to totalDays) foreach { 
-    	  t => Await.result( universe.tick(t), 1 seconds)
-        }
-        fh.flush(); fh.close()
-        println("finished "+fname)
-        system.poisonPill(universe)
-        
-    }
-    
-    as.shutdown
-  }
+//  def main(args:Array[String]) : Unit = {
+//    val as = ActorSystem("Demo")
+//    val system = TypedActor(as)
+//    
+//    import MontrealProps._
+//    import SimulationProperties._  
+//        
+//    for (
+//      agentN <- minAgents to maxAgents by 2;
+//      plotPeriod <- maxPlotPeriod to minPlotPeriod by -10;
+//      meetLocs <- 1 to maxMeetLocations;
+//      agentP = dailyVisitProb - 1.0/plotPeriod;
+//      sample <- 1 to sampleSize;
+//      fname = f"./simdata/$agentN-$plotPeriod-$meetLocs-$sample%03d"
+//    ) {
+//      
+//        val fh = CSVLogger.makeFH(fname)
+//        val universe = system.typedActorOf(Universe.props(plotPeriod, agentN, uniqueLocs, meetLocs, agentP, avgLocs, fh))
+//        (startDay to totalDays) foreach { 
+//    	  t => Await.result( universe.tick(t), 1 seconds)
+//        }
+//        fh.flush(); fh.close()
+//        println("finished "+fname)
+//        system.poisonPill(universe)
+//        
+//    }
+//    
+//    as.shutdown
+//  }
 }

### 2015-10-15 src/main/scala/edu/cap10/cora/demo/MontrealReprocess.scala

@@ -30,30 +30,30 @@ object MontrealReprocess {
 
   import java.io.{File, FilenameFilter}
   
-  def main(args:Array[String]) : Unit = {
-    val baseStream = parseFile("/Users/carlpearson/Dropbox/montreal/merged.o").toStream // args(0)
-    val mergeSrc = "/Users/carlpearson/scala-commsim/simdata" // args(1)
-    val dir = new File(mergeSrc)
-    for (src <- dir.list().filter( f => f.endsWith(".sim") )) {
-      val fw = new java.io.PrintWriter(mergeSrc+"/"+src.replace(".sim", ".cu"))
-      implicit val recorder = (obs:List[PairObs]) => {
-        obs foreach {
-          p => fw.println(p.userA+" "+p.userB+" "+p.start+" "+p.end)
-        }
-        fw.flush
-      }
-      parseFile(mergeSrc+"/"+src).foreach( o => parseOne(o, baseStream.filter(b => o.start < b.start )) )
-      fw.flush()
-      fw.close()
-    }
-    
-    // iterate over file in mergeSrc that match .sim
-    // parse(merge(basefile, simfile)) // with fw corresponding to .sim filename, but .sim replaced w/ .uu
-
-    // TODO receive a margin-of-error separate for overlaps
-    // val oname = "/Users/carlpearson/Dropbox/epidemics4share/paired.o" // args(2)
-    
-  }
+//  def main(args:Array[String]) : Unit = {
+//    val baseStream = parseFile("/Users/carlpearson/Dropbox/montreal/merged.o").toStream // args(0)
+//    val mergeSrc = "/Users/carlpearson/scala-commsim/simdata" // args(1)
+//    val dir = new File(mergeSrc)
+//    for (src <- dir.list().filter( f => f.endsWith(".sim") )) {
+//      val fw = new java.io.PrintWriter(mergeSrc+"/"+src.replace(".sim", ".cu"))
+//      implicit val recorder = (obs:List[PairObs]) => {
+//        obs foreach {
+//          p => fw.println(p.userA+" "+p.userB+" "+p.start+" "+p.end)
+//        }
+//        fw.flush
+//      }
+//      parseFile(mergeSrc+"/"+src).foreach( o => parseOne(o, baseStream.filter(b => o.start < b.start )) )
+//      fw.flush()
+//      fw.close()
+//    }
+//    
+//    // iterate over file in mergeSrc that match .sim
[…]

### 2015-10-15 src/test/scala/edu/cap10/actormodels/covert/SynthUserTests.scala

@@ -19,8 +19,35 @@ class SynthUserTests extends FunSuite {
   )
   
   test("a synth user with one location just uses that location pdf") {
-    val userA = SynthUser(-1, 1, 1, Array(testLocOne), Array(1.0), .9)
+    val userA = SynthUser(-1, 1, 1, .9, Array(testLocOne), Array(1.0))
     assert(userA.pdfHour.length === testLocOne.pdf.length)
     for (i <- 0 until 24) assert(userA.pdfHour(i) === testLocOne.pdf(i) +- 1e-8)
   }
+  
+  test("a synth user with two locations uses the combined pdf, weighted by pref") {
+    val userA = SynthUser(-1, 1, 1, .9, Array(testLocOne, testLocTwo), Array(0.5, 0.5))
+    val ref = (List.fill(12)(1d/48) ++ List.fill(12)(3d/48)).toArray
+    for (i <- 0 until 24) assert(userA.pdfHour(i) === ref(i) +- 1e-8)
+    val userB = SynthUser(-1, 1, 1, .9, Array(testLocOne, testLocTwo), Array(1d/4, 3d/4))
+    val refB = (List.fill(12)(1d/96) ++ List.fill(12)(7d/96)).toArray
+    for (i <- 0 until 24) assert(userB.pdfHour(i) === refB(i) +- 1e-8, "problem in i = "+i)
+  }
+  
+  test("all hours have location pdfs that sum to 1") {
+    val userB = SynthUser(-1, 1, 1, .9, Array(testLocOne, testLocTwo), Array(1d/4, 3d/4))
+    val refB = (List.fill(12)(1d/96) ++ List.fill(12)(7d/96)).toArray
+    for (i <- 0 until 24) assert(userB.pdfLocs(i).sum === 1d +- 1e-8, "problem in i = "+i)
+  }
+  
+  test("draws work properly") {
+    val ref = (List.fill(12)(1d/48) ++ List.fill(12)(3d/48)).toArray
+    val cref = ref.clone
+    val drawOne = 1d/49
+    assert(pdfFind(cref, drawOne) === 0)
+    val left = 1 - cref(0)
+    cref(0) = 0d
+    assert(cref(0) !== ref(0))
+    assert(pdfFind(cref, drawOne*left) === 1)
+  }
+  
 }

### 2015-10-23 rscripts/mk_users.R
tweaks to user script

@@ -1,17 +1,14 @@
-#!/usr/bin/Rscript
+#!/bin/env Rscript
 
 ## invoke from command line, w/ rscripts as wd
-## follow invocation w/
-## mv covert-set-*.csv ../input/$SUBDIRFORTHISSET/
-## mv matched.txt ../input/$SUBDIRFORTHISSET/
-## mv unmatched.txt ../input/$SUBDIRFORTHISSET/
 
 args <- commandArgs(trailingOnly = T)
 
-if (length(args) == 0) args <- c('1000', '6', 'mid', 'med', 'middle')
+if (length(args) == 0) args <- c('1000', '6', 'mid', 'med', 'middle','.')
 
 sets <- as.integer(args[1])
 count <- as.integer(args[2])
+basedir <- args[6]
 
 suppressPackageStartupMessages({
   require(data.table)
@@ -61,11 +58,11 @@ ressrc = user_rows[user_id %in% allpotentialusers,
 matched <- compute(sets, args[3], args[4], args[5], T)
 unmatched <- anticompute(sets, args[3], args[4], args[5], T)
 
-write(matched, file="matchedlocs.txt", ncolumns = 1)
-write(unmatched, file="unmatchedlocs.txt", ncolumns = 1)
+write(matched, file=paste0(basedir, "/matchedlocs.txt"), ncolumns = 1)
+write(unmatched, file=paste0(basedir, "/unmatchedlocs.txt"), ncolumns = 1)
 
 for (i in 1:sets) {
-  nm <- sprintf("covert-set-%d.csv",i)
+  nm <- sprintf("%s/covert-set-%d.csv",basedir,i)
   if (file.create(nm)) config <- file(nm, open = "w")
   users <- sample(allpotentialusers, count)
   ressrc[user_id %in% users][,{

### 2015-10-23 rscripts/supercomp.R

@@ -1,3 +1,4 @@
+#!/apps/R/3.2.0/bin/Rscript
 # TODO receive arg(s)
 #  - a config file
 #  - a series of files that provides lists to combine
@@ -10,7 +11,7 @@ size <- c(seq(5, 20, 5),30)
 k <- 2
 meetFrequency <- c(7, 14, 28)
 
-write.table(expand.grid(lifetime, variability, peakTime, size, k, meetFrequency), "../params.src",
+write.table(expand.grid(lifetime, variability, peakTime, size, k, meetFrequency), "../sim-src.pars",
   row.names = F, col.names = F, quote = F)
 
 ## script

### 2015-10-23 rscripts/mk_users.R
fix (anti)compute

@@ -4,7 +4,7 @@
 
 args <- commandArgs(trailingOnly = T)
 
-if (length(args) == 0) args <- c('1000', '6', 'mid', 'med', 'middle','.')
+if (length(args) == 0) args <- c('10', '6', 'mid', 'med', 'middle','.')
 
 sets <- as.integer(args[1])
 count <- as.integer(args[2])
@@ -22,12 +22,12 @@ censor.dt <- readRDS("../input/censor.RData")
 locs <- readRDS("../input/locClusters.RData")
 
 compute <- function(n, lcat, pcat, vcat, rep=F) sample(
-  locs[lifetime_cat == lcat & pwr_clust == pcat & vMFcluster == vcat],
+  locs[lifetime_cat == lcat & pwr_clust == pcat & vMFcluster == vcat, location_id],
   n, replace = rep
 )
 
 anticompute <- function(n, lcat, pcat, vcat, rep=F) sample(
-  locs[lifetime_cat != lcat | pwr_clust != pcat | vMFcluster != vcat],
+  locs[lifetime_cat != lcat | pwr_clust != pcat | vMFcluster != vcat, location_id],
   n, replace = rep
 )
 

### 2015-10-23 src/main/scala/edu/cap10/actormodels/covert/Locations.scala
ongoing hpc tweaks

@@ -5,6 +5,8 @@ import scala.util.Random
 
 import breeze.stats.distributions._
 
+import util.Properties.lineSeparator
+
 case class UseEvent(startDaySecs:Long, endDaySecs:Long)
 
 case class Location(id: HotSpotID, val pdf:Array[Double], means:Array[Double], ks:Array[Double]) {
@@ -23,9 +25,9 @@ case class Location(id: HotSpotID, val pdf:Array[Double], means:Array[Double], k
   }
   
   override val toString = {
-    id.toString() + System.lineSeparator() +
-    "probabilities ("+pdf.size+"): "+ pdf.mkString(", ") +System.lineSeparator() + //
-    "means("+means.size+"): "+ means.mkString(", ") +System.lineSeparator() // 
+    id.toString() + lineSeparator +
+    "probabilities ("+pdf.size+"): "+ pdf.mkString(", ") + lineSeparator + //
+    "means("+means.size+"): "+ means.mkString(", ") + lineSeparator // 
   }
 }
 

### 2015-10-23 rscripts/stow_location_data.R
update location storage

@@ -37,13 +37,13 @@ output <- censor.dt[, c(as.list({
 # output[shapes == NaN] returns ~10% of rows, however only ~.1 of usage.  So: elimate.
 # discard times w/ login durations cannot be determined
 
-filled_cdf <- dcast.data.table(output[!is.nan(shapes),{
+filled_pdf <- dcast.data.table(output[!is.nan(shapes),{
   res <- rep(0, length.out=24)
   res[login_hour+1] <- usage / sum(usage)
-  list(hour=0:23, prop = cumsum(res))
+  list(hour=0:23, prop = res)
 }, keyby=location_id], location_id ~ hour, value.var = "prop")
 
-write.table(filled_cdf, file="../input/loc_cdf.csv", sep=",", row.names = F, col.names = F)
+write.table(filled_pdf, file="../input/loc_probs.csv", sep=",", row.names = F, col.names = F)
 
 filled_means <- dcast.data.table(output[!is.nan(shapes),{
   res <- rep(0, length.out=24)

### 2015-10-23 rscripts/mk_users.R
fix location bugs

@@ -19,7 +19,9 @@ suppressPackageStartupMessages({
 refusers <- readRDS("../input/user.RData")
 user_rows <- readRDS("../input/userPrefs.RData")
 censor.dt <- readRDS("../input/censor.RData")
-locs <- readRDS("../input/locClusters.RData")
+locsub <- fread("../input/loc_probs.csv")$V1
+locs <- readRDS("../input/locClusters.RData")[location_id %in% locsub]
+
 
 compute <- function(n, lcat, pcat, vcat, rep=F) sample(
   locs[lifetime_cat == lcat & pwr_clust == pcat & vMFcluster == vcat, location_id],

### 2015-10-23 src/main/scala/edu/cap10/actormodels/covert/Locations.scala

@@ -46,13 +46,20 @@ object Locations {
     
     val stream = (stream_pdf, stream_means, stream_ks).zipped.toList
     
-    stream.map { case (pline, mline, kline) =>
+    val prelim = stream.map { case (pline, mline, kline) =>
       val means = strsToDoubles(mline.split(",").tail)
       val ks    = strsToDoubles(kline.split(",").tail)
       val other = pline.split(",")
       val cdf   = strsToDoubles(other.tail)
       Location(other.head.trim.toInt, cdf, means, ks)
     }.toArray
+    
+    val maxid = prelim.map(_.id.toInt).max
+    val res = Array.ofDim[Location](maxid)
+    prelim.foreach { l => 
+      res(l.id.toInt-1) = l
+    }
+    res
   }
 
   def get(l:Iterable[Int]) = l.map(alllocs(_)).toArray

### 2015-10-23 src/main/scala/edu/cap10/actormodels/covert/SynthUser.scala

@@ -24,6 +24,7 @@ case class SynthUser(id:Int,
   
   assert(locations.length == prefPDF.length)
   assert(Math.abs(prefPDF.sum - 1) <= 1e-6)
+  assert(locations.find({ _ == null }).isEmpty, id.toString + ":" + locations.indexOf(null) + "\n" + (locations mkString "\n"))
   
   val pdfHour = {
     val lp = locations.zip(prefPDF)
@@ -68,7 +69,7 @@ case class SynthUser(id:Int,
       var pdfs = pdfHour.clone
       var left = 1.0
       val hours = Array.ofDim[Int](n)
-      (0 until n).map { i =>
+      (0 until n).foreach { i =>
         var draw = rng.nextDouble() * left
         val hour = pdfFind(pdfs, draw)
         left = left - pdfs(hour)

### 2016-04-26 rscripts/location_clusters.Rmd
updates to drawing scripts

@@
 
-```{r, echo=FALSE, include=FALSE}
+```{r, loadpkgs, echo=FALSE, include=FALSE}
 req.packages <- c("data.table", "ggplot2", "reshape2", "grid", "scales", "lubridate", "movMF")
@@
 
-```{r, echo=FALSE}
+```{r, loadData, echo=FALSE}
 max_hours <- 24
@@
 # censor.dt[,list(lifetime = max(logout)-min(login)), by=list(location_id, user_id)][lifetime > 60*60*24*365, unique(user_id)]
+```
+
+Let's look at some quick views of some components of the data.  First, how many hotspots are in use?  We can look at the life of all hotspots:
+
+```{r, location_counts, echo=F, fig.cap="Location Counts"}
+loclifetimes <- censor.dt[,
+  list(start=min(login), end=max(logout)),
+  keyby=location_id
+]
+p <- ggplot(loclifetimes) + theme_bw() + aes(y=location_id, yend=location_id, x=start/3600/24, xend=end/3600/24) + geom_segment(alpha=0.5) + labs(x="start / end day", y="location") + scale_x_continuous(breaks=(0:6)*365)
+ggsave("location-liftimes.png", p, width=7, height=4)
+p
+```
+
+as well as the net count of locations over time:
+
+```{r, location_counts, echo=F, fig.cap="Location Counts"}
+locCounts <- melt.data.table(loclifetimes, id.vars = "location_id")[,
+  list(change=sum(c(1,-1)[variable])), keyby=value
+]
+p <- ggplot(locCounts[,list(time=value/3600/24, net=cumsum(change))][net>0]) + theme_bw() + aes(x=time, y=net) + geom_step() + coord_cartesian(xlim=c(1,2000), ylim=c(1,300), expand = F) + scale_y_log10("net locations", breaks=c(1,2,5,10,20,50,100,200)) + scale_x_continuous(breaks=(0:6)*365)
+ggsave("location-net.png", p, width=7, height=4)
+p
+```
+
+There's a similar trend in users:
+
+```{r, location_counts, echo=F, fig.cap="Location Counts"}
+userCounts <- melt.data.table(lifetimes <- censor.dt[,
+  list(start=min(login), end=max(logout)),
+  keyby=user_id
+], id.vars = "user_id")[,
+  list(change=sum(c(1,-1)[variable])), keyby=value
+]
+p <- ggplot(userCounts[,list(time=value/3600/24, net=cumsum(change))][net>0]) + theme_bw() + aes(x=time, y=net) + geom_step() + coord_cartesian(xlim=c(1,2000), ylim=c(1,20000), expand = F) + scale_x_continuous(breaks=(0:6)*365) + scale_y_log10("net users", breaks=c(1,2,5,10,20,50,100,200,500,1000,2000,5000,10000,20000))
+ggsave("user-net.png", p, width=7, height=4)
+p
 ```
@@
 ```{r, echo=F, fig.cap="Raw proportions by hour"}
-ggplot(temporal_data_slice) + theme_bw() + aes(group = location_id, x = hour, y = value) + facet_grid(measure ~ .) + geom_line(alpha=0.2) + scale_x_discrete(breaks=0:23, expand=c(0,0)) +
+p <- ggplot(temporal_data_slice) + theme_bw() + aes(group = location_id, x = hour, y = value) + facet_grid(measure ~ .) + geom_line(alpha=0.2) + scale_x_discrete(breaks=0:23, expand=c(0,0)) +
   scale_y_continuous("proportion", expand=c(0,0))
+ggsave("time-sig.png", p, width=7, height=4)
+p
 ```
@@
 peaks <- temporal_data_slice[, list(peak=which.max(value)-1), by=list(location_id, measure)]
-ggplot(peaks) + theme_bw() + aes(x = peak) + facet_grid(measure ~ .) + geom_bar(binwidth=1)
+p <- ggplot(peaks) + theme_bw() + aes(x = peak) + facet_grid(measure ~ .) +
+  geom_bar() +
+  scale_x_discrete("peak hour", limits=0:23)
+p
 ```
@@
 
-ggplot(pwr[,
+p <- ggplot(pwr[,
   list(power=value, frequency=as.numeric(sub("_perday","", variable))), by=list(location_id, measure)
@@
 # pwr[, list(peak_freq = variable[which.max(value)]), by=list(location_id, measure)][,.N,by=list(measure, peak_freq)]
+p
 ```
@@
 ```{r, echo=F, fig.cap="Login, Logout, Usage Peak Clusters"}
-ggplot(peaks[temporal_cluster]) + theme_bw() + aes(x = peak, fill=vMFcluster) + facet_grid(measure ~ .) + geom_bar(binwidth=1) + labs(fill="von Mises-Fisher\n3-Sphere\nCluster\nPeak Time")
+p <- ggplot(peaks[temporal_cluster]) + theme_bw() + aes(x = peak, fill=vMFcluster) + facet_grid(measure ~ .) + geom_bar(binwidth=1) + labs(fill="von Mises-Fisher\n3-Sphere\nCluster\nPeak Time")
+ggsave("time-peaks.png", p, width=7, height=4)
+p
 ```
@@
 ]
-ggplot(pwr[pwr_cluster][,
+p <- ggplot(pwr[pwr_cluster][,
   list(power=value, frequency=as.numeric(sub("_perday","", variable)), peak_freq, pwr_cat),
@@
 ]) + theme_bw() + aes(y = power, x=frequency, group=location_id, alpha=peak_freq, color = pwr_cat) + facet_grid(measure ~ .) + geom_line() + scale_x_continuous("per day frequency", breaks=c(1,2,3,4,6,8,12))
+ggsave("freq-powers.png", p, width=7, height=4)
+p
 ```
@@
   scale_alpha_manual("power\ncluster", limits=c("low","med","hi"), values=(2^0.5)/c(2,5,8)^0.5)
+ggsave("locationclusters.png", p, width=7, height=4)
 p
@@
 setkey(pwr_cluster, location_id, measure)
-ggplot(temporal_data_slice[temporal_cluster][pwr_cluster]) + theme_bw() +
+p <- ggplot(temporal_data_slice[temporal_cluster][pwr_cluster]) + theme_bw() +
   aes(group = location_id, x = hour, y = value, color=pwr_cat) +
@@
   scale_y_continuous("proportion", expand=c(0,0))
+p
 ```
@@
 )
-ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1) + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
+p <- ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1) + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
+p
 ```
@@
 ```{r, echo=F, fig.width=11.25, fig.height=7, fig.cap="normalized distribution of main location type usage"}
-ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1, position="fill") + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
+p <- ggplot(digests) + facet_grid(type ~ .) + theme_bw() + aes(x=log10(switch_rate), fill=main, alpha=factor(floor(share*10))) + geom_bar(binwidth=0.1, position="fill") + scale_alpha_manual(breaks=0:10, values=seq(0.5,1,length.out = 11))
+p
 ```
@@
 #thing[, list(user_id, mains=interaction(lifetime_main, pwr_main, peak_main))][,.N,by=mains]
-ggplot(user_src.dt[, list(y=log10(usage), x=log10(life), pwr_main, lifetime_main, peak_main, membership = lifetime_share * pwr_share * peak_share)]) +facet_grid(pwr_main ~ lifetime_main, labeller = function(var, val){
-  if (var == "pwr_main") {
-    ifelse(val == "lo", "smooth usage", ifelse(val=="med","some usage variability","highly variable usage"))
+p <- ggplot(user_src.dt[, list(y=log10(usage), x=log10(life), pwr_main, lifetime_main, peak_main, membership = lifetime_share * pwr_share * peak_share)]) +facet_grid(pwr_main ~ lifetime_main, labeller = function(vardt){
+  if (!is.null(vardt$pwr_main)) {
+    list(c(lo="smooth usage",med="some usage variability",hi="highly variable usage")[as.character(vardt$pwr_main)])
   } else {
-    ifelse(val == "low", "low usage per time", ifelse(val=="mid","typical usage per time","high usage per time"))
+    list(c(low="low usage per time", mid="typical usage per time",high="high usage per time")[as.character(vardt[[1]])])
   }
 }) + aes(x=x,y=y, color=factor(peak_main), alpha = membership) + geom_point() + theme_bw() + labs(y="log(usage)", x="log(life)", color="usage\npeak")
+ggsave("users.png", p)
+p
 ```
@@
 
+```{r}
+slice <- user_src.dt[lifetime_main=="mid" & pwr_main == "med" & peak_main == "middle"]
+mlt.slice <- melt.data.table(slice[,list(life,ulog=log10(usage), unilocs=ul),by=user_id], id.vars = "user_id", measure.vars = c("life","ulog","unilocs"))
+ggsave("mmm_life.png", ggplot(mlt.slice[variable == "life"]) + theme_bw() + aes(x=value, y=..density..) + geom_histogram() + labs(x="lifetime (end - start)", y="density"), width=7, height=4)
+ggsave("mmm_usage.png", ggplot(mlt.slice[variable == "ulog"]) + theme_bw() + aes(x=value, y=..density..) + geom_histogram() + labs(x="log10(usage)", y="density"), width=7, height=4)
+ggsave("mmm_ulocs.png", ggplot(mlt.slice[variable == "unilocs"]) + theme_bw() + aes(x=value, y=..density..) + geom_histogram() + labs(x="unique locations", y="density"), width=7, height=4)
+
+slice <- user_src.dt[lifetime_main=="mid" & pwr_main == "lo" & peak_main == "early"]
+mlt.slice <- melt.data.table(slice[,list(life,ulog=log10(usage), unilocs=ul),by=user_id], id.vars = "user_id", measure.vars = c("life","ulog","unilocs"))
+ggsave("mle_life.png", ggplot(mlt.slice[variable == "life"]) + theme_bw() + aes(x=value, y=..density..) + geom_histogram() + labs(x="lifetime (end - start)", y="density"), width=7, height=4)
+ggsave("mle_usage.png", ggplot(mlt.slice[variable == "ulog"]) + theme_bw() + aes(x=value, y=..density..) + geom_histogram() + labs(x="log10(usage)", y="density"), width=7, height=4)
+ggsave("mle_ulocs.png", ggplot(mlt.slice[variable == "unilocs"]) + theme_bw() + aes(x=value, y=..density..) + geom_histogram() + labs(x="unique locations", y="density"), width=7, height=4)
+```
+
+```{r}
+mmm.cc <- loclifetimes[setkey(fread("../../muri-overall/output/matched/mid/med/middle/10/001-covert-0-cc.csv", col.names = c("user.a","user.b","location_id","start","end","other")), location_id)][start < i.end & i.start < end, list(start=i.start, end=i.end, user.a, user.b, location_id)]
+mmm.cu <- loclifetimes[setkey(fread("../../muri-overall/output/matched/mid/med/middle/10/001-covert-0-cu.csv", col.names = c("user.a","user.b","location_id","start","end","other")), location_id)][start < i.end & i.start < end, list(start=i.start, end=i.end, user.a, user.b, location_id)]
+mmm.all <- rbind(mmm.cc, mmm.cu)
+mmm.both <- rbind(
+  mmm.all[user.a < 0, list(mx=max(end),mn=min(start),sm=sum(end-start),ul=length(unique(location_id))), by=list(user_id=user.a)],
+  mmm.all[user.b < 0, list(mx=max(end),mn=min(start),sm=sum(end-start),ul=length(unique(location_id))), by=list(user_id=user.b)]
+)[,list(life=max(mx)-min(mn), usage=sum(sm), ul=sum(ul)),by=user_id]
+```
+
+```{r}
+mle.cc <- loclifetimes[setkey(fread("../../muri-overall/output/matched/mid/lo/late/10/001-covert-0-cc.csv", col.names = c("user.a","user.b","location_id","start","end","other")), location_id)][start < i.end & i.start < end, list(start=i.start, end=i.end, user.a, user.b, location_id)]
+mle.cu <- loclifetimes[setkey(fread("../../muri-overall/output/matched/mid/lo/late/10/001-covert-0-cu.csv", col.names = c("user.a","user.b","location_id","start","end","other")), location_id)][start < i.end & i.start < end, list(start=i.start, end=i.end, user.a, user.b, location_id)]
+mle.all <- rbind(mle.cc, mle.cu)
+mle.both <- rbind(
+  mle.all[user.a < 0, list(mx=max(end),mn=min(start),sm=sum(end-start),ul=length(unique(location_id))), by=list(user_id=user.a)],
+  mle.all[user.b < 0, list(mx=max(end),mn=min(start),sm=sum(end-start),ul=length(unique(location_id))), by=list(user_id=user.b)]
+)[,list(life=max(mx)-min(mn), usage=sum(sm), ul=sum(ul)),by=user_id]
+```
+
 show swtiching rate normalized by visit rate on a per user basis
@@
 show 27 categories
+
+group sizes 5-20 by 5, also a 30
+meeting frequency - once a week, once every two weeks, once every month
+try
+ - meeting at less probable times
+ - meeting location be inconsistent with normal behavior
+
+detection based on cycles

### 2016-05-18 mkusergroup.R
make up covert group sample targets

@@ -1,0 +1,125 @@
+#!/usr/bin/env Rscript
+rm(list=ls())
+## invoke from command line, w/ rscripts as wd
+
+# input/user.RData
+# input/censor.RData
+# input/locClusters.RData
+# input/userPrefs.RData
+# input/loc_probs.csv
+
+suppressPackageStartupMessages({
+  require(data.table)
+  require(stats4)
+  require(methods)
+  require(optparse)
+})
+
+getseed <- function(chr) set.seed(as.integer(chr))
+
+parse_args <- function(argv = commandArgs(trailingOnly = T)) {
+  parser <- optparse::OptionParser(
+    usage = "usage: %prog [data srcs] (low|mid|high) (lo|med|hi) (early|middle|late) N I",
+    description = "visualize persistence community results",
+    option_list = list(
+      optparse::make_option(
+        c("--verbose","-v"),  action="store_true", default = FALSE,
+        help="verbose?"
+      ),
+      optparse::make_option(
+        c("--target","-t"), default = "input",
+        help="target base directory; default input"
+      )
+    )
+  )
+  req_pos <- list(
+    users.dt=readRDS, censor.dt=readRDS, locrefs.dt=readRDS,
+    loc_probs = fread, userPrefs.dt = readRDS,
+    lft_cat=function(ar){
+      stopifnot(grepl("(low|mid|high)", ar))
+      ar
+    },
+    pwr_cat=function(ar){
+      stopifnot(grepl("(lo|med|hi)", ar))
+      ar
+    },
+    tm_cat=function(ar){
+      stopifnot(grepl("(early|middle|late)", ar))
+      ar
+    },
+    count=as.integer, seed=getseed
+  )
+  parsed <- optparse::parse_args(parser, argv, positional_arguments = length(req_pos))
+  parsed$options$help <- NULL
+  result <- c(mapply(function(f,c) f(c), req_pos, parsed$args, SIMPLIFY = F), parsed$options)
+  if(result$verbose) print(result)
+  result
+}
+
+cat(with(parse_args(
+# c("input/digest/clustering/userrefs.rds", "input/digest/filter/detail_input.rds", "input/digest/clustering/locrefs.rds", "input/digest/filter/location_pdf.csv", "input/digest/clustering/uprefs.rds", "high", "hi", "late", "20", "001")
+), {
[…]

### 2016-05-18 sample_pbs.sh

@@ -1,0 +1,18 @@
+cat <<EOF
+#!/bin/bash
+#PBS -r n
+#PBS -N $1
+#PBS -o $1.o
+#PBS -e $1.err
+#PBS -m a
+#PBS -M cap10@ufl.edu
+#PBS -l walltime=00:30:00
+#PBS -l nodes=1:ppn=1
+#PBS -l pmem=1gb
+#PBS -t 1-$3
+
+module load gcc/5.2.0 R/3.2.2
+cd /scratch/lfs/cap10/scala-commsim
+tar=\$(printf 'input/digest/covert/$2/%03d.csv' \$PBS_ARRAYID)
+make \$tar
+EOF

### 2016-05-18 rscripts/location_clusters.Rmd
make targets for processed samples

@@
 ```{r, echo=F}
-process_hour_rle <- function(prefix) {
-  hrsind <- 0:23
-  nms <- sapply(hrsind, function(h) sprintf("%s_hr_%.2d", prefix, h))
-  function(hrrle) with(hrrle, {
-    missing <- hrsind[!(hrsind %in% values)]
-    res <- c(lengths, rep.int(0, length(missing)))[order(c(values, missing))] / sum(lengths)
-    names(res) <- nms
-    res
-  })
-}
-
 day.abb <- c("Sun", "Mon", "Tues", "Wed","Thurs","Fri","Sat")
@@
 )
+
+process_hour_rle <- function(prefix) {
+  hrsind <- 0:23
+  nms <- sapply(hrsind, function(h) sprintf("%s_hr_%.2d", prefix, h))
+  function(hrrle) with(hrrle, {
+    missing <- hrsind[!(hrsind %in% values)]
+    res <- c(lengths, rep.int(0, length(missing)))[order(c(values, missing))] / sum(lengths)
+    names(res) <- nms
+    res
+  })
+}
 
@@
 })
+
 digestor <- function(dt, preds, training, cache = "../input/initPred.RData") {

### 2016-05-18 translate_pbs.sh

@@ -1,0 +1,18 @@
+cat <<EOF
+#!/bin/bash
+#PBS -r n
+#PBS -N $1
+#PBS -o $1.o
+#PBS -e $1.err
+#PBS -m a
+#PBS -M cap10@ufl.edu
+#PBS -l walltime=01:00:00
+#PBS -l nodes=1:ppn=1
+#PBS -l pmem=1gb
+#PBS -t 1-$3
+
+module load gcc/5.2.0 R/3.2.2
+cd /scratch/lfs/cap10/scala-commsim
+tar=\$(printf 'input/simulate/covert/$2/%03d/cc.csv' \$PBS_ARRAYID)
+make \$tar
+EOF

### 2016-05-18 translate.R
mv translate.R to scala-commsim

@@ -1,0 +1,18 @@
+# translate.R raw sim output => new user_ids (negatives), original location ids, correct times
+
+args <- commandArgs(trailingOnly = T)
+# args <- c("input/remap-location-ids.rds", "output/matched/mid/lo/late/10/001-covert-out.csv", "20649600")
+
+
+require(data.table)
+
+location_remap <- readRDS(args[1])[, location_id, keyby=new_location_id]
+orig <- fread(args[2])
+
+orig$V2 <- location_remap[orig$V2, location_id]
+orig$V1 <- -orig$V1
+offset <- as.integer(args[3])
+orig$V3 <- orig$V3 + offset
+orig$V4 <- orig$V4 + offset
+
+write.table(orig, file = stdout(), row.names = F, col.names = F, sep = ",", quote = F)

### 2016-05-18 translate_pbs.sh
corrections to translation targets

@@ -11,7 +11,7 @@ cat <<EOF
 #PBS -l pmem=1gb
 #PBS -t 1-$3
 
-module load gcc/5.2.0 R/3.2.2
+module load gcc/5.2.0 R/3.2.2 scala
 cd /scratch/lfs/cap10/scala-commsim
 tar=\$(printf 'input/simulate/covert/$2/%03d/cc.csv' \$PBS_ARRAYID)
 make \$tar

### 2016-05-18 src/main/scala/edu/cap10/actormodels/covert/package.scala
change source files for locations

@@ -25,10 +25,9 @@ package object covert {
     in
   }
   
-  val locationMeanSrc  = "./input/loc_means.csv"
-  val locationShapeSrc = "./input/loc_shapes.csv"
-  val locationCDFSrc   = "./input/loc_cdf.csv"
-  val locationPDFSrc   = "./input/loc_probs.csv"
+  val locationMeanSrc  = "./input/digest/filter/location_means.csv"
+  val locationShapeSrc = "./input/digest/filter/location_shapes.csv"
+  val locationPDFSrc   = "./input/digest/filter/location_pdf.csv"
   
   def strsToDoubles(ss:Array[String]) = ss.map(_.trim).map(_.toDouble)
   

### 2016-05-18 translate_pbs.sh

@@ -11,7 +11,7 @@ cat <<EOF
 #PBS -l pmem=1gb
 #PBS -t 1-$3
 
-module load gcc/5.2.0 R/3.2.2 scala
+module load gcc/5.2.0 R/3.2.2 java/1.8.0_31 scala
 cd /scratch/lfs/cap10/scala-commsim
 tar=\$(printf 'input/simulate/covert/$2/%03d/cc.csv' \$PBS_ARRAYID)
 make \$tar

### 2016-05-18 src/main/scala/edu/cap10/actormodels/covert/Locations.scala
fix locations use in simulator + preserve intermediates in makefile

@@ -54,11 +54,9 @@ object Locations {
       Location(other.head.trim.toInt, cdf, means, ks)
     }.toArray
     
-    val maxid = prelim.map(_.id.toInt).max
+    val maxid = prelim.map(_.id.toInt).max + 1
     val res = Array.ofDim[Location](maxid)
-    prelim.foreach { l => 
-      res(l.id.toInt-1) = l
-    }
+    prelim.foreach { l => res(l.id.toInt) = l }
     res
   }
 

### 2016-05-18 src/main/scala/edu/cap10/actormodels/covert/SynthUser.scala

@@ -102,7 +102,7 @@ object SynthUser {
     val Array(shape, mean, binop) = same.map(_.toDouble)
     val len = varying.length / 2
     val (locs, prefs) = varying.splitAt(len)
-    SynthUser(in._2+1, Gamma(shape, mean/shape), binop, Locations.get(locs.map(_.toInt - 1)), prefs.map(_.toDouble))
+    SynthUser(in._2+1, Gamma(shape, mean/shape), binop, Locations.get(locs.map(_.toInt)), prefs.map(_.toDouble))
   }
 }
 

### 2016-05-18 translate.R
do not translate location ids

@@ -1,17 +1,15 @@
 # translate.R raw sim output => new user_ids (negatives), original location ids, correct times
 
 args <- commandArgs(trailingOnly = T)
-# args <- c("input/remap-location-ids.rds", "output/matched/mid/lo/late/10/001-covert-out.csv", "20649600")
+# args <- c("output/matched/mid/lo/late/10/001-covert-out.csv", "20649600")
 
 
 require(data.table)
 
-location_remap <- readRDS(args[1])[, location_id, keyby=new_location_id]
-orig <- fread(args[2])
+orig <- fread(args[1])
 
-orig$V2 <- location_remap[orig$V2, location_id]
 orig$V1 <- -orig$V1
-offset <- as.integer(args[3])
+offset <- as.integer(args[2])
 orig$V3 <- orig$V3 + offset
 orig$V4 <- orig$V4 + offset
 
